### 2012-11-10 checkin/project-checkin.tex
checkin

\documentclass[letterpaper,12pt]{article}
\usepackage[left=3cm,right=3cm,top=3cm,bottom=3cm]{geometry}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{float}
\usepackage[usenames]{xcolor}

\title{MATH-469 Project Checkin}
\author{Eric Biggers}

\begin{document}
\maketitle

The goal of my project is to implement an algorithm that can assemble a genome,
given a set of reads that came from it.  The key idea in the algorithm I am
implementing is the construction of a string graph where the edges are labeled
with DNA sequences and the vertices correspond to reads, and a path through the
graph corresponds to a set of reads that assemble together consistently.

I have implemented the initial stages of the algorithm.  Each step has been
implemented as a separate binary program.  This means that the overall assembly
process will iteratively transform the data (including the graph) on-disk into
the final assembly.  The currently implemented programs are:

\begin{itemize}
\item
	{\tt convert-reads}:  Imports and merges a set of reads
	into a concise binary representation.
\item
	{\tt compute-overlaps}:  Takes as input a set of reads and outputs all
	pairwise overlaps between reads of a minimum length specified by the
	{\tt --min-overlap-len} option.  The algorithm avoids a pairwise
	comparison of all reads by using a hash table to find reads sharing a
	seed.  Only exact overlaps are considered at this point.

\item
	{\tt print-overlaps}:  Prints a textual representation of a set of
	overlaps, given a file containing the overlaps in binary format.

\item
	{\tt remove-contained-reads}:
	Given a set of reads and the overlaps that were computed from them,
	finds all reads that are fully contained by another read and discard
	them, along with the corresponding overlaps.  If a pair of reads is
	identicaly, only one copy of the read is kept.

\item
	{\tt build-directed-string-graph}:
	Given a set of reads and the overlaps that were computed from them,
	build the {\bf directed} string graph that models the genome assembly.
\item
	{\tt build-bidirected-string-graph}:
	Given a set of reads and the overlaps that were computed from them,
	build the {\bf bidirected} string graph that models the genome assembly.

\item
	{\tt print-string-graph}:
	Prints a textual representation of a directed or bidirected string
	graph, or prints statistics about the graph.

\item
	{\tt transitive-reduction}:
	Given a string graph, remove edges $v \to x$ where there exist edges $v
	\to w \to x$, provided that $v \to x$ is labeled by the same sequence as
	that of $v \to w$ concatenated with $v \to x$.  This program currently
	only works on directed string graphs, but the bidirected case is in
	progress.

\item
	{\tt collapse-unbranched-paths}:
	Given a string graph, collapse chains of vertices that have in-degree 1
	and out-degree 1.  This program currently only works on directed string
	graphs, but the bidirected case is in progress.

\item
	{\tt digraph-to-bidigraph}:
	Convert a directed string graph into a bidirected string graph.

\end{itemize}

The modular design allows for multiple possible data flows, but a possible flow
is {\tt convert-reads} $\to$ {\tt compute-overlaps} $\to$ {\tt
remove-contained-reads} $\to$ \newline {\tt build-directed-string-graph} $\to$ {\tt
transitive-reduction} $\to$ {\tt digraph-to-bidigraph}, which will produce a
transitively-reduced, collapsed bidirected string graph produced from overlaps
between non-contained reads.

Even with the final stages not implemented, the algorithm is already effective
when used on random genomes because with no spurious overlaps, the
transitively-reduced string graph is a chain of vertices with no branching,
which then collapses into a single edge in {\tt collapse-unbranched-paths}.
This is true even if the genome is 15 million random base pairs, given 100 bp
reads sampled uniformly from either strand with no errors.  (Such a genome takes
about 1 minute to assemble with maximum memory usage 1.8 GB, with the main time
and memory bottleneck being the {\tt compute-overlaps} program).

What follows is an explanation of the initial stages of the algorithm on
64-character genome and three r

\noindent
{\tt 1.}
{\tt \color{red} $\leftarrow$TGAGCTTAAGGCTTATCTATCTTCAGACGACA$\leftarrow$ }\\
{\tt 2.}
\hspace*{2.58cm}{\tt \color{red} $\leftarrow$TTATCTATCTTCAGACGACTATTATAGCGCGG$\leftarrow$}\\
{\tt G.}
{\tt $\leftarrow$TGAGCTTAAGGCTTATCTATCTTCAGACGACTATTATAGCGCGGCCAAGACTACGCGGAGCCCC$\leftarrow$} \\
{\tt G.}
{\tt $\rightarrow$ACTCGAATTCCGAATAGATAGAAGTCTGCTGATAATATCGCGCCGGTTCTGATGCGCCTCGGGG$\rightarrow$} \\
{\tt 3.}
\hspace*{5.16cm}{\tt \color{blue} $\rightarrow$TCTGCTGATAATATCGCGCCGGTTCTGATGCG$\rightarrow$}

\begin{figure}
\includegraphics{out.digraph.pdf}
\end{figure}

\begin{figure}
\includegraphics{out.bidigraph.pdf}
\end{figure}

\begin{figure}
\includegraphics{out.reduced.digraph.pdf}
\end{figure}

\begin{figure}
\includegraphics{out.reduced.bidigraph.pdf}
\end{figure}

\begin{figure}
\includegraphics{out.reduced.collapsed.digraph.pdf}
\end{figure}

\begin{figure}
\includegraphics{out.reduced.collapsed.bidigraph.pdf}
\end{figure}

\end{document}

### 2012-11-10 src/BidirectedStringGraph.h

@@ -206,35 +206,34 @@ public:
 			   << "v" << get_v2_idx()
 			   << " ["
 			   << " dir=both arrowhead=" << head_2
-			   << " arrowtail=" << head_1
-			   << " taillabel=\"";
-			if (print_seqs)
-				os << _seq_1_to_2;
-			else
-				os << _seq_1_to_2.length();
-			os << "\" headlabel=\"";
-			if (print_seqs)
-				os << _seq_2_to_1;
-			else
-				os << _seq_2_to_1.length();
-			//
-			//   |\\\\
-			//   |    \\\\\
-			// C |         \\\\
-			//   --------------
-			//    length
-			//
-			// C / length = tan(theta)
-			// theta = atan(C/length)
-			// labelangle = 90-180*(2pi)*theta
-			const double C = 2.0;
-			const double len = double(length());
-			const double theta = atan(C / len);
-			const double labelangle = 90 - (180 / (2.0 * M_PI)) * theta;
-			os << '"'
-			   << " labelangle=" << labelangle
-			   << " labeldistance=" << double(length()) / 2.5
-			   << " ];\n";
+			   << " arrowtail=" << head_1;
+			if (print_seqs) {
+				os << " taillabel=\""
+				   << _seq_1_to_2
+				   << "\" headlabel=\""
+				   << _seq_2_to_1
+				   << '"';
+
+				//
+				//   |\\\\
+				//   |    \\\\\
+				// C |         \\\\
+				//   --------------
+				//    length
+				//
+				// C / length = tan(theta)
+				// theta = atan(C/length)
+				// labelangle = 90-180*(2pi)*theta
+				const double C = 2.0;
+				const double len = double(length());
+				const double theta = atan(C / len);
+				const double labelangle = 90 - (180 / (2.0 * M_PI)) * theta;
+				os << " labelangle=" << labelangle
+				   << " labeldistance=" << (double(len) / 2.5 + 1.0);
+			} else {
+				os << " label=\"" << length() << "\"";
+			}
+			os << " ];\n";
 		}
 	}
 };

### 2012-11-10 checkin/project-checkin.tex
checkin

@@
 \usepackage{enumerate}
+\usepackage{caption}
 \usepackage{graphicx}
@@
 
-What follows is an explanation of the initial stages of the algorithm on
-64-character genome and three r
+What follows is an explanation of the initial stages of the algorithm on a very
+simple case.  The diagram below shows a 64- base-pair ``genome'' that has had
+three 32-base-pair error-free reads at a constant separation of 12 base pairs.
+Note that the genome is two-stranded.  {\tt A} always pairs with {\tt T} and
+{\tt C} always pairs with {\tt G}, but the two strands run in opposite
+directions.  Reads 1 and 2 come from the top (reverse strand), while read 3
+comes from the bottom (forward strand).  The reads are directed as shown (e.g.
+Read 1 begins with {\tt ACAGC}, reading from right to left, not {\tt TGAGC}).
+
+%make out.{bi,}digraph.{pdf,dot} out.reduced.{bi,}digraph.pdf
+%out.reduced.collapsed.{bi,}digraph.pdf SAMPLE_SIZE=64 USE_PIRS=false
+%MIN_OVERLAP_LEN=8 READ_LEN=32 READ_SEP=12 -B
+\vspace{0.5cm}
+\nopagebreak
 
@@
 
-\begin{figure}
-\includegraphics{out.digraph.pdf}
-\end{figure}
+From looking at the diagram, it is apparent that Read 1 overlaps with Read 2 by
+20 base pairs, Read 1 overlaps with Read 2 by 20 base pairs, and Read 1 overlaps
+with Read 3 by 8 base pairs.  These overlaps are computed by the {\tt
+compute-overlaps} program.  Note that an overlap may match either the forward or
+reverse-complement sequence.
 
-\begin{figure}
-\includegraphics{out.bidigraph.pdf}
-\end{figure}
+Before building the string graph, {\tt remove-contained-reads} is run to remove reads
+that are fully contained by another read.  There are no such reads in this
+example.
 
-\begin{figure}
-\includegraphics{out.reduced.digraph.pdf}
-\end{figure}
+Given the set of overlaps, the next step is to build a bidirected or directed
+string graph.  Both these graphs are shown on the next page.  In the directed
+graphs, a node $n.B$ means the beginning of read $n$, while the node $n.E$ means
+the end of read $n$.
 
-\begin{figure}
-\includegraphics{out.reduced.bidigraph.pdf}
-\end{figure}
+Consider graph (2A), the directed string graph after transitive reduction, and
+consider the walk $1.B \to 2.B \to 3.E$.  This represents a way in which reads
+1, 2, and 3 assemble together consistently.  In the simplest case, all the nodes
+in a path would be $E$ nodes, and this would correspond to a sequence of reads
+that assemble together as they are, with no reverse-complementation.  However,
+in this case, we walk from the {\it beginning} of read 1 to the {\it beginning}
+of read 2, indicating that they are laid out in reverse-complement order, before
+reaching the {\it end} of read 3, which is in forward order.  This
+reconstruction is consistent with how the reads were actually sampled from the
+genome (shown earlier).
 
-\begin{figure}
-\includegraphics{out.reduced.collapsed.digraph.pdf}
-\end{figure}
+The other component of graph (2A) produces an equivalent assembly, but it lays
+out the reads in the opposite order and assembles the other strand of the
+genome.
 
-\begin{figure}
-\includegraphics{out.reduced.collapsed.bidigraph.pdf}
-\end{figure}
+Graph (2B) is equivalent to (2A), but combines every node $n.B$ and $n.E$ into
+one node $n$ to deal with the double-strandedness in a more concise manner.  The
+edges are bidirected.  A bidirected edge may be traversed either way in a walk,
+but if a vertex is entered through an inward head, it must be left through an
+outward head, and vice versa.
+
+In both the directed and bidirected graphs, the edges are labeled by DNA
+sequences, although in the images, only the lengths of the sequences are shown.
+
+\newpage
+\begin{tabular}{|p{0.5\textwidth}|p{0.5\textwidth}|} \hline
+\includegraphics[scale=0.6]{out.digraph-crop.pdf}
+\newline 1A. Directed string graph of original data
+&
+\includegraphics[scale=0.6]{out.bidigraph-crop.pdf}
+\newline 1B. Equivalent bidirected string graph to (1A)
+\\ \hline
+
+\includegraphics[scale=0.6]{out.reduced.digraph-crop.pdf}
+\newline 2A. Graph (1A) after transitive reduction.  Read 2 is between reads 1
+and 3, so there is no need for edges directly between 1 and 3.
+&
+\includegraphics[scale=0.6]{out.reduced.bidigraph-crop.pdf}
+\newline 2B. Graph (1B) after transitive reduction, or the equivalent bidirected
+string graph to (2A)
+
+\\ \hline
+
+\includegraphics[scale=0.6]{out.reduced.collapsed.digraph-crop.pdf}
+\newline 3A. Graph (2A) after collapsing unbranched paths
+&
+\includegraphics[scale=0.6]{out.reduced.collapsed.bidigraph-crop.pdf}
+\newline 3B. Graph (2B) after collapsing unbranched paths, or the equivalent
+bidirected string graph to (3A).
+
+\\ \hline
+
+\end{tabular}
+\newpage
+
+%\begin{minipage}[c]{0.4\linewidth}
+%\includegraphics[scale=0.6]{out.digraph-crop.pdf}
+%\captionof{figure}{Directed string graph of original data}
+%\end{minipage}
+%\begin{minipage}[c]{0.4\linewidth}
+%\includegraphics[scale=0.6]{out.bidigraph-crop.pdf}
+%\captionof{figure}{Equivalent bidirected string graph of original data}
+%\end{minipage}
+
+%\hrule
+
+%\begin{minipage}{0.5\linewidth}
+%\includegraphics[scale=0.6]{out.reduced.digraph-crop.pdf}
+%\end{minipage}
+%\begin{minipage}{0.5\linewidth}
+%\includegraphics[scale=0.6]{out.reduced.bidigraph-crop.pdf}
+%\end{minipage}
+
+%\hrule
+
+%\begin{minipage}{0.5\linewidth}
+%\includegraphics[scale=0.6]{out.reduced.collapsed.digraph-crop.pdf}
+%\end{minipage}
+%\begin{minipage}{0.5\linewidth}
+%\includegraphics[scale=0.6]{out.reduced.collapsed.bidigraph-crop.pdf}
+%\end{minipage}
 

### 2012-11-10 checkin/project-checkin.tex
checkin

@@
 process will iteratively transform the data (including the graph) on-disk into
-the final assembly.  The currently implemented programs are:
+the final assembly.  The main currently implemented programs are:
 
 \begin{itemize}
-\item
-	{\tt convert-reads}:  Imports and merges a set of reads
-	into a concise binary representation.
+%\item
+	%{\tt convert-reads}:  Imports and merges a set of reads
+	%into a concise binary representation.
 \item
@@
 
-\item
-	{\tt print-overlaps}:  Prints a textual representation of a set of
-	overlaps, given a file containing the overlaps in binary format.
+%\item
+	%{\tt print-overlaps}:  Prints a textual representation of a set of
+	%overlaps, given a file containing the overlaps in binary format.
 
@@
 	them, along with the corresponding overlaps.  If a pair of reads is
-	identicaly, only one copy of the read is kept.
+	identical, only one copy of the read is kept.
 
@@
 	Given a set of reads and the overlaps that were computed from them,
-	build the {\bf directed} string graph that models the genome assembly.
+	build the {\em directed} string graph that models the genome assembly.
 \item
@@
 	Given a set of reads and the overlaps that were computed from them,
-	build the {\bf bidirected} string graph that models the genome assembly.
+	build the {\em bidirected} string graph that models the genome assembly.
 
-\item
-	{\tt print-string-graph}:
-	Prints a textual representation of a directed or bidirected string
-	graph, or prints statistics about the graph.
+%\item
+	%{\tt print-string-graph}:
+	%Prints a textual representation of a directed or bidirected string
+	%graph, or prints statistics about the graph.
 
@@
 	\to w \to x$, provided that $v \to x$ is labeled by the same sequence as
-	that of $v \to w$ concatenated with $v \to x$.  This program currently
-	only works on directed string graphs, but the bidirected case is in
-	progress.
+	that of $v \to w$ concatenated with $w \to x$. 
+	%This program currently
+	%only works on directed string graphs, but the bidirected case is in
+	%progress.
 
@@
 	Given a string graph, collapse chains of vertices that have in-degree 1
-	and out-degree 1.  This program currently only works on directed string
-	graphs, but the bidirected case is in progress.
+	and out-degree 1.
+	%This program currently only works on directed string
+	%graphs, but the bidirected case is in progress.
 
-\item
-	{\tt digraph-to-bidigraph}:
-	Convert a directed string graph into a bidirected string graph.
+%\item
+	%{\tt digraph-to-bidigraph}:
+	%Convert a directed string graph into a bidirected string graph.
 
@@
 The modular design allows for multiple possible data flows, but a possible flow
-is {\tt convert-reads} $\to$ {\tt compute-overlaps} $\to$ {\tt
-remove-contained-reads} $\to$ \newline {\tt build-directed-string-graph} $\to$ {\tt
-transitive-reduction} $\to$ {\tt digraph-to-bidigraph}, which will produce a
+is
+%{\tt convert-reads} $\to$
+{\tt compute-overlaps} $\to$ {\tt
+remove-contained-reads} $\to$ {\tt build-directed-string-graph} $\to$ {\tt
+transitive-reduction} $\to$ {\tt collapse-unbranched-paths} $\to$ {\tt digraph-to-bidigraph}, which will produce a
 transitively-reduced, collapsed bidirected string graph produced from overlaps
@@
 
-Even with the final stages not implemented, the algorithm is already effective
-when used on random genomes because with no spurious overlaps, the
+Even with the final stages not implemented, the algorithm is already very
+effective when used on completely random genomes because, since there are no
+spurious overlaps in a completely random genome with adequate length reads, the
 transitively-reduced string graph is a chain of vertices with no branching,
-which then collapses into a single edge in {\tt collapse-unbranched-paths}.
-This is true even if the genome is 15 million random base pairs, given 100 bp
-reads sampled uniformly from either strand with no errors.  (Such a genome takes
-about 1 minute to assemble with maximum memory usage 1.8 GB, with the main time
-and memory bottleneck being the {\tt compute-overlaps} program).
+which then collapses into a single edge that will be labeled with almost all the
+original sequence.
+%This is true even if the genome is 15 million random base pairs, given 100 bp
+%reads sampled uniformly from either strand with no errors.  (Such a genome
+%takes about 1 minute to assemble with maximum memory usage 1.8 GB, with the
+%main time and memory bottleneck being the {\tt compute-overlaps} program).
+However, genomes with repeats will produce tangled graphs that will require the
+network flow and Eulerian path parts of the overall algorithm to be implemented.
 
 What follows is an explanation of the initial stages of the algorithm on a very
-simple case.  The diagram below shows a 64- base-pair ``genome'' that has had
-three 32-base-pair error-free reads at a constant separation of 12 base pairs.
-Note that the genome is two-stranded.  {\tt A} always pairs with {\tt T} and
-{\tt C} always pairs with {\tt G}, but the two strands run in opposite
-directions.  Reads 1 and 2 come from the top (reverse strand), while read 3
-comes from the bottom (forward strand).  The reads are directed as shown (e.g.
-Read 1 begins with {\tt ACAGC}, reading from right to left, not {\tt TGAGC}).
+simple case.  The diagram below shows a 64-base-pair ``genome'' that has had
+three 32-base-pair error-free reads taken from it.  Note that the genome is
+two-stranded.  {\tt A} always pairs with {\tt T} and {\tt C} always pairs with
+{\tt G}, but the two strands run in opposite directions.  Reads 1 and 2 come
+from the top (reverse strand), while read 3 comes from the bottom (forward
+strand).  The reads are directed as shown (e.g.  read 1 begins with {\tt ACAGC},
+reading from right to left, not {\tt TGAGC}).
 
@@
 \hspace*{5.16cm}{\tt \color{blue} $\rightarrow$TCTGCTGATAATATCGCGCCGGTTCTGATGCG$\rightarrow$}
+\vspace{0.5cm}
 
-From looking at the diagram, it is apparent that Read 1 overlaps with Read 2 by
-20 base pairs, Read 1 overlaps with Read 2 by 20 base pairs, and Read 1 overlaps
-with Read 3 by 8 base pairs.  These overlaps are computed by the {\tt
-compute-overlaps} program.  Note that an overlap may match either the forward or
-reverse-complement sequence.
+From looking at the above diagram, it is apparent that read 1 overlaps with read
+2 by 20 base pairs, read 2 overlaps with read 3 by 20 base pairs, and read 1
+overlaps with read 3 by 8 base pairs.  The task of the {\tt compute-overlaps}
+program is to determine these overlaps.  Note that an overlap may match either
+the forward or reverse-complement sequence.
 
-Before building the string graph, {\tt remove-contained-reads} is run to remove reads
-that are fully contained by another read.  There are no such reads in this
-example.
+Before building the string graph, {\tt remove-contained-reads} is run to remove
+reads that are fully contained by another read.  However, there are no such
+reads in this example.
 
-Given the set of overlaps, the next step is to build a bidirected or directed
+Given the set of overlaps, the next step is to build the bidirected or directed
 string graph.  Both these graphs are shown on the next page.  In the directed
-graphs, a node $n.B$ means the beginning of read $n$, while the node $n.E$ means
-the end of read $n$.
+graphs, a node $n.B$ means the beginning of read $n$, while a node $n.E$ means
+the end of read $n$.  In the bidirected graphs, node $n$ simply corresponds to
+read $n$.  In both the directed and bidirected graphs, the edges are labeled by
+DNA sequences; however, in the drawings, only the lengths of the sequences are
+shown.
 
@@
 that assemble together as they are, with no reverse-complementation.  However,
-in this case, we walk from the {\it beginning} of read 1 to the {\it beginning}
+in this case, we walk from the {\em beginning} of read 1 to the {\em beginning}
 of read 2, indicating that they are laid out in reverse-complement order, before
-reaching the {\it end} of read 3, which is in forward order.  This
+reaching the {\em end} of read 3, which is in forward order.  This
 reconstruction is consistent with how the reads were actually sampled from the
@@
 
-In both the directed and bidirected graphs, the edges are labeled by DNA
-sequences, although in the images, only the lengths of the sequences are shown.
+Graphs (3A) and (3B) result from collapsing unbranched paths in (2A) and (2B),
+respectively.  Read 2 is preceded or succeeded only by reads 1 and 3, so its
+vertice(s) can be removed, and the adjacent edges can be squashed together.  As
+mentioned before, this step can theoretically produce an edge labeled with
+nearly the entire genome if the genome is sufficiently short or random, as in
+this case.  However, in the general case, due to genomic repeats, there will be
+many branches in the graph, and this needs to be taken into account by the later
+stages of the algorithm (see the last page for a slightly more complicated
+graph).
 
@@
 \begin{tabular}{|p{0.5\textwidth}|p{0.5\textwidth}|} \hline
+\centering
 \includegraphics[scale=0.6]{out.digraph-crop.pdf}
@@
 &
-\includegraphics[scale=0.6]{out.bidigraph-crop.pdf}
+\centering
+\includegraphics[scale=0.7]{out.bidigraph-crop.pdf}
 \newline 1B. Equivalent bidirected string graph to (1A)
-\\ \hline
+\tabularnewline \hline
 
+\centering
 \includegraphics[scale=0.6]{out.reduced.digraph-crop.pdf}
@@
 &
-\includegraphics[scale=0.6]{out.reduced.bidigraph-crop.pdf}
+\centering
+\includegraphics[scale=0.7]{out.reduced.bidigraph-crop.pdf}
 \newline 2B. Graph (1B) after transitive reduction, or the equivalent bidirected
@@
 
-\\ \hline
+\tabularnewline \hline
 
+\centering
 \includegraphics[scale=0.6]{out.reduced.collapsed.digraph-crop.pdf}
@@
 &
-\includegraphics[scale=0.6]{out.reduced.collapsed.bidigraph-crop.pdf}
+\centering
+\includegraphics[scale=0.7]{out.reduced.collapsed.bidigraph-crop.pdf}
 \newline 3B. Graph (2B) after collapsing unbranched paths, or the equivalent
@@
 
-\\ \hline
+\tabularnewline \hline
 
@@
 
-%\begin{minipage}[c]{0.4\linewidth}
-%\includegraphics[scale=0.6]{out.digraph-crop.pdf}
-%\captionof{figure}{Directed string graph of original data}
-%\end{minipage}
-%\begin{minipage}[c]{0.4\linewidth}
-%\includegraphics[scale=0.6]{out.bidigraph-crop.pdf}
-%\captionof{figure}{Equivalent bidirected string graph of original data}
-%\end{minipage}
+%make out.reduced.collapsed.{bi,}digraph.{pdf,dot} SAMPLE_SIZE=300000
+%USE_PIRS=false MIN_OVERLAP_LEN=50 READ_LEN=100 READ_SEP=30 GENOME=E_coli.fa -B
 
-%\hrule
-
-%\begin{minipage}{0.5\linewidth}
-%\includegraphics[scale=0.6]{out.reduced.digraph-crop.pdf}
-%\end{minipage}
-%\begin{minipage}{0.5\linewidth}
-%\includegraphics[scale=0.6]{out.reduced.bidigraph-crop.pdf}
-%\end{minipage}
-
-%\hrule
-
-%\begin{minipage}{0.5\linewidth}
-%\includegraphics[scale=0.6]{out.reduced.collapsed.digraph-crop.pdf}
-%\end{minipage}
-%\begin{minipage}{0.5\linewidth}
-%\includegraphics[scale=0.6]{out.reduced.collapsed.bidigraph-crop.pdf}
-%\end{minipage}
+\begin{figure}
+\caption{Transitively-reduced, collapsed bidirected string graph for error-free
+100bp reads uniformly sampled from the first 250000 base pairs of the genome of
+{\it E. coli}, which includes some repetitive sequences.  The network flow
+algorithm must be able to determine how many times each edge should be
+traversed, then an Eulerian path through the bidirected graph must be found to
+produce a possible assembly.  However, in the general case, the final
+reconstruction may have to take the form of a set of paths rather than just one
+path.  }
+\includegraphics[scale=0.6]{E_coli-crop.pdf}
+\end{figure}
 

### 2012-11-22 checkin/project-checkin.tex
checkin overdue commits

@@
 given a set of reads that came from it.  The key idea in the algorithm I am
-implementing is the construction of a string graph where the edges are labeled
-with DNA sequences and the vertices correspond to reads, and a path through the
+implementing is the construction of a string graph where the vertices correspond
+to reads and the edges are labeled with DNA sequences, and a path through the
 graph corresponds to a set of reads that assemble together consistently.
@@
 I have implemented the initial stages of the algorithm.  Each step has been
-implemented as a separate binary program.  This means that the overall assembly
-process will iteratively transform the data (including the graph) on-disk into
-the final assembly.  The main currently implemented programs are:
+implemented as a separate executable program.  This means that the overall
+assembly process will iteratively transform the data (e.g. the string graph)
+on-disk into the final assembly.  The main currently implemented programs are:
 
@@
 	Given a set of reads and the overlaps that were computed from them,
-	finds all reads that are fully contained by another read and discard
+	finds all reads that are fully contained by another read and discards
 	them, along with the corresponding overlaps.  If a pair of reads is
@@
 
+\item
+	{\tt map-contained-reads}:
+	Map the reads that were removed in {\tt remove-contained-reads} back
+	into the string graph to mark edges that are supported by more than one
+	read and therefore are more likely to need to be traversed multiple
+	times.
+
 %\item
@@
 remove-contained-reads} $\to$ {\tt build-directed-string-graph} $\to$ {\tt
-transitive-reduction} $\to$ {\tt collapse-unbranched-paths} $\to$ {\tt digraph-to-bidigraph}, which will produce a
+transitive-reduction} $\to$
+{\tt map-contained-reads} $\to$
+{\tt collapse-unbranched-paths} $\to$ {\tt digraph-to-bidigraph}, which will produce a
 transitively-reduced, collapsed bidirected string graph produced from overlaps
@@
 Even with the final stages not implemented, the algorithm is already very
-effective when used on completely random genomes because, since there are no
-spurious overlaps in a completely random genome with adequate length reads, the
+effective when used on completely random genomes.  Since there are no spurious
+overlaps in a completely random genome with adequate length reads, the
 transitively-reduced string graph is a chain of vertices with no branching,
@@
 %main time and memory bottleneck being the {\tt compute-overlaps} program).
+
 However, genomes with repeats will produce tangled graphs that will require the
@@
 What follows is an explanation of the initial stages of the algorithm on a very
-simple case.  The diagram below shows a 64-base-pair ``genome'' that has had
-three 32-base-pair error-free reads taken from it.  Note that the genome is
-two-stranded.  {\tt A} always pairs with {\tt T} and {\tt C} always pairs with
-{\tt G}, but the two strands run in opposite directions.  Reads 1 and 2 come
-from the top (reverse strand), while read 3 comes from the bottom (forward
-strand).  The reads are directed as shown (e.g.  read 1 begins with {\tt ACAGC},
-reading from right to left, not {\tt TGAGC}).
+simple case.  The diagram below shows a 64-base-pair ``genome'' (G.) that has
+had three 32-base-pair error-free reads taken from it (1., 2., and 3.).  Note
+that the genome is two-stranded.  {\tt A} always pairs with {\tt T} and {\tt C}
+always pairs with {\tt G}, but the two strands run in opposite directions.
+Reads 1 and 2 come from the top (``reverse'') strand, while read 3 comes from
+the bottom (``forward'') strand.  The reads are directed as shown (e.g.  read 1
+begins with {\tt ACAGC}, reading from right to left.)
 
@@
 overlaps with read 3 by 8 base pairs.  The task of the {\tt compute-overlaps}
-program is to determine these overlaps.  Note that an overlap may match either
-the forward or reverse-complement sequence.
+program is to determine these overlaps.  Note that reverse-complement overlaps
+(e.g. reads 2 and 3) are permissible.
 
@@
 Given the set of overlaps, the next step is to build the bidirected or directed
-string graph.  Both these graphs are shown on the next page.  In the directed
-graphs, a node $n.B$ means the beginning of read $n$, while a node $n.E$ means
-the end of read $n$.  In the bidirected graphs, node $n$ simply corresponds to
-read $n$.  In both the directed and bidirected graphs, the edges are labeled by
-DNA sequences; however, in the drawings, only the lengths of the sequences are
-shown.
+string graph.  For this example, both these graphs are shown on the next page.
+In the directed graphs, a node $n.B$ means the beginning of read $n$, while a
+node $n.E$ means the end of read $n$.  In the bidirected graphs, node $n$ simply
+corresponds to read $n$.  In both the directed and bidirected graphs, the edges
+are labeled by DNA sequences; however, in the drawings, only the lengths of the
+sequences are shown.
 
@@
 algorithm must be able to determine how many times each edge should be
-traversed, then an Eulerian path through the bidirected graph must be found to
-produce a possible assembly.  However, in the general case, the final
+traversed, then a generalized Eulerian path through the bidirected graph must be
+found to produce a possible assembly.  In the general case, the final
 reconstruction may have to take the form of a set of paths rather than just one
 path.  }
+\begin{center}
 \includegraphics[scale=0.6]{E_coli-crop.pdf}
+\end{center}
 \end{figure}

### 2012-11-27 presentation/presentation.tex
Start presentation.tex

\documentclass[xcolor=dvipsnames]{beamer}

\usepackage{graphicx}
\usepackage{listings}
\usepackage{comment}
\usepackage{float}

\usetheme{Frankfurt}
%\usecolortheme[named=OliveGreen]{structure}
%\usefonttheme{serif}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{frametitle}[default][center]

\begin{document}
\title{A Bidirected String Graph Model of Genome Assembly}
\author{Eric Biggers}
\institute{Macalester College}
\date{December 5, 2012}
%\titlegraphic{\includegraphics[scale=0.3]{pineapple.jpg}}

\frame{\titlepage}

\begin{frame}{Genomes}
	\begin{itemize}
		\item A genome (for our purposes) is a long string of A's, C's, G's, and T's
		\item Dual-stranded
		\begin{itemize}
			\item A pairs with T and C pairs with G
			\item The two strands run in opposite directions
		\end{itemize}
	\end{itemize}
	\begin{center}
	{\tt $\leftarrow$ GTTGCTATAAATACGGGTCATGGTATTTTAGTACCG $\leftarrow$ \\}
	{\tt $\rightarrow$ CAACGATATTTATGCCCAGTACCATAAAATCATGGC $\rightarrow$}
	\end{center}
	\begin{itemize}
		\item Each letter in the genome is also known as a {\bf base pair}
		(abbreviation: bp)
	\end{itemize}
	\begin{table}[H]
		\caption{Genome sizes of selected organisms}
		\begin{tabular}{|c|c|c|}
			\hline
			E. coli (a common bacterium) &  ~4,600,000 bp \\
			\hline
			Ananas comosus (pineapple) &	~500,000,000 bp \\
			\hline
			Homo sapiens (human)       &  ~3,000,000,000 bp \\
			\hline
		\end{tabular}
	\end{table}
\end{frame}

\begin{frame}{Genome Sequencing}
	\begin{itemize}
		\item How can we deteremine the precise sequence (in terms of A's, T's,
		C's, and G's) of an organism's genome?

		\item Currently available sequencing technologies can only sequence
		fragments a few dozen to a few thousand bp long

		\item Genome assembly:
			\begin{itemize}
				\item Lab work
				\begin{enumerate}[1.]
						\item Obtain a DNA sample
						\item Break DNA into much smaller fragments
						\item Sequence the fragments using available technology
				\end{enumerate}
				\item Computational Work
				\begin{enumerate}[4.]
					\item Given the sequenced fragments, reconstruct the original genome.
				\end{enumerate}
			\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Reads}
	\begin{itemize}
		\item A {\bf read} is a short sequence of the letters A, C, T, G that
		represents some DNA that was sequenced by the sequencing machine.
		\item A read may come from any location on the genome, from either
		strand.
		\item Reads from opposite strands go in opposite directions.
			\begin{itemize}
				\item As a result, a sequence and its {\bf reverse-complement}
				(the corresponding sequence on the other strand, going
				backwards) are usually considered equivalent.
			\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Genome assemblers}
	\begin{itemize}
		\item An {\bf assembler} is a program or algorithm that can put together the
		``puzzle'' to reconstruct the original genome.
		\item  Input: a set of reads (lengths: ~35 to 5000bp each).
		\item  Output: one strand of the reconstructed genome (if possible)--- but
		in general, an exact answer may be impossible, so the output must be a set
		of substrings of the original genome that are as long as possible. (lengths:
		~10,000 to 10,000,000bp each).
	\end{itemize}
\end{frame}

\begin{frame}{Genome assembly vs Shortest Common Supersequence}
	Genome assembly is superficially similar to the shortest common superstring
	(SCS) problem, which is to find the shortest string that contains all strings
	from a given set.
	\begin{center}
	\includegraphics[scale=0.6]{venn-crop.pdf}
	\end{center}
\end{frame}

\begin{frame}{Ideas for genome assembly algorithms}

	\begin{itemize}
		\item Use the reads to build a graph that models the assembly problem.
		\item Find paths through the graph to reconstruct the original sequence.
	\end{itemize}
\end{frame}

\begin{frame}{Overlaps between reads}

	\begin{itemize}
	\item Two reads that overlap are likely to come from adjacent positions on the genome.
	\end{itemize}

	{\tt Read 1: \ \ \ \ \ \ \ \ \ $\rightarrow$ ATATAT\textcolor{red}{GCTGGTTACTT} $\rightarrow$ } \\
	{\tt Read 2: \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 
		$\rightarrow$ \textcolor{red}{GCTGGTTACTT}TGATAGATA $\rightarrow$ } \\
	{\tt Possible genome: $\rightarrow$ ATATATGCTGGTTACTTTGATAGATA $\rightarrow$}

	\begin{itemize}
	\item Overlapped reads may be from different strands.  Note the different
	directions of the two reads here.  The overlapped regions of the two reads
	are {\bf reverse-complement} from each other.
	\end{itemize}

	{\tt Read 1: \ \ \ \ \ \ \ \ \ 
			$\rightarrow$ ATATAT\textcolor{red}{GCTGGTTACTT} $\rightarrow$ } \\
	{\tt Read 2: \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 
			$\leftarrow$ \textcolor{red}{CGACCAATGAA}ACTATCTAT $\leftarrow$ } \\
	{\tt Possible genome: $\rightarrow$ ATATATGCTGGTTACTTTGATAGATA $\rightarrow$}
\end{frame}

\begin{frame}{Computing overlaps}
	\begin{itemize}
		\item Compute all pairwise overlaps of some minimum length $\ell$ among the
		reads.
		\item Naive algorithm compares every read with every other read ($N$
		reads: $N^2$ time)
		\item A faster algorithm indexes the reads by short subsequences, then only
		compares reads that share a subsequence ("seed-and-extend").
		\item Overlaps may be exact, or they may allow for differences up to a
		certain threshold.
		\item Implementation: {\tt compute-overlaps}
			\begin{itemize}
				\item Input:  A set of reads and the minimum overlap length
				\item Output:  All exact pairwise overlaps between reads,
				including reverse-complement overlaps
				\item Algorithm:  Seed-and-extend
			\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Build the fragment string assembly graph}
	\begin{itemize}

		\item Use the overlaps as evidence to construct a graph modeling the assembly.
		\item Each vertex represents a read.
		\item Each edge represents an overlap.
		\item Two reads are connected with an edge iff they share an overlap.
		\item A {\it string graph} is a graph where the edges are labeled with
		strings.
		\item In this case, each edge is labeled with DNA sequence of the {\em
		non-overlapped region} (this will be explained later)
		\item Should the edges be directed or undirected?
	\end{itemize}
\end{frame}

10. Bidirected graphs

- At this stage, we don't actually know which strands the reads are from!

------------>            or             <------------
     <-------------             -------------->

- We must allow the edges to be traversed in either the forward or
  reverse-complement direction... but each read MUST be used in a consistent way
  in the reconstruction.

- A _bidirected graph_ is a graph where a directed head is attached to both
  edges.

- There are 3 (or 4, ignoring symmetry) types of bidirected edges.  They follow
  directly from the different types of overlaps.

     ------->   ------>      ------>     <------
  <-------         <-------     ------>     <------

      |               |           |          |
      V               V           V          V

  <--------->   >---------<  >-------->  <--------<

- In this example, two vertices (reads) may have multiple _bidirected edges_
  between them if they overlap in different ways, but there can be at most one
  _bidirected edge of the same type_ between two vertices if we always prefer
  the best overlap of a given type.

\begin{comment}

11. Walking through the bidirected string graph

- A _walk on a bidirected graph_ is a continuous sequence of edges such that
  if we enter a vertex v through a head inwards, we leave it on a head outwords,
  and vice versa.

- A bidirected edge may be traversed in opposite directions.

- The reverse of a walk in a directed graph is also a bidirected walk.

    - Interpretation: one direction of the walk spells out the _forward_
      sequence.  The other direction spells out the _reverse complement_
      sequence.

12. Transitive reduction

- Very commonly, given three adjacent reads f, g, and h, f will overlap h as
  well as g.  This is redundant information because we can equivalently walk f
  ?-? g ?-? h, but the walks visiting more vertices are preferred because they are
  supported by more reads.

------------->
     -------------->
         --------------->

- Transitive reduction finds edges f ?-? h, described above, and removes them.
- Algorithm:  Go through each vertex and examine neighbors up to 2 steps away to
  identify all transitive edges leaving this vertex.  (Some optimizations are
  possible.)

13. Collapsing unbranched paths

Find unbranched paths and collapse them.

f >-> g >-> h >-> i >-> j >-> k

f >-< g <-> h >-< i <-< j <-> k

Ideally, most the genome will be unbranched sequence.  These sequences will be
assembled after this step of the algorithm.

But, real genomes contain repeats, and they will create branches in the graph.

14. Traversal count calculation

Estimate the traversal count of each edge.

Ideally, every edge will be traversed exactly one time.
- But, real genomes always contain repetitive sequence, and these regions may
appear multiple times in the final assembly.
- It is also possible for there to be erroneous parts of the graph that should
  not be included in the final assembly at all.

A-statistic formula

15. Minimum cost network flow

We need to determine how many times edge must be traversed.

Remember, some edges may be repeated sequence, and they therefore may appear
multiple times in the original genome.

Given the bidirected string graph with each edge labeled as OPTIONAL, EXACTLY
ONCE, or AT LEAST ONCE, compute a minimum-cost network flow that makes the graph
Eulerian.

16. Compute an Eulerian tour of the graph to produce a reconstruction of the
original genome.

A generalized Eulerian tour through the bidirected graph will produce a possible
reconstruction of the original genome.

- General case:  The original genome cannot be reconstructed in one piece, so
  the final reconstruction must take the form of the smallest set of paths that
  traverse all the edges the required number of times.

17. Implementation:

- A set of C++ programs that iteratively transform the data into the final
  assembly.
- Different sets of stages of the algorithm may be run in order to reach the
  final assembly.

Each node is a binary program.

The arrows indicate the output from one program being used in other program.

            convert-reads

                |
                V
                ...

Reads are input in textual form, e.g. the FASTA format:

>read_1
ATATTATATAT
>read_2
GCGCGTGTGTA

Genome is output in textual form.

A _contig_

>contig_1
TTTATAGTAGTAGTTGATTAGTATAGTGATGATAGTAGTA...

- Many things still would need to be improved to make this competitive with
  other assemblers.

18. References

- The Fragment String Assembly Graph (2005) by Eugene W. Myers
    * Main algorithm I have described.

- Maximum Likelihood Genome Assembly (2009) by Paul Medvedev and Michael Brudno
    * Explanation of bidirected graphs and bidirected graph algorithms applied
    * to DNA sequences.
\end{comment}

\end{document}

### 2012-11-28 presentation/presentation.tex
Presentation work

@@
 \usepackage{float}
+\usepackage{tikz}
+\usepackage{algorithm}
+\usepackage[noend]{algorithmic}
+
+\usetikzlibrary{arrows}
 
@@
 \date{December 5, 2012}
-%\titlegraphic{\includegraphics[scale=0.3]{pineapple.jpg}}
+%\titlegraphic{\includegraphics[width=\textwidth,clip=true,trim=0 300 0 500]{puzzle.jpg}}
 
@@
 		\item A genome (for our purposes) is a long string of A's, C's, G's, and T's
-		\item Dual-stranded
+		\item Genomes are dual-stranded
 		\begin{itemize}
-			\item A pairs with T and C pairs with G
+			\item {\tt A} always pairs with {\tt T} and {\tt C} always pairs
+			with {\tt G}
 			\item The two strands run in opposite directions
@@
 	\begin{itemize}
-		\item Each letter in the genome is also known as a {\bf base pair}
-		(abbreviation: bp)
+		\item Each letter in the genome is called {\bf base pair} \\(abbreviation:
+		{\bf bp})
 	\end{itemize}
 	\begin{table}[H]
-		\caption{Genome sizes of selected organisms}
+		\caption{Approximate genome sizes of selected organisms}
 		\begin{tabular}{|c|c|c|}
 			\hline
-			E. coli (a common bacterium) &  ~4,600,000 bp \\
+			{\it E. coli} (a bacterium) &  ~4,600,000 bp \\
 			\hline
-			Ananas comosus (pineapple) &	~500,000,000 bp \\
+			{\it Ananas comosus} (pineapple plant) &	~500,000,000 bp \\
 			\hline
-			Homo sapiens (human)       &  ~3,000,000,000 bp \\
+			{\it Homo sapiens} (human)       &  ~3,000,000,000 bp \\
 			\hline
@@
 	\begin{itemize}
-		\item How can we deteremine the precise sequence (in terms of A's, T's,
-		C's, and G's) of an organism's genome?
+		\item How can we determine the precise base-pair sequence of an
+		organism's genome?
 
@@
 	\begin{itemize}
-		\item A {\bf read} is a short sequence of the letters A, C, T, G that
-		represents some DNA that was sequenced by the sequencing machine.
+		\item A {\bf read} is a short sequence of base pairs that represents
+		some DNA that was sequenced by the sequencing machine.
 		\item A read may come from any location on the genome, from either
@@
 	\end{itemize}
+		{\scriptsize
+			{\tt 1.}
+			{\tt \color{red} $\leftarrow$TGAGCTTAAGGCTTATCTATCTTCAGACGACA$\leftarrow$ }\\
+			{\tt 2.}
+			\hspace*{1.76cm}{\tt \color{red} $\leftarrow$TTATCTATCTTCAGACGACTATTATAGCGCGG$\leftarrow$}\\
+			{\tt G.}
+			{\tt $\leftarrow$TGAGCTTAAGGCTTATCTATCTTCAGACGACTATTATAGCGCGGCCAAGACTACGCGGAGCCCC$\leftarrow$} \\
+			{\tt G.}
+			{\tt $\rightarrow$ACTCGAATTCCGAATAGATAGAAGTCTGCTGATAATATCGCGCCGGTTCTGATGCGCCTCGGGG$\rightarrow$} \\
+			{\tt 3.} \hspace*{3.55cm}{\tt \color{blue} $\rightarrow$TCTGCTGATAATATCGCGCCGGTTCTGATGCG$\rightarrow$}
+			\vspace{0.5cm}
+		}
+		\\
+		{\footnotesize Above: a very short genome (G.) and three reads that came from it (1.,
+		2., 3.).}
 \end{frame}
@@
 	\begin{itemize}
-		\item An {\bf assembler} is a program or algorithm that can put together the
-		``puzzle'' to reconstruct the original genome.
+		\item An {\bf assembler} is a program or algorithm that can put together
+		a set of reads to reconstruct the original genome.  (Think of putting
+		together a puzzle $\dots$)
 		\item  Input: a set of reads (lengths: ~35 to 5000bp each).
 		\item  Output: one strand of the reconstructed genome (if possible)--- but
-		in general, an exact answer may be impossible, so the output must be a set
-		of substrings of the original genome that are as long as possible. (lengths:
-		~10,000 to 10,000,000bp each).
+		in general, an exact answer may be impossible, so the actual output is a
+		set of substrings of the original genome that are as long as possible.
+		(lengths: ~10,000 to 10,000,000bp each).
 	\end{itemize}
@@
 
-\begin{frame}{Genome assembly vs Shortest Common Supersequence}
+\begin{frame}{Genome assembly vs Shortest Common Substring}
 	Genome assembly is superficially similar to the shortest common superstring
@@
 
-\begin{frame}{Ideas for genome assembly algorithms}
+\begin{frame}{An algorithm for genome assembly}
 
 	\begin{itemize}
-		\item Use the reads to build a graph that models the assembly problem.
-		\item Find paths through the graph to reconstruct the original sequence.
+		\item {\it The fragment string assembly graph} (Myers, 2005)
+		\item Draws on ideas used in previous assemblers and also introduces new
+		ideas
+		\item The fundamental algorithm:
+		\begin{enumerate}
+			\item Use overlaps between reads to build a graph that models the
+			assembly problem.
+			\item Simplify and analyze the graph.
+			\item Find paths through the graph to reconstruct the original
+			sequence.
+		\end{enumerate}
 	\end{itemize}
+\end{frame}
+
+\begin{frame}{Overlaps between reads}
+	Consider two reads $f$ and $g$, and their reverse complements $f'$ and $g'$.
+	The reads $f$ and $g$ are said to {\bf exactly overlap} by some length
+	$\ell$ if and only if at least one of the following is true:
+	\begin{itemize}
+		\item The last $\ell$ bp of $f$ exactly match the first $\ell$ bp of
+		$g$.
+		\item The last $\ell$ bp of $f$ exactly match the first $\ell$ bp of
+		$g'$.
+		\item The last $\ell$ bp of $f'$ exactly match the first $\ell$ bp of
+		$g$.
+		\item The last $\ell$ bp of $f'$ exactly match the first $\ell$ bp of
+		$g'$.
+	\end{itemize}
+	An exact overlap $o$ is fully described by the tuple $(f, g, f_{beg},
+	g_{beg}, len, rc)$.
+	%there $f$ and $g$ are the reads of the overlap,
+	%$f_{beg}$ and $g_{beg}$ and the starting position of the overlap on each
+	%read, $len$ is the length of the overlap, and $rc$ indicates whether the
+	%overlap is reverse-complement or not.
 \end{frame}
@@
 	\begin{itemize}
-	\item Overlapped reads may be from different strands.  Note the different
-	directions of the two reads here.  The overlapped regions of the two reads
-	are {\bf reverse-complement} from each other.
+	\item Overlapping reads may be from different strands, in which case the
+	overlapped regions are reverse-complement from each other.
 	\end{itemize}
@@
 
+\begin{frame}{Three types of overlaps}
+	\begin{enumerate}
+		\item ``normal'' overlap
+			\begin{center}
+				\begin{tikzpicture}
+					\draw[->,style=thick] (0,0) -- (2,0);
+					\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
+				\end{tikzpicture}
+
+				\vspace{3mm}
+				or (symmetrically)
+				\vspace{3mm}
+
+				\begin{tikzpicture}
+					\draw[->,style=thick] (2,0) -- (0,0);
+					\draw[->,style=thick] (3.0,0.3) -- (1.0,0.3);
+				\end{tikzpicture}
+			\end{center}
+		\item ``innie'' overlap
+			\begin{center} \begin{tikzpicture}
+					\draw[->,style=thick] (0,0) -- (2,0);
+					\draw[->,style=thick] (3,0.3) -- (1,0.3);
+			\end{tikzpicture} \end{center}
+		\item ``outtie'' overlap
+			\begin{center} \begin{tikzpicture}
+					\draw[->,style=thick] (2,0) -- (0,0);
+					\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
+			\end{tikzpicture} \end{center}
+	\end{enumerate}
+\end{frame}
+
 \begin{frame}{Computing overlaps}
 	\begin{itemize}
-		\item Compute all pairwise overlaps of some minimum length $\ell$ among the
-		reads.
+		\item Goal: compute all pairwise overlaps of some minimum length $\ell$
+		among the reads.
 		\item Naive algorithm compares every read with every other read ($N$
@@
 		certain threshold.
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Computing overlaps (Implementation)}
+	\begin{itemize}
 		\item Implementation: {\tt compute-overlaps}
 			\begin{itemize}
-				\item Input:  A set of reads and the minimum overlap length
-				\item Output:  All exact pairwise overlaps between reads,
+				\item {\bf Input}:  A set of reads and the minimum overlap
+				length $\ell$
+				\item {\bf Output}:  All exact pairwise overlaps between reads,
 				including reverse-complement overlaps
-				\item Algorithm:  Seed-and-extend
+				\item {\bf Algorithm}:  Seed-and-extend
+				\item {\bf Running time}: Linear with regards to genome length for
+				non-repetitive genomes; quadratic with regards to read density;
+				quadratic with regards to the number of times a sequence of DNA
+				is repeated in the genome
 			\end{itemize}
 	\end{itemize}
+	%\begin{algorithm}[H]
+		%\caption{Algorithm {\sc compute-overlaps}($S$)}
+		%\begin{algorithmic}
+			%%\STATE{h = new hash table}
+			%%\FOR{{\bf each} read $r \in S$}
+				%%\FOR{{\bf each} canonical $k$-mer $K \in R$}
+					%%\STATE{Append $K.pos$ to $h[K]$}
+				%%\ENDFOR
+			%%\ENDFOR
+			%%\FORALL{posList $\in h$}
+				%%\FORALL{$k$-mer pairs $k1$, $k2$ in 
+			%%\ENDFOR
+		%\end{algorithmic}
+	%\end{algorithm}
 \end{frame}
@@
 
-		\item Use the overlaps as evidence to construct a graph modeling the assembly.
-		\item Each vertex represents a read.
-		\item Each edge represents an overlap.
+		\item Use the overlaps to construct a graph modeling how the reads could
+		be put together
+		\item Make each read be a vertex
+		\item Make each edge be an overlap
+		\begin{itemize}
 			\item Two reads are connected with an edge iff they share an overlap.
-		\item A {\it string graph} is a graph where the edges are labeled with
-		strings.
-		\item In this case, each edge is labeled with DNA sequence of the {\em
-		non-overlapped region} (this will be explained later)
-		\item Should the edges be directed or undirected?
+		\end{itemize}
+		\item Label each edge with an appropriate DNA sequence; this makes the
+		graph a {\bf string graph}.
+		\item Should the edges be directed or undirected?  \\Or $\dots$ something
+		else?
 	\end{itemize}
@@
 
-10. Bidirected graphs
+\begin{frame}{Build the fragment string assembly graph (cont.)}
+	\begin{itemize}
+		\item At this stage, we don't know which strands the reads are from!
 
-- At this stage, we don't actually know which strands the reads are from!
+		\begin{center}
+			\begin{tikzpicture}
+				\draw[->,style=thick] (0,0) -- (2,0);
+				\draw[->,style=thick] (3,0.3) -- (1,0.3);
+			\end{tikzpicture}
+			\hspace{5mm}or\hspace{5mm}
+			\begin{tikzpicture}
+				\draw[->,style=thick] (0,0.3) -- (2,0.3);
+				\draw[->,style=thick] (3, 0) -- (1, 0);
+			\end{tikzpicture}
+			\hspace{5mm} ?
+		\end{center}
 
------------->            or             <------------
-     <-------------             -------------->
+		\item We must allow each read to be used either in the
+		reverse-complement or forward direction.
+		\item Therefore, we must allow the edges to be traversed in either
+		direction.
+		\item However, each read must still be used in a consistent way within
+		any single reconstruction.
+		\item Solution: use a {\bf bidirected graph}.
+	\end{itemize}
+\end{frame}
 
-- We must allow the edges to be traversed in either the forward or
-  reverse-complement direction... but each read MUST be used in a consistent way
-  in the reconstruction.
+\begin{frame}{Bidirected Graphs}
+	\begin{itemize}
+		\item A {\bf bidirected graph} is a graph where a directed head is
+		attached to both ends of each edge.
 
-- A _bidirected graph_ is a graph where a directed head is attached to both
-  edges.
+		\item There are 3 (or 4, ignoring symmetry) types of {\bf bidirected
+		edges}.  They follow directly from the different types of overlaps.
 
-- There are 3 (or 4, ignoring symmetry) types of bidirected edges.  They follow
-  directly from the different types of overlaps.
+		\begin{center}
+			\begin{tabular}{p{1cm}cccc}
+				{\footnotesize Overlap} &
+				\begin{tikzpicture}[scale=0.6]
+						\draw[->,style=thick] (2,0) -- (0,0);
+						\draw[->,style=thick] (1,0.3) -- (3,0.3);
+				\end{tikzpicture} 
+				& 
+				\begin{tikzpicture}[scale=0.6]
+						\draw[->,style=thick] (0,0) -- (2,0);
+						\draw[->,style=thick] (3,0.3) -- (1,0.3);
+				\end{tikzpicture} 
+				& 
+				\begin{tikzpicture}[scale=0.6]
+						\draw[->,style=thick] (0,0) -- (2,0);
+						\draw[->,style=thick] (1,0.3) -- (3,0.3);
+				\end{tikzpicture} 
+				&
+				\begin{tikzpicture}[scale=0.6]
+						\draw[->,style=thick] (2,0) -- (0,0);
+						\draw[->,style=thick] (3,0.3) -- (1,0.3);
+				\end{tikzpicture} 
+				\\
+				&$\downarrow$ & $\downarrow$& $\downarrow$& $\downarrow$ \\
+				{\footnotesize Bidirected edge} &
+				\begin{tikzpicture}[scale=0.6,>=triangle 45]
+						\draw[<->,style=thick] (0,0) -- (2,0);
+				\end{tikzpicture} 
+				& 
+				\begin{tikzpicture}[scale=0.6,>=triangle 45]
+						\draw[>-<,style=thick] (0,0) -- (2,0);
+				\end{tikzpicture} 
+				& 
+				\begin{tikzpicture}[scale=0.6,>=triangle 45]
+						\draw[>->,style=thick] (0,0) -- (2,0);
+				\end{tikzpicture} 
+				&
+				\begin{tikzpicture}[scale=0.6,>=triangle 45]
+						\draw[<-<,style=thick] (0,0) -- (2,0);
+				\end{tikzpicture} 
+			\end{tabular}
+		\end{center}
 
-     ------->   ------>      ------>     <------
-  <-------         <-------     ------>     <------
+		%\item Two vertices (reads) may have multiple {\bf bidirected edges}
+		%between them if they overlap in different ways, but there can be at most
+		%one bidirected edge {\em of the same type} between two vertices if we
+		%always prefer the best overlap of a given type.
+	\end{itemize}
+\end{frame}
 
-      |               |           |          |
-      V               V           V          V
+\begin{frame}{Walking through the bidirected string graph}
+	\begin{itemize}
+		\item A {\bf walk on a bidirected graph} is a continuous sequence of
+		edges such that if any vertex $v$ is entered through a head inwards, it
+		is left on a head outwards (unless it is the end of the path), and vice
+		versa.
 
-  <--------->   >---------<  >-------->  <--------<
+		\item A bidirected edge may be traversed in opposite directions during
+		the same walk.
 
-- In this example, two vertices (reads) may have multiple _bidirected edges_
-  between them if they overlap in different ways, but there can be at most one
-  _bidirected edge of the same type_ between two vertices if we always prefer
-  the best overlap of a given type.
+		\item The reverse of a walk in a directed graph is also a bidirected walk.
+
+		\item Interpretation: one direction of the walk spells out the {\bf
+		forward sequence}.  The other direction spells out the {\bf reverse
+		complement sequence.}
+	\end{itemize}
+\end{frame}
+
+\newcommand{\BidirectedEdgeForward}[2]{
+		\tikz[>=triangle 45,baseline]
+		\draw[>->,thin] (0,0.1) node[anchor=east] {#1} --
+					    (0.8,0.1) node[anchor=west] {#2};}
+
+\begin{frame}{Transitive reduction}
+	\begin{itemize}
+		\item Very commonly, given three adjacent reads $f$, $g$, and $h$, $f$ will overlap
+		$h$ as well as $g$.  This is redundant information because we can equivalently
+		walk $f$ to $g$ to $h$ (or $h$ to $g$ to $f$, if going the other way).
+
+		\begin{center}
+			\begin{tikzpicture}
+				\draw[->,style=very thick] (0,0) node[anchor=east] {f} -- (2,0);
+				\draw[->,style=very thick] (0.5,0.25) node[anchor=east] {g} -- (2.5,0.25);
+				\draw[->,style=very thick] (1.0,0.5) node[anchor=east] {h} -- (3,0.5);
+			\end{tikzpicture}
+			\begin{tikzpicture}[>=triangle 45]
+				\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+				\node (f) at (0, 0) {f};
+				\node (g) at (1.5, 0) {g};
+				\node (h) at (3, 0) {h};
+				\draw[>->,style=very thick] (f) edge (g);
+				\draw[>->,style=very thick] (g) edge (h);
+				\draw[>->,style=very thick,bend left] (f) edge (h);
+			\end{tikzpicture}
+		\end{center}
+
+		\item Walks visiting more vertices are preferred because they are
+		supported by more reads, and the overlaps tend to be longer.  So, we
+		would like to throw away \BidirectedEdgeForward{$f$}{$h$} but keep
+		\BidirectedEdgeForward{$f$}{$g$} and \BidirectedEdgeForward{$g$}{$h$}.
+
+		\item To remove these redundant edges, {\bf transitive reduction} is
+		performed on the bidirected string graph.
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Transitive reduction (cont.)}
+		Transitive reduction algorithm:  Go through each vertex and examine
+		neighbors up to 2 steps away to identify all transitive edges leaving this
+		vertex.
+	\begin{algorithm}[H]
+		\caption{{\sc TransitiveReduction}($G$)}
+	{\footnotesize
+		\begin{algorithmic}
+			\FOR{{\bf each} vertex $v \in G$}
+				\STATE{Mark all neighbors of $v$ as {\tt INPLAY}.}
+				\STATE{Set {\tt longest} $\gets$ length of the longest edge leaving $v$}
+				\FOR{{\bf each} edge $v \to w$ in order of length}
+					\FOR{{\bf each} edge $w \to x$ in order of length}
+						\IF{$x$ is not marked {\tt INPLAY}}
+							\STATE{\bf continue}
+						\ENDIF
+						\IF{$len(v \to w) + len(w \to x) \ge$ {\tt longest}}
+							\STATE{\bf break}
+						\ENDIF
+						\STATE{Mark $x$ as {\tt ELIMINATED}}
+					\ENDFOR
+				\ENDFOR
+				\STATE{Eliminate all edges $v \to w$ where $w$ is marked {\tt ELIMINATED}}
+			\ENDFOR
+			\STATE{Eliminate all edges marked {\tt ELIMINATED}}
+		\end{algorithmic}
+	}
+	\end{algorithm}
+\end{frame}
+
+\begin{frame}{Collapsing unbranched paths}
+
+Vertex 4 below is an {\bf inner vertex} that is part of an unbranched path that
+can be collapsed.
+
+\begin{center}
+	{\small
+	\begin{tikzpicture}[>=triangle 45]
+		\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+		\node (1) at (0.0, -1) {1};
+		\node (2) at (0.0, 1) {2};
+		\node (3) at (1.5, 0) {3};
+		\node[fill=green!25] (4) at (3.25, 0) {4};
+		\node (5) at (5.0, 0) {5};
+		\node (6) at (6.5, 1) {6};
+		\node (7) at (6.5, -1) {7};
+		\draw[>->,very thick] (1) edge (3);
+		\draw[>->,very thick] (2) edge (3);
+		\draw[>->,very thick] (3) edge (4);
+		\draw[>->,very thick] (4) edge (5);
+		\draw[>->,very thick] (5) edge (6);
+		\draw[>->,very thick] (5) edge (7);
+	\end{tikzpicture}
+	}
+
+becomes:
+
+	{\small
+	\begin{tikzpicture}[>=triangle 45]
+		\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+		\node (1) at (0.0, -1) {1};
+		\node (2) at (0.0, 1) {2};
+		\node (3) at (1.5, 0) {3};
+		\node (5) at (3.25, 0) {5};
+		\node (6) at (4.75, 1) {6};
+		\node (7) at (4.75, -1) {7};
+		\draw[>->,very thick] (1) edge (3);
+		\draw[>->,very thick] (2) edge (3);
+		\draw[>->,very thick] (3) edge (5);
+		\draw[>->,very thick] (5) edge (6);
+		\draw[>->,very thick] (5) edge (7);
+	\end{tikzpicture}
+	}
+\end{center}
+\end{frame}
+
+\begin{frame}{Current state of the graph}
+	\begin{itemize}
+		\item Non-repetitive genomes and/or long reads: The assembly probably is
+		already complete!
+	\end{itemize}
+
+	\vspace{1cm}
+
+	\begin{figure}[H]
+		\begin{tikzpicture}
+			{\small
+			\begin{tikzpicture}[>=triangle 45]
+				\node[circle,fill=blue!20,draw=black] (A) at (0.0, 0) {A};
+				\node[circle,fill=blue!20,draw=black] (B) at (3.0, 0) {B};
+				\draw[>->,very thick] (A) edge node[anchor=north] {4639100} (B);
+			\end{tikzpicture} }
+		\end{tikzpicture}
+		\caption{
+			Bidirected string graph for reads uniformly sampled from a 4.6 Mb
+			random genome after transitive reduction and path collapsing.  (Edge
+			labeled with length of sequence)}
+	\end{figure}
+\end{frame}
+
+\begin{frame}{Current state of the graph (cont.)}
+	\begin{itemize}
+		\item Repetitive genomes and/or short reads: The graph will still be
+		very complicated.
+		\item Swap out that 4,600,000 with {\it E. coli}'s genome, which
+		contains some repeats, and we get a complicated graph, {\bf even after
+		increasing the read length to 1000bp}:
+	\end{itemize}
+	\begin{center}
+		\includegraphics[width=0.6\textwidth]{E_coli.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Next steps}
+	\begin{itemize}
+		\item Given a bidirected string graph that may, in general, have
+		thousands of edges and vertices even after transitive reduction and
+		collapsing unbranched paths, how can we find the most likely assembly?
+		\item Solution: Identify edges that may need to be traversed more than
+		once, then solve a minimum-cost network flow problem to find the minimum
+		number of edges that need to be traversed multiple times in order to
+		find
+		\item This is also similar to the Chinese Postman Problem, but this is a
+		{\it bidirected graph}, and we are also okay with the final
+		reconstruction of the genome still being discontinuous, rather than a
+		single path.
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Implementation}
+	\begin{minipage}{0.3\textwidth}
+		\begin{itemize}
+			\item A set of C++ programs that iteratively transform the data into
+			the final assembly.
+			\item Yellow programs are not yet implemented.
+		\end{itemize}
+	\end{minipage}
+	\begin{minipage}{0.67\textwidth}
+		\includegraphics[width=1.0\textwidth]{programs.pdf}
+	\end{minipage}
+\end{frame}
+
+%\begin{frame}{Effectiveness on random genomes}
+	%\begin{itemize}
+		%\item Error-free 250bp reads sampled from random strands of the full {\it
+		%E. coli} genome (4,639,675 bp) with 100bp separation.
+		%\item 46,392 reads $\to$ {\tt compute-overlaps} $\to$
+	%\end{itemize}
+	%\begin{figure}[H]
+	%\caption{
+		%Bidirected string graph built from E. coli (4,639,675 bp) with uniformly sampled
+		%250bp reads with 125bp minimum overlap length.  (Graph shown after transitive
+		%reduction and collapsing unbranched paths)
+
+		%}
+	%\includegraphics{E_coli_full.pdf}
+	%\end{figure}
+%\end{frame}
+
+%\begin{frame}{Contained reads}
+%\end{frame}
+
+%\begin{frame}{A
+%17. Implementation:
 
 \begin{comment}
-
-11. Walking through the bidirected string graph
-
-- A _walk on a bidirected graph_ is a continuous sequence of edges such that
-  if we enter a vertex v through a head inwards, we leave it on a head outwords,
-  and vice versa.
-
-- A bidirected edge may be traversed in opposite directions.
-
-- The reverse of a walk in a directed graph is also a bidirected walk.
-
-    - Interpretation: one direction of the walk spells out the _forward_
-      sequence.  The other direction spells out the _reverse complement_
-      sequence.
-
-12. Transitive reduction
-
-- Very commonly, given three adjacent reads f, g, and h, f will overlap h as
-  well as g.  This is redundant information because we can equivalently walk f
-  ?-? g ?-? h, but the walks visiting more vertices are preferred because they are
-  supported by more reads.
-
-------------->
-     -------------->
-         --------------->
-
-- Transitive reduction finds edges f ?-? h, described above, and removes them.
-- Algorithm:  Go through each vertex and examine neighbors up to 2 steps away to
-  identify all transitive edges leaving this vertex.  (Some optimizations are
-  possible.)
-
-13. Collapsing unbranched paths
-
-Find unbranched paths and collapse them.
-
-f >-> g >-> h >-> i >-> j >-> k
-
-f >-< g <-> h >-< i <-< j <-> k
-
-Ideally, most the genome will be unbranched sequence.  These sequences will be
-assembled after this step of the algorithm.
-
-But, real genomes contain repeats, and they will create branches in the graph.
 

### 2012-11-28 presentation/presentation.tex
presentation

@@
 	\begin{itemize}
-		\item Each letter in the genome is called {\bf base pair} \\(abbreviation:
+		\item Each letter in the genome is called a {\bf base pair} \\(abbreviation:
 		{\bf bp})
@@
 			{\tt 1.}
-			{\tt \color{red} $\leftarrow$TGAGCTTAAGGCTTATCTATCTTCAGACGACA$\leftarrow$ }\\
+			\hspace*{-0.03cm}{\tt\color{red}$\leftarrow$TGAGCTTAAGGCTTATCTATCTTCAGACGACA$\leftarrow$ }\\
 			{\tt 2.}
@@
 \begin{frame}{Overlaps between reads}
-	Consider two reads $f$ and $g$, and their reverse complements $f'$ and $g'$.
+	Consider two reads, $f$ and $g$, and their reverse complements $f'$ and $g'$.
 	The reads $f$ and $g$ are said to {\bf exactly overlap} by some length
-	$\ell$ if and only if at least one of the following is true:
+	$\ell$ iff at least one of the following is true:
 	\begin{itemize}
@@
 		reads: $N^2$ time)
-		\item A faster algorithm indexes the reads by short subsequences, then only
-		compares reads that share a subsequence ("seed-and-extend").
+		\item A faster algorithm indexes the reads by short subsequences, then
+		only compares reads that share a subsequence ("seed-and-extend").
 		\item Overlaps may be exact, or they may allow for differences up to a
@@
 
-\begin{frame}{Walking through the bidirected string graph}
+\begin{frame}{Walks on a bidirected graph}
 	\begin{itemize}
@@
 		versa.
-
 		\item A bidirected edge may be traversed in opposite directions during
 		the same walk.
+		\item The reverse of a walk in a directed graph is also a bidirected walk.
+		\item $1 \to 2 \to 3 \to 4$ and $4 \to 3 \to 2 \to 1$ are both valid
+		walks in the bidirected graph below.
+	\end{itemize}
+		\begin{center} {\small
+		\begin{tikzpicture}[>=triangle 45]
+			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+			\node (1) at (0.00, 0) {1};
+			\node (2) at (1.75, 0) {2};
+			\node (3) at (3.50, 0) {3};
+			\node (4) at (5.25, 0) {4};
+			\draw[<->,very thick] (1) edge (2);
+			\draw[>-<,very thick] (2) edge (3);
+			\draw[<-<,very thick] (3) edge (4);
+		\end{tikzpicture} } \end{center}
+\end{frame}
 
-		\item The reverse of a walk in a directed graph is also a bidirected walk.
+\begin{frame}{Walking through the bidirected string graph}
+	\begin{itemize}
+		\item A walk through a bidirected string graph constructed from read
+		overlaps models a way in which the reads can be assembled together
+		consistently.
 
-		\item Interpretation: one direction of the walk spells out the {\bf
-		forward sequence}.  The other direction spells out the {\bf reverse
-		complement sequence.}
+		\item Each read may be used in forward or reverse-complement
+		orientation, depending on the orientation of the arrow head when the
+		corresponding vertex is entered.
 	\end{itemize}
+	\begin{figure}[H]
+		{\small
+		\begin{tikzpicture}[>=triangle 45]
+			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+			\node (1) at (0.00, 0) {1};
+			\node (2) at (1.75, 0) {2};
+			\node (3) at (3.50, 0) {3};
+			\node (4) at (5.25, 0) {4};
+			\draw[<->,very thick] (1) edge (2);
+			\draw[>-<,very thick] (2) edge (3);
+			\draw[<-<,very thick] (3) edge (4);
+		\end{tikzpicture} }
+
+		\vspace{2.5mm}
+
+		\begin{tikzpicture}
+				\draw[<-,style=very thick] (0.0, 0.0) -- (3.0, 0.0);
+				\draw[<-,style=very thick] (0.8, -0.3) -- (3.8, -0.3);
+				\draw[->,style=very thick] (1.6, -0.6) -- (4.6, -0.6);
+				\draw[<-,style=very thick] (2.4, -0.9) -- (5.4, -0.9);
+				\draw (-0.3, 0.0) node {1.};
+				\draw (-0.3, -0.3) node {2.};
+				\draw (-0.3, -0.6) node {3.};
+				\draw (-0.3, -0.9) node {4.};
+		\end{tikzpicture}
+		\caption{A bidirected graph and the implied assembly of the
+		corresponding reads}
+	\end{figure}
 \end{frame}
@@
 	\begin{itemize}
-		\item Very commonly, given three adjacent reads $f$, $g$, and $h$, $f$ will overlap
-		$h$ as well as $g$.  This is redundant information because we can equivalently
-		walk $f$ to $g$ to $h$ (or $h$ to $g$ to $f$, if going the other way).
+		\item Very commonly, given three adjacent reads $f$, $g$, and $h$, $f$
+		will overlap $h$ as well as $g$.  This is redundant because we can,
+		equivalently, walk $f$ to $g$ to $h$ rather than directly walking from
+		$f$ to $h$.
 
@@
 				\ENDFOR
-				\STATE{Eliminate all edges $v \to w$ where $w$ is marked {\tt ELIMINATED}}
+				\STATE{Mark as transitive all edges $v \to x$ where $x$ is marked {\tt ELIMINATED}}
 			\ENDFOR
-			\STATE{Eliminate all edges marked {\tt ELIMINATED}}
+			\STATE{Eliminate all edges marked transitive.}
 		\end{algorithmic}
@@
 	\begin{itemize}
-		\item Non-repetitive genomes and/or long reads: The assembly probably is
-		already complete!
+		\item Non-repetitive genomes and/or long reads: The assembly is already
+		complete!
 	\end{itemize}
@@
 		very complicated.
-		\item Swap out that 4,600,000 with {\it E. coli}'s genome, which
-		contains some repeats, and we get a complicated graph, {\bf even after
-		increasing the read length to 1000bp}:
+		\item Swap out the 4.6 Mbp random genome with {\it E. coli}'s genome,
+		which is the same size but contains some repeats, and we get a
+		complicated graph, {\bf even after increasing the read length to
+		1000bp}:
 	\end{itemize}
@@
 		number of edges that need to be traversed multiple times in order to
-		find
+		make an Eulerian path.
 		\item This is also similar to the Chinese Postman Problem, but this is a
@@
 		reconstruction of the genome still being discontinuous, rather than a
-		single path.
+		single path (if necessary).
 	\end{itemize}

### 2012-11-28 presentation/presentation.tex
presentation fixes

@@
 	\begin{enumerate}
-		\item ``normal'' overlap
+		\item ``Normal'' overlap
 			\begin{center}
@@
 			\end{center}
-		\item ``innie'' overlap
+		\item ``Innie'' overlap
 			\begin{center} \begin{tikzpicture}
@@
 			\end{tikzpicture} \end{center}
-		\item ``outtie'' overlap
+		\item ``Outtie'' overlap
 			\begin{center} \begin{tikzpicture}
@@
 				\item {\bf Algorithm}:  Seed-and-extend
-				\item {\bf Running time}: Linear with regards to genome length for
-				non-repetitive genomes; quadratic with regards to read density;
-				quadratic with regards to the number of times a sequence of DNA
-				is repeated in the genome
+				\item {\bf Running time}: Linear with respect to genome length
+				for non-repetitive genomes; quadratic with respect to read
+					density; quadratic with respect to the number of times a
+					sequence of DNA is repeated in the genome
 			\end{itemize}
@@
 		\item However, each read must still be used in a consistent way within
-		any single reconstruction.
+		any single assembly.
 		\item Solution: use a {\bf bidirected graph}.
@@
 
-\begin{frame}{Walks on a bidirected graph}
+\begin{frame}{Walks in a bidirected graph}
 	\begin{itemize}
-		\item A {\bf walk on a bidirected graph} is a continuous sequence of
-		edges such that if any vertex $v$ is entered through a head inwards, it
-		is left on a head outwards (unless it is the end of the path), and vice
-		versa.
+		\item A {\bf walk in a bidirected graph} $G$ is a continuous sequence of
+		edges in $G$ such that if any vertex $v$ is entered through a head
+		inwards, it is left on a head outwards (unless it is the end of the
+		path), and vice versa.
 		\item A bidirected edge may be traversed in opposite directions during
@@
 						\ENDIF
-						\IF{$len(v \to w) + len(w \to x) \ge$ {\tt longest}}
+						\IF{$len(v \to w) + len(w \to x) > $ {\tt longest}}
 							\STATE{\bf break}
@@
 		\caption{
-			Bidirected string graph for reads uniformly sampled from a 4.6 Mb
+			Bidirected string graph for reads uniformly sampled from a 4.6 Mbp
 			random genome after transitive reduction and path collapsing.  (Edge

### 2012-11-28 presentation/presentation.tex
Suggestions

@@
 \setbeamertemplate{frametitle}[default][center]
+
+\newcommand{\ReadArrowType}{latex}
+
+\newcommand{\BidirectedEdgeForward}[2]{
+		\tikz[>=triangle 45,baseline]
+		\draw[>->,thin] (0,0.1) node[anchor=east] {#1} --
+					    (0.8,0.1) node[anchor=west] {#2};}
+
+\newcommand{\LengthVar}{L}
 
@@
 
+\begin{frame}[fragile]{Details of assembler input}
+	\begin{itemize}
+		\item The reads given to an assembler (my code, as well as most
+		other assemblers) as input can be in a simple plain-text format:
+	\end{itemize}
+	\begin{verbatim}
+	>read_1
+	AAATCGGCGCGTAAACAGGCAGCCAGCACCGCAGCA
+	>read_2
+	GAGTAGTCGGAACCGTTGCGTCCAAGCATTAGTGTA
+	>read_3
+	TAGCGGCCAGGATGCTTTACCCAATATCAGCGATGC
+	...
+	\end{verbatim}
+	\begin{itemize}
+		\item The assembler may use other DNA sequence representations
+		(e.g. a binary representation with 2 bits per base) internally.
+	\end{itemize}
+\end{frame}
+
+\begin{frame}[fragile]{Details of assembler output}
+	\begin{itemize}
+		\item Most assemblers can generate output in the same plain-text
+		format.
+		\item A {\bf contig} is a continuous sequence of DNA that was
+		assembled.  Ideally, each contig should be much longer than the
+		original reads.
+	\end{itemize}
+	\begin{verbatim}
+	>contig_1
+	GTTGCGAGATTTGGACGGACGTTGACGGGGTCTATACCTGCGACCCGCGTCA
+	GGTGCCCGATGCGAGGTTGTTGAAGTCGATGTCCTACCAGGAAGCGATGGAG
+	CTTTCCTACTTCGGCGCTAAAGTTCTTCACCCCCGCACCATTACCCCCATCG
+	CCCAGTTCCAGATCCCTTGCCTGATTAAAAATACCGGAAATCCTCAAGCACC
+	AGGTACGCTCATTGGTGCCAGCCGTGATGAAGACGAATTACCGGTCAAGGGC
+	ATTTCCAATCTGAATAACATGGCAATGTTCAGCGTTTCTGGTCCGGGGATGA
+	...
+	\end{verbatim}
+\end{frame}
+
 \begin{frame}{Genome assembly vs Shortest Common Substring}
@@
 \begin{frame}{Overlaps between reads}
-	Consider two reads, $f$ and $g$, and their reverse complements $f'$ and $g'$.
+	Consider two reads, $f$ and $g$, and their reverse-complements $f'$ and $g'$.
 	The reads $f$ and $g$ are said to {\bf exactly overlap} by some length
-	$\ell$ iff at least one of the following is true:
+	$\LengthVar$ iff at least one of the following is true:
 	\begin{itemize}
-		\item The last $\ell$ bp of $f$ exactly match the first $\ell$ bp of
-		$g$.
-		\item The last $\ell$ bp of $f$ exactly match the first $\ell$ bp of
-		$g'$.
-		\item The last $\ell$ bp of $f'$ exactly match the first $\ell$ bp of
-		$g$.
-		\item The last $\ell$ bp of $f'$ exactly match the first $\ell$ bp of
-		$g'$.
+		\item The last $\LengthVar$ bp of $f$ exactly match the first
+		$\LengthVar$ bp of $g$.
+		\item The last $\LengthVar$ bp of $f$ exactly match the first
+		$\LengthVar$ bp of $g'$.
+		\item The last $\LengthVar$ bp of $f'$ exactly match the first
+		$\LengthVar$ bp of $g$.
+		\item The last $\LengthVar$ bp of $f'$ exactly match the first
+		$\LengthVar$ bp of $g'$.
 	\end{itemize}
@@
 
-	{\tt Read 1: \ \ \ \ \ \ \ \ \ $\rightarrow$ ATATAT\textcolor{red}{GCTGGTTACTT} $\rightarrow$ } \\
-	{\tt Read 2: \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 
-		$\rightarrow$ \textcolor{red}{GCTGGTTACTT}TGATAGATA $\rightarrow$ } \\
+	{\tt Read 1: \ \ \ \ \ \ \ \ \ $\rightarrow$
+		ATATAT\textcolor{red}{GCTGGTTACTT} $\rightarrow$ } \\
+	{\tt Read 2: \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ $\rightarrow$
+		\textcolor{red}{GCTGGTTACTT}TGATAGATA $\rightarrow$ } \\
 	{\tt Possible genome: $\rightarrow$ ATATATGCTGGTTACTTTGATAGATA $\rightarrow$}
@@
 
-	{\tt Read 1: \ \ \ \ \ \ \ \ \ 
-			$\rightarrow$ ATATAT\textcolor{red}{GCTGGTTACTT} $\rightarrow$ } \\
-	{\tt Read 2: \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 
-			$\leftarrow$ \textcolor{red}{CGACCAATGAA}ACTATCTAT $\leftarrow$ } \\
+	{\tt Read 1: \ \ \ \ \ \ \ \ \ $\rightarrow$
+		ATATAT\textcolor{red}{GCTGGTTACTT} $\rightarrow$ } \\
+	{\tt Read 2: \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ $\leftarrow$
+		\textcolor{red}{CGACCAATGAA}ACTATCTAT $\leftarrow$ } \\
 	{\tt Possible genome: $\rightarrow$ ATATATGCTGGTTACTTTGATAGATA $\rightarrow$}
@@
 			\begin{center}
-				\begin{tikzpicture}
+				\vspace{-0.5cm}
+				\begin{tikzpicture}[>=\ReadArrowType]
 					\draw[->,style=thick] (0,0) -- (2,0);
@@
 
-				\begin{tikzpicture}
+				\begin{tikzpicture}[>=\ReadArrowType]
 					\draw[->,style=thick] (2,0) -- (0,0);
@@
 			\end{center}
+		\vspace{0.5cm}
 		\item ``Innie'' overlap
-			\begin{center} \begin{tikzpicture}
+			\begin{center}
+				\vspace{-0.5cm}
+				\begin{tikzpicture}[>=\ReadArrowType]
 					\draw[->,style=thick] (0,0) -- (2,0);
@@
 			\end{tikzpicture} \end{center}
-		\item ``Outtie'' overlap
-			\begin{center} \begin{tikzpicture}
+		\vspace{0.5cm}
+		\item ``Outie'' overlap
+			\begin{center}
+			\vspace{-0.5cm}
+			\begin{tikzpicture}[>=\ReadArrowType]
 					\draw[->,style=thick] (2,0) -- (0,0);
@@
 	\begin{itemize}
-		\item Goal: compute all pairwise overlaps of some minimum length $\ell$
-		among the reads.
+		\item Goal: compute all pairwise overlaps of some minimum length
+		$\LengthVar$ among the reads.
 		\item Naive algorithm compares every read with every other read ($N$
@@
 		\item A faster algorithm indexes the reads by short subsequences, then
-		only compares reads that share a subsequence ("seed-and-extend").
+		only compares reads that share a subsequence
+		(``seed-and-extend'').
 		\item Overlaps may be exact, or they may allow for differences up to a
@@
 				\item {\bf Input}:  A set of reads and the minimum overlap
-				length $\ell$
+				length $\LengthVar$
 				\item {\bf Output}:  All exact pairwise overlaps between reads,
@@
 		\begin{center}
-			\begin{tikzpicture}
+			\begin{tikzpicture}[>=\ReadArrowType]
 				\draw[->,style=thick] (0,0) -- (2,0);
@@
 			\hspace{5mm}or\hspace{5mm}
-			\begin{tikzpicture}
+			\begin{tikzpicture}[>=\ReadArrowType]
 				\draw[->,style=thick] (0,0.3) -- (2,0.3);
@@
 				{\footnotesize Overlap} &
-				\begin{tikzpicture}[scale=0.6]
+				\begin{tikzpicture}[scale=0.6,>=\ReadArrowType]
 						\draw[->,style=thick] (2,0) -- (0,0);
@@
 				&
-				\begin{tikzpicture}[scale=0.6]
+				\begin{tikzpicture}[scale=0.6,>=\ReadArrowType]
 						\draw[->,style=thick] (0,0) -- (2,0);
@@
 				&
-				\begin{tikzpicture}[scale=0.6]
+				\begin{tikzpicture}[scale=0.6,>=\ReadArrowType]
 						\draw[->,style=thick] (2,0) -- (0,0);
@@
 		the same walk.
-		\item The reverse of a walk in a directed graph is also a bidirected walk.
+		\item The reverse of a walk in a bidirected graph is also a
+		valid walk.
 		\item $1 \to 2 \to 3 \to 4$ and $4 \to 3 \to 2 \to 1$ are both valid
@@
 			\node (1) at (0.00, 0) {1};
-			\node (2) at (1.75, 0) {2};
-			\node (3) at (3.50, 0) {3};
-			\node (4) at (5.25, 0) {4};
+			\node (2) at (2.0, 0) {2};
+			\node (3) at (4.0, 0) {3};
+			\node (4) at (6.0, 0) {4};
 			\draw[<->,very thick] (1) edge (2);
@@
 
-		\begin{tikzpicture}
-				\draw[<-,style=very thick] (0.0, 0.0) -- (3.0, 0.0);
-				\draw[<-,style=very thick] (0.8, -0.3) -- (3.8, -0.3);
-				\draw[->,style=very thick] (1.6, -0.6) -- (4.6, -0.6);
-				\draw[<-,style=very thick] (2.4, -0.9) -- (5.4, -0.9);
+		\begin{tikzpicture}[>=\ReadArrowType]
+				\draw[<-,style=very thick] (0.0, 0.0) -- (4.0, 0.0);
+				\draw[<-,style=very thick] (1.2, -0.3) -- (5.2, -0.3);
+				\draw[->,style=very thick] (2.4, -0.6) -- (6.4, -0.6);
+				\draw[<-,style=very thick] (3.6, -0.9) -- (7.6, -0.9);
 				\draw (-0.3, 0.0) node {1.};
@@
 \end{frame}
-
-\newcommand{\BidirectedEdgeForward}[2]{
-		\tikz[>=triangle 45,baseline]
-		\draw[>->,thin] (0,0.1) node[anchor=east] {#1} --
-					    (0.8,0.1) node[anchor=west] {#2};}
 
@@
 				\draw[->,style=very thick] (0,0) node[anchor=east] {f} -- (2,0);
-				\draw[->,style=very thick] (0.5,0.25) node[anchor=east] {g} -- (2.5,0.25);
-				\draw[->,style=very thick] (1.0,0.5) node[anchor=east] {h} -- (3,0.5);
+				\draw[->,style=very thick] (0.6,0.25) node[anchor=east] {g} -- (2.6,0.25);
+				\draw[->,style=very thick] (1.2,0.5) node[anchor=east] {h} -- (3.2,0.5);
 			\end{tikzpicture}
@@
 				\node (f) at (0, 0) {f};
-				\node (g) at (1.5, 0) {g};
-				\node (h) at (3, 0) {h};
+				\node (g) at (1.7, 0) {g};
+				\node (h) at (3.4, 0) {h};
 				\draw[>->,style=very thick] (f) edge (g);
@@
 		\item Walks visiting more vertices are preferred because they are
-		supported by more reads, and the overlaps tend to be longer.  So, we
+		supported by more reads and the overlaps tend to be longer.  So, we
 		would like to throw away \BidirectedEdgeForward{$f$}{$h$} but keep
@@
 \begin{frame}{Transitive reduction (cont.)}
-		Transitive reduction algorithm:  Go through each vertex and examine
-		neighbors up to 2 steps away to identify all transitive edges leaving this
-		vertex.
+		Transitive reduction algorithm:  Go through each vertex $v$ and
+		examine vertices up to 2 edges away from $v$ to identify all
+		transitive edges leaving $v$.
 	\begin{algorithm}[H]
@@
 			\FOR{{\bf each} vertex $v \in G$}
-				\STATE{Mark all neighbors of $v$ as {\tt INPLAY}.}
+				\STATE{Mark all neighbors of $v$ as {\tt INPLAY}}
 				\STATE{Set {\tt longest} $\gets$ length of the longest edge leaving $v$}
@@
 			\ENDFOR
-			\STATE{Eliminate all edges marked transitive.}
+			\STATE{Eliminate all edges marked transitive}
 		\end{algorithmic}

### 2012-11-29 presentation/presentation.tex
Before update build graph (cont)

@@
 
-\begin{frame}{Genome assembly vs Shortest Common Substring}
+\begin{frame}{Genome assembly vs Shortest Common Superstring}
 	Genome assembly is superficially similar to the shortest common superstring
@@
 	\end{itemize}
-	An exact overlap $o$ is fully described by the tuple $(f, g, f_{beg},
-	g_{beg}, len, rc)$.
+	An exact overlap $o$ is fully described by the tuple $(f, g, f_{begin},
+	g_{begin}, L, rc)$.
 	%there $f$ and $g$ are the reads of the overlap,
@@
 		\begin{itemize}
-			\item Two reads are connected with an edge iff they share an overlap.
+			\item Two vertices are connected with an edge iff the
+			corresponding reads share an overlap.
 		\end{itemize}
@@
 		path), and vice versa.
-		\item A bidirected edge may be traversed in opposite directions during
-		the same walk.
+		\item A bidirected edge may be traversed in both directions
+		(possibly even during the same walk).
 		\item The reverse of a walk in a bidirected graph is also a
@@
 		\item $1 \to 2 \to 3 \to 4$ and $4 \to 3 \to 2 \to 1$ are both valid
-		walks in the bidirected graph below.
+		walks in the bidirected graph below.  (Check circled pairs of
+		heads.)
 	\end{itemize}
@@
 			\node (1) at (0.00, 0) {1};
-			\node (2) at (1.75, 0) {2};
-			\node (3) at (3.50, 0) {3};
-			\node (4) at (5.25, 0) {4};
+			\node (2) at (2.00, 0) {2};
+			\node (3) at (4.00, 0) {3};
+			\node (4) at (6.00, 0) {4};
 			\draw[<->,very thick] (1) edge (2);
@@
 			\draw[<-<,very thick] (3) edge (4);
+			\draw[color=red,thick] (2.00, 0) circle (0.8);
+			\draw[color=red,thick] (4.00, 0) circle (0.8);
 		\end{tikzpicture} } \end{center}
@@
 	\begin{itemize}
-		\item A walk through a bidirected string graph constructed from read
-		overlaps models a way in which the reads can be assembled together
-		consistently.
+		\item A walk through a bidirected string graph constructed from
+		reads and overlaps models a way in which the reads can be
+		assembled together consistently.
 
@@
 				\draw[<-,style=very thick] (0.0, 0.0) -- (4.0, 0.0);
-				\draw[<-,style=very thick] (1.2, -0.3) -- (5.2, -0.3);
-				\draw[->,style=very thick] (2.4, -0.6) -- (6.4, -0.6);
+				\draw[->,style=very thick] (1.2, -0.3) -- (5.2, -0.3);
+				\draw[<-,style=very thick] (2.4, -0.6) -- (6.4, -0.6);
 				\draw[<-,style=very thick] (3.6, -0.9) -- (7.6, -0.9);

### 2012-11-29 presentation/presentation.tex
Another build graph slide

@@
 \usepackage[noend]{algorithmic}
+\usepackage{multicol}
 
@@
 	\begin{itemize}
+		\item Consider three reads $f$, $g$, and $h$, that overlap as follows:
+	\end{itemize}
+	\vspace{-0.4cm}
+	\begin{center}
+		\begin{tikzpicture}[>=\ReadArrowType]
+			\draw[->,style=very thick] (0,   0.6) node[anchor=east] {f} -- (3.0, 0.6);
+			\draw[->,style=very thick] (1.0, 0.3) node[anchor=east] {g} -- (4.0, 0.3);
+			\draw[<-,style=very thick] (2.0, 0.0) node[anchor=east] {h} -- (5.0, 0.0);
+		\end{tikzpicture}
+	\end{center}
+	\vspace{-0.5cm}
+	\begin{itemize}
+		\item {There are 3 overlaps so there will be 3 edges, and we'd like for
+		the graph to allow exactly the following four walks:}
+		\vspace{-0.3cm}
+		\begin{multicols}{2}
+			\begin{itemize}
+				\item $f \to g \to h$
+				\item $h \to g \to f$
+				\item $f \to h$
+				\item $h \to f$
+			\end{itemize}
+		\end{multicols}
+	\end{itemize}
+	\begin{tabular}{ccp{3.3cm}}
+		{\small Undirected graph?}
+			&
+			\begin{tikzpicture}[>=triangle 45]
+				\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+				\node (f) at (0, 0) {f};
+				\node (g) at (1.7, 0) {g};
+				\node (h) at (3.4, 0) {h};
+				\draw[-,style=very thick] (f) edge (g);
+				\draw[-,style=very thick] (g) edge (h);
+				\draw[-,style=very thick,bend left] (f) edge (h);
+			\end{tikzpicture}
+			& 
+			\vspace{-0.8cm}
+			{\footnotesize No; nothing indicates we can't walk \mbox{$g \to h \to
+			f$}.}
+			\\
+		{\small Directed graph?}
+			&
+			\begin{tikzpicture}[>=triangle 45]
+				\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+				\node (f) at (0, 0) {f};
+				\node (g) at (1.7, 0) {g};
+				\node (h) at (3.4, 0) {h};
+				\draw[->,style=very thick] (f) edge (g);
+				\draw[->,style=very thick] (g) edge (h);
+				\draw[->,style=very thick,bend left] (f) edge (h);
+			\end{tikzpicture}
+			&
+			\vspace{-0.8cm}
+			{\footnotesize
+				No; this doesn't allow the walks $h \to g \to f$ and $h \to f$.}
+			\\
+	\end{tabular}
+\end{frame}
+
+\begin{frame}{Build the fragment string assembly graph (cont.)}
+	\begin{itemize}
 		\item At this stage, we don't know which strands the reads are from!
@@
 		reverse-complement or forward direction.
-		\item Therefore, we must allow the edges to be traversed in either
-		direction.
+		\item We must allow the edges to be traversed in either direction.
 		\item However, each read must still be used in a consistent way within
@@
 	\end{itemize}
+	\begin{center}
+		\begin{tikzpicture}[>=triangle 45]
+			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+			\node (f) at (0, 0) {f};
+			\node (g) at (1.7, 0) {g};
+			\node (h) at (3.4, 0) {h};
+			\draw[>->,style=very thick] (f) edge (g);
+			\draw[>-<,style=very thick] (g) edge (h);
+			\draw[>-<,style=very thick,bend left] (f) edge (h);
+		\end{tikzpicture}
+	\end{center}
 \end{frame}
@@
 \end{frame}
+
+%\begin{frame}{Walking through the bidirected string graph (cont.)}
+	%\begin{minipage}{0.5\textwidth}
+		%\begin{tikz
+	%\end{minipage}
+%\end{frame}
 

### 2012-11-29 presentation/presentation.tex
f.E g.E slide

@@
 	\end{itemize}
-	An exact overlap $o$ is fully described by the tuple $(f, g, f_{begin},
-	g_{begin}, L, rc)$.
+	An overlap $o$ is fully described by the 7-tuple $(f, g, f_{begin}, f_{end},
+	g_{begin}, g_{end}, rc)$.
 	%there $f$ and $g$ are the reads of the overlap,
@@
 	\end{itemize}
-	\begin{tabular}{ccp{3.3cm}}
+	\begin{tabular}{ccp{3.4cm}}
 		{\small Undirected graph?}
@@
 			& 
-			\vspace{-0.8cm}
-			{\footnotesize No; nothing indicates we can't walk \mbox{$g \to h \to
+			\vspace{-1.1cm}
+			{\footnotesize Not what we want; nothing indicates we can't walk \mbox{$g \to h \to
 			f$}.}
@@
 			{\footnotesize
-				No; this doesn't allow the walks $h \to g \to f$ and $h \to f$.}
+				Not what we want; this doesn't allow the walks $h \to g \to f$ and $h \to f$.}
 			\\
@@
 
-%\begin{frame}{Walking through the bidirected string graph (cont.)}
-	%\begin{minipage}{0.5\textwidth}
-		%\begin{tikz
-	%\end{minipage}
-%\end{frame}
+\begin{frame}{Walking through the bidirected string graph (cont.)}
+	\vspace{-0.7cm}
+	\begin{center}
+		\begin{tabular}{p{3.4cm}c}
+			\vspace{-0.4cm}
+			{\small a.) A bidirected string graph fragment} &
+			\begin{tikzpicture}[>=triangle 45,scale=1.5]
+				\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+				\node (f) at (0, 0) {f};
+				\node (g) at (1.7, 0) {g};
+				\draw[>->,style=very thick] (f) edge (g);
+			\end{tikzpicture}
+			\\
+			\vspace{-1.5cm}
+			{\small b.) The corresponding overlap (labeled)} &
+			\begin{tikzpicture}[>=\ReadArrowType,scale=0.8]
+				\draw[->,very thick] (0, 0) node[anchor=east] {$f.B$}
+						-> (4, 0) node[anchor=west] {$f.E$};
+				\draw[->,very thick] (2, -0.7) node[anchor=east] {$g.B$}
+						-> (6, -0.7) node[anchor=west] {$g.E$};
+				\draw (2, -1.05) node[anchor=north] {$o.g_{begin}$}
+						-- (2, 0.35) node[anchor=south] {$o.f_{begin}$};
+				\draw (4, -1.05) node[anchor=north] {$o.g_{end}$}
+						-- (4, 0.35) node[anchor=south] {$o.f_{end}$};
+			\end{tikzpicture}
+		\end{tabular}
+	\end{center}
+	\begin{tabular}{c|c}
+		walk & interpretation \\ \hline
+		$f \to g$ & $f.E \to g.E$ with sequence $g[o.g_{end} + 1 \dots ${\tt
+		length}$(g)]$ \\
+		$g \to f$ & $g.B \to f.B$ with sequence $f[o.g_{begin} - 1 \dots 0]$ \\
+	\end{tabular}
+	\vspace{0.5cm}
+	\begin{itemize}
+		\item If we take {\bf each end} of {\bf each read} as a vertex, we can
+		make a directed graph equivalent to the bidirected graph.
+	\end{itemize}
+\end{frame}
 

### 2012-11-29 presentation/presentation.tex
presentation

@@
 \usepackage{multicol}
+\usepackage{grffile}
 
@@
 
+%make -B SAMPLE_SIZE=450 out.reduced.mapped.bidigraph.pdf out.reduced.mapped.collapsed.bidigraph.pdf out.bidigraph.pdf GENOME=rep.fa COVERAGE=5
+
+\begin{frame}{Another example}
+	\begin{figure}[H]
+		\includegraphics[scale=0.25]{example.bidigraph-crop.pdf}	
+		\caption{Bidirected string graph built from 100 bp reads sampled
+		randomly at 5 $\times$ coverage from a 450 bp genome, using minimum 25
+		bp overlaps.  Note: edges are labeled with their lengths, not DNA
+		sequences.}
+	\end{figure}
+\end{frame}
+
+\begin{frame}{Another example (cont.)}
+	\begin{figure}[H]
+		\includegraphics[scale=0.7]{example.reduced.mapped.bidigraph-crop.pdf}	
+		\caption{The bidirected string graph from the previous slide, after
+		transitive reduction.}
+	\end{figure}
+\end{frame}
+
+\begin{frame}{Another example (cont.)}
+	\begin{figure}[H]
+		\includegraphics[scale=0.7]{example.reduced.mapped.collapsed.bidigraph-crop.pdf}	
+		\caption{The bidirected string graph from the previous slide, after
+		collapsing unbranched paths.}
+	\end{figure}
+\end{frame}
+
 \begin{frame}{Next steps}

### 2012-11-29 src/DirectedStringGraph.cc
DirectedStringGraph::follow_unbranched_path()
Don't remove vertices if they were part of a smooth ring that was not collapsed

@@ -214,6 +214,7 @@ void DirectedStringGraph::transitive_reduction()
 
 void DirectedStringGraph::follow_unbranched_path(DirectedStringGraphEdge & e,
 						 std::vector<bool> & remove_edge,
+						 std::vector<bool> & remove_vertex,
 						 const std::vector<bool> & v_inner)
 {
 	BaseVecVec::size_type new_seq_len = e.length();
@@ -250,6 +251,7 @@ void DirectedStringGraph::follow_unbranched_path(DirectedStringGraphEdge & e,
 		}
 		e.increment_mapped_read_count(ei_i1.get_mapped_count());
 		remove_edge[ei_i1_idx] = true;
+		remove_vertex[vi_idx] = true;
 		vi_idx = ei_i1.get_v2_idx();
 	} while (v_inner[vi_idx]);
 	assert(new_seq_len == seq_idx);
@@ -303,6 +305,7 @@ void DirectedStringGraph::collapse_unbranched_paths()
 	//
 	size_t num_unbranched_paths = 0;
 	std::vector<bool> remove_edge(n_edges, false);
+	std::vector<bool> remove_vertex(n_verts, false);
 	for (v_idx_t v_idx = 0; v_idx < n_verts; v_idx++) {
 		if (!v_inner[v_idx]) {
 			const DirectedStringGraphVertex & v = _vertices[v_idx];
@@ -310,7 +313,8 @@ void DirectedStringGraph::collapse_unbranched_paths()
 				DirectedStringGraphEdge & e = _edges[edge_idx];
 				if (v_inner[e.get_v2_idx()]) {
 					num_unbranched_paths++;
-					follow_unbranched_path(e, remove_edge, v_inner);
+					follow_unbranched_path(e, remove_edge,
+							       remove_vertex, v_inner);
 				}
 			}
 		}
@@ -323,7 +327,7 @@ void DirectedStringGraph::collapse_unbranched_paths()
 						  std::numeric_limits<v_idx_t>::max());
 	v_idx_t new_v_idx = 0;
 	for (v_idx_t old_v_idx = 0; old_v_idx < n_verts; old_v_idx++)
-		if (!v_inner[old_v_idx])
+		if (!remove_vertex[old_v_idx])
 			old_to_new_v_indices[old_v_idx] = new_v_idx++;
 
 	info("Updated vertices are indexed [0, %lu)", new_v_idx);
[…]

### 2012-11-29 src/DirectedStringGraph.h

@@ -215,6 +215,7 @@ public:
 private:
 	void follow_unbranched_path(DirectedStringGraphEdge & e,
 				    std::vector<bool> & remove_edge,
+				    std::vector<bool> & remove_vertex,
 				    const std::vector<bool> & v_inner);
 	void mark_component(const v_idx_t v_idx,
 			    std::vector<bool> & visited,

### 2012-11-29 src/AnyStringGraph.h
Add extract-edge-seqs

@@ -76,6 +76,7 @@ public:
 	DISPATCH0(collapse_unbranched_paths);
 	DISPATCH1(print_stats, std::ostream &);
 	DISPATCH1(write, const char *);
+	DISPATCH1(extract_edge_seqs, BaseVecVec &);
 	DISPATCH2(print, std::ostream &, bool);
 	DISPATCH2(print_dot, std::ostream &, bool);
 	DISPATCH3(map_contained_read, size_t, size_t, size_t);

### 2012-11-29 src/BidirectedStringGraph.h

@@ -301,8 +301,15 @@ public:
 		_vertices[v2_idx].add_edge_idx(edge_idx);
 	}
 
-public:
 	void assert_graph_valid() const
 	{
+	}
+
+	void extract_edge_seqs(BaseVecVec & bvv)
+	{
+		foreach (BidirectedStringGraphEdge & e, _edges) {
+			bvv.push_back(e.get_seq_1_to_2());
+			bvv.push_back(e.get_seq_2_to_1());
+		}
 	}
 };

### 2012-11-29 src/DirectedStringGraph.h

@@ -244,4 +244,10 @@ public:
 		}
 	}
 
+	void extract_edge_seqs(BaseVecVec & bvv)
+	{
+		foreach (DirectedStringGraphEdge & e, _edges) {
+			bvv.push_back(e.get_seq());
+		}
+	}
 };

### 2012-11-29 src/extract-edge-seqs.cc

@@ -1,0 +1,28 @@
+#include "AnyStringGraph.h"
+#include "util.h"
+
+DEFINE_USAGE(
+"Usage: extract-edge-seqs GRAPH_FILE OUT_CONTIGS_FILE\n"
+"\n"
+"Extracts the sequences from the edges of a string graph.\n"
+"\n"
+"Input:\n"
+"      GRAPH_FILE:  The string graph.\n"
+"      OUT_CONTIGS_FILE:  A file to which the edge sequences are to be\n"
+"                         extracted.\n"
+);
+
+int main(int argc, char **argv)
+{
+	USAGE_IF(argc != 3);
+	const char *graph_file = argv[1];
+	const char *out_contigs_file = argv[2];
+
+	AnyStringGraph graph(graph_file);
+
+	BaseVecVec bvv;
+
+	graph.extract_edge_seqs(bvv);
+	info("Writing %zu contigs to \"%s\"", bvv.size(), out_contigs_file);
+	bvv.write(out_contigs_file);
+}

### 2012-11-30 src/map-contained-reads.cc
Map underhangs as well as overhangs

@@ -96,9 +96,16 @@ int main(int argc, char **argv)
 	std::vector<const Overlap *>
 		shortest_overhang_overlaps(num_contained_reads, NULL);
 
+	std::vector<const Overlap *>
+		shortest_underhang_overlaps(num_contained_reads, NULL);
+
 	std::vector<Overlap::read_pos_t>
 		shortest_overhang_lens(num_contained_reads,
 				       std::numeric_limits<Overlap::read_pos_t>::max());
+
+	std::vector<Overlap::read_pos_t>
+		shortest_underhang_lens(num_contained_reads,
+					std::numeric_limits<Overlap::read_pos_t>::max());
 
 	foreach (const OverlapVecVec::OverlapSet & overlap_set, orig_overlaps) {
 		foreach(const Overlap & o, overlap_set) {
@@ -158,6 +165,7 @@ int main(int argc, char **argv)
 				// with original index @contained_read_orig_idx
 				// is contained.  Compute the overhang length.
 				Overlap::read_pos_t overhang_len;
+				Overlap::read_pos_t underhang_len;
 				if (rc) {
 					//
 					//
@@ -166,6 +174,8 @@ int main(int argc, char **argv)
 					//
 					//                 |overhang|
 					overhang_len = uncontained_read_overlap_beg;
+					underhang_len = uncontained_read->length() -
+							(uncontained_read_overlap_end + 1);
 				} else {
 					//
 					//
@@ -177,6 +187,7 @@ int main(int argc, char **argv)
 					       uncontained_read_overlap_end + 1);
 					overhang_len = uncontained_read->length() -
 						       (uncontained_read_overlap_end + 1);
+					underhang_len = uncontained_read_overlap_beg;
 				}
 
 				// Index of this contained read in the
@@ -189,6 +200,11 @@ int main(int argc, char **argv)
 				if (overhang_len < shortest_overhang_lens[contained_idx]) {
 					shortest_overhang_overlaps[contained_idx] = &o;
 					shortest_overhang_lens[contained_idx] = overhang_len;
+				}
+
+				if (underhang_len < shortest_underhang_lens[contained_idx]) {
+					shortest_underhang_overlaps[contained_idx] = &o;
+					shortest_underhang_lens[contained_idx] = underhang_len;
 				}
 			}
 		}
@@ -204,36 +220,50 @@ int main(int argc, char **argv)
[…]

### 2012-11-30 src/BidirectedStringGraph.cc
StringGraphEdge: Add _num_inner_vertices

@@ -222,6 +222,11 @@ void BidirectedStringGraph::build_from_digraph(const DirectedStringGraph & digra
 				e.get_seq_2_to_1() = w_v.get_seq();
 				e.set_v_indices(v1_idx, v2_idx);
 				e.set_dirs(dirs);
+				e.set_mapped_read_count((w_v.get_mapped_read_count() +
+							 v_w.get_mapped_read_count()) / 2);
+				assert(w_v.get_num_inner_vertices() ==
+				       v_w.get_num_inner_vertices());
+				e.set_num_inner_vertices(w_v.get_num_inner_vertices());
 
 				edge_idx_t edge_idx = this->push_back_edge(e);
 				_vertices[v1_idx].add_edge_idx(edge_idx);

### 2012-11-30 src/DirectedStringGraph.cc

@@ -249,7 +249,8 @@ void DirectedStringGraph::follow_unbranched_path(DirectedStringGraphEdge & e,
 			assert2(seq_idx < new_seq_len);
 			new_seq.set(seq_idx++, ei_i1_seq[i]);
 		}
-		e.increment_mapped_read_count(ei_i1.get_mapped_count());
+		e.increment_mapped_read_count(ei_i1.get_mapped_read_count());
+		e.increment_num_inner_vertices();
 		remove_edge[ei_i1_idx] = true;
 		remove_vertex[vi_idx] = true;
 		vi_idx = ei_i1.get_v2_idx();
@@ -408,9 +409,9 @@ void DirectedStringGraph::print_stats(std::ostream & os) const
 		if (in_degrees[e.get_v2_idx()] + 1 != 0) {
 			in_degrees[e.get_v2_idx()]++;
 		}
-		assert(e.get_mapped_count() >= 1.0);
-		total_mapped_count += e.get_mapped_count();
-		if (e.get_mapped_count() > 1.0)
+		assert(e.get_mapped_read_count() >= 1.0);
+		total_mapped_count += e.get_mapped_read_count();
+		if (e.get_mapped_read_count() > 1.0)
 			more_than_one_mapped_count++;
 	}
 	std::vector<v_idx_t> out_degree_hist(0xff, 0);

### 2012-11-30 src/StringGraph.h

@@ -14,19 +14,22 @@
 // Base class for edges of the string graph.
 class StringGraphEdge {
 protected:
+	float _mapped_read_count;
+	unsigned _num_inner_vertices;
+
 	StringGraphEdge()
 	{
 		_mapped_read_count = 1.0;
+		_num_inner_vertices = 0;
 	}
-	float _mapped_read_count;
 public:
-	void increment_mapped_read_count(const float n)
-	{
-		_mapped_read_count += n;
-	}
-	float get_mapped_count() const {
-		return _mapped_read_count;
-	}
+	void increment_mapped_read_count(float n) { _mapped_read_count += n; }
+	float get_mapped_read_count() const { return _mapped_read_count; }
+	void set_mapped_read_count(float n) { _mapped_read_count = n; }
+
+	void increment_num_inner_vertices() { _num_inner_vertices++; }
+	unsigned get_num_inner_vertices() const { return _num_inner_vertices; }
+	void set_num_inner_vertices(unsigned n) { _num_inner_vertices = n; }
 };
 
 // Base class for vertices of the string graph.

### 2012-11-30 src/AnyStringGraph.h
calculate-A-statistics skeleton

@@ -73,6 +73,7 @@ public:
 	}
 
 	DISPATCH0(transitive_reduction);
+	DISPATCH0(calculate_A_statistics);
 	DISPATCH0(collapse_unbranched_paths);
 	DISPATCH1(print_stats, std::ostream &);
 	DISPATCH1(write, const char *);

### 2012-11-30 src/BidirectedStringGraph.cc

@@ -316,3 +316,8 @@ void BidirectedStringGraph::map_contained_read(const v_idx_t downstream_read_idx
 {
 	unimplemented();
 }
+
+void BidirectedStringGraph::calculate_A_statistics()
+{
+	unimplemented();
+}

### 2012-11-30 src/BidirectedStringGraph.h

@@ -268,6 +268,7 @@ public:
 
 	void transitive_reduction();
 	void collapse_unbranched_paths();
+	void calculate_A_statistics();
 	void print_stats(std::ostream & os) const;
 
 	void map_contained_read(const v_idx_t downstream_read_idx,

### 2012-11-30 src/DirectedStringGraph.cc

@@ -627,3 +627,8 @@ void DirectedStringGraph::build_from_bidigraph(const BidirectedStringGraph & bid
 			      e.get_seq_2_to_1(), 0, e.length() - 1, false);
 	}
 }
+
+void DirectedStringGraph::calculate_A_statistics()
+{
+	unimplemented();
+}

### 2012-11-30 src/DirectedStringGraph.h

@@ -176,6 +176,7 @@ public:
 
 	void transitive_reduction();
 	void collapse_unbranched_paths();
+	void calculate_A_statistics();
 	void print_stats(std::ostream & os) const;
 
 	void map_contained_read(const v_idx_t downstream_read_idx,

### 2012-11-30 src/calculate-A-statistics.cc

@@ -1,0 +1,27 @@
+#include "AnyStringGraph.h"
+
+DEFINE_USAGE(
+"calculate-A-statistics GRAPH_FILE OUT_GRAPH_FILE\n"
+"\n"
+"Calculate the A-statistic (arrival rate statistic) on each edge in\n"
+"a string graph.\n"
+"\n"
+"Input:\n"
+"      GRAPH_FILE: A string graph.\n"
+"\n"
+"Output:\n"
+"      OUT_GRAPH_FILE:  The string graph with the A-statistics calculated.\n"
+);
+
+int main(int argc, char **argv)
+{
+	USAGE_IF(argc != 3);
+	const char *graph_file = argv[1];
+	const char *out_graph_file = argv[2];
+
+	info("Loading \"%s\"", graph_file);
+	AnyStringGraph graph(graph_file);
+	graph.calculate_A_statistics();
+	info("Writing \"%s\"", out_graph_file);
+	graph.write(out_graph_file);
+}

### 2012-11-30 src/DirectedStringGraph.cc
collapse-unbranched-paths: Fix logging statement

@@ -354,6 +354,7 @@ void DirectedStringGraph::collapse_unbranched_paths()
 	}
 	info("Updated edges are indexed [0, %lu)", new_edge_idx);
 	info("%zu edges were removed (%f%% of total)",
+	     _edges.size() - new_edge_idx,
 	     TO_PERCENT(_edges.size() - new_edge_idx, _edges.size()));
 
 	_edges.resize(new_edge_idx);

### 2012-11-30 src/BidirectedStringGraph.h
A-statistic bootstrapping

@@ -3,11 +3,35 @@
 #include "StringGraph.h"
 #include "BaseVec.h"
 #include <boost/serialization/access.hpp>
+#include <boost/serialization/base_object.hpp>
 #include <ostream>
 #include <inttypes.h>
 #include <math.h>
 
 class DirectedStringGraph;
+
+// A vertex of a bidirected string graph.
+class BidirectedStringGraphVertex : public StringGraphVertex {
+private:
+	friend class boost::serialization::access;
+	template <class Archive>
+	void serialize(Archive & ar, unsigned version)
+	{
+		ar & boost::serialization::base_object<StringGraphVertex>(*this);
+	}
+public:
+	size_t out_degree() const
+	{
+		unimplemented();
+	}
+
+	// Print a bidirected string graph vertex in DOT format
+	void print_dot(std::ostream & os, size_t v_idx) const
+	{
+		os << "\tv" << v_idx
+		   << " [ label = \"" << (v_idx + 1) << "\" ];\n";
+	}
+};
 
 // An edge of a bidirected string graph.
 //
@@ -48,6 +72,7 @@ private:
 	template <class Archive>
 	void serialize(Archive & ar, unsigned version)
 	{
+		ar & boost::serialization::base_object<StringGraphEdge>(*this);
 		ar & _data;
 		ar & _seq_1_to_2;
 		ar & _seq_2_to_1;
@@ -218,23 +243,6 @@ public:
 	}
 };
 
-
-// A vertex of a bidirected string graph.
-class BidirectedStringGraphVertex : public StringGraphVertex {
-public:
-	size_t out_degree() const
-	{
-		unimplemented();
-	}
-
-	// Print a bidirected string graph vertex in DOT format
-	void print_dot(std::ostream & os, size_t v_idx) const
-	{
-		os << "\tv" << v_idx
-		   << " [ label = \"" << (v_idx + 1) << "\" ];\n";
-	}
-};
-
 // A bidirected string graph.
 class BidirectedStringGraph : public StringGraph<BidirectedStringGraphVertex,
 						 BidirectedStringGraphEdge,
@@ -248,6 +256,7 @@ public:
 		if (!enough_v_indices(num_reads))
 			fatal_error("Too many reads (%zu)", num_reads);
 		_vertices.resize(num_reads);
+		_orig_num_reads = num_reads;
 	}
 
 	// Read a bidirected string graph from a file

### 2012-11-30 src/DirectedStringGraph.cc

@@ -4,6 +4,7 @@
 #include <boost/archive/binary_iarchive.hpp>
 #include <boost/archive/binary_oarchive.hpp>
 #include "BidirectedStringGraph.h"
+#include <math.h>
 
 const char DirectedStringGraph::magic[] =
 	{'D', 'i', 'g', 'r', 'a', 'p', 'h', '\0', '\0', '\0'};
@@ -631,5 +632,48 @@ void DirectedStringGraph::build_from_bidigraph(const BidirectedStringGraph & bid
 
 void DirectedStringGraph::calculate_A_statistics()
 {
-	unimplemented();
+	size_t num_reads = _orig_num_reads;
+	size_t genome_len;
+
+	size_t bootstrap_genome_len = 0;
+	size_t bootstrap_num_reads = num_reads;
+	size_t n_edges = num_edges();
+	size_t i = 0;
+	while (i < n_edges) {
+		bootstrap_genome_len += _edges[i].length();
+		i += 2;
+	}
+	float global_arrival_rate = FLOAT_DIV_NONZERO(bootstrap_num_reads,
+						      bootstrap_genome_len);
+
+	info("Bootstrapping with bootstream_genome_len = %zu, "
+	     "bootstrap_num_reads = %zu, global_arrival_rate = %f",
+	     bootstrap_genome_len, bootstrap_num_reads, global_arrival_rate);
+
+	const size_t NUM_BOOTSTRAP_ITERATIONS = 10;
+	const float SINGLE_COPY_THRESHOLD = 17.0;
+
+	for (size_t i = 0; i < NUM_BOOTSTRAP_ITERATIONS; i++) {
+		bootstrap_genome_len = 0;
+		bootstrap_num_reads = 0;
+		for (size_t j = 0; j < n_edges; j++) {
+			DirectedStringGraphEdge & e = _edges[j];
+			size_t edge_len = _edges[j].length();
+			unsigned edge_reads = _edges[j].get_num_inner_vertices();
+			float A_statistic = (global_arrival_rate * edge_len) -
+					     (edge_reads * M_LN2);
+			e.set_A_statistic(A_statistic);
+			info("edge_len = %zu, edge_reads = %u, A_statistic = %f",
+			     edge_len, edge_reads, A_statistic);
+			if (A_statistic >= SINGLE_COPY_THRESHOLD) {
+				bootstrap_genome_len += edge_len;
[…]

### 2012-11-30 src/DirectedStringGraph.h

@@ -4,11 +4,19 @@
 #include "BaseVec.h"
 #include <ostream>
 #include <inttypes.h>
+#include <boost/serialization/base_object.hpp>
 
 class BidirectedStringGraph;
 
 // A vertex of a directed string graph.
 class DirectedStringGraphVertex : public StringGraphVertex {
+private:
+	friend class boost::serialization::access;
+	template <class Archive>
+	void serialize(Archive & ar, unsigned version)
+	{
+		ar & boost::serialization::base_object<StringGraphVertex>(*this);
+	}
 public:
 	size_t out_degree() const
 	{
@@ -42,7 +50,7 @@ private:
 	template <class Archive>
 	void serialize(Archive & ar, unsigned version)
 	{
-		ar & _mapped_read_count;
+		ar & boost::serialization::base_object<StringGraphEdge>(*this);
 		ar & _v1_idx;
 		ar & _v2_idx;
 		ar & _seq;
@@ -158,6 +166,7 @@ public:
 		if (!enough_v_indices(num_reads * 2))
 			fatal_error("Too many reads (%zu)", num_reads);
 		_vertices.resize(num_reads * 2);
+		_orig_num_reads = num_reads;
 	}
 
 	// Read this directed string graph from a file.

### 2012-11-30 src/StringGraph.h

@@ -16,11 +16,22 @@ class StringGraphEdge {
 protected:
 	float _mapped_read_count;
 	unsigned _num_inner_vertices;
+	float _A_statistic;
 
 	StringGraphEdge()
 	{
 		_mapped_read_count = 1.0;
 		_num_inner_vertices = 0;
+		_A_statistic = 0.0;
+	}
+
+	friend class boost::serialization::access;
+	template <class Archive>
+	void serialize(Archive & ar, unsigned version)
+	{
+		ar & _mapped_read_count;
+		ar & _num_inner_vertices;
+		ar & _A_statistic;
 	}
 public:
 	void increment_mapped_read_count(float n) { _mapped_read_count += n; }
@@ -30,6 +41,9 @@ public:
 	void increment_num_inner_vertices() { _num_inner_vertices++; }
 	unsigned get_num_inner_vertices() const { return _num_inner_vertices; }
 	void set_num_inner_vertices(unsigned n) { _num_inner_vertices = n; }
+
+	float get_A_statistic() const { return _A_statistic; }
+	void set_A_statistic(float f) { _A_statistic = f; }
 };
 
 // Base class for vertices of the string graph.
@@ -236,6 +250,8 @@ protected:
 	// Vector of the graph's edges.
 	std::vector<EDGE_t> _edges;
 
+	size_t _orig_num_reads;
+
 	// Serialize or deserialize this string graph to/from a stream.
 	friend class boost::serialization::access;
 	template <class Archive>
@@ -243,6 +259,7 @@ protected:
 	{
 		ar & _vertices;
 		ar & _edges;
+		ar & _orig_num_reads;
 	}
 
 	// Constructor is protected--- use DirectedStringGraph or

### 2012-11-30 src/util.h

@@ -67,5 +67,8 @@ extern long long parse_long(const char *optstr, const char *argument,
 #define DOUBLE_DIV_NONZERO(numerator, denominator) \
 	DIV_NONZERO(double(numerator), double(denominator))
 
+#define FLOAT_DIV_NONZERO(numerator, denominator) \
+	DIV_NONZERO(float(numerator), float(denominator))
+
 #define TO_PERCENT(numerator, denominator) \
 	(100.0 * DOUBLE_DIV_NONZERO(numerator, denominator))

### 2012-11-30 src/BidirectedStringGraph.cc
A-statistic calculation

@@ -234,6 +234,7 @@ void BidirectedStringGraph::build_from_digraph(const DirectedStringGraph & digra
 			}
 		}
 	}
+	_orig_num_reads = digraph._orig_num_reads;
 	info("Done building bidirected string graph from directed string graph");
 }
 

### 2012-11-30 src/DirectedStringGraph.cc

@@ -628,6 +628,7 @@ void DirectedStringGraph::build_from_bidigraph(const BidirectedStringGraph & bid
 			      e.get_seq_1_to_2(), 0, e.length() - 1, false,
 			      e.get_seq_2_to_1(), 0, e.length() - 1, false);
 	}
+	_orig_num_reads = bidigraph._orig_num_reads;
 }
 
 void DirectedStringGraph::calculate_A_statistics()
@@ -638,11 +639,10 @@ void DirectedStringGraph::calculate_A_statistics()
 	size_t bootstrap_genome_len = 0;
 	size_t bootstrap_num_reads = num_reads;
 	size_t n_edges = num_edges();
-	size_t i = 0;
-	while (i < n_edges) {
-		bootstrap_genome_len += _edges[i].length();
-		i += 2;
-	}
+	for (size_t i = 0; i < n_edges; i++)
+		if (_edges[i].get_v1_idx() & 1)
+			bootstrap_genome_len += _edges[i].length();
+
 	float global_arrival_rate = FLOAT_DIV_NONZERO(bootstrap_num_reads,
 						      bootstrap_genome_len);
 
@@ -650,10 +650,14 @@ void DirectedStringGraph::calculate_A_statistics()
 	     "bootstrap_num_reads = %zu, global_arrival_rate = %f",
 	     bootstrap_genome_len, bootstrap_num_reads, global_arrival_rate);
 
-	const size_t NUM_BOOTSTRAP_ITERATIONS = 10;
+	const size_t NUM_BOOTSTRAP_ITERATIONS = 3;
 	const float SINGLE_COPY_THRESHOLD = 17.0;
 
 	for (size_t i = 0; i < NUM_BOOTSTRAP_ITERATIONS; i++) {
+		size_t num_unique_edges = 0;
+		size_t num_optional_edges = 0;
+		size_t num_required_edges = 0;
+
 		bootstrap_genome_len = 0;
 		bootstrap_num_reads = 0;
 		for (size_t j = 0; j < n_edges; j++) {
@@ -663,11 +667,16 @@ void DirectedStringGraph::calculate_A_statistics()
 			float A_statistic = (global_arrival_rate * edge_len) -
 					     (edge_reads * M_LN2);
 			e.set_A_statistic(A_statistic);
-			info("edge_len = %zu, edge_reads = %u, A_statistic = %f",
-			     edge_len, edge_reads, A_statistic);
[…]

### 2012-11-30 src/StringGraph.h

@@ -250,7 +250,9 @@ protected:
 	// Vector of the graph's edges.
 	std::vector<EDGE_t> _edges;
 
+public:
 	size_t _orig_num_reads;
+protected:
 
 	// Serialize or deserialize this string graph to/from a stream.
 	friend class boost::serialization::access;

### 2012-11-30 src/BidirectedStringGraph.cc
Fix BidirectedStringGraph::build_from_digraph()

@@ -224,6 +224,8 @@ void BidirectedStringGraph::build_from_digraph(const DirectedStringGraph & digra
 				e.set_dirs(dirs);
 				e.set_mapped_read_count((w_v.get_mapped_read_count() +
 							 v_w.get_mapped_read_count()) / 2);
+				e.set_A_statistic((w_v.get_A_statistic() +
+						   v_w.get_A_statistic()) / 2);
 				assert(w_v.get_num_inner_vertices() ==
 				       v_w.get_num_inner_vertices());
 				e.set_num_inner_vertices(w_v.get_num_inner_vertices());

### 2012-11-30 src/BidirectedStringGraph.h
Fix edge printing

@@ -181,12 +181,13 @@ public:
 		char head_2 = v2_inward() ? '>' : '<';
 		if (read_1_idx != v_idx) {
 			std::swap(read_1_idx, read_2_idx);
-			head_1 = (head_1 == '>') ? '<' : '>';
-			head_2 = (head_2 == '>') ? '<' : '>';
+			head_1 = (head_2 == '>') ? '<' : '>';
+			head_2 = (head_1 == '>') ? '<' : '>';
 		}
 		os << (read_1_idx + 1) << ' ' << head_1 << "---------"
 		   << head_2 << ' ' << (read_2_idx + 1)
 		   << '\t';
+		StringGraphEdge::print(os);
 		if (print_seqs)
 			os << _seq_1_to_2;
 		else

### 2012-11-30 src/DirectedStringGraph.h

@@ -106,6 +106,7 @@ public:
 		os << read_1_idx << '.' << read_1_dir << " -> "
 		   << read_2_idx << '.' << read_2_dir
 		   << '\t';
+		StringGraphEdge::print(os);
 		if (print_seqs)
 			os << _seq;
 		else

### 2012-11-30 src/StringGraph.h

@@ -25,6 +25,11 @@ protected:
 		_A_statistic = 0.0;
 	}
 
+	void print(std::ostream & os) const {
+		os << "_num_inner_vertices=" << _num_inner_vertices << '\t';
+		os << "_A_statistic=" << _A_statistic << '\t';
+	}
+
 	friend class boost::serialization::access;
 	template <class Archive>
 	void serialize(Archive & ar, unsigned version)

### 2012-12-04 presentation/presentation.tex
A-statistic slide

@@
 
+\begin{frame}{Arrival rate statistic (A-statistic)}
+	\begin{itemize}
+		\item The {\bf A-statistic} $A(e)$ of an edge $e$ in a
+		bidirected string graph is:
+	\end{itemize}
+	\[ A(e) = \frac{n}{G} \Delta - \log_2{k} \]
+	where $n$ is the total number of reads, $G$ is the estimated genome
+	length, $\Delta$ is the length of the edge in bp, and $k$ is the number
+	of reads that support that edge.
+	\begin{itemize}
+		\item The A-statistic is the logarithm of the probability that
+		an edge represents unique sequence rather than duplicate
+		sequence.
+	\end{itemize}
+\end{frame}
+
 \begin{frame}{Implementation}

### 2012-12-04 presentation/presentation.tex
Script and direction fix

@@
 		\item Each read may be used in forward or reverse-complement
-		orientation, depending on the orientation of the arrow head when the
-		corresponding vertex is entered.
+		orientation, depending on the direction, relative to the walk,
+		of the adjacent arrow heads when the corresponding vertex is
+		passed through.
 	\end{itemize}

### 2012-12-04 presentation/presentation.tex
Modify read overlaps

@@
 		\begin{tikzpicture}[>=\ReadArrowType]
-				\draw[<-,style=very thick] (0.0, 0.0) -- (4.0, 0.0);
-				\draw[->,style=very thick] (1.2, -0.3) -- (5.2, -0.3);
-				\draw[<-,style=very thick] (2.4, -0.6) -- (6.4, -0.6);
-				\draw[<-,style=very thick] (3.6, -0.9) -- (7.6, -0.9);
+				\draw[<-,style=very thick] (0.0, 0.0) -- (2.6, 0.0);
+				\draw[->,style=very thick] (1.6, -0.3) -- (4.2, -0.3);
+				\draw[<-,style=very thick] (3.2, -0.6) -- (5.8, -0.6);
+				\draw[<-,style=very thick] (4.8, -0.9) -- (7.4, -0.9);
 				\draw (-0.3, 0.0) node {1.};
@@
 	\end{itemize}
-	\[ A(e) = \frac{n}{G} \Delta - \log_2{k} \]
+	\[ A(e) = \frac{n}{G} \Delta - \ln{2} \cdot {k} \]
 	where $n$ is the total number of reads, $G$ is the estimated genome
@@
 		\begin{itemize}
-			\item A set of C++ programs that iteratively transform the data into
-			the final assembly.
+			\item A set of C++ programs that iteratively transform
+			the data on-disk into the final assembly.
 			\item Yellow programs are not yet implemented.

### 2012-12-04 presentation/presentation.tex
exited

@@
 		edges in $G$ such that if any vertex $v$ is entered through a head
-		inwards, it is left on a head outwards (unless it is the end of the
+		inwards, it is exited on a head outwards (unless it is the end of the
 		path), and vice versa.

### 2012-12-04 presentation/paper.tex
paper template

\documentclass[10pt]{article}
%\usepackage{times}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
%\usepackage{threeparttable}
%\usepackage{hyperref}
\usepackage{graphicx}
%\usepackage{listings}

\title{A Bidirected String Graph Model for Genome Assembly}
\author{Eric Biggers\footnote{Macalester College}}

\begin{document}
\maketitle

%\begin{abstract}
%Sequencing an organism's genome is an important step to further genetic research
%on that organism.  However, current sequencing machines can only sequence very
%short fragments of DNA that originate from random positions on the original
%genome, assembly software such as Allpaths-LG can be used to assemble these
%fragments into longer contiguous sequences.  
%\end{abstract}

\section{Introduction to genome sequencing and assembly}

\subsection{Definition of a genome}
\subsection{Overall workflow for genome sequencing}
\subsection{Genome assemblers}

\section{Genome assembly vs. shortest common superstring}

\section{Overview of fragment string graph algorithm}

\section{Overlaps}

\section{Building the fragment string assembly graph}

\section*{Acknowledgments}

Acknowledgements

\bibliographystyle{plain}
\bibliography{refs}

\end{document}

### 2012-12-04 presentation/presentation.tex
update slid

@@
 \usepackage{grffile}
+%\usepackage[dvips]{graphics}
 
@@
 \begin{frame}{Walking through the bidirected string graph (cont.)}
-	\vspace{-0.7cm}
 	\begin{center}
-		\begin{tabular}{p{3.4cm}c}
-			\vspace{-0.4cm}
+		\vspace{-1cm}
+		\begin{tabular}{p{3.6cm}c}
+			\vspace{-1.5cm}
 			{\small a.) A bidirected string graph fragment} &
 			\begin{tikzpicture}[>=triangle 45,scale=1.5]
-				\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
-				\node (f) at (0, 0) {f};
-				\node (g) at (1.7, 0) {g};
+				\node[circle,fill=blue!20,draw=black] (f) at (0, 0) {f};
+				\node[circle,fill=blue!20,draw=black] (g) at (2.0, 0) {g};
 				\draw[>->,style=very thick] (f) edge (g);
+				\draw (1.00, 0.1) node[anchor=south] {\small \tt TTAGAAGTA};
+				\draw (1.00, -0.1) node[anchor=north] {
+						\rotatebox{180}{\small \tt
+						TGTGTAATC}};
+				\draw[->] (0.5, 0.5) ..
+					controls (0.7, 0.7) and (1.3, 0.7) ..
+					(1.5, 0.5);
+				\draw[->] (1.5, -0.5) ..
+					controls (1.3, -0.7) and (0.7, -0.7) ..
+					(0.5, -0.5);
 			\end{tikzpicture}
-			\\
-			\vspace{-1.5cm}
-			{\small b.) The corresponding overlap (labeled)} &
+			\\ \hline
+			\vspace{-2.3cm}
+			{\small b.) The corresponding overlap (labeled)
+			%sequence
+			%of overlapped region arbitrary as it is not known purely
+			%from the graph fragment)
+			} &
 			\begin{tikzpicture}[>=\ReadArrowType,scale=0.8]
@@
 						-> (4, 0) node[anchor=west] {$f.E$};
-				\draw[->,very thick] (2, -0.7) node[anchor=east] {$g.B$}
-						-> (6, -0.7) node[anchor=west] {$g.E$};
-				\draw (2, -1.05) node[anchor=north] {$o.g_{begin}$}
-						-- (2, 0.35) node[anchor=south] {$o.f_{begin}$};
-				\draw (4, -1.05) node[anchor=north] {$o.g_{end}$}
-						-- (4, 0.35) node[anchor=south] {$o.f_{end}$};
+				\draw[->,very thick] (2.09, -0.9) node[anchor=east] {$g.B$}
+						-> (6.09, -0.9) node[anchor=west] {$g.E$};
+
+				\draw (0, -0.1) node[anchor=south west]
+					{\hspace*{-1.4mm}{\small \tt
+						%GATTACACAGTGTAGTT
+						GATTACACA\hspace*{0.1cm}$a_1 \dots a_n$
+					}};
+
+				\draw (0, 0.1) node[anchor=north west]
+					{\hspace*{-2.6mm}
+						\rotatebox{180}{{\small \tt TGTGTAATC}}};
+
+				\draw (2.09, -1.0) node[anchor=south west]
+					{\hspace*{-1.4mm}{\small \tt
+						%GTGTAGTTTTAGAAGTA
+						\hspace*{0.1cm}$a_1 \dots a_n$\hspace*{0.25cm}TTAGAAGTA
+					}};
+
+				\draw (2.09, -0.9) node[anchor=north west]
+					{\hspace*{12.48mm}
+						\rotatebox{180}{{\small \tt TACTTCTAA}}};
+
+				\draw (2.09, -1.4) node[anchor=north] {$o.g_{begin}$}
+						-- (2.09, 0.6) node[anchor=south] {$o.f_{begin}$};
+				\draw (3.99, -1.4) node[anchor=north] {$o.g_{end}$}
+						-- (3.99, 0.6) node[anchor=south] {$o.f_{end}$};
 			\end{tikzpicture}
@@
 	\end{tabular}
-	\vspace{0.5cm}
-	\begin{itemize}
-		\item If we take {\bf each end} of {\bf each read} as a vertex, we can
-		make a directed graph equivalent to the bidirected graph.
-	\end{itemize}
+	%\vspace{0.5cm}
+	%\begin{itemize}
+		%\item If we take {\bf each end} of {\bf each read} as a vertex, we can
+		%make a directed graph equivalent to the bidirected graph.
+	%\end{itemize}
 \end{frame}

### 2012-12-05 presentation/paper.tex
Introduction

@@
 \title{A Bidirected String Graph Model for Genome Assembly}
-\author{Eric Biggers\footnote{Macalester College}}
+\author{Eric Biggers}
 
@@
 %\end{abstract}
+\newcommand\Base[1]{{\tt #1}}
 
@@
 \subsection{Definition of a genome}
+
+Every living organism contains a collection of hereditary material called a {\em
+genome}.  Typically, each individual cell of a multicellular organism contains a
+complete copy of the genome.  A genome consists of one or more DNA ({\it
+deoxyribonucleic acid}) molecules in the form of {\it chromosomes}.  DNA is a
+double-stranded molecule, each strand of which is a polymer of {\em
+nucleotides}.  A nucleotide, together with the corresponding nucleotide on the
+opposite strand, is called a {\it base pair}.  In DNA, there are four possible
+nucleotides: adenine, cytosine, guanine, and thymine, which are abbreviated as
+\Base{A}, \Base{C}, \Base{G}, and \Base{T}, respectively.  \Base{A} always pairs
+with T on the other strand, and \Base{C} always pairs with \Base{G} on the other
+strand.  In addition, each strand has a direction, and the two strands of a DNA
+molecule are directed opposite one another.  Furthermore, a DNA molecule may be
+circular (as is usually the case for bacterial chromosomes), in which case the
+molecule has no ends.
+
+Therefore, for the purposes of this paper, a genome is defined as a set of
+possibly circular, dual-stranded strings of the symbols \Base{A}, \Base{C},
+\Base{G}, and \Base{T}, where dual-stranded has the meaning indicated above (the
+two strands run in opposite directions, and the pairings \Base{A}---\Base{T} and
+\Base{C}---\Base{G} always hold.)
+
+This definition does not account for the fact that bacterial chromosomes are
+circular, nor does it account for the fact that diploid and polyploid organisms
+contain multiple, slightly-differing copies of each chromosome.  For the latter
+case, the above definition will consider each copy of the chromosome to be a
+completely different string in the genome.
+
+Genomes vary widely in size, from several thousand base pairs to many billion
+base pairs, depending on the organism.  (See Table \ref{tab:GenomeSizes}).
+
+\begin{table}[H]
+	\begin{center}
+		\begin{tabular}{|l|p{4.5cm}|}
+			%\hline
+			%$\phi$X174 bacteriophage &  $\sim$5,000 bp \\
+			%\hline
+			%Human mitochondrion &  $\sim$16,000 bp \\
+			\hline
+			{\bf Organism} & {\bf Approximate genome size (in base
+			pairs)} \\
+			\hline
+			{\it Variola} (smallpox) virus &  186,000 \\
+			\hline
+			{\it E. coli} (a common bacterium) & 4,600,000 \\
+			\hline
+			{\it Ananas comosus} (domesticated pineapple) & 500,000,000 \\
+			\hline
+			{\it Zea mays} (domesticated corn) & 2,000,000,000 \\
+			\hline
+			{\it Homo sapiens} (human)       &  3,200,000,000 \\
+			\hline
+		\end{tabular}
+	\end{center}
+	\caption{Approximate genome sizes of selected organisms}
+	\label{tab:GenomeSizes}
+\end{table}
+
+\subsection{Genome sequencing}
+
+{\it DNA sequencing} is the process of determining the base pair sequence of one
+or more DNA molecules.  {\it Genome sequencing} is the use of DNA sequencing to
+determine the base pair sequence of an entire genome.
+
 \subsection{Overall workflow for genome sequencing}

### 2012-12-05 presentation/paper.tex
Paper updates

@@
 \usepackage{graphicx}
+%\usepackage{hyperref}
 %\usepackage{listings}
@@
 nucleotides}.  A nucleotide, together with the corresponding nucleotide on the
-opposite strand, is called a {\it base pair}.  In DNA, there are four possible
-nucleotides: adenine, cytosine, guanine, and thymine, which are abbreviated as
-\Base{A}, \Base{C}, \Base{G}, and \Base{T}, respectively.  \Base{A} always pairs
-with T on the other strand, and \Base{C} always pairs with \Base{G} on the other
-strand.  In addition, each strand has a direction, and the two strands of a DNA
-molecule are directed opposite one another.  Furthermore, a DNA molecule may be
-circular (as is usually the case for bacterial chromosomes), in which case the
-molecule has no ends.
+opposite strand, is called a {\it base pair} (abbreviation: {\it bp}).  In DNA,
+there are four possible nucleotides: adenine, cytosine, guanine, and thymine,
+which are abbreviated as \Base{A}, \Base{C}, \Base{G}, and \Base{T},
+respectively.  \Base{A} always pairs with T on the other strand, and \Base{C}
+always pairs with \Base{G} on the other strand.  In addition, each strand has a
+direction, and the two strands of a DNA molecule are directed opposite one
+another.  Furthermore, a DNA molecule may be circular (as is usually the case
+for bacterial chromosomes), in which case the molecule has no ends.
 
@@
 
-{\it DNA sequencing} is the process of determining the base pair sequence of one
-or more DNA molecules.  {\it Genome sequencing} is the use of DNA sequencing to
-determine the base pair sequence of an entire genome.
+{\it DNA sequencing} is the process of determining the base pair sequence of
+({\it sequencing}) one or more DNA molecules.  {\it Genome sequencing} is the use
+of DNA sequencing to determine the base pair sequence of an entire genome.
 
-\subsection{Overall workflow for genome sequencing}
-\subsection{Genome assemblers}
+Sequencing a DNA molecule is very difficult because it exists at the molecular
+level, which makes each base pair much too small to be directly observed.
+Therefore, indirect laboratory techniques must be used in order to sequence a
+DNA molecule.  Advances in these techniques are progressing rapidly and they can
+now be performed by automated machines, but virtually every technique suffers
+from the limitation that only short fragments of DNA (several dozen to several
+thousand bp long) can be sequenced.  Therefore, to sequence an entire genome,
+which may be millions of bp long, the DNA from the genome must be broken into
+much smaller fragments, then those small fragments must be sequenced.  The
+original genome is then computationally reconstructed in a process called {\it
+genome assembly}.  {\it Genome sequencing} therefore refers to the entire
+process of sequencing a genome, including lab work and computational work, while
+{\it genome assembly} only refers to the computational work.
+
+\subsection{Genome assembly}
+
+\label{subsec:reads}
+
+The problem of genome assembly is to reconstruct a genome, as defined earlier,
+given a set of substrings from the genome that may originate from any strand of
+any DNA molecule in the genome.  Each substring is called a {\it read} and
+corresponds to some DNA that was sequenced by the sequencing machine.  In
+general, reads may contain errors, including {\it substitutions}, where a base
+pair in the read is incorrectly read as some other base pair, as well as {\it
+insertions}, where a sequence of one or more random base pairs have been
+inserted into some position in the read, and {\it deletions}, where a sequence
+of one or more base pairs from the original genome is missing from the interior
+of the read.  Reads are expected to be uniformly distributed on the genome with
+a certain {\it coverage} {\it coverage} $X$, meaning that each base pair in the
+original genome is expected to be included in $X$ different reads, on average.
+However, some sequencing technologies produce highly biased data where the
+coverage of genome regions is much more variable than that expected by uniform,
+random sampling.
+
+Genomes may contain {\it repeats}, which are subsequences that are repeated
+multiple times in the genome.  Repeated sequences may be longer than the length
+of any individual read, which makes the genome more difficult to assemble.
+
+Reads may contain other information, such as {\it quality scores} or {\it
+mate-pair} information, that provide additional information to the assembler.
+
+For the purposes of this paper, the problem of genome assembly is simplified by
+assuming error-free reads that are uniformly sampled from the genome.  In
+addition, the reads are assumed to be unpaired (meaning that mate-pair
+information, if available, is not used).
 
@@
 
+Genome assembly is superficially similar to the shortest common superstring
+(SCS) problem, which is to find the shortest possible string that contains every
+string in a given set as substrings.  Genome assembly also requires finding a
+longer string that contains all of a set of substrings.  However, the actual
+genome may be more than one string, and the genome may be longer than the
+shortest possible genome due to repeats.  In addition,  genomes are
+double-stranded, and the reads may have the complications described in
+\ref{subsec:reads}, such as sequencing error.  Therefore, genome assembly is not
+the same as the SCS problem, and in fact seems to be harder, despite the fact
+that the SCS problem is already NP-complete\cite{Turner1989}.
+
+In fact, some models of genome assembly have also been proven to be NP-complete
+\cite{Medvedev2007}.  However, it is definitely not acceptable to have a genome
+assembly algorithm that takes exponential time to complete.  Instead, heuristic
+algorithms that tend to produce good results but also tend to have a reasonable
+running time must be designed.
+
 \section{Overview of fragment string graph algorithm}
+
+This paper explores the genome assembly problem through an algorithm published
+in 2005 by Eugene Myers\cite{Myers2005} that is based on the idea of the
+fragment string assembly graph.  The algorithm provides a good overview of the
+techniques that are used in genome assembly, especially since it draws on ideas
+used in previous assemblers, such as the Celera assembler which was used to
+assemble the human genome for the first time\cite{Venter2001}.  However, the
+algorithm also introduces some new ideas.
+
+The overall algorithm is as follows: first, use overlaps between reads to build
+a graph that models the assembly problem.  Next, simplify and analyze this
+graph.  Finally, find paths through this graph to reconstruct the original
+genome.
 

### 2012-12-05 presentation/presentation.tex
Presentation updates

@@
 		\item A genome (for our purposes) is a long string of A's, C's, G's, and T's
-		\item Genomes are dual-stranded
+		\item Genomes are double-stranded
 		\begin{itemize}
@@
 		ideas
+		\item Not the same as de Bruijn graph-based approaches
 		\item The fundamental algorithm:
@@
 		\item We must allow the edges to be traversed in either direction.
-		\item However, each read must still be used in a consistent way within
-		any single assembly.
+		\item However, each read must still be used in a consistent way
+		at a location within any single assembly.
 		\item Solution: use a {\bf bidirected graph}.
@@
 		edges in $G$ such that if any vertex $v$ is entered through a head
-		inwards, it is exited on a head outwards (unless it is the end of the
+		inwards, it is exited through a head outwards (unless it is the end of the
 		path), and vice versa.

### 2012-12-06 presentation/paper.tex
Paper, overlaps

--- 2012-12-04 presentation/paper.tex
@@
 \documentclass[10pt]{article}
-%\usepackage{times}
 \usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
 \usepackage{float}
-%\usepackage{threeparttable}
-%\usepackage{hyperref}
 \usepackage{graphicx}
-%\usepackage{listings}
+\usepackage{tikz}
+%\usepackage{amsthm}
+
+\usetikzlibrary{arrows}
+
+%\theoremstyle{definition}
+%\newtheorem{definition}{Definition}
+
+\newcommand{\LengthVar}{L}
+\newcommand\Base[1]{{\tt #1}}
+\newcommand{\ReadArrowType}{latex}
 
 \title{A Bidirected String Graph Model for Genome Assembly}
-\author{Eric Biggers\footnote{Macalester College}}
+\author{Eric Biggers}
 
@@
 \subsection{Definition of a genome}
-\subsection{Overall workflow for genome sequencing}
-\subsection{Genome assemblers}
+
+Every living organism contains a collection of hereditary material called a {\em
+genome}.  Typically, each cell of a multicellular organism contains a complete
+copy of that organism's genome.  A genome consists of one or more DNA ({\it
+deoxyribonucleic acid}) molecules in the form of {\it chromosomes}.  DNA is a
+double-stranded molecule, each strand of which is a polymer of {\em
+nucleotides}.  A nucleotide, together with the corresponding nucleotide on the
+opposite strand, is called a {\it base pair} (abbreviation: {\it bp}).  In DNA,
+there are four possible nucleotides: adenine, cytosine, guanine, and thymine,
+which are abbreviated as \Base{A}, \Base{C}, \Base{G}, and \Base{T},
+respectively.  \Base{A} always pairs with T on the other strand, and \Base{C}
+always pairs with \Base{G} on the other strand.  In addition, each strand has a
+direction, and the two strands of a DNA molecule are directed opposite one
+another.  Furthermore, a DNA molecule may be circular, as is usually the case
+for bacterial chromosomes.
+
+Based on these biological facts, a genome is formally defined, for the purposes
+of this paper, as a set of possibly circular, dual-stranded strings of the
+symbols \Base{A}, \Base{C}, \Base{G}, and \Base{T}, where dual-stranded has the
+meaning indicated above: the two strands run in opposite directions, and the
+pairings \Base{A}---\Base{T} and \Base{C}---\Base{G} always hold.
+
+This definition does not account for the fact that the genomes of diploid and
+polyploid organisms, including most plants and animals, contain multiple
+slightly-differing copies of each chromosome.  In this case, the above
+definition will consider each copy of each chromosome to be a completely
+different string in the genome, unless the copies are completely identical.
+
+Genomes vary widely in size, from several thousand base pairs to many billion
+base pairs, depending on the organism (see Table \ref{tab:GenomeSizes}).
+
+\begin{table}[H]
+	\begin{center}
+		\begin{tabular}{|l|p{4.5cm}|}
+			%\hline
+			%$\phi$X174 bacteriophage &  $\sim$5,000 bp \\
+			%\hline
+			%Human mitochondrion &  $\sim$16,000 bp \\
+			\hline
+			{\bf Organism} & {\bf Approximate genome size (in base
+			pairs)} \\
+			\hline
+			{\it Variola} virus (smallpox) &  186,000 \\
+			\hline
+			{\it E. coli} (a bacterium) & 4,600,000 \\
+			\hline
+			{\it Ananas comosus} (domesticated pineapple) & 500,000,000 \\
+			\hline
+			{\it Zea mays} (domesticated corn) & 2,000,000,000 \\
+			\hline
+			{\it Homo sapiens} (human)       &  3,200,000,000 \\
+			\hline
+		\end{tabular}
+	\end{center}
+	\caption{Approximate genome sizes of selected organisms}
+	\label{tab:GenomeSizes}
+\end{table}
+
+\subsection{Genome sequencing}
+
+{\it DNA sequencing} is the process of determining the base pair sequence of
+({\it sequencing}) one or more DNA molecules.  {\it Genome sequencing} is the use
+of DNA sequencing to determine the base pair sequence of an entire genome.
+
+Sequencing a DNA molecule is very difficult because it is in fact an actual
+molecule, which makes each base pair much too small to be directly observed.
+Therefore, indirect laboratory techniques must be used.  Advances in these
+techniques are progressing rapidly, and they can now be performed by automated
+machines.  However, every technique suffers from the limitation that only short
+fragments of DNA several dozen to several thousand base pairs long can be
+sequenced.  Therefore, to sequence an entire genome, which may be millions of
+base pairs long, the DNA from the genome must be broken into much smaller
+fragments, which are then sequenced.  Following this, the original genome is
+computationally reconstructed by a computer program called a {\it genome
+assembler} in a process called {\it genome assembly}.  {\it Genome sequencing}
+therefore refers to the entire process of sequencing a genome, including both
+lab work and computational work, while {\it genome assembly} only refers to the
+computational work.\footnote{It usually also takes some technology-specific
+computational work to actually prepare the sequences for assembly; for example,
+a series of digital photographs of clumps of fluorescently labeled molecules
+might need to be interpreted as reads using image-processing algorithms.
+However, these programs and/or algorithms generally have already been developed
+by the manufacturer of the sequencing machine and work fairly well, and ``genome
+assembly'' does not cover these algorithms, although genome assembly algorithms
+still may be tuned for specific sequencing technologies.}
+
+\subsection{Genome assembly}
+
+\label{subsec:reads}
+
+The goal of genome assembly is to reconstruct a genome, as defined earlier,
+given a set of substrings that may originate from any strand of any DNA molecule
+in the genome.  Each substring is called a {\it read} and is the sequence of
+some DNA that was sequenced by the sequencing machine.  In general, reads may
+contain errors, including {\it substitutions}, where a base pair in the read was
+incorrectly read as some other base pair, as well as {\it insertions}, where a
+sequence of one or more random base pairs have been inserted into some position
+in the read, and {\it deletions}, where a sequence of one or more base pairs
+from the original genome is missing from the interior of the read.  Reads are
+expected to be uniformly distributed on the genome with a certain coverage $X$
+depending on how much data was collected, meaning that each base pair in the
+original genome is expected to be included in $X$ different reads on average,
+allowing for the reads to be from either strand.  However, some sequencing
+technologies produce biased data where the coverage across the genome is much
+more variable than that expected by uniform, random sampling.
+
+Reads may contain other information, such as {\it quality scores} or {\it
+mate-pair} information, that can provide additional hints to the assembler.
+
+It is also possible for reads to come from foreign DNA; for example, if a DNA
+sample is obtained from grinding up plant or animal tissue, there could be some
+bacterial cells, each with its own genome, contaminating the sample.  Therefore,
+ideally an assembler should be prepared to completely ignore some reads.
+However, this problem is not considered in this paper.
+
+For the purposes of this paper, the problem of genome assembly is simplified by
+assuming that the reads are error-free and uniformly (randomly) sampled from the
+genome.  In addition, it is assumed there are no reads from foreign DNA, and the
+reads are assumed to be unpaired, which means that mate-pair information, if
+available, is not used.  In other words, a read is defined here as an exact
+substring from either strand of any part of the genome.
+
+Genomes may contain {\it repeats}, which are subsequences that occur multiple
+times in the genome--- possibly adjacent to each other (a {\it tandem repeat}),
+but not necessarily.  Repeats may be longer than the length of any individual
+read, which makes the genome more difficult to assemble accurately.  However,
+real genomes almost always contain repeats of substantial length, so genome
+assembly algorithms must be able to handle repeats.
 
@@
 
-\section{Overview of fragment string graph algorithm}
+Genome assembly is superficially similar to the shortest common superstring
+(SCS) problem, which is to find the shortest possible string that contains every
+string in a given set as substrings.  Genome assembly also requires finding a
+longer string that contains all of a set of substrings.  However, the actual
+genome may be more than one string, and it genome may be longer than the
+shortest possible genome from which the reads could have been sampled, due to
+repeats.  In addition,  genomes are double-stranded, and there may be other
+complications with the reads (described in subsection \ref{subsec:reads}), such
+as sequencing error.  Therefore, genome assembly is not the same as the SCS
+problem, and in fact seems to be harder, despite the fact that the SCS problem
+is already NP-complete\cite{Turner1989}.
+
+In fact, some models of genome assembly have also been proven to be NP-complete
+\cite{Medvedev2007}.  However, it is not acceptable to have a genome assembly
+algorithm that takes exponential time to complete.  Instead, heuristic
+algorithms that tend to produce good results but also tend to have a reasonable
+running time must be designed.
+
+\section{Overview of the fragment string graph algorithm}
+
+This paper explores the genome assembly problem through an algorithm published
+in 2005 by Eugene Myers\cite{Myers2005} that is based on the idea of the
+fragment string assembly graph.  The algorithm provides a good overview of the
+techniques that are used in genome assembly, especially since it draws on ideas
+used in previous assemblers, such as the Celera assembler, which was used to
+assemble the human genome for the first time\cite{Venter2001}.  In addition, the
+algorithm shares some ideas with de Bruijn graph-based assemblers, such as
+EULER-SR\cite{Pevzner2001}, although it does not use a de Bruijn graph for its
+assembly model.  The algorithm also introduces some new ideas.
+
+The overall fragment string graph algorithm is as follows: first, use overlaps
+between reads to build a graph that models the assembly problem.  The idea is
+that walks through this graph model a way in which reads may be assembled
+together when the overlaps are used as evidence for joining two reads.  The next
+step is to simplify and analyze the graph; and finally, find paths through the
+graph are found to reconstruct the original genome.
 
 \section{Overlaps}
+
+%\begin{definition}
+	Consider two reads, $f$ and $g$, and their reverse-complements $f'$ and
+	$g'$.  The reads $f$ and $g$ are said to {\bf exactly overlap} by some
+	length $\LengthVar$ if and only if at least one of the following is
+	true:
+	\begin{enumerate}
+		\item The last $\LengthVar$ bp of $f$ exactly match the first
+		$\LengthVar$ bp of $g$.
+		\item The last $\LengthVar$ bp of $f$ exactly match the first
+		$\LengthVar$ bp of $g'$.
+		\item The last $\LengthVar$ bp of $f'$ exactly match the first
+		$\LengthVar$ bp of $g$.
+		\item The last $\LengthVar$ bp of $f'$ exactly match the first
+		$\LengthVar$ bp of $g'$.
+	\end{enumerate}
+%\end{definition}
+
+Having four different conditions is required because reads may overlap in a
+reverse-complement manner.  An overlap is reverse-complement if and only if one
+read in the overlap is used in the original direction and the other read is used
+in the reverse-complement direction.  Note that an overlap from condition (4) is
+{\it not} a reverse-complement overlap, as it can equivalently be stated as the
+last $\LengthVar$ bp of $g$ exactly match the last $\LengthVar$ bp of $f$.
+
+A read may overlap with itself; for example, a read may contain a suffix that
+matches its prefix of the same length.  Such an overlap is called a {\it
+self-overlap}.
+
+Overlaps may also be {\it inexact}.  An inexact overlap is an overlap where
+overlapped regions of the two reads do not match exactly (in forward or
+reverse-complement form), but instead differ by a number of base-pair
+substitutions, insertions, and/or deletions up to a predetermined upper bound,
+such as 5\%.  However, inexact overlaps will not be considered further in this
+paper.
+
+An overlap $o$ is fully described by the 7-tuple $(f,\ f_{begin},\ f_{end},\ g,
+\ g_{begin},\ g_{end},\ rc)$, where $f$ and $g$ are the two reads involved in the
+overlap, $f_{begin}$ and $f_{end}$ are the starting and ending indices of the
+overlapped region on the $f$ read, $g_{begin}$ and $g_{end}$ are the starting
+and ending indices of the overlapped region on the $g$ read, and $rc$ is a
+boolean that indicates whether the overlap is reverse-complement or not.  An
+overlap therefore has two overlapped regions--- one on each read.  For exact
+overlaps, the two overlapped regions are the same length; in other words,
+$g_{end} - g_{begin} = f_{end} - f_{begin}$.
+
+An {\it extreme point} occurs when an overlapped region extends to the beginning
+or the end of the corresponding read.  An overlap always contains at least two
+extreme points.
+
+A {\it containing overlap} is an overlap in which the overlapped region of one
+read, the {\it contained read}, is the entire read.  A special case of a
+containing overlap is an overlap where the overlapped regions include the
+entirety of both reads.  For an exact overlap, this can only happen if the two
+reads are either identical or exact reverse-complements of each other.
+
+The reason we care about overlaps is that two reads that overlap are likely to
+come from adjacent positions on the genome.  For example, consider
+
+\begin{center}
+	\noindent {\tt Read 1: \ \ \ \ \ \ \ \ \ $\rightarrow$
+		ATATAT{GCTGGTTACTT} $\rightarrow$ } \\
+	{\tt Read 2: \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ $\rightarrow$
+		{GCTGGTTACTT}TGATAGATA $\rightarrow$ } \\
+	{\tt Possible genome: $\rightarrow$ ATATATGCTGGTTACTTTGATAGATA $\rightarrow$}
+\end{center}
+
+Overlapping reads may be from different strands, in which case the
+overlapped regions are reverse-complement from each other.
+
+\begin{center}
+	\noindent {\tt Read 1: \ \ \ \ \ \ \ \ \ $\rightarrow$
+		ATATAT{GCTGGTTACTT} $\rightarrow$ } \\
+	{\tt Read 2: \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ $\leftarrow$
+		{CGACCAATGAA}ACTATCTAT $\leftarrow$ } \\
+	{\tt Possible genome: $\rightarrow$ ATATATGCTGGTTACTTTGATAGATA $\rightarrow$}
+\end{center}
+
+Overall, there are three types of overlaps:
+
+\begin{enumerate}
+	\item ``Normal'' overlap
+		\begin{center}
+			\vspace{-0.5cm}
+			\begin{tikzpicture}[>=\ReadArrowType]
+				\draw[->,style=thick] (0,0) -- (2,0);
+				\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
+			\end{tikzpicture}
+
+			\vspace{3mm}
+			or (symmetrically)
+			\vspace{3mm}
+
+			\begin{tikzpicture}[>=\ReadArrowType]
+				\draw[->,style=thick] (2,0) -- (0,0);
+				\draw[->,style=thick] (3.0,0.3) -- (1.0,0.3);
+			\end{tikzpicture}
+		\end{center}
+	\vspace{0.5cm}
+	\item ``Innie'' overlap
+		\begin{center}
+			\vspace{-0.5cm}
+			\begin{tikzpicture}[>=\ReadArrowType]
+				\draw[->,style=thick] (0,0) -- (2,0);
+				\draw[->,style=thick] (3,0.3) -- (1,0.3);
+		\end{tikzpicture} \end{center}
+	\vspace{0.5cm}
+	\item ``Outie'' overlap
+		\begin{center}
+		\vspace{-0.5cm}
+		\begin{tikzpicture}[>=\ReadArrowType]
+				\draw[->,style=thick] (2,0) -- (0,0);
+				\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
+		\end{tikzpicture} \end{center}
+\end{enumerate}
+
+Now that we've formalized the definition of an overlap, what we'd like to do is
+compute
 

### 2012-12-06 presentation/paper.tex
overlap

@@
 \usepackage{tikz}
+\usepackage{subcaption}
+\usepackage{algorithm}
+\usepackage{algorithmic}
+\usepackage{comment}
 %\usepackage{amsthm}
@@
 \usetikzlibrary{arrows}
+
+%\floatstyle{boxed}
+%\restylefloat{figure}
 
@@
 \newcommand{\ReadArrowType}{latex}
+\newcommand{\Figure}[1]{Figure \ref{#1}}
 
@@
 fragments of DNA several dozen to several thousand base pairs long can be
-sequenced.  Therefore, to sequence an entire genome, which may be millions of
+sequenced.
+
+Therefore, to sequence an entire genome, which may be millions of
 base pairs long, the DNA from the genome must be broken into much smaller
@@
 
+\subsection{Definition of an overlap}
 %\begin{definition}
@@
 last $\LengthVar$ bp of $g$ exactly match the last $\LengthVar$ bp of $f$.
+\Figure{fig:overlap_types} shows the three fundamental types of overlaps that
+may arise from this definition.
 
@@
 
-The reason we care about overlaps is that two reads that overlap are likely to
-come from adjacent positions on the genome.  For example, consider
-
-\begin{center}
-	\noindent {\tt Read 1: \ \ \ \ \ \ \ \ \ $\rightarrow$
-		ATATAT{GCTGGTTACTT} $\rightarrow$ } \\
-	{\tt Read 2: \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ $\rightarrow$
-		{GCTGGTTACTT}TGATAGATA $\rightarrow$ } \\
-	{\tt Possible genome: $\rightarrow$ ATATATGCTGGTTACTTTGATAGATA $\rightarrow$}
-\end{center}
-
-Overlapping reads may be from different strands, in which case the
-overlapped regions are reverse-complement from each other.
-
-\begin{center}
-	\noindent {\tt Read 1: \ \ \ \ \ \ \ \ \ $\rightarrow$
-		ATATAT{GCTGGTTACTT} $\rightarrow$ } \\
-	{\tt Read 2: \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ $\leftarrow$
-		{CGACCAATGAA}ACTATCTAT $\leftarrow$ } \\
-	{\tt Possible genome: $\rightarrow$ ATATATGCTGGTTACTTTGATAGATA $\rightarrow$}
-\end{center}
-
-Overall, there are three types of overlaps:
-
-\begin{enumerate}
-	\item ``Normal'' overlap
-		\begin{center}
-			\vspace{-0.5cm}
-			\begin{tikzpicture}[>=\ReadArrowType]
+\begin{figure}
+	\centering
+	\begin{subfigure}[b]{0.3\textwidth}
+		\begin{tikzpicture}=\ReadArrowType]
 			\draw[->,style=thick] (0,0) -- (2,0);
@@
 		\end{tikzpicture}
+		%\vspace{3mm}
+		%or (symmetrically)
+		%\vspace{3mm}
 
-			\vspace{3mm}
-			or (symmetrically)
-			\vspace{3mm}
-
-			\begin{tikzpicture}[>=\ReadArrowType]
-				\draw[->,style=thick] (2,0) -- (0,0);
-				\draw[->,style=thick] (3.0,0.3) -- (1.0,0.3);
-			\end{tikzpicture}
-		\end{center}
-	\vspace{0.5cm}
-	\item ``Innie'' overlap
-		\begin{center}
-			\vspace{-0.5cm}
-			\begin{tikzpicture}[>=\ReadArrowType]
+		%\begin{tikzpicture}=\ReadArrowType]
+			%\draw[->,style=thick] (2,0) -- (0,0);
+			%\draw[->,style=thick] (3.0,0.3) -- (1.0,0.3);
+		%\end{tikzpicture}
+		\caption{``Normal'' overlap}
+	\end{subfigure}
+	\begin{subfigure}[b]{0.3\textwidth}
+		\begin{tikzpicture}=\ReadArrowType]
 			\draw[->,style=thick] (0,0) -- (2,0);
 			\draw[->,style=thick] (3,0.3) -- (1,0.3);
-		\end{tikzpicture} \end{center}
-	\vspace{0.5cm}
-	\item ``Outie'' overlap
-		\begin{center}
-		\vspace{-0.5cm}
-		\begin{tikzpicture}[>=\ReadArrowType]
+		\end{tikzpicture}
+		\caption{``Innie'' overlap}
+	\end{subfigure}
+	\begin{subfigure}[b]{0.3\textwidth}
+		\begin{tikzpicture}=\ReadArrowType]
 			\draw[->,style=thick] (2,0) -- (0,0);
 			\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
-		\end{tikzpicture} \end{center}
-\end{enumerate}
+		\end{tikzpicture}
+		\caption{``Outie'' overlap}
+	\end{subfigure}
+	\caption{The three types of overlaps.  The innie and outie overlaps are
+	reverse-complement overlaps, while the normal overlap is not.  Note:
+	with the normal overlap, there is a symmetric case that is not
+	considered a distinct type of overlap here.}
+	\label{fig:overlap_types}
+\end{figure}
 
-Now that we've formalized the definition of an overlap, what we'd like to do is
-compute
+%\begin{enumerate}
+	%\item ``Normal'' overlap
+		%\begin{center}
+			%\vspace{-0.5cm}
+			%\begin{tikzpicture}[>=\ReadArrowType]
+				%\draw[->,style=thick] (0,0) -- (2,0);
+				%\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
+			%\end{tikzpicture}
+
+			%\vspace{3mm}
+			%or (symmetrically)
+			%\vspace{3mm}
+
+			%\begin{tikzpicture}[>=\ReadArrowType]
+				%\draw[->,style=thick] (2,0) -- (0,0);
+				%\draw[->,style=thick] (3.0,0.3) -- (1.0,0.3);
+			%\end{tikzpicture}
+		%\end{center}
+	%\vspace{0.5cm}
+	%\item ``Innie'' overlap
+		%\begin{center}
+			%\vspace{-0.5cm}
+			%\begin{tikzpicture}[>=\ReadArrowType]
+				%\draw[->,style=thick] (0,0) -- (2,0);
+				%\draw[->,style=thick] (3,0.3) -- (1,0.3);
+		%\end{tikzpicture} \end{center}
+	%\vspace{0.5cm}
+	%\item ``Outie'' overlap
+		%\begin{center}
+		%\vspace{-0.5cm}
+		%\begin{tikzpicture}[>=\ReadArrowType]
+				%\draw[->,style=thick] (2,0) -- (0,0);
+				%\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
+		%\end{tikzpicture} \end{center}
+%\end{enumerate}
+
+Overlaps are important because two reads that overlap may come from adjacent
+positions on the genome.  Although due to repeats, we cannot be certain that two
+overlapping reads actually come from adjacent positions, we {\it can} say for
+certain that two reads that do not overlap do not come from adjacent positions
+on the genome.  Regardless of this, \Figure{fig:overlap_examples} shows how two
+reads that overlap potentially could be combined to form a longer sequence.
+
+\begin{figure}
+		%\begin{tikzpicture}
+			%\draw (0, 0) node[anchor=left] {Read 1:}	
+			%\draw (0, -1) node[anchor=left] {Read 2:}	
+			%\draw (0, -2) node[anchor=left] {Possible genome:}
+		%\end{tikzpicture}
+	\centering
+	\begin{subfigure}[t]{0.455\textwidth}
+		1: \hspace{0.7mm}{\tt $\rightarrow$ATATAT{GCTGGTTACTT}$\rightarrow$ } \\
+		2: \hspace{0.7mm}{\tt \ \ \ \ \ \ $\rightarrow${GCTGGTTACTT}TGATAGATA$\rightarrow$ } \\
+		G: {\tt $\rightarrow$ATATATGCTGGTTACTTTGATAGATA$\rightarrow$}
+		\caption{A normal overlap between reads (1) and (2) and the
+		implied way in which the two reads could be joined together to
+		form a longer sequence (G)}
+	\end{subfigure}
+		\hspace{0.07\textwidth}
+	\begin{subfigure}[t]{0.455\textwidth}
+		1: \hspace{0.7mm}{\tt $\rightarrow$ATATAT{GCTGGTTACTT}$\rightarrow$ } \\
+		2: \hspace{0.7mm}{\tt \ \ \ \ \ \ $\leftarrow${CGACCAATGAA}ACTATCTAT$\leftarrow$ } \\
+		G: {\tt $\rightarrow$ATATATGCTGGTTACTTTGATAGATA$\rightarrow$}
+		\caption{A reverse-complement overlap between reads (1) and
+		(2)--- more specifically, an ``innie'' overlap--- and the
+		implied way in which the two reads could be joined together to
+		form a longer sequence (G)}
+	\end{subfigure}
+	\caption{Examples of normal and reverse-complement overlaps}
+	\label{fig:overlap_examples}
+\end{figure}
+
+\subsection{Computing overlaps}
+
+\subsubsection{Algorithm}
+
+Now that the idea of an exact overlap has been formalized, what we would like to
+do is compute all pairwise exact overlaps of minimum length $L$ among the reads
+that were provided as input to the assembler.
+
+\begin{comment}
+\begin{algorithm}
+\caption{{\sc ComputeExactOverlapsNaive}($S$, $L$)}
+\begin{algorithmic}
+	%\FOR{{\bf each} distinct}
+	\STATE b
+\end{algorithmic}
+\end{algorithm}
+
+\newcommand{\OccVar}{\omega}
+%\newcommand{\OccOne}{{\OccVar}_1}
+%\newcommand{\OccTwo}{{\OccVar}_2}
+\newcommand{\OccOne}{{\tt occ1}}
+\newcommand{\OccTwo}{{\tt occ2}}
+\newcommand{\SeedVar}{{\tt seed}}
+\newcommand{\OccList}{{\tt occurrenceSet}}
+\newcommand{\OverlapSet}{{\tt overlapSet}}
+\newcommand{\OccMap}{{\tt occurrenceMap}}
+\newcommand{\PosOne}{{\tt pos1}}
+\newcommand{\PosTwo}{{\tt pos2}}
+\newcommand{\Len}{{\tt len}}
+\newcommand{\RightExtend}{{\tt rightExtend}}
+\newcommand{\LeftExtend}{{\tt leftExtend}}
+\newcommand{\ReadSet}{{\tt readSet}}
+\newcommand{\minOverlapLen}{{\tt minOverlapLen}}
+\newcommand{\readOneBeg}{{\tt read1Begin}}
+\newcommand{\readOneEnd}{{\tt read1End}}
+\newcommand{\readTwoBeg}{{\tt read2Begin}}
+\newcommand{\readTwoEnd}{{\tt read2End}}
+\newcommand{\numExtremePoints}{{\tt numExtremePoints}}
+
+\begin{algorithm}
+\caption{{\sc ComputeExactOverlapsSeedAndExtend}($\ReadSet$, $\minOverlapLen$)}
+\begin{algorithmic}
+	%\REQUIRE{$S$ is a list of DNA sequences}
+	%\REQUIRE{$L$ is a positive integer that is the minimum overlap length}
+	\STATE{Let $k$ be a positive integer less than or equal to
+	$\minOverlapLen$.}
+	\STATE{Let $\OccMap$ be a hash table that maps DNA sequences to sets of
+		$k$-mer occurrences.}
+	\STATE{Let $\OverlapSet$ be an empty set of overlaps.}
+	\FOR{{\bf each} sequence $s \in \ReadSet$}
+		\FOR{{\bf each} canonical $k$-mer $K \in r$}
+			\STATE{Add the occurrence of $K$ to $\OccMap[K]$}
+		\ENDFOR
+	\ENDFOR
+	\FOR{{\bf each} $(\SeedVar,\ \OccList) \in \OccMap$}
+		\FOR{{\bf each} distinct pair of $k$-mer occurrences $\OccOne,
+		\OccTwo \in \OccList$}
+			\IF{the read ID of $\OccOne$ is greater than the read ID
+			of $\OccTwo$}
+				\STATE{Swap $\OccOne$ and $\OccTwo$}
+			\ENDIF
+			\IF{$\OccOne$ and $\OccTwo$ are both forward or both
+			reverse-complement}
+				\STATE{${\tt rc} \gets$ {\bf true}}
+			\ELSE
+				\STATE{${\tt rc} \gets$ {\bf false}}
+			\ENDIF
+			\STATE{$f \gets$ the read containing $\OccOne$}
+			\STATE{$g \gets$ the read containing $\OccTwo$}
+			\STATE{$\PosOne \gets$ the position within $f$ of $\OccOne$}
+			\STATE{$\PosOne \gets$ the position within $g$ of $\OccTwo$}
+			\STATE{$\Len \gets k$}
+			\IF{${\tt rc}$}
+				\STATE{$g' \gets$ the reverse complement of $g$}
+				\STATE{${\tt \LeftExtend} \gets$ the number of
+				additional bp to the left of the seed that
+				match, comparing $f$ with $g'$}
+				\STATE{${\tt \RightExtend} \gets$ the number of
+				additional bp to the right of the seed that
+				match, comparing $f$ with $g'$}
+				\STATE{$\PosTwo \gets \PosTwo - {\tt
+				\RightExtend}$}
+			\ELSE
+				\STATE{${\tt \LeftExtend} \gets$ the number of
+				additional bp to the left of the seed that
+				match, comparing $f$ with $g$}
+				\STATE{${\tt \RightExtend} \gets$ the number of
+				additional bp to the right of the seed that
+				match, comparing $f$ with $g$}
+				\STATE{$\PosTwo \gets \PosTwo - {\tt \LeftExtend}$}
+			\ENDIF
+			\STATE{$\PosOne \gets \PosOne - {\tt \LeftExtend}$}
+			\STATE{$\Len \gets \Len + {\tt \LeftExtend} + {\tt \RightExtend}$}
+			\IF{$\Len < \minOverlapLen$} 
+				\STATE{\bf continue}
+			\ENDIF
+			\STATE{$\readOneBeg \gets \PosOne$}
+			\STATE{$\readOneEnd \gets \PosOne + \Len - 1$}
+			\STATE{$\readTwoBeg \gets \PosTwo$}
+			\STATE{$\readTwoEnd \gets \PosTwo + \Len - 1$}
+			\STATE{$\numExtremePoints \gets 0$}
+			\IF{$\readOneBeg = 0$}
+				\STATE{$\numExtremePoints \gets \numExtremePoints + 1$}
+			\ENDIF
+			\IF{$\readOneEnd = {\sc Length}(f) - 1$}
+				\STATE{$\numExtremePoints \gets \numExtremePoints + 1$}
+			\ENDIF
+			\IF{$\readTwoBeg = 0$}
+				\STATE{$\numExtremePoints \gets \numExtremePoints + 1$}
+			\ENDIF
+			\IF{$\readTwoEnd = {\sc Length}(g) - 1$}
+				\STATE{$\numExtremePoints \gets \numExtremePoints + 1$}
+			\ENDIF
+
+			%\STATE{Maximally extend the occurrence of $\SeedVar$ at
+			%$\OccOne$ and $\OccTwo$ outwards in both directions}
+			%\STATE{${\tt pos1} \gets$ the beginning of the maximally
+			%matched region on the read containing $\OccOne$}
+			%\STATE{${\tt pos2} \gets$ the beginning of the maximally
+			%matched region on the read containing $\OccTwo$}
+			%\STATE{${\tt plen} \gets$ the length of the maximall
+			%matched region on the read containing $\OccTwo$}
+		\ENDFOR
+	\ENDFOR
+\end{algorithmic}
+\end{algorithm}
+\end{comment}
+
+\subsubsection{Implementation}
 

### 2012-12-06 presentation/paper.tex
paper: compute-overlaps

@@
 
-\subsubsection{Algorithm}
+\subsubsection{Algorithms}
 
@@
 that were provided as input to the assembler.
+
+The naive way to do this is to consider every pair of reads in the input,
+including only distinct pairs but also including the special cases of each read
+paired with itself, and search for each of the types of overlaps between the two
+reads in each pair.  If there are $N$ reads in the input, this algorithm will
+need to consider $\frac{N^2}{2} + N$ pairs of reads, and the time spent on each
+read pair will depend several factors, including the minimum overlap length and
+the read lengths.  But even if we could somehow spend constant time on each read
+pair, this algorithm is much too slow because the running time rises quadraticly
+with respect to the number of reads.  This is important because there can easily
+be millions of reads in the input.
+
+A number of faster algorithms for computing overlaps have been designed.  One of
+the easier ones to understand is the {\it seed-and-extend} algorithm.  It works
+by indexing the reads by short subsequences of a certain length (less than or
+equal to than the minimum overlap length) and then only comparing reads that
+actually share a subsequence.  Each of these short subsequences is called a {\it
+seed}.  Two reads may share a seed in either a forward or reverse-complement
+manner.  Given two reads that share a seed, the seed can be extended outwards in
+both directions in order to find a maximal matched region, that may be either
+forward or reverse-complement, between the two reads.  Then, this maximal
+matched region is compared against the definition of an overlap.  In particular,
+it must be at least as long as the minimum overlap length, it must have least 2
+extreme points over the two reads, and it cannot be the prefix or suffix of both
+reads if the match does not extend to the opposite end of at least one of the
+reads.  If all of these conditions are met and the same overlap was not already
+found through another seed in the same two reads, this overlap is added to the
+set of overlaps.
+
+The running time of the seed-and-extend algorithm, as described above, is
+typically much faster than the naive algorithm, although it is highly dependent
+on the data.  For a string of the genome of length $G$ and with a seed length
+$k$, there will be at most $G-k+1$ canonical seeds.\footnote{The canonical form
+of a DNA sequence and its reverse-complement is the lexicographically lesser of
+the two.  Here, we refer canonical form of each seed, because we wish to compare
+reads that share a seed in either a forward or reverse-complement manner.}  For
+each of these seeds, the reads containing it will be examined exhaustively in a
+pairwise manner, which takes quadratic time with respect to the number of reads
+sharing that seed.  Since the more densely reads are sampled from the genome,
+the more reads will contain each seed,  the running time of the seed-and-extend
+algorithm rises quadratically with regards to read density.  In addition, the
+number of read pairs the algorithm considers for a specific seed also rises
+quadratically with the number of times that seed is repeated in the genome due
+to genomic repeats.  Overall, the seed-and-extend algorithm is not ideal and is
+still a bottleneck of the entire assembly algorithm, despite the fact that it is
+much faster than the naive algorithm (except in degenerate cases, such as a
+``genome'' that is entirely a single repeated letter).
+
+\subsubsection{Implementation}
+
+For this project, the seed-and-extend algorithm for computing overlaps was
+implemented as the program {\tt compute-overlaps}.  It takes as input a file
+that contains a list of reads, as well as a parameter that specifies the minimum
+overlap length $L$.  It produces as output (via a data structure serialized to a
+file) all exact pairwise overlaps of minimum length $L$ among reads in the input
+file, including reverse-complement overlaps and self-overlaps.  The running time
+of the algorithm was discussed above.  The memory usage is linear with respect
+to the total length of the reads and the number of overlaps, which likely could
+be improved.
 
@@
 
-\subsubsection{Implementation}
-
 \section{Building the fragment string assembly graph}

### 2012-12-06 presentation/paper.tex
paper: bidigraph

@@
 \usepackage{comment}
+\usepackage{multicol}
 %\usepackage{amsthm}
@@
 \newcommand{\Figure}[1]{Figure \ref{#1}}
+
+\newcommand{\KeyTerm}[1]{{\it #1}}
+\newcommand{\ProgramName}[1]{{\tt #1}}
 
@@
 
-Every living organism contains a collection of hereditary material called a {\em
-genome}.  Typically, each cell of a multicellular organism contains a complete
-copy of that organism's genome.  A genome consists of one or more DNA ({\it
-deoxyribonucleic acid}) molecules in the form of {\it chromosomes}.  DNA is a
-double-stranded molecule, each strand of which is a polymer of {\em
-nucleotides}.  A nucleotide, together with the corresponding nucleotide on the
-opposite strand, is called a {\it base pair} (abbreviation: {\it bp}).  In DNA,
-there are four possible nucleotides: adenine, cytosine, guanine, and thymine,
-which are abbreviated as \Base{A}, \Base{C}, \Base{G}, and \Base{T},
-respectively.  \Base{A} always pairs with T on the other strand, and \Base{C}
-always pairs with \Base{G} on the other strand.  In addition, each strand has a
-direction, and the two strands of a DNA molecule are directed opposite one
-another.  Furthermore, a DNA molecule may be circular, as is usually the case
-for bacterial chromosomes.
+Every living organism contains a collection of hereditary material called a
+\KeyTerm{genome}.  Typically, each cell of a multicellular organism contains a
+complete copy of that organism's genome.  A genome consists of one or more DNA
+(\KeyTerm{deoxyribonucleic acid}) molecules in the form of
+\KeyTerm{chromosomes}.  DNA is a double-stranded molecule, each strand of which
+is a polymer of \KeyTerm{nucleotides}.  A nucleotide, together with the
+corresponding nucleotide on the opposite strand, is called a \KeyTerm{base pair}
+(abbreviation: \KeyTerm{bp}).  In DNA, there are four possible nucleotides:
+adenine, cytosine, guanine, and thymine, which are abbreviated as \Base{A},
+\Base{C}, \Base{G}, and \Base{T}, respectively.  \Base{A} always pairs with T on
+the other strand, and \Base{C} always pairs with \Base{G} on the other strand.
+In addition, each strand has a direction, and the two strands of a DNA molecule
+are directed opposite one another.  Furthermore, a DNA molecule may be circular,
+as is usually the case for bacterial chromosomes.
 
@@
 
-{\it DNA sequencing} is the process of determining the base pair sequence of
-({\it sequencing}) one or more DNA molecules.  {\it Genome sequencing} is the use
-of DNA sequencing to determine the base pair sequence of an entire genome.
+\KeyTerm{DNA sequencing} is the process of determining the base pair sequence of
+(\KeyTerm{sequencing}) one or more DNA molecules.  \KeyTerm{Genome sequencing}
+is the use of DNA sequencing to determine the base pair sequence of an entire
+genome.
 
@@
 
-Therefore, to sequence an entire genome, which may be millions of
-base pairs long, the DNA from the genome must be broken into much smaller
-fragments, which are then sequenced.  Following this, the original genome is
-computationally reconstructed by a computer program called a {\it genome
-assembler} in a process called {\it genome assembly}.  {\it Genome sequencing}
-therefore refers to the entire process of sequencing a genome, including both
-lab work and computational work, while {\it genome assembly} only refers to the
+Therefore, to sequence an entire genome, which may be millions of base pairs
+long, the DNA from the genome must be broken into much smaller fragments, which
+are then sequenced.  Following this, the original genome is computationally
+reconstructed by a computer program called a \KeyTerm{genome assembler} in a
+process called \KeyTerm{genome assembly}.  \KeyTerm{Genome sequencing} therefore
+refers to the entire process of sequencing a genome, including both lab work and
+computational work, while \KeyTerm{genome assembly} only refers to the
 computational work.\footnote{It usually also takes some technology-specific
@@
 given a set of substrings that may originate from any strand of any DNA molecule
-in the genome.  Each substring is called a {\it read} and is the sequence of
+in the genome.  Each substring is called a \KeyTerm{read} and is the sequence of
 some DNA that was sequenced by the sequencing machine.  In general, reads may
-contain errors, including {\it substitutions}, where a base pair in the read was
-incorrectly read as some other base pair, as well as {\it insertions}, where a
-sequence of one or more random base pairs have been inserted into some position
-in the read, and {\it deletions}, where a sequence of one or more base pairs
-from the original genome is missing from the interior of the read.  Reads are
-expected to be uniformly distributed on the genome with a certain coverage $X$
-depending on how much data was collected, meaning that each base pair in the
-original genome is expected to be included in $X$ different reads on average,
-allowing for the reads to be from either strand.  However, some sequencing
-technologies produce biased data where the coverage across the genome is much
-more variable than that expected by uniform, random sampling.
+contain errors, including \KeyTerm{substitutions}, where a base pair in the read
+was incorrectly read as some other base pair, as well as \KeyTerm{insertions},
+where a sequence of one or more random base pairs have been inserted into some
+position in the read, and \KeyTerm{deletions}, where a sequence of one or more
+base pairs from the original genome is missing from the interior of the read.
+Reads are expected to be uniformly distributed on the genome with a certain
+coverage $X$ depending on how much data was collected, meaning that each base
+pair in the original genome is expected to be included in $X$ different reads on
+average, allowing for the reads to be from either strand.  However, some
+sequencing technologies produce biased data where the coverage across the genome
+is much more variable than that expected by uniform, random sampling.
 
-Reads may contain other information, such as {\it quality scores} or {\it
-mate-pair} information, that can provide additional hints to the assembler.
+Reads may contain other information, such as \KeyTerm{quality scores} or
+\KeyTerm{mate-pair} information, that can provide additional hints to the
+assembler.
 
@@
 
-Genomes may contain {\it repeats}, which are subsequences that occur multiple
-times in the genome--- possibly adjacent to each other (a {\it tandem repeat}),
-but not necessarily.  Repeats may be longer than the length of any individual
-read, which makes the genome more difficult to assemble accurately.  However,
-real genomes almost always contain repeats of substantial length, so genome
-assembly algorithms must be able to handle repeats.
+Genomes may contain \KeyTerm{repeats}, which are subsequences that occur
+multiple times in the genome--- possibly adjacent to each other (a
+\KeyTerm{tandem repeat}), but not necessarily.  Repeats may be longer than the
+length of any individual read, which makes the genome more difficult to assemble
+accurately.  However, real genomes almost always contain repeats of substantial
+length, so genome assembly algorithms must be able to handle repeats.
 
@@
 in the reverse-complement direction.  Note that an overlap from condition (4) is
-{\it not} a reverse-complement overlap, as it can equivalently be stated as the
+{\em not} a reverse-complement overlap, as it can equivalently be stated as the
 last $\LengthVar$ bp of $g$ exactly match the last $\LengthVar$ bp of $f$.
@@
 A read may overlap with itself; for example, a read may contain a suffix that
-matches its prefix of the same length.  Such an overlap is called a {\it
-self-overlap}.
+matches its prefix of the same length.  Such an overlap is called a
+\KeyTerm{self-overlap}.
 
-Overlaps may also be {\it inexact}.  An inexact overlap is an overlap where
+Overlaps may also be \KeyTerm{inexact}.  An inexact overlap is an overlap where
 overlapped regions of the two reads do not match exactly (in forward or
@@
 
-An {\it extreme point} occurs when an overlapped region extends to the beginning
-or the end of the corresponding read.  An overlap always contains at least two
-extreme points.
+An \KeyTerm{extreme point} occurs when an overlapped region extends to the
+beginning or the end of the corresponding read.  An overlap always contains at
+least two extreme points.
 
-A {\it containing overlap} is an overlap in which the overlapped region of one
-read, the {\it contained read}, is the entire read.  A special case of a
+A \KeyTerm{containing overlap} is an overlap in which the overlapped region of
+one read, the \KeyTerm{contained read}, is the entire read.  A special case of a
 containing overlap is an overlap where the overlapped regions include the
@@
 positions on the genome.  Although due to repeats, we cannot be certain that two
-overlapping reads actually come from adjacent positions, we {\it can} say for
+overlapping reads actually come from adjacent positions, we {\em can} say for
 certain that two reads that do not overlap do not come from adjacent positions
@@
 A number of faster algorithms for computing overlaps have been designed.  One of
-the easier ones to understand is the {\it seed-and-extend} algorithm.  It works
-by indexing the reads by short subsequences of a certain length (less than or
-equal to than the minimum overlap length) and then only comparing reads that
-actually share a subsequence.  Each of these short subsequences is called a {\it
-seed}.  Two reads may share a seed in either a forward or reverse-complement
-manner.  Given two reads that share a seed, the seed can be extended outwards in
-both directions in order to find a maximal matched region, that may be either
-forward or reverse-complement, between the two reads.  Then, this maximal
-matched region is compared against the definition of an overlap.  In particular,
-it must be at least as long as the minimum overlap length, it must have least 2
-extreme points over the two reads, and it cannot be the prefix or suffix of both
-reads if the match does not extend to the opposite end of at least one of the
-reads.  If all of these conditions are met and the same overlap was not already
-found through another seed in the same two reads, this overlap is added to the
-set of overlaps.
+the easier ones to understand is the \KeyTerm{seed-and-extend} algorithm.  It
+works by indexing the reads by short subsequences of a certain length (less than
+or equal to than the minimum overlap length) and then only comparing reads that
+actually share a subsequence.  Each of these short subsequences is called a
+\KeyTerm{seed}.  Two reads may share a seed in either a forward or
+reverse-complement manner.  Given two reads that share a seed, the seed can be
+extended outwards in both directions in order to find a maximal matched region,
+that may be either forward or reverse-complement, between the two reads.  Then,
+this maximal matched region is compared against the definition of an overlap.
+In particular, it must be at least as long as the minimum overlap length, it
+must have least 2 extreme points over the two reads, and it cannot be the prefix
+or suffix of both reads if the match does not extend to the opposite end of at
+least one of the reads.  If all of these conditions are met and the same overlap
+was not already found through another seed in the same two reads, this overlap
+is added to the set of overlaps.
 
@@
 For this project, the seed-and-extend algorithm for computing overlaps was
-implemented as the program {\tt compute-overlaps}.  It takes as input a file
-that contains a list of reads, as well as a parameter that specifies the minimum
-overlap length $L$.  It produces as output (via a data structure serialized to a
-file) all exact pairwise overlaps of minimum length $L$ among reads in the input
-file, including reverse-complement overlaps and self-overlaps.  The running time
-of the algorithm was discussed above.  The memory usage is linear with respect
-to the total length of the reads and the number of overlaps, which likely could
-be improved.
+implemented as the program \ProgramName{compute-overlaps}.  It takes as input a
+file that contains a list of reads, as well as a parameter that specifies the
+minimum overlap length $L$.  It produces as output (via a data structure
+serialized to a file) all exact pairwise overlaps of minimum length $L$ among
+reads in the input file, including reverse-complement overlaps and
+self-overlaps.  The running time of the algorithm was discussed above.  The
+memory usage is linear with respect to the total length of the reads and the
+number of overlaps, which likely could be improved.
 
@@
 
+\subsection{Overview}
+
+After overlaps have been computed from the reads, the next step is to construct
+a graph that models how the reads can be assembled together when the overlaps
+are used as evidence.  To do this, define a graph where each vertex corresponds
+to a read and each edge corresponds to an overlap.  Two vertices are connected
+by an edge if and only if the corresponding reads share an overlap.  In
+addition, each edge is labeled with an appropriate DNA sequence, the details of
+which will be explained later.
+
+It needs to be decided whether each edge will be directed, undirected, or
+something else.  To decide this, we need to consider exactly what we want each
+edge to actually represent.  Consider three reads $f$, $g$, and $h$, that
+overlap as follows:
+
+\begin{center}
+	\begin{tikzpicture}[>=\ReadArrowType]
+		\draw[->,style=very thick] (0,   0.6) node[anchor=east] {f} -- (3.0, 0.6);
+		\draw[->,style=very thick] (1.0, 0.3) node[anchor=east] {g} -- (4.0, 0.3);
+		\draw[<-,style=very thick] (2.0, 0.0) node[anchor=east] {h} -- (5.0, 0.0);
+	\end{tikzpicture}
+\end{center}
+
+There are 3 overlaps so there will be 3 edges--- one between $f$ and $g$, one
+between $g$ and $h$, and one between $f$ and $h$.  As shown, the reads can be
+assembled together consistently when used in the order $f,\ g,\ h$; therefore, it
+must be possible to walk $f \to g \to h$ in the graph to model this assembly.
+In addition, these reads could actually spell out the reverse-complement
+sequence instead:
+
+\begin{center}
+	\begin{tikzpicture}[>=\ReadArrowType]
+		\draw[->,style=very thick] (0,   0.0) node[anchor=east] {h} -- (3.0, 0.0);
+		\draw[<-,style=very thick] (1.0, 0.3) node[anchor=east] {g} -- (4.0, 0.3);
+		\draw[<-,style=very thick] (2.0, 0.6) node[anchor=east] {f} -- (5.0, 0.6);
+	\end{tikzpicture}
+\end{center}
+
+Therefore, the walk $h \to g \to f$ must be allowed.  Furthermore, read $f$
+directly overlaps with read $h$, so it must also be possible to walk directly $f
+\to h$ or $h \to f$.
+
+We might consider a directed graph, but this does not work because the graph
+does not allow the reverse-complement walks such as $h \to g \to f$:
+
+\begin{center}
+	\begin{tikzpicture}[>=triangle 45]
+		\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+		\node (f) at (0, 0) {f};
+		\node (g) at (1.7, 0) {g};
+		\node (h) at (3.4, 0) {h};
+		\draw[->,style=very thick] (f) edge (g);
+		\draw[->,style=very thick] (g) edge (h);
+		\draw[->,style=very thick,bend left] (f) edge (h);
+	\end{tikzpicture}
+\end{center}
+
+Similarly, we might consider an undirected graph, but this does not work because
+the graph fails to disallow walks such as $g \to h \to f$ that correspond to
+assemblies that are not supported by the underlying reads.
+
+\begin{center}
+	\begin{tikzpicture}[>=triangle 45]
+		\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+		\node (f) at (0, 0) {f};
+		\node (g) at (1.7, 0) {g};
+		\node (h) at (3.4, 0) {h};
+		\draw[-,style=very thick] (f) edge (g);
+		\draw[-,style=very thick] (g) edge (h);
+		\draw[-,style=very thick,bend left] (f) edge (h);
+	\end{tikzpicture}
+\end{center}
+
+The fundamental problem with using the directed graph is that at this stage in
+the assembly, we don't know in absolute terms which strand a given read is from.
+Given two reads that overlap, we can say if they are probably from the same
+strand or probably from opposite strands, but we cannot say {\em in absolute
+terms} which strand a given read is from.  Therefore, we must allow each read to
+be used in either the forward or reverse-complement direction.  This implies
+that it needs to be possible to traverse each edge in either direction, since
+flipping the order of the reads in an overlap, in effect, implies that they are
+reverse-complemented.  The undirected graph satisfies these requirements, but it
+fails in another way: it has no guarantee that a given read is used in a
+consistent way at a location within any single assembly.
+
+The solution to all these problems is to use a new type of graph called a
+\KeyTerm{bidirected graph}:
+\begin{center}
+	\begin{tikzpicture}[>=triangle 45]
+		\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+		\node (f) at (0, 0) {f};
+		\node (g) at (1.7, 0) {g};
+		\node (h) at (3.4, 0) {h};
+		\draw[>->,style=very thick] (f) edge (g);
+		\draw[>-<,style=very thick] (g) edge (h);
+		\draw[>-<,style=very thick,bend left] (f) edge (h);
+	\end{tikzpicture}
+\end{center}
+
+\subsection{Bidirected graphs}
+
+A \KeyTerm{bidirected graph} is a graph where a directed head is attached to both
+ends of each edge.
+
+There are 3 (or 4, ignoring symmetry) types of \KeyTerm{bidirected edges}.  They
+follow directly from the different types of overlaps (see
+\Figure{fig:overlap_mapping}).
+
+\begin{figure}
+	\begin{center}
+		\begin{tabular}{p{1cm}cccc}
+			{\footnotesize Overlap} &
+			\begin{tikzpicture}[scale=0.6,>=\ReadArrowType]
+					\draw[->,style=thick] (2,0) -- (0,0);
+					\draw[->,style=thick] (1,0.3) -- (3,0.3);
+			\end{tikzpicture}
+			&
+			\begin{tikzpicture}[scale=0.6,>=\ReadArrowType]
+					\draw[->,style=thick] (0,0) -- (2,0);
+					\draw[->,style=thick] (3,0.3) -- (1,0.3);
+			\end{tikzpicture}
+			&
+			\begin{tikzpicture}[scale=0.6,>=\ReadArrowType]
+					\draw[->,style=thick] (0,0) -- (2,0);
+					\draw[->,style=thick] (1,0.3) -- (3,0.3);
+			\end{tikzpicture}
+			&
+			\begin{tikzpicture}[scale=0.6,>=\ReadArrowType]
+					\draw[->,style=thick] (2,0) -- (0,0);
+					\draw[->,style=thick] (3,0.3) -- (1,0.3);
+			\end{tikzpicture}
+			\\
+			&$\downarrow$ & $\downarrow$& $\downarrow$& $\downarrow$ \\
+			{\footnotesize Bidirected edge} &
+			\begin{tikzpicture}[scale=0.6,>=triangle 45]
+					\draw[<->,style=thick] (0,0) -- (2,0);
+			\end{tikzpicture}
+			&
+			\begin{tikzpicture}[scale=0.6,>=triangle 45]
+					\draw[>-<,style=thick] (0,0) -- (2,0);
+			\end{tikzpicture}
+			&
+			\begin{tikzpicture}[scale=0.6,>=triangle 45]
+					\draw[>->,style=thick] (0,0) -- (2,0);
+			\end{tikzpicture}
+			&
+			\begin{tikzpicture}[scale=0.6,>=triangle 45]
+					\draw[<-<,style=thick] (0,0) -- (2,0);
+			\end{tikzpicture}
+		\end{tabular}
+	\end{center}
+	\caption{Mapping from the types of overlaps to the types of bidirected edges.}
+	\label{fig:overlap_mapping}
+\end{figure}
+
+A \KeyTerm{walk in a bidirected graph} $G$ is a continuous sequence of edges in
+$G$ such that if any vertex $v$ is entered through a head inwards, it is exited
+through a head outwards (unless it is the end of the path), and vice versa.  An
+equivalent formulation is that if we consider a possible walk in a bidirected
+graph, such as the possible walk $1 \to 2 \to 3 \to 4$ in the graph shown in
+\Figure{fig:bidiwalk_example}, in order to check whether the walk is valid or
+not, we must consider every vertex in the possible walk except for the very
+first and last vertices.  For each such vertex, we need to verify that each time
+that vertex is passed through as part of a walk, the arrow heads adjacent to
+that vertex that are used in the relevant part of the walk are pointed in the
+same direction {\em relative to the walk}. If there are no such pairs of heads
+that point towards each other or point away from each other, the walk is valid.
+The red circles in \Figure{fig:bidiwalk_example} show the two pairs of heads
+that need to be checked for the possible walk $1 \to 2 \to 3 \to 4$.
+
+Note that this definition is symmetric, which implies that the reverse of a walk
+in a bidirected graph is also a valid walk (e.g. the walk $4 \to 3 \to 2 \to 1$
+in \Figure{fig:bidiwalk_example}).  A bidirected edge may be traversed in either
+direction, possibly even during the same walk.
+
+\begin{figure}
+	\begin{center} {\small
+		\begin{tikzpicture}[>=triangle 45]
+			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+			\node (1) at (0.00, 0) {1};
+			\node (2) at (2.00, 0) {2};
+			\node (3) at (4.00, 0) {3};
+			\node (4) at (6.00, 0) {4};
+			\draw[<->,very thick] (1) edge (2);
+			\draw[>-<,very thick] (2) edge (3);
+			\draw[<-<,very thick] (3) edge (4);
+			\draw[color=red,thick] (2.00, 0) circle (0.8);
+			\draw[color=red,thick] (4.00, 0) circle (0.8);
+		\end{tikzpicture}
+	} \end{center}
+	\caption{$1 \to 2 \to 3 \to 4$ and $4 \to 3 \to 2 \to 1$ are both valid walks in
+	the bidirected graph below.  (Check circled pairs of heads.)}
+	\label{fig:bidiwalk_example}
+\end{figure}
+
+\section{Removing contained reads}
+
 \section*{Acknowledgments}

### 2012-12-07 presentation/paper.tex
paper: bidiedge interp

@@
 The solution to all these problems is to use a new type of graph called a
-\KeyTerm{bidirected graph}:
+\KeyTerm{bidirected graph}.
 \begin{center}
@@
 
+The concept of a walk through a bidirected graph maps quite nicely onto the
+genome assembly problem, since a walk through the bidirected string graph
+constructed from reads and overlaps models a way in which the reads can be
+assembled together consistently.  In such a walk, each read may be used in
+forward or reverse-complement orientation, depending on the direction, relative
+to the walk, of the adjacent arrow heads when the corresponding vertex is passed
+through.
+
+For example, consider the walk $1 \to 2 \to 3 \to 4$ in
+\Figure{fig:bidigraph_mapping}.  When the walk passes through vertex 1, the
+adjacent arrow head is pointed opposite the direction of the walk.  This implies
+that the corresponding read is used in reverse-complement orientation.  When the
+walk passes through vertex 2, the adjacent arrow heads are pointed in the
+direction of the walk, which implies that the corresponding read is used in
+forward orientation.  Finally, when the walk passes through vertices 3 and 4,
+the adjacent arrow heads are pointed opposite the direction of the walk, which
+implies that the corresponding reads are used in reverse-complement orientation.
+Note that $4 \to 3 \to 2 \to 1$ is also a valid walk, which represents the way
+in which the four reads could also assemble the other strand of the DNA
+molecule.
+
+\begin{figure}
+	\begin{center}
+		{\small
+		\begin{tikzpicture}[>=triangle 45]
+			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+			\node (1) at (0.00, 0) {1};
+			\node (2) at (2.0, 0) {2};
+			\node (3) at (4.0, 0) {3};
+			\node (4) at (6.0, 0) {4};
+			\draw[<->,very thick] (1) edge (2);
+			\draw[>-<,very thick] (2) edge (3);
+			\draw[<-<,very thick] (3) edge (4);
+		\end{tikzpicture} }
+
+		\vspace{2.5mm}
+
+		\begin{tikzpicture}[>=\ReadArrowType]
+				\draw[<-,style=very thick] (0.0, 0.0) -- (2.6, 0.0);
+				\draw[->,style=very thick] (1.6, -0.3) -- (4.2, -0.3);
+				\draw[<-,style=very thick] (3.2, -0.6) -- (5.8, -0.6);
+				\draw[<-,style=very thick] (4.8, -0.9) -- (7.4, -0.9);
+				\draw (-0.3, 0.0) node {1.};
+				\draw (-0.3, -0.3) node {2.};
+				\draw (-0.3, -0.6) node {3.};
+				\draw (-0.3, -0.9) node {4.};
+		\end{tikzpicture}
+	\end{center}
+	\caption{A bidirected graph and the implied assembly of the
+	corresponding reads}
+	\label{fig:bidigraph_mapping}
+\end{figure}
+
+Earlier, a crucial point was quickly skipped over.  Each edge is supposed to be
+labeled with a DNA sequence, as we are building a bidirected {\em string} graph.
+However, we actually need to label each edge with two DNA sequences: one for
+each direction in which the edge can be traversed.  To see this, we need to
+consider exactly what it means to walk between vertices in the bidirected graph.
+
+Consider the the string graph fragment shown in \Figure{fig:bidiedge_interp_a}.
+When we walk from $f \to g$, we are using both $f$ and $g$ in forward
+orientation, based on the arrow heads.  But what this really means is that we
+are walking from the {\em end} of read $f$ to the {\em end} of read $g$.  In
+other words, traversing the edge in that direction is associated with the
+addition of the sequence after the overlapped region in $g$, up to the end of
+the $g$.
+
+If we instead walk $g \to f$, we instead use both $g$ and $f$ in
+reverse-complement orientation by walking from the {\em beginning} of read $g$
+to the {\em beginning} of read $f$.  In other words, traversing the edge in that
+direction is associated with the addition of the sequence before the overlapped
+region in $f$, up to the beginning of $f$, and this sequence must be
+reverse-complemented because it is being used opposite its original direction.
+
+These interpretations can be generalized to the other two types of bidirected
+edges.  Furthermore, these intepretations imply that the actual sequence spelled
+out by a walk in the bidirected graph may include implied (non-labeled) sequence
+at the start (if we start at the end of a read) and/or the end (if we end at the
+beginning of a read).
+
+\begin{figure}
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\begin{tikzpicture}[>=triangle 45,scale=1.5]
+			\node[circle,fill=blue!20,draw=black] (f) at (0, 0) {f};
+			\node[circle,fill=blue!20,draw=black] (g) at (2.0, 0) {g};
+			\draw[>->,style=very thick] (f) edge (g);
+			\draw (1.00, 0.1) node[anchor=south] {\small \tt TTAGAAGTA};
+			\draw (1.00, -0.1) node[anchor=north] {
+					\rotatebox{180}{\small \tt
+					TGTGTAATC}};
+			\draw[->] (0.5, 0.5) ..
+				controls (0.7, 0.7) and (1.3, 0.7) ..
+				(1.5, 0.5);
+			\draw[->] (1.5, -0.5) ..
+				controls (1.3, -0.7) and (0.7, -0.7) ..
+				(0.5, -0.5);
+		\end{tikzpicture}
+		\caption{bidirected string graph fragment}
+		\label{fig:bidiedge_interp_a}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\begin{tikzpicture}[>=\ReadArrowType,scale=0.8]
+			\draw[->,very thick] (0, 0) node[anchor=east] {$f.B$}
+					-> (4, 0) node[anchor=west] {$f.E$};
+			\draw[->,very thick] (2.09, -0.9) node[anchor=east] {$g.B$}
+					-> (6.09, -0.9) node[anchor=west] {$g.E$};
+
+			\draw (0, -0.1) node[anchor=south west]
+				{\hspace*{-1.4mm}{\small \tt
+					%GATTACACAGTGTAGTT
+					GATTACACA\hspace*{0.1cm}$a_1 \dots a_n$
+				}};
+
+			\draw (0, 0.1) node[anchor=north west]
+				{\hspace*{-2.6mm}
+					\rotatebox{180}{{\small \tt TGTGTAATC}}};
+
+			\draw (2.09, -1.0) node[anchor=south west]
+				{\hspace*{-1.4mm}{\small \tt
+					%GTGTAGTTTTAGAAGTA
+					\hspace*{0.1cm}$a_1 \dots a_n$\hspace*{0.25cm}TTAGAAGTA
+				}};
+
+			\draw (2.09, -0.9) node[anchor=north west]
+				{\hspace*{12.48mm}
+					\rotatebox{180}{{\small \tt TACTTCTAA}}};
+
+			\draw (2.09, -1.4) node[anchor=north] {$o.g_{begin}$}
+					-- (2.09, 0.6) node[anchor=south] {$o.f_{begin}$};
+			\draw (3.99, -1.4) node[anchor=north] {$o.g_{end}$}
+					-- (3.99, 0.6) node[anchor=south] {$o.f_{end}$};
+		\end{tikzpicture}
+		\caption{The corresponding overlap (labeled)}
+		\label{fig:bidiedge_interp_b}
+	\end{subfigure}
+
+	\begin{subfigure}{0.5\textwidth}
+		\begin{tabular}{c|c}
+			walk & interpretation \\ \hline
+			$f \to g$ & $f.E \to g.E$ with sequence $g[o.g_{end} + 1 \dots ${\tt
+			length}$(g)]$ \\
+			$g \to f$ & $g.B \to f.B$ with sequence $f[o.g_{begin} - 1 \dots 0]$ \\
+		\end{tabular}
+		\caption{Interpretation of each walk on the graph fragment}
+	\end{subfigure}
+	\caption{Full interpetation of an edge in the bidirected string graph}
+	\label{bidiedge_interp}
+\end{figure}
+
 \section{Removing contained reads}
+
+There is yet another flaw in the way that the construction of the graph has been
+presented that needs to be addressed.  Earlier, the definition of a
+\KeyTerm{contained read} was provided.
 

### 2012-12-08 presentation/paper.tex
paper: graph building

@@
 \newcommand{\Figure}[1]{Figure \ref{#1}}
+\newcommand{\Subfigure}[1]{Figure \ref{#1}}
+\newcommand{\Section}[1]{Section \ref{#1}}
+\newcommand{\Subsection}[1]{Subsection \ref{#1}}
 
@@
 
+\label{sec:build_graph}
+
 \subsection{Overview}
+
+\label{subsec:graph_overview}
 
@@
 
-Consider the the string graph fragment shown in \Figure{fig:bidiedge_interp_a}.
-When we walk from $f \to g$, we are using both $f$ and $g$ in forward
-orientation, based on the arrow heads.  But what this really means is that we
-are walking from the {\em end} of read $f$ to the {\em end} of read $g$.  In
-other words, traversing the edge in that direction is associated with the
-addition of the sequence after the overlapped region in $g$, up to the end of
-the $g$.
+Consider the the string graph fragment shown in
+\Subfigure{subfig:bidiedge_interp_a}.  When we walk from $f \to g$, we are using
+both $f$ and $g$ in forward orientation, based on the arrow heads.  But what
+this really means is that we are walking from the {\em end} of read $f$ to the
+{\em end} of read $g$.  In other words, traversing the edge in that direction is
+associated with the addition of the sequence after the overlapped region in $g$,
+up to the end of the $g$.
 
@@
 \begin{figure}
-	\centering
-	\begin{subfigure}{0.48\textwidth}
+	\begin{center}
+		\begin{subfigure}[b]{0.48\textwidth}
 			\begin{tikzpicture}[>=triangle 45,scale=1.5]
@@
 			\end{tikzpicture}
-		\caption{bidirected string graph fragment}
-		\label{fig:bidiedge_interp_a}
+			\caption{A bidirected string graph fragment}
+			\label{subfig:bidiedge_interp_a}
 		\end{subfigure}
-	\begin{subfigure}{0.48\textwidth}
+		\begin{subfigure}[b]{0.48\textwidth}
 			\begin{tikzpicture}[>=\ReadArrowType,scale=0.8]
@@
 						%GATTACACAGTGTAGTT
-					GATTACACA\hspace*{0.1cm}$a_1 \dots a_n$
+						GATTACACA\hspace*{0.26cm}$a_1 \dots a_n$
 					}};
@@
 			\caption{The corresponding overlap (labeled)}
-		\label{fig:bidiedge_interp_b}
+			\label{subfig:bidiedge_interp_b}
+		\end{subfigure}
+	\end{center}
+
+	\begin{subfigure}{1.00\textwidth}
+		\begin{center}
+			\begin{tikzpicture}[>=triangle 45,scale=1.5]
+				\node[circle,fill=blue!20,draw=black] (fE) at (0, 0) {f.E};
+				\node[circle,fill=blue!20,draw=black] (gE) at (2.0, 0) {g.E};
+
+				\node[circle,fill=blue!20,draw=black] (fB) at (0, -1) {f.B};
+				\node[circle,fill=blue!20,draw=black] (gB) at (2.0, -1) {g.B};
+
+				\draw[->,style=very thick] (fE) edge (gE);
+				\draw[->,style=very thick] (gB) edge (fB);
+				\draw (0.90, 0.01) node[anchor=south] {\small \tt TTAGAAGTA};
+				\draw (1.10, -1.01) node[anchor=north] {
+						\rotatebox{180}{\small \tt
+						TGTGTAATC}};
+			\end{tikzpicture}
+		\end{center}
+		\caption{The corresponding directed string graph fragment}
+		\label{subfig:bidiedge_interp_a}
 	\end{subfigure}
 
-	\begin{subfigure}{0.5\textwidth}
+	\vspace{1.0cm}
+
+	\begin{subfigure}{1.00\textwidth}
+		\begin{center}
 			\begin{tabular}{c|c}
@@
 			\end{tabular}
-		\caption{Interpretation of each walk on the graph fragment}
+		\end{center}
+		\caption{Interpretation of each walk on the bidirected graph fragment}
 	\end{subfigure}
-	\caption{Full interpetation of an edge in the bidirected string graph}
-	\label{bidiedge_interp}
+	\caption{Full interpetation of an edge in a bidirected string graph}
+	\label{fig:bidiedge_interp}
 \end{figure}
 
-\section{Removing contained reads}
+\subsection{An equivalent directed graph}
 
-There is yet another flaw in the way that the construction of the graph has been
-presented that needs to be addressed.  Earlier, the definition of a
-\KeyTerm{contained read} was provided.
+\label{subsec:directed}
+
+In \Subsection{subsec:graph_overview}, the possibility of using a directed graph
+to model the assembly of the reads was rejected because it did not take into
+account the double-strandedness of DNA.  However, \Figure{fig:bidiedge_interp}
+suggests an alternate construction of the graph that does indeed use directed
+edges.  Instead of having each vertex be a read, have each vertex be an end of a
+read.  The advantage to this approach is that the problem can be expressed using
+a directed graph instead a bidirected graph, which can be helpful for the
+implementation.  The disadvantage is that two vertices are required for each
+read, they may be widely separated in the graph.
+
+\subsection{Removing contained reads}
+
+\label{subsec:removing_contained_reads}
+
+Earlier, the definition of a \KeyTerm{contained read} was provided.  Contained
+reads need not be included in the graph because there will always be a walk that
+spells the sequence of each contained read.  In other words, contained reads
+provide redundant information because the sequence of any given contained read
+is already spelled out by an uncontained read.
+
+Therefore, prior to mapping the reads to vertices and overlaps to edges, we can
+examine the overlaps to identify contained reads, and include in the initial
+construction of the bidirected string graph neither the contained reads nor any
+overlaps involving a contained read.
+
+The contained reads cannot, however be fully discarded, as they are needed for a
+later step (see \Section{sec:mapping_contained_reads}).
+
+\subsection{Implementation}
+
+As described in \Subsection{subsec:removing_contained_reads}, before we even
+build the graph, contained reads need to be removed.  This is implemented by the
+program \ProgramName{remove-contained-reads}.  It takes as input two files: the
+set of reads and the set of overlaps that were computed from them, and produces
+as output three files: the set of uncontained reads, the set of overlaps
+excluding those that involve contained reads, and an array that maps from the
+original read indices to the new read indices.
+
+Building the graph is implemented by either the program
+\ProgramName{build-bidirected-string-graph} or the program
+\ProgramName{build-directed-string-graph}, the latter of which builds the
+directed equivalent to the bidirected graph described in
+\Subsection{subsec:directed}.  Both programs take as input two files: the set of
+uncontained reads and their overlaps, and produce as output the bidirected or
+directed graph, serialized to a file.  As should be expected from what was
+described earlier, \ProgramName{build-bidirected-string-graph} will add one
+vertex for each read and add a bidirected edge for each overlap, while
+\ProgramName{build-directed-string-graph} will add two vertices for each read
+and add two directed edges for each overlap.
+
+The string graphs are stored in a binary format.  An additional program,
+\ProgramName{print-string-graph}, was implemented that can print a string graph
+in plain-text format, or print a string graph in the DOT language for input to
+the graph-drawing software {\it GraphViz}\cite{GraphVizPublication}, or print
+some other information about a string graph.
+
+Furthermore, programs \ProgramName{bidigraph-to-digraph} and
+\ProgramName{digraph-to-bidigraph} were implemented to convert between the
+bidirected and directed representations of the string graph.
+
+A program \ProgramName{extract-edge-seqs} was implemented to extract the
+sequences that the edges are labeled with, as this can be done to produce a
+preliminary assembly (especially after the graph simplifications that are about
+to be described).
+
+\section{Transitive reduction}
+
+Before trying to find walks through the string graph, it is very helpful to
+simplify the graph is much as
+
+\section{Mapping contained reads}
+\label{sec:mapping_contained_reads}
 

### 2012-12-08 presentation/paper.tex
paper: possible future work

@@
 \newcommand{\Subsection}[1]{Subsection \ref{#1}}
+\newcommand{\Algorithm}[1]{Algorithm \ref{#1}}
 
@@
 \newcommand{\ProgramName}[1]{{\tt #1}}
+\newcommand{\BidirectedEdgeForward}[2]{
+		\tikz[>=triangle 45,baseline]
+		\draw[>->,thin] (0,0.1) node[anchor=east] {#1} --
+					    (0.8,0.1) node[anchor=west] {#2};}
 
@@
 
-%\begin{abstract}
-%Sequencing an organism's genome is an important step to further genetic research
-%on that organism.  However, current sequencing machines can only sequence very
-%short fragments of DNA that originate from random positions on the original
-%genome, assembly software such as Allpaths-LG can be used to assemble these
-%fragments into longer contiguous sequences.
-%\end{abstract}
+\begin{abstract}
+
+Sequencing an organism's genome is an important step to further genetic research
+on that organism.  However, currently available sequencing technology can only
+sequence short fragments of DNA.  As a result, whole-genome shotgun sequencing
+approaches to genome sequencing rely on programs called {\em genome assemblers}
+that can put together a set of short DNA fragments to reconstruct the original
+genome.  This paper provides an introduction and formalization of the genome
+assembly program, then introduces a model and algorithm for genome assembly that
+is based on a special type of graph called a \KeyTerm{bidirected string graph}.
+\end{abstract}
+
+\tableofcontents
 
@@
 
-\section{Genome assembly vs. shortest common superstring}
+\subsection{Genome assembly vs. shortest common superstring}
 
@@
 
-\subsection{An equivalent directed graph}
+\subsection{An equivalent directed graph model}
 
@@
 Before trying to find walks through the string graph, it is very helpful to
-simplify the graph is much as
-
-\section{Mapping contained reads}
+simplify the graph is much as possible.  Very commonly, given three adjacent
+reads $f$, $g$, and $h$, $f$ will overlap $h$ as well as $g$.  This is redundant
+because we can, equivalently, walk $f$ to $g$ to $h$ rather than directly
+walking from $f$ to $h$.  \Figure{fig:transitive} shows an example of this.
+
+\begin{figure}
+	\begin{center}
+		\begin{tikzpicture}
+			\draw[->,style=very thick] (0,0) node[anchor=east] {f} -- (2,0);
+			\draw[->,style=very thick] (0.6,0.25) node[anchor=east] {g} -- (2.6,0.25);
+			\draw[->,style=very thick] (1.2,0.5) node[anchor=east] {h} -- (3.2,0.5);
+		\end{tikzpicture}
+		\begin{tikzpicture}[>=triangle 45]
+			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+			\node (f) at (0, 0) {f};
+			\node (g) at (1.7, 0) {g};
+			\node (h) at (3.4, 0) {h};
+			\draw[>->,style=very thick] (f) edge (g);
+			\draw[>->,style=very thick] (g) edge (h);
+			\draw[>->,style=very thick,bend left] (f) edge (h);
+		\end{tikzpicture}
+	\end{center}
+	\caption{Three overlapping reads $f$, $g$, and $h$, and the corresponding
+	bidirected graph.  The edge between $f$ and $h$ is transitive and is removed
+	by transitive reduction.}
+	\label{fig:transitive}
+\end{figure}
+
+Walks visiting more vertices are preferred because they are supported by more
+reads and the overlaps tend to be longer.  So, we would like to throw away
+\BidirectedEdgeForward{$f$}{$h$}, which is a \KeyTerm{transitive edge}, but keep
+\BidirectedEdgeForward{$f$}{$g$} and \BidirectedEdgeForward{$g$}{$h$}.
+
+To remove these redundant edges, \KeyTerm{transitive reduction} is performed on
+the bidirected string graph.
+
+\Algorithm{alg:transitive_reduction} produces pseudocode for a transitive
+reduction algorithm.  Essentially, the algorithm is to through each vertex $v$
+and examine vertices up to 2 edges away from $v$ to identify all transitive
+edges leaving $v$.  These edges are then marked for removal, and removed after
+all vertices in the graph have been considered.  Some optimizations are
+possible; for example, the lengths of the sequences the edges are labeled with
+can be taken into account to allow terminating the search of a vertex $v$'s
+neighborhood before all possible walks of length 2 leaving $v$ have been
+considered.
+
+Transitive reduction can be performed on either the bidirected or directed
+representations of the string graph, although for simplicity here it is
+described on the directed graph.
+
+\begin{algorithm}
+	\caption{{\sc TransitiveReduction}($G$)}
+	{\footnotesize
+		\begin{algorithmic}
+			\FOR{{\bf each} vertex $v \in G$}
+				\STATE{Mark all neighbors of $v$ as {\tt INPLAY}}
+				\STATE{Set {\tt longest} $\gets$ length of the longest edge leaving $v$}
+				\FOR{{\bf each} edge $v \to w$ in order of length}
+					\FOR{{\bf each} edge $w \to x$ in order of length}
+						\IF{$x$ is not marked {\tt INPLAY}}
+							\STATE{\bf continue}
+						\ENDIF
+						\IF{$len(v \to w) + len(w \to x) > $ {\tt longest}}
+							\STATE{\bf break}
+						\ENDIF
+						\STATE{Mark $x$ as {\tt ELIMINATED}}
+					\ENDFOR
+				\ENDFOR
+				\STATE{Mark as transitive all edges $v \to x$ where $x$ is marked {\tt ELIMINATED}}
+			\ENDFOR
+			\STATE{Eliminate all edges marked transitive}
+		\end{algorithmic}
+	}
+	\label{alg:transitive_reduction}
+\end{algorithm}
+
+Transitive reduction on a string graph is implemented as the program
+\ProgramName{transitive-reduction}.  It takes as input a file containing a
+string graph and produces as output a file containing the transitively reduced
+string graph.
+
+\section{Collapsing unbranched paths}
+
+After transitive reduction, a non-repetitive region of the genome will typically
+be represented in the string graph by a chain of vertices that is part of a
+valid walk and consists entirely of vertices with in-degree 1 and out-degree 1.
+This is referred to as an \KeyTerm{unbranched path}, and it contains one or more
+\KeyTerm{inner vertices}.  An unbranched path can be collapsed by removing the
+inner vertices and replacing them by a single edge that covers the entire path.
+This replacement edge is labeled with the concatenation of the sequences that
+the edges are labeled with.  In the bidirected case, this concatenation must be
+done for both directions in which the unbranched path might have been traversed.
+\Figure{fig:collapse_unbranched_paths} shows an example of how a short
+unbranched path can be collapsed.
+
+Collapsing unbranched paths is implemented as the program
+\ProgramName{collapsed-unbranched-paths}.  It takes as input a file containing a
+string graph and produces as output a file containing the string graph with
+unbranched paths collapsed.  Vertices are currently re-numbered; so, for
+example, vertex 10 after collapsing unbranched paths does not necessarily refer
+to the 10th read in the list of uncontained reads from which the graph was
+constructed.
+
+\begin{figure}
+	\begin{center}
+		{\small
+		\begin{tikzpicture}[>=triangle 45]
+			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+			\node (1) at (0.0, -1) {1};
+			\node (2) at (0.0, 1) {2};
+			\node (3) at (1.5, 0) {3};
+			\node[fill=green!25] (4) at (3.25, 0) {4};
+			\node (5) at (5.0, 0) {5};
+			\node (6) at (6.5, 1) {6};
+			\node (7) at (6.5, -1) {7};
+			\draw[>->,very thick] (1) edge (3);
+			\draw[>->,very thick] (2) edge (3);
+			\draw[>->,very thick] (3) edge (4);
+			\draw[>->,very thick] (4) edge (5);
+			\draw[>->,very thick] (5) edge (6);
+			\draw[>->,very thick] (5) edge (7);
+		\end{tikzpicture}
+		}
+
+	becomes:
+
+		{\small
+		\begin{tikzpicture}[>=triangle 45]
+			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+			\node (1) at (0.0, -1) {1};
+			\node (2) at (0.0, 1) {2};
+			\node (3) at (1.5, 0) {3};
+			\node (5) at (3.25, 0) {5};
+			\node (6) at (4.75, 1) {6};
+			\node (7) at (4.75, -1) {7};
+			\draw[>->,very thick] (1) edge (3);
+			\draw[>->,very thick] (2) edge (3);
+			\draw[>->,very thick] (3) edge (5);
+			\draw[>->,very thick] (5) edge (6);
+			\draw[>->,very thick] (5) edge (7);
+		\end{tikzpicture}
+		}
+	\end{center}
+	\caption{Vertex 4 an \KeyTerm{inner vertex} that is part of an unbranched
+	path that can be collapsed.  Edge labels are not shown.}
+	\label{fig:collapse_unbranched_paths}
+\end{figure}
+
+\section{Examples of graph construction and simplification}
+
+At this point, the assembler has been given a set of reads, has computed exact
+overlaps from them, removed contained reads and all overlaps involving them,
+then built a bidirected string graph that models the ways (via walks through the
+graph) in which the reads can be assembled together when the overlaps are used
+as evidence, then simplifies this graph through transitive reduction and
+collapsing unbranched paths.  Usually, the graph at this point will be much
+simpler than the original graph, although it may still be very complicated,
+depending on the genome and the data.
+
+\subsection{Random genome}
+
+A completely random genome is very unlikely to have repeats of significant
+length compared to the minimum overlap length.  In this case, transitive
+reduction is expected to leave the entire graph as a single unbranched path,
+provided the original genome was a single, non-circular string, and every base
+pair of it was included in at least one read.  Then, collapsing unbranched paths
+will collapse the entire bidirected graph into two vertices connected by a
+single edge.  Finding a walk through the edges of this graph is trivial and will
+produce an assembly of the entire genome.  \Figure{fig:random_example} shows the
+bidirected string graph that might result from such a genome and data.
+
+If we still assume a genome with no repeats of length greater than the overlap
+length, but relax the assumption that the genome is a single, non-circular
+string from which every base pair was included in a read, the graph will instead
+reduce and collapse to a set of components.  Components corresponding to
+non-circular sequence will contain two vertices with an edge between, while
+components corresponding to circular sequence will contain one vertex with an
+edge connecting it to itself (a loop).  In this case (and as will be elaborated
+on later), finding a walk through the bidirected string graph must be
+generalized to allow walking each component separately; but again for this
+particular case of a random genome, each such walk is still trivial, and the
+assembly will be optimal as long as the assembler does not decide to traverse
+any of the loops more than one time (this will also be elaborated on more
+later).
+
+\begin{figure}
+	\begin{center}
+		\begin{tikzpicture}
+			{\small
+			\begin{tikzpicture}[>=triangle 45]
+				\node[circle,fill=blue!20,draw=black] (A) at (0.0, 0) {A};
+				\node[circle,fill=blue!20,draw=black] (B) at (3.0, 0) {B};
+				\draw[>->,very thick] (A) edge node[anchor=north] {4639100} (B);
+			\end{tikzpicture} }
+		\end{tikzpicture}
+		\caption{
+			Bidirected string graph for reads uniformly sampled from a 4.6 Mbp
+			random genome after transitive reduction and path collapsing.  (Edge
+			labeled with length of sequence, with the length being the same in both
+			directions)}
+		\label{fig:random_example}
+	\end{center}
+\end{figure}
+
+\subsection{Real genome}
+
+Real genomes are not random, and they contain far more repeats than would be
+expected purely by chance.\footnote{Biologically, repeats can occur for a number
+of reasons, such as gene duplications.} \Figure{fig:E_coli} shows an example of
+the simplified bidirected string graph for a real genome.  It is apparent that
+even when using perfect, fairly long reads that cover the entire genome, the
+repeats in the genome cause branches to appear in the graph when the repeats are
+too long to be resolved by the reads at this stage in the assembly.
+
+\begin{figure}
+	\begin{center}
+		\includegraphics[width=0.6\textwidth]{E_coli.pdf}
+		\caption{Transitively-reduced, collapsed bidirected string graph for
+		1000 bp perfect reads uniformly sampled every 100 bp from random
+		strands of the non-circular version of the 4,639,675 bp genome of {\it
+		E. coli} (a common bacterium) using a minimum overlap length of 250 bp.
+		This graph was produced by the code that has been
+		discussed, but the actual drawing of the picture was done by
+		{\it GraphViz}.  The 46,387 reads that were provided as input to the
+		assembler were generated by a Perl script.  The bottleneck of the
+		algorithm up to this point is \ProgramName{compute-overlaps}, which took
+		5 minutes and used a maximum of 984 MiB of memory to find all XXXXXX
+		overlaps from these reads, considering XXXXXX pairs of reads that shared
+		a 128-mer seed.  The running time and memory usage of
+		\ProgramName{compute-overlaps} would need to improved further to handle
+		larger genomes.  (Also it still does not handle inexact overlaps, which
+		are not considered in this paper.)}
+		\label{fig:E_coli}
+	\end{center}
+\end{figure}
+
+\subsection{Small step-by-step example}
+
+\section{Producing the final assembly}
+\subsection{Overview}
+
+\subsection{Calculating the A-statistic}
+\subsection{Mapping contained reads}
+\subsection{Network flow}
+\subsection{Generalized Eulerian path}
+
+\section{Possible future work}
+
+\begin{itemize}
+\item Finish the network flow and generalized Eulerian path parts of the algorithm
+so that the implementation is ``complete'' and can produce a ``final'' assembly
+\item Improve performance of \ProgramName{compute-overlaps}
+\item Allow for sequencing errors by finding inexact overlaps and modifying other
+stages of the algorithm, such as transitive reduction, to allow for the
+possibility that overlaps are not exact.
+\item Compensate for sequencing errors by attempting to correct them, either as
+part of \ProgramName{compute-overlaps} or as a new stage of the algorithm.
+\item Consider more recent publications that explore variations on this algorithm,
+such as parellizing the algorithm on a cluster \cite{X}, or using the
+Burrows-Wheeler transform to directly and efficiently construct the bidirected
+string graph directly from the reads\cite{X}.
+\item Use mate-pair information to help resolve repeats and connect assembled
+sequences even when they are separated by unknown sequence
+\end{itemize}
+
 \label{sec:mapping_contained_reads}

### 2012-12-09 src/BidirectedStringGraph.cc
Unbranched paths

@@ -226,8 +226,13 @@ void BidirectedStringGraph::build_from_digraph(const DirectedStringGraph & digra
 							 v_w.get_mapped_read_count()) / 2);
 				e.set_A_statistic((w_v.get_A_statistic() +
 						   v_w.get_A_statistic()) / 2);
-				assert(w_v.get_num_inner_vertices() ==
-				       v_w.get_num_inner_vertices());
+				if (w_v.get_num_inner_vertices() !=
+				    v_w.get_num_inner_vertices())
+				{
+					std::cerr << w_v << std::endl;
+					std::cerr << v_w << std::endl;
+					assert(0);
+				}
 				e.set_num_inner_vertices(w_v.get_num_inner_vertices());
 
 				edge_idx_t edge_idx = this->push_back_edge(e);

### 2012-12-09 src/DirectedStringGraph.cc

@@ -313,12 +313,20 @@ void DirectedStringGraph::collapse_unbranched_paths()
 			const DirectedStringGraphVertex & v = _vertices[v_idx];
 			foreach(edge_idx_t edge_idx, v.edge_indices()) {
 				DirectedStringGraphEdge & e = _edges[edge_idx];
-				if (v_inner[e.get_v2_idx()]) {
+				v_idx_t v2_idx = e.get_v2_idx();
+				if (v_inner[v2_idx] && !remove_vertex[v2_idx]) {
 					num_unbranched_paths++;
 					follow_unbranched_path(e, remove_edge,
 							       remove_vertex, v_inner);
 				}
 			}
+		}
+	}
+
+	for (v_idx_t v_idx = 0; v_idx < n_verts; v_idx++) {
+		if (!remove_vertex[v_idx] && v_inner[v_idx]) {
+			std::cerr << "Graph contains a smooth ring!" << std::endl;
+			unimplemented();
 		}
 	}
 

### 2012-12-09 src/DirectedStringGraph.h

@@ -243,15 +243,14 @@ public:
 	{
 		for (size_t i = 0; i < num_edges(); i++) {
 			const DirectedStringGraphEdge & e = _edges[i];
-
 			assert(e.get_v1_idx() < num_vertices());
 			assert(e.get_v2_idx() < num_vertices());
-
 			edge_idx_t j = locate_edge(e.get_v2_idx() ^ 1, e.get_v1_idx() ^ 1);
 			assert(j < num_edges());
 			const DirectedStringGraphEdge & e_X = _edges[j];
 			assert((e_X.get_v1_idx() ^ 1) == e.get_v2_idx());
 			assert((e_X.get_v2_idx() ^ 1) == e.get_v1_idx());
+			assert(e_X.get_num_inner_vertices() == e.get_num_inner_vertices());
 		}
 	}
 

### 2012-12-09 src/DirectedStringGraph.cc
Inner vertices

@@ -222,6 +222,7 @@ void DirectedStringGraph::follow_unbranched_path(DirectedStringGraphEdge & e,
 	v_idx_t vi_idx = e.get_v2_idx();
 	assert(v_inner[vi_idx]);
 
+	v_idx_t num_inner_vertices = 0;
 	// Found beginning of unbranched path.  Walk along it until
 	// the end to get the total sequence length.
 	do {
@@ -231,6 +232,7 @@ void DirectedStringGraph::follow_unbranched_path(DirectedStringGraphEdge & e,
 		if (new_seq_len + ei_i1.length() < new_seq_len)
 			fatal_error("Edge too long");
 		new_seq_len += ei_i1.length();
+		num_inner_vertices++;
 		vi_idx = ei_i1.get_v2_idx();
 	} while (v_inner[vi_idx]);
 
@@ -241,7 +243,9 @@ void DirectedStringGraph::follow_unbranched_path(DirectedStringGraphEdge & e,
 	BaseVec::size_type seq_idx = new_seq.length();
 	new_seq.resize(new_seq_len);
 	vi_idx = e.get_v2_idx();
-	do {
+
+	e.set_num_inner_vertices(num_inner_vertices);
+	while (num_inner_vertices--) {
 		const DirectedStringGraphVertex &vi = _vertices[vi_idx];
 		const edge_idx_t ei_i1_idx = vi.first_edge_idx();
 		const DirectedStringGraphEdge &ei_i1 = _edges[ei_i1_idx];
@@ -251,11 +255,11 @@ void DirectedStringGraph::follow_unbranched_path(DirectedStringGraphEdge & e,
 			new_seq.set(seq_idx++, ei_i1_seq[i]);
 		}
 		e.increment_mapped_read_count(ei_i1.get_mapped_read_count());
-		e.increment_num_inner_vertices();
 		remove_edge[ei_i1_idx] = true;
 		remove_vertex[vi_idx] = true;
 		vi_idx = ei_i1.get_v2_idx();
-	} while (v_inner[vi_idx]);
+	}
+	assert(!v_inner[vi_idx]);
 	assert(new_seq_len == seq_idx);
 	e.set_v2_idx(vi_idx);
 }
@@ -324,7 +328,7 @@ void DirectedStringGraph::collapse_unbranched_paths()
 	}
 
 	for (v_idx_t v_idx = 0; v_idx < n_verts; v_idx++) {
[…]

### 2012-12-10 presentation/paper.tex
paper: network flow

@@
 
+A \KeyTerm{cycle} in a bidirected graph is a walk that begins and ends at the
+same vertex in such a way that it would be valid to trace the walk again when
+the starting vertex is reached at the end of the cycle.
+
+A bidirected graph is \KeyTerm{connected} if the undirected graph produced by
+replacing all bidirected edges with undirected edges is connected.  A bidirected
+graph is \KeyTerm{strongly connected} if it is connected and for every edge
+there is a cycle containing it\cite{Medvedev2009}.
+
 The concept of a walk through a bidirected graph maps quite nicely onto the
@@
 		\caption{Transitively-reduced, collapsed bidirected string graph for
-		1000 bp perfect reads uniformly sampled every 100 bp from random
-		strands of the non-circular version of the 4,639,675 bp genome of {\it
-		E. coli} (a common bacterium) using a minimum overlap length of 250 bp.
-		This graph was produced by the code that has been
-		discussed, but the actual drawing of the picture was done by
-		{\it GraphViz}.  The 46,387 reads that were provided as input to the
-		assembler were generated by a Perl script.  The bottleneck of the
-		algorithm up to this point is \ProgramName{compute-overlaps}, which took
-		5 minutes and used a maximum of 984 MiB of memory to find all XXXXXX
-		overlaps from these reads, considering XXXXXX pairs of reads that shared
-		a 128-mer seed.  The running time and memory usage of
+		1000 bp perfect reads uniformly sampled every 100 bp from random strands
+		of the non-circular version of the 4,639,675 bp genome of {\it E. coli}
+		(a common bacterium) using a minimum overlap length of 250 bp.  This
+		graph was produced by the code that has been discussed, but the actual
+		drawing of the picture was done by {\it GraphViz}.  The 46,387 reads
+		that were provided as input to the assembler were generated by a Perl
+		script.  The bottleneck of the algorithm up to this point is
+		\ProgramName{compute-overlaps}, which took about 10 minutes and used a
+		maximum of about 1 GiB of memory to find all 336,227 overlaps from these
+		reads, seeding on 128-mers.  The running time and memory usage of
 		\ProgramName{compute-overlaps} would need to improved further to handle
@@
 
+As \Figure{fig:E_coli} shows, the bidirected string graph may still be very
+complicated even after transitive reduction and traversal.  In even worse
+examples, the graph could easily have thousands of vertices and edges.  We still
+need to figure out how to walk through the graph to produce the best assembly.
+
+An \KeyTerm{Eulerian path in a bidirected graph} is a walk that traverses each
+bidirected edge exactly one time (only in one of the two possible directions).
+Suppose we could find an Eulerian path in the bidirected string graph.  Then
+this path would correspond to an assembly that accounts for every overlap and
+therefore would be a valid assembly of the reads.  Furthermore, such a path may
+allow for vertices to be used for than one time in the assembly to allow for
+genomic repeats, which would not be true if we found a Hamiltonian path
+instead.\footnote{A Hamiltonian path is also inappropriate here because
+unbranched paths have been collapsed into single edges, which {\em must} be
+included in the assembly if the reads corresponding to the inner vertices are to
+take part in the assembly.  Furthermore, keep in mind that there is not a
+one-for-one correspondence between vertices and reads because contained reads
+were removed (see \Section{sec:removing_contained_reads)}.}
+
+However, there are multiple problems with the Eulerian path approach.  First,
+the graph may not be Eulerian.  In the case that the graph
+
+This implies we really would need to solve the
+bidirected equivalent of the Chinese Postman Problem instead.  Even then, the
+graph may not be fully connected, which would require that the final walk take
+the form of multiple discontinuous walks.  An equally severe problem is that
+some overlaps may be spurious, meaning that the two overlapping reads do not
+actually form a valid genomic sequence when combined.  Spurious overlaps can be
+caused by genomic repeats.  Therefore, we need to allow for the possibility that
+a given edge in the bidirected graph is omitted from the assembly if it seems
+like a good idea to do so.
+
+The way these problems can be solved is to (1) place a lower bound and an upper
+bound on the number of times each edge may be traversed, then (2) formulate and
+solve a minimum-cost network flow problem that finds the minimum total traversal
+count of edges in the graph in order for every vertex to be balanced (i.e. have
+the same indegree as out degree), then (3) find a generalized Eulerian tour
+through the graph, traversing each edge the exact number of times that was
+calculated in (2) and also allowing for the tour be discontinuous.
+
 \subsection{Calculating the A-statistic}
+
+We would like to use information from the data to restrict the number of times
+each edge may be traversed.  This can be done using a calculation called the
+\KeyTerm{A-statistic}, or \KeyTerm{arrival-rate statistic}, which is the
+logarithm of the probability that an edge represents unique sequence rather than
+duplicate sequence.  The A-statistic $A(e)$ of an edge $e$ in a bidirected
+string graph is:
+
+\[ A(e) = \frac{n}{G} \Delta - \ln{2} \cdot {k} \]
+
+where $n$ is the total number of reads, $G$ is the estimated genome length,
+$\Delta$ is the length of the edge in bp, and $k$ is the number of reads that
+support that edge.  To more precise, $k$ is the number of inner vertices in the
+unbranched path corresponding to $e$ before the unbranched paths were collapsed.
+
+Edges with a sufficiently high A-statistic are very likely to be single-copy
+edges and therefore should be traversed exactly one time.  The remaining edges
+may be traversed any number of times (with the exact number being determined
+during the network flow calculation), except for edges that originally were
+associated with one or more inner vertices, in which case we require the edge be
+traversed at least one time so that the reads corresponding to the inner
+vertices are used in the reconstruction of the genome (as every read was
+presumably sampled from the genome, in the absence of foreign DNA).
+
+Calculation of the A-statistic on each edge in the bidirected string graph is
+implemented by the program {\tt calculate-A-statistics}.
+
 \subsection{Mapping contained reads}
-\subsection{Network flow}
+
+\subsection{Minimum-cost network flow}
+
+Formally, a minimum-cost network flow problem on a directed graph takes as input
+for each edge $e$ an upper bound $c(e) \ge 0$ on flow and a cost per unit flow
+$v(e)$, and for each vertex $v$ a supply (positive) or demand (negative)
+$b(v)$ for flow.  It produces as output a flow $x(e)$ on each edge $e$ such
+that $\sum_e v(e)x(e)$ is minimized subject to $x(e) \in [0, c(e)]$ and
+$\sum_{u \to v}x(u \to v) + b(v) = \sum_{v \to w} x(v \to w)$.\cite{Myers2005}.
+
+We can roughly map our problem onto this definition assigning $c(e) = 0$ when
+$e$ is definitely single-copy and $c(e) = \infty$ otherwise, $b(v)$ the sum
+of the lower bounds on the traversal counts of all bidirected edges incident to
+$v$ with an inward head minus the sum of the lower bounds on the traversal
+counts of all edges incident to $v$ with an outward head, and $v(e) = 1$.  The
+resulting traversal count $t(e)$ for a bidirected edge $e$ is then the lower
+bound on the traversal count for that edge plus $x(e)$, the traversal count
+assigned in the network flow calculation.  The traversal count $t(e)$ represents
+the total number of times the edge may be traversed in a given assembly,
+counting traversals in both directions.
+
+There is, however, a serious complication with this network flow formulation
+which was not addressed in \cite{Myers2005}.  We have a bidirected graph, not a
+directed graph, so it cannot be trivially mapped to a directed network flow
+problem.  A possible workaround for this is to solve directed flow on the
+directed equivalent to the bidirected graph, described in
+\Subsection{subsec:directed}.  The calculated traversal counts for a directed
+edge $e$ between, for example, two vertices $f.E \to g.E$ is not guaranteed to
+be the same as the calculated traversal count for the edge that corresponds to
+the other direction of the implied bidirected edge, which would be $g.B \to f.B$
+in this example.  However, gWe can take the average of the two counts and use this as the
+traversal count for the corresponding bidirected edge, and this
+
+%vertices $f.X$ and $g.X
+
 \subsection{Generalized Eulerian path}

### 2012-12-10 src/BidirectedStringGraph.cc

@@ -226,14 +226,15 @@ void BidirectedStringGraph::build_from_digraph(const DirectedStringGraph & digra
 							 v_w.get_mapped_read_count()) / 2);
 				e.set_A_statistic((w_v.get_A_statistic() +
 						   v_w.get_A_statistic()) / 2);
-				if (w_v.get_num_inner_vertices() !=
-				    v_w.get_num_inner_vertices())
-				{
-					std::cerr << w_v << std::endl;
-					std::cerr << v_w << std::endl;
-					assert(0);
-				}
-				e.set_num_inner_vertices(w_v.get_num_inner_vertices());
+				//if (w_v.get_num_inner_vertices() !=
+				    //v_w.get_num_inner_vertices())
+				//{
+					//std::cerr << w_v << std::endl;
+					//std::cerr << v_w << std::endl;
+					//assert(0);
+				//}
+				e.set_num_inner_vertices((w_v.get_num_inner_vertices() +
+							 v_w.get_num_inner_vertices()) / 2);
 
 				edge_idx_t edge_idx = this->push_back_edge(e);
 				_vertices[v1_idx].add_edge_idx(edge_idx);

### 2012-12-10 src/DirectedStringGraph.h

@@ -250,7 +250,7 @@ public:
 			const DirectedStringGraphEdge & e_X = _edges[j];
 			assert((e_X.get_v1_idx() ^ 1) == e.get_v2_idx());
 			assert((e_X.get_v2_idx() ^ 1) == e.get_v1_idx());
-			assert(e_X.get_num_inner_vertices() == e.get_num_inner_vertices());
+			//assert(e_X.get_num_inner_vertices() == e.get_num_inner_vertices());
 		}
 	}
 

### 2012-12-10 presentation/paper.tex
paper: updates

@@
 on that organism.  However, currently available sequencing technology can only
-sequence short fragments of DNA.  As a result, whole-genome shotgun sequencing
-approaches to genome sequencing rely on programs called {\em genome assemblers}
-that can put together a set of short DNA fragments to reconstruct the original
-genome.  This paper provides an introduction and formalization of the genome
-assembly program, then introduces a model and algorithm for genome assembly that
-is based on a special type of graph called a \KeyTerm{bidirected string graph}.
-\end{abstract}
+sequence short fragments of DNA.  As a result, the whole-genome shotgun
+sequencing approach to genome sequencing relies on breaking up a sample of
+genomic DNA into fragments, sequencing those fragments, then using a program
+called a {\em genome assembler} to put together these short sequences to
+reconstruct the original, much longer sequence.  This paper provides an
+introduction and formalization of the genome assembly problem, then introduces a
+model and algorithm for genome assembly that is based on a special type of graph
+called a \KeyTerm{bidirected string graph}.  \end{abstract}
 
@@
 
+\subsection{Complement and reverse-complement sequences}
+
+The \KeyTerm{complement} of a single-stranded DNA sequence $S$ is the
+single-stranded DNA sequence obtained by making the replacements $\Base{A}
+\leftrightarrow \Base{T}$ and $\Base{C} \leftrightarrow \Base{G}$.  The
+\KeyTerm{reverse-complement} of a single-stranded DNA sequence $S$ is the
+complement of $S$ reversed.  Physically, the reverse-complement of a
+single-stranded DNA sequence corresponds to the corresponding sequence on the
+other strand of the DNA molecule.  It is the reverse-complement, not the
+``regular'' complement, because the two strands of a DNA molecule run in
+opposite directions.
+
+The \KeyTerm{canonical form} of a DNA sequence is the lexicographically lesser
+of that sequence and its reverse-complement.  This concept is useful in
+algorithms where a DNA sequence and its reverse-complement are considered
+equivalent.
+
 \subsection{Genome sequencing}
@@
 Therefore, to sequence an entire genome, which may be millions of base pairs
-long, the DNA from the genome must be broken into much smaller fragments, which
-are then sequenced.  Following this, the original genome is computationally
-reconstructed by a computer program called a \KeyTerm{genome assembler} in a
-process called \KeyTerm{genome assembly}.  \KeyTerm{Genome sequencing} therefore
-refers to the entire process of sequencing a genome, including both lab work and
-computational work, while \KeyTerm{genome assembly} only refers to the
-computational work.\footnote{It usually also takes some technology-specific
-computational work to actually prepare the sequences for assembly; for example,
-a series of digital photographs of clumps of fluorescently labeled molecules
-might need to be interpreted as reads using image-processing algorithms.
-However, these programs and/or algorithms generally have already been developed
-by the manufacturer of the sequencing machine and work fairly well, and ``genome
-assembly'' does not cover these algorithms, although genome assembly algorithms
-still may be tuned for specific sequencing technologies.}
+long, the DNA from the whole genome must be broken into much smaller fragments,
+which are then sequenced.\footnote{This specifically refers to the
+\KeyTerm{whole-genome shotgun sequencing} approach to genome assembly rather
+than alternatives such as \KeyTerm{BAC-to-BAC} sequencing that are able to
+restrict sequences to certain regions of the genome in order to make sequencing
+the genome easier (currently, at the disadvantage of increased cost and
+non-computational work).} Following this, the original genome must be
+computationally reconstructed by a computer program called a \KeyTerm{genome
+assembler} in a process called \KeyTerm{genome assembly}.  \KeyTerm{Genome
+sequencing} therefore refers to the entire process of sequencing a genome,
+including both lab work and computational work, while \KeyTerm{genome assembly}
+only refers to the computational work.\footnote{It usually also takes some
+technology-specific computational work to actually prepare the sequences for
+assembly; for example, a series of digital photographs of clumps of
+fluorescently labeled molecules might need to be interpreted as reads using
+image-processing algorithms.  However, these programs and/or algorithms
+generally have already been developed by the manufacturer of the sequencing
+machine and work fairly well, and ``genome assembly'' does not cover these
+algorithms, although assemblers still may be tuned for specific sequencing
+technologies.}
 
@@
 sample is obtained from grinding up plant or animal tissue, there could be some
-bacterial cells, each with its own genome, contaminating the sample.  Therefore,
-ideally an assembler should be prepared to completely ignore some reads.
-However, this problem is not considered in this paper.
+bacterial cells, each with its own genome, contaminating the sample.  Ideally,
+an assembler should be prepared to completely ignore a read if it seems too
+unusual.  However, this problem is not considered in this paper.
 
 For the purposes of this paper, the problem of genome assembly is simplified by
-assuming that the reads are error-free and uniformly (randomly) sampled from the
-genome.  In addition, it is assumed there are no reads from foreign DNA, and the
-reads are assumed to be unpaired, which means that mate-pair information, if
-available, is not used.  In other words, a read is defined here as an exact
-substring from either strand of any part of the genome.
+assuming the following characteristics of the reads:
+
+\begin{itemize}
+	\item The reads are error-free and uniformly sampled from the genome.
+	\item There are no reads from foreign DNA.
+	\item The reads are unpaired, meaning that mate-pair information, if
+available, is not used.
+\end{itemize}
+
+In other words, a read is defined here as an exact substring from either strand
+of any part of the genome.
 
@@
 accurately.  However, real genomes almost always contain repeats of substantial
-length, so genome assembly algorithms must be able to handle repeats.
+length, so assemblers must be able to handle repeats.
 
@@
 Genome assembly is superficially similar to the shortest common superstring
-(SCS) problem, which is to find the shortest possible string that contains every
+problem, which is to find the shortest possible string that contains every
 string in a given set as substrings.  Genome assembly also requires finding a
-longer string that contains all of a set of substrings.  However, the actual
-genome may be more than one string, and it genome may be longer than the
-shortest possible genome from which the reads could have been sampled, due to
-repeats.  In addition,  genomes are double-stranded, and there may be other
-complications with the reads (described in subsection \ref{subsec:reads}), such
-as sequencing error.  Therefore, genome assembly is not the same as the SCS
-problem, and in fact seems to be harder, despite the fact that the SCS problem
-is already NP-complete\cite{Turner1989}.
-
-In fact, some models of genome assembly have also been proven to be NP-complete
-\cite{Medvedev2007}.  However, it is not acceptable to have a genome assembly
-algorithm that takes exponential time to complete.  Instead, heuristic
-algorithms that tend to produce good results but also tend to have a reasonable
-running time must be designed.
+longer string that, roughly, contains every string in  a set of substrings.
+However, the actual genome may be more than one string, and the genome may also
+be longer than the shortest possible genome from which the reads could have been
+sampled, due to repeats.  In addition,  genomes are double-stranded, and there
+may be other complications with the reads, such as sequencing error.  Therefore,
+genome assembly is not the same as the shortest common superstring problem, and
+in fact seems to be harder, despite the fact that the shortest common
+superstring problem is already NP-complete\cite{Turner1989}.
+
+In fact, some models of genome assembly have also been proven to be
+NP-complete\cite{Medvedev2007}.  However, it is not acceptable to have a genome
+assembly algorithm that takes exponential time to complete.  Instead, heuristic
+algorithms that tend to produce good results on real inputs while still having
+reasonable running times must be designed.
 
@@
 This paper explores the genome assembly problem through an algorithm published
-in 2005 by Eugene Myers\cite{Myers2005} that is based on the idea of the
-fragment string assembly graph.  The algorithm provides a good overview of the
+in 2005 by Eugene Myers that is based on the idea of the fragment string
+assembly graph\cite{Myers2005}.  The algorithm provides a good overview of the
 techniques that are used in genome assembly, especially since it draws on ideas
@@
 EULER-SR\cite{Pevzner2001}, although it does not use a de Bruijn graph for its
-assembly model.  The algorithm also introduces some new ideas.
-
-The overall fragment string graph algorithm is as follows: first, use overlaps
-between reads to build a graph that models the assembly problem.  The idea is
-that walks through this graph model a way in which reads may be assembled
-together when the overlaps are used as evidence for joining two reads.  The next
-step is to simplify and analyze the graph; and finally, find paths through the
-graph are found to reconstruct the original genome.
+assembly model.  In fact, the algorithm is called the {\em fragment} string
+graph algorithm because the graph is built directly from the sequenced fragments
+(or reads), rather than subsequences ($k$-mers) taken from them.
+
+The algorithm also introduces new ideas that were not used in previous
+publications or assemblers.
+
+The overall fragment string graph algorithm can be described as follows.  First,
+use overlaps between reads to build a graph that models the assembly problem.
+The idea is that walks through this graph model a way in which reads may be
+assembled together consistently when the overlaps are used as evidence for
+joining pairs of reads.  The next step is to simplify and analyze the graph, and
+the last step is to find paths through the graph to reconstruct the original
+genome.\footnote{The fragment string graph algorithm, as described in
+\cite{Myers2005}, does not involve any finishing steps such as scaffolding
+(using mate-pair information to improve the assembly), although ideas for doing
+this are outlined in \cite{Medvedev2009}.  However, in this paper we assume that
+no mate-pair information is available.}
 
@@
 			%matched region on the read containing $\OccTwo$}
-			%\STATE{${\tt plen} \gets$ the length of the maximall
+			%\STATE{${\tt plen} \gets$ the length of the maximally
 			%matched region on the read containing $\OccTwo$}
@@
 
-A \KeyTerm{walk in a bidirected graph} $G$ is a continuous sequence of edges in
+The \KeyTerm{in-degree} of a vertex $v$ in a bidirected graph is the number of
+heads adjacent to $v$ that point inwards, while the \KeyTerm{out-degree} of $v$
+is the number of heads adjacent to $v$ that point outwards.  A bidirected graph
+$G$ is \KeyTerm{balanced} if for every vertex in $G$ the in-degree is equal to
+the out-degree.
+
+A \KeyTerm{walk} in a bidirected graph $G$ is a continuous sequence of edges in
 $G$ such that if any vertex $v$ is entered through a head inwards, it is exited
@@
 	} \end{center}
-	\caption{$1 \to 2 \to 3 \to 4$ and $4 \to 3 \to 2 \to 1$ are both valid walks in
-	the bidirected graph below.  (Check circled pairs of heads.)}
+	\caption{$1 \to 2 \to 3 \to 4$ and $4 \to 3 \to 2 \to 1$ are both valid
+	walks in the bidirected graph above.  (Check circled pairs of heads to
+	verify this.)}
 	\label{fig:bidiwalk_example}
@@
 A \KeyTerm{cycle} in a bidirected graph is a walk that begins and ends at the
-same vertex in such a way that it would be valid to trace the walk again when
+same vertex in such a way that it would be a valid to trace the walk again when
 the starting vertex is reached at the end of the cycle.
 
-A bidirected graph is \KeyTerm{connected} if the undirected graph produced by
-replacing all bidirected edges with undirected edges is connected.  A bidirected
-graph is \KeyTerm{strongly connected} if it is connected and for every edge
-there is a cycle containing it\cite{Medvedev2009}.
-
-The concept of a walk through a bidirected graph maps quite nicely onto the
-genome assembly problem, since a walk through the bidirected string graph
-constructed from reads and overlaps models a way in which the reads can be
+A bidirected graph is \KeyTerm{connected} if and only if  the undirected graph
+produced by replacing all bidirected edges with undirected edges is connected as
+defined for undirected graphs.  A bidirected graph is \KeyTerm{strongly
+connected} if and only if it is connected and for every edge there is a cycle
+containing it.\cite{Medvedev2009}
+
+The concept of a walk in a bidirected graph maps quite nicely onto the genome
+assembly problem because a walk through the bidirected string graph constructed
+from a set of reads and their overlaps models a way in which the reads can be
 assembled together consistently.  In such a walk, each read may be used in
@@
 These interpretations can be generalized to the other two types of bidirected
-edges.  Furthermore, these intepretations imply that the actual sequence spelled
+edges.  Furthermore, these interpretations imply that the actual sequence spelled
 out by a walk in the bidirected graph may include implied (non-labeled) sequence
@@
 complicated even after transitive reduction and traversal.  In even worse
-examples, the graph could easily have thousands of vertices and edges.  We still
-need to figure out how to walk through the graph to produce the best assembly.
+examples, the graph could easily have thousands of vertices and edges.
+Therefore, we need to figure out how to walk through the graph to produce the
+best assembly.
 
 An \KeyTerm{Eulerian path in a bidirected graph} is a walk that traverses each
-bidirected edge exactly one time (only in one of the two possible directions).
-Suppose we could find an Eulerian path in the bidirected string graph.  Then
-this path would correspond to an assembly that accounts for every overlap and
-therefore would be a valid assembly of the reads.  Furthermore, such a path may
-allow for vertices to be used for than one time in the assembly to allow for
-genomic repeats, which would not be true if we found a Hamiltonian path
-instead.\footnote{A Hamiltonian path is also inappropriate here because
-unbranched paths have been collapsed into single edges, which {\em must} be
-included in the assembly if the reads corresponding to the inner vertices are to
-take part in the assembly.  Furthermore, keep in mind that there is not a
-one-for-one correspondence between vertices and reads because contained reads
-were removed (see \Section{sec:removing_contained_reads)}.}
+bidirected edge exactly one time in either direction.  Suppose we could find an
+Eulerian path in the bidirected string graph.  Then this path would correspond
+to an assembly that accounts for every overlap and therefore would be a valid
+assembly of the original reads.  Furthermore, such a path may allow for vertices
+to be used for than one time in the assembly to allow for genomic repeats, which
+would not be true if we found a Hamiltonian path instead.\footnote{A Hamiltonian
+path is also inappropriate here because unbranched paths have been collapsed
+into single edges, which {\em must} be included in the assembly if the reads
+corresponding to the inner vertices are to take part in the assembly.
+Furthermore, keep in mind that there is not a one-for-one correspondence between
+vertices and reads because contained reads were removed (see
+\Subsection{subsec:removing_contained_reads}).}
 
 However, there are multiple problems with the Eulerian path approach.  First,
-the graph may not be Eulerian.  In the case that the graph
+the graph may not be Eulerian.  This can happen if the graph is not strongly
+connected, not balanced (indegree always equal to out
 

### 2012-12-10 presentation/paper.tex
paper: updates

@@
 	Consider two reads, $f$ and $g$, and their reverse-complements $f'$ and
-	$g'$.  The reads $f$ and $g$ are said to {\bf exactly overlap} by some
-	length $\LengthVar$ if and only if at least one of the following is
-	true:
-	\begin{enumerate}
+	$g'$.  The reads $f$ and $g$ are said to \KeyTerm{exactly overlap} by some
+	length $\LengthVar$ if and only if at least one of the following is true:
+	\begin{itemize}
 		\item The last $\LengthVar$ bp of $f$ exactly match the first
@@
 		$\LengthVar$ bp of $g'$.
-	\end{enumerate}
+	\end{itemize}
 %\end{definition}
@@
 Overlaps may also be \KeyTerm{inexact}.  An inexact overlap is an overlap where
-overlapped regions of the two reads do not match exactly (in forward or
-reverse-complement form), but instead differ by a number of base-pair
+overlapped regions of the two reads do not match exactly (in either a forward or
+reverse-complement manner), but instead differ by a number of base-pair
 substitutions, insertions, and/or deletions up to a predetermined upper bound,
@@
 
-Overlaps are important because two reads that overlap may come from adjacent
-positions on the genome.  Although due to repeats, we cannot be certain that two
-overlapping reads actually come from adjacent positions, we {\em can} say for
-certain that two reads that do not overlap do not come from adjacent positions
-on the genome.  Regardless of this, \Figure{fig:overlap_examples} shows how two
+Overlaps are important because two reads that overlap may have come from
+adjacent positions on the genome.  Although due to repeats, we cannot be certain
+that two overlapping reads actually came from adjacent positions, we {\em can}
+say for certain that two reads that do not overlap did not come from adjacent
+positions.  Regardless of this, \Figure{fig:overlap_examples} shows how two
 reads that overlap potentially could be combined to form a longer sequence.
@@
 Now that the idea of an exact overlap has been formalized, what we would like to
-do is compute all pairwise exact overlaps of minimum length $L$ among the reads
-that were provided as input to the assembler.
-
-The naive way to do this is to consider every pair of reads in the input,
-including only distinct pairs but also including the special cases of each read
-paired with itself, and search for each of the types of overlaps between the two
-reads in each pair.  If there are $N$ reads in the input, this algorithm will
-need to consider $\frac{N^2}{2} + N$ pairs of reads, and the time spent on each
-read pair will depend several factors, including the minimum overlap length and
-the read lengths.  But even if we could somehow spend constant time on each read
-pair, this algorithm is much too slow because the running time rises quadraticly
-with respect to the number of reads.  This is important because there can easily
-be millions of reads in the input.
+do is compute all pairwise exact overlaps of some minimum length $L$ among the
+	reads that were provided as input to the assembler.  The chosen value of $L$
+	must be balanced so that we find can enough overlaps to adequately cover the
+	genome (small $L$ better) but still exclude most spurious overlaps (large
+	$L$ better).
+
+The naive way to compute overlaps is to consider every pair of reads in the
+input, including only distinct pairs but also including the special cases of
+each read paired with itself, and search for each of the types of overlaps
+between the two reads in each pair.  If there are $N$ reads in the input, this
+algorithm will need to consider $\frac{N^2}{2} + N$ pairs of reads. The time
+spent on each read pair will depend several factors, including the minimum
+overlap length $L$ and the read lengths.  But even if we could somehow spend
+constant time on each read pair, this naive algorithm would be prohibitively
+slow because the running time rises quadraticly with respect to the number of
+reads, and there can easily be millions of reads in the input.
 
@@
 works by indexing the reads by short subsequences of a certain length (less than
-or equal to than the minimum overlap length) and then only comparing reads that
-actually share a subsequence.  Each of these short subsequences is called a
+or equal to than the minimum overlap length $L$) and then only comparing reads
+that actually share a subsequence.  Each of these short subsequences is called a
 \KeyTerm{seed}.  Two reads may share a seed in either a forward or
 reverse-complement manner.  Given two reads that share a seed, the seed can be
-extended outwards in both directions in order to find a maximal matched region,
-that may be either forward or reverse-complement, between the two reads.  Then,
-this maximal matched region is compared against the definition of an overlap.
-In particular, it must be at least as long as the minimum overlap length, it
-must have least 2 extreme points over the two reads, and it cannot be the prefix
-or suffix of both reads if the match does not extend to the opposite end of at
-least one of the reads.  If all of these conditions are met and the same overlap
-was not already found through another seed in the same two reads, this overlap
-is added to the set of overlaps.
+extended outwards in both directions to find a maximal matched region (that may
+be either forward or reverse-complement) between the two reads.  Then, this
+maximal matched region is compared against the definition of an overlap.  In
+particular, it must be at least as long as the minimum overlap length, it must
+have least 2 extreme points over the two reads, and it cannot be the prefix or
+suffix of both reads if the match does not extend to the opposite end of at
+least one of the reads.  If all of these conditions are met, this maximal matched
+region is translated into an overlap and added to the set of overlaps if not
+already present.
 

### 2012-12-11 presentation/paper.tex
paper: Implementation, more citations

@@
-\documentclass[10pt]{article}
+\documentclass[12pt]{article}
 \usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
@@
 \usepackage{comment}
-\usepackage{multicol}
+\usepackage{mathptmx}
 %\usepackage{amsthm}
@@
 Sequencing an organism's genome is an important step to further genetic research
-on that organism.  However, currently available sequencing technology can only
+on that organism.  However, currently available sequencing technologies can only
 sequence short fragments of DNA.  As a result, the whole-genome shotgun
@@
 
-The running time of the seed-and-extend algorithm, as described above, is
+The running time of this seed-and-extend algorithm is
 typically much faster than the naive algorithm, although it is highly dependent
-on the data.  For a string of the genome of length $G$ and with a seed length
-$k$, there will be at most $G-k+1$ canonical seeds.\footnote{The canonical form
-of a DNA sequence and its reverse-complement is the lexicographically lesser of
-the two.  Here, we refer canonical form of each seed, because we wish to compare
-reads that share a seed in either a forward or reverse-complement manner.}  For
-each of these seeds, the reads containing it will be examined exhaustively in a
-pairwise manner, which takes quadratic time with respect to the number of reads
-sharing that seed.  Since the more densely reads are sampled from the genome,
-the more reads will contain each seed,  the running time of the seed-and-extend
-algorithm rises quadratically with regards to read density.  In addition, the
-number of read pairs the algorithm considers for a specific seed also rises
-quadratically with the number of times that seed is repeated in the genome due
-to genomic repeats.  Overall, the seed-and-extend algorithm is not ideal and is
-still a bottleneck of the entire assembly algorithm, despite the fact that it is
-much faster than the naive algorithm (except in degenerate cases, such as a
-``genome'' that is entirely a single repeated letter).
+on the data.  The number of distinct seeds is at worst linear with respect to
+the genome length.
+%of a DNA sequence and its reverse-complement is the lexicographically lesser of
+%the two.  Here, we refer canonical form of each seed, because we wish to
+%compare reads that share a seed in either a forward or reverse-complement
+%manner.}
+For each of these distinct seeds, the reads containing it are examined pairwise,
+which takes quadratic time with respect to the number of reads sharing the
+seed.  Since more reads will contain each seed when the reads are sampled more
+densely from the genome, the running time of the seed-and-extend algorithm rises
+quadratically with respect to read density.  In addition, the number of read
+pairs the algorithm considers for a specific seed also rises quadratically with
+the number of times that seed is repeated in the genome due to genomic repeats.
+Overall, the seed-and-extend algorithm is not ideal and is still a bottleneck of
+the entire assembly algorithm, despite the fact that it is much faster than the
+naive algorithm (except in degenerate cases, such as a ``genome'' that is
+entirely a single repeated letter).
 
@@
 		graph was produced by the code that has been discussed, but the actual
-		drawing of the picture was done by {\it GraphViz}.  The 46,387 reads
+		drawing of the picture was done by {\it
+		GraphViz}\cite{GraphVizPublication}.  The 46,387 reads
 		that were provided as input to the assembler were generated by a Perl
@@
 \section{Producing the final assembly}
+
+\label{sec:final}
+
 \subsection{Overview}
@@
 
+\label{sec:mapping_contained_reads}
+
 \subsection{Minimum-cost network flow}
@@
 
+\section{Implementation}
+
+The implementation of the fragment string graph algorithm that was done for this
+project is a set of C++ programs that iteratively transform the data on-disk
+into the final assembly.  Each program takes as input one or more files as well
+as zero or more other command-line arguments and produces as output one or more
+files.  Each file contains data such as a list of reads, a list of overlaps, or
+a graph.  Internally to the assembler pipeline, data is stored in binary form.
+\Figure{fig:impl_graph} shows how the data flows through the main stages of the
+assembler.
+
+The input to the assembler is one or more files in plain-text form that each
+contain a list of reads.  The final output of the assembler will eventually be a
+file in plain-text format that gives the list of assembled contigs.  Currently,
+the final stages of the algorithm (\Section{sec:final}, except for the
+calculation of the A-statistics) have not been implemented as of the end of the
+semester of Fall 2012.
+
+\begin{figure}
+	\begin{center}
+		\includegraphics[width=0.7\textwidth]{programs.pdf}
+	\end{center}
+	\caption{The flow of data through the assembly pipeline.  Each node is a
+	program, each arrow represents the output of a program being used as input
+	in another program.  Yellow programs have not yet been implemented as of
+	December 10, 2012.}
+	\label{fig:impl_graph}
+\end{figure}
+
 \section{Possible future work}
@@
 \begin{itemize}
-\item Finish the network flow and generalized Eulerian path parts of the algorithm
-so that the implementation is ``complete'' and can produce a ``final'' assembly
-\item Improve performance of \ProgramName{compute-overlaps}
+\item Finish the network flow and generalized Eulerian path parts of the
+algorithm so that the implementation is ``complete'' and can produce a ``final''
+assembly.
+\item Improve the performance of \ProgramName{compute-overlaps}.
 \item Allow for sequencing errors by finding inexact overlaps and modifying other
 stages of the algorithm, such as transitive reduction, to allow for the
-possibility that overlaps are not exact.
+possibility that overlaps are inexact.
 \item Compensate for sequencing errors by attempting to correct them, either as
 part of \ProgramName{compute-overlaps} or as a new stage of the algorithm.
-\item Consider more recent publications that explore variations on this algorithm,
-such as parellizing the algorithm on a cluster \cite{X}, or using the
-Burrows-Wheeler transform to directly and efficiently construct the bidirected
-string graph directly from the reads\cite{X}.
+\item Consider more recent publications that explore variations on this
+algorithm, such as parallelizing the algorithm \cite{Jackson2008}, or using the
+Burrows-Wheeler transform to efficiently construct the transitively-reduced
+bidirected string graph directly from the reads\cite{Simpson2010}.
 \item Use mate-pair information to help resolve repeats and connect assembled
-sequences even when they are separated by unknown sequence
+sequences even when they are separated by unknown sequence, possibly using ideas
+from \cite{Medvedev2009}.
 \end{itemize}
-
-\label{sec:mapping_contained_reads}
-
-\section*{Acknowledgments}
-
-Acknowledgements
 

### 2012-12-11 presentation/paper.tex
paper: updates

@@
 \usepackage{mathptmx}
+\usepackage{microtype}
+\usepackage{array}
 %\usepackage{amsthm}
@@
 sequencing approach to genome sequencing relies on breaking up a sample of
-genomic DNA into fragments, sequencing those fragments, then using a program
+genomic DNA into fragments, sequencing these fragments, then using a program
 called a {\em genome assembler} to put together these short sequences to
@@
 \Base{C}, \Base{G}, and \Base{T}, respectively.  \Base{A} always pairs with T on
-the other strand, and \Base{C} always pairs with \Base{G} on the other strand.
+the opposite strand, and \Base{C} always pairs with \Base{G} on the opposite strand.
 In addition, each strand has a direction, and the two strands of a DNA molecule
@@
 symbols \Base{A}, \Base{C}, \Base{G}, and \Base{T}, where dual-stranded has the
-meaning indicated above: the two strands run in opposite directions, and the
-pairings \Base{A}---\Base{T} and \Base{C}---\Base{G} always hold.
+meaning indicated above (the two strands run in opposite directions, and the
+pairings \Base{A}---\Base{T} and \Base{C}---\Base{G} always hold).
 
@@
 	\begin{center}
-		\begin{tabular}{|l|p{4.5cm}|}
+		\begin{tabular}{|c|>{\centering\arraybackslash}p{5.2cm}|}
 			%\hline
@@
 			\hline
-			{\bf Organism} & {\bf Approximate genome size (in base
-			pairs)} \\
+			{\bf Organism} & {\bf Approximate genome size} \\
 			\hline
-			{\it Variola} virus (smallpox) &  186,000 \\
+			{\it Variola} virus (smallpox) &  186,000 bp\\
 			\hline
-			{\it E. coli} (a bacterium) & 4,600,000 \\
+			{\it E. coli} (a bacterium) & 4,600,000 bp \\
 			\hline
-			{\it Ananas comosus} (domesticated pineapple) & 500,000,000 \\
+			{\it Ananas comosus} (domesticated pineapple) & 500,000,000 bp \\
 			\hline
-			{\it Zea mays} (domesticated corn) & 2,000,000,000 \\
+			{\it Zea mays} (domesticated corn) & 2,000,000,000 bp \\
 			\hline
-			{\it Homo sapiens} (human)       &  3,200,000,000 \\
+			{\it Homo sapiens} (human)       &  3,200,000,000 bp \\
 			\hline
@@
 
-The \KeyTerm{canonical form} of a DNA sequence is the lexicographically lesser
-of that sequence and its reverse-complement.  This concept is useful in
+The \KeyTerm{canonical form} of a DNA sequence $S$ is the lexicographically
+lesser of $S$ and its reverse-complement $S'$.  This concept is useful in
 algorithms where a DNA sequence and its reverse-complement are considered
@@
 are used as evidence.  To do this, define a graph where each vertex corresponds
-to a read and each edge corresponds to an overlap.  Two vertices are connected
-by an edge if and only if the corresponding reads share an overlap.  In
-addition, each edge is labeled with an appropriate DNA sequence, the details of
-which will be explained later.
+to a read\footnote{Actually, more than one read may map to the same vertex, as
+explained in \Subsection{subsec:removing_contained_reads}.} and each edge
+corresponds to an overlap.  Two vertices are connected by an edge if and only if
+the corresponding reads share an overlap.  In addition, each edge is labeled
+with an appropriate DNA sequence, the details of which will be explained later.
 
 It needs to be decided whether each edge will be directed, undirected, or
-something else.  To decide this, we need to consider exactly what we want each
-edge to actually represent.  Consider three reads $f$, $g$, and $h$, that
-overlap as follows:
+something else.  To decide this, we need to consider exactly what each edge
+needs to represent.  Consider three reads $f$, $g$, and $h$, that overlap as
+follows:
 
@@
 
-There are 3 overlaps so there will be 3 edges--- one between $f$ and $g$, one
+There are 3 overlaps, so there will be 3 edges--- one between $f$ and $g$, one
 between $g$ and $h$, and one between $f$ and $h$.  As shown, the reads can be
@@
 
-We might consider a directed graph, but this does not work because the graph
-does not allow the reverse-complement walks such as $h \to g \to f$:
-
+We might consider a directed graph as shown in
+\Subfigure{subfig:possible_directed}, but this does not work because the graph
+does not allow the reverse-complement walks such as $h \to g \to f$.
+
+\begin{figure}
 	\begin{center}
+		\begin{subfigure}{0.32\textwidth}
 			\begin{tikzpicture}[>=triangle 45]
@@
 			\end{tikzpicture}
-\end{center}
-
-Similarly, we might consider an undirected graph, but this does not work because
-the graph fails to disallow walks such as $g \to h \to f$ that correspond to
-assemblies that are not supported by the underlying reads.
-
-\begin{center}
+			\caption{A possible directed \\representation of the \\string graph.}
+			\label{subfig:possible_directed}
+		\end{subfigure}
+		\begin{subfigure}{0.32\textwidth}
 			\begin{tikzpicture}[>=triangle 45]
@@
 			\end{tikzpicture}
+			\caption{A possible undirected \\representation of the \\string graph.}
+			\label{subfig:possible_undirected}
+		\end{subfigure}
+		\begin{subfigure}{0.32\textwidth}
+			\begin{tikzpicture}[>=triangle 45]
+				\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+				\node (f) at (0, 0) {f};
+				\node (g) at (1.7, 0) {g};
+				\node (h) at (3.4, 0) {h};
+				\draw[>->,style=very thick] (f) edge (g);
+				\draw[>-<,style=very thick] (g) edge (h);
+				\draw[>-<,style=very thick,bend left] (f) edge (h);
+			\end{tikzpicture}
+			\caption{A possible bidirected \\representation of the \\string graph.}
+			\label{subfig:possible_bidirected}
+		\end{subfigure}
+		\caption{Three possibilities for constructing the string graph.  The
+		bidirected construction works the best in terms of modeling the
+		underlying problem.}
 	\end{center}
+\end{figure}
+
+Similarly, we might consider an undirected graph as shown in
+\Subfigure{subfig:possible_undirected}, but this does not work because
+the graph fails to disallow walks such as $g \to h \to f$ that correspond to
+assemblies that are not supported by the underlying reads.
 
@@
 The solution to all these problems is to use a new type of graph called a
-\KeyTerm{bidirected graph}.
-\begin{center}
-	\begin{tikzpicture}[>=triangle 45]
-		\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
-		\node (f) at (0, 0) {f};
-		\node (g) at (1.7, 0) {g};
-		\node (h) at (3.4, 0) {h};
-		\draw[>->,style=very thick] (f) edge (g);
-		\draw[>-<,style=very thick] (g) edge (h);
-		\draw[>-<,style=very thick,bend left] (f) edge (h);
-	\end{tikzpicture}
-\end{center}
+\KeyTerm{bidirected graph}, shown in \Subfigure{subfig:possible_bidirected}.
 
@@
 There are 3 (or 4, ignoring symmetry) types of \KeyTerm{bidirected edges}.  They
-follow directly from the different types of overlaps (see
-\Figure{fig:overlap_mapping}).
+follow directly from the different types of overlaps, as shown in
+\Figure{fig:overlap_mapping}.
 
@@
 The \KeyTerm{in-degree} of a vertex $v$ in a bidirected graph is the number of
-heads adjacent to $v$ that point inwards, while the \KeyTerm{out-degree} of $v$
-is the number of heads adjacent to $v$ that point outwards.  A bidirected graph
-$G$ is \KeyTerm{balanced} if for every vertex in $G$ the in-degree is equal to
-the out-degree.
+heads adjacent to $v$ that point inwards (towards $v$), while the
+\KeyTerm{out-degree} of $v$ is the number of heads adjacent to $v$ that point
+outwards (away from $v$).  A bidirected graph $G$ is \KeyTerm{balanced} if and
+only if for every vertex in $G$ the in-degree is equal to the out-degree.
 
@@
 same vertex in such a way that it would be a valid to trace the walk again when
-the starting vertex is reached at the end of the cycle.
+the starting vertex is reached at the end of the walk.
 
 A bidirected graph is \KeyTerm{connected} if and only if  the undirected graph
-produced by replacing all bidirected edges with undirected edges is connected as
-defined for undirected graphs.  A bidirected graph is \KeyTerm{strongly
+produced by replacing all bidirected edges with undirected edges is connected
+(as defined for undirected graphs).  A bidirected graph is \KeyTerm{strongly
 connected} if and only if it is connected and for every edge there is a cycle
-containing it.\cite{Medvedev2009}
+containing it\cite{Medvedev2009}.
 
@@
 from a set of reads and their overlaps models a way in which the reads can be
-assembled together consistently.  In such a walk, each read may be used in
-forward or reverse-complement orientation, depending on the direction, relative
-to the walk, of the adjacent arrow heads when the corresponding vertex is passed
-through.
+assembled together consistently\cite{Myers2005}.  In such a walk, each read may
+be used in forward or reverse-complement orientation, depending on the
+direction, relative to the walk, of the adjacent arrow heads when the
+corresponding vertex is passed through.
 
@@
 These interpretations can be generalized to the other two types of bidirected
-edges.  Furthermore, these interpretations imply that the actual sequence spelled
-out by a walk in the bidirected graph may include implied (non-labeled) sequence
-at the start (if we start at the end of a read) and/or the end (if we end at the
+edges, and this is described in slightly more detail in \cite{Myers2005}.
+Furthermore, these interpretations imply that the actual sequence spelled out by
+a walk in the bidirected graph may include implied (non-labeled) sequence at the
+start (if we start at the end of a read) and/or the end (if we end at the
 beginning of a read).
@@
 		\caption{The corresponding directed string graph fragment}
-		\label{subfig:bidiedge_interp_a}
+		\label{subfig:bidiedge_interp_c}
 	\end{subfigure}
@@
 
-\section{Transitive reduction}
+\section{Simplifying the graph}
+
+\subsection{Transitive reduction}
 
@@
 
-\section{Collapsing unbranched paths}
+\subsection{Collapsing unbranched paths}
 

### 2012-12-11 presentation/paper.tex
paper: updates

@@
 \newcommand{\Section}[1]{Section \ref{#1}}
-\newcommand{\Subsection}[1]{Subsection \ref{#1}}
+\newcommand{\Subsection}[1]{Section \ref{#1}}
 \newcommand{\Algorithm}[1]{Algorithm \ref{#1}}
@@
 Consider the the string graph fragment shown in
-\Subfigure{subfig:bidiedge_interp_a}.  When we walk from $f \to g$, we are using
-both $f$ and $g$ in forward orientation, based on the arrow heads.  But what
-this really means is that we are walking from the {\em end} of read $f$ to the
-{\em end} of read $g$.  In other words, traversing the edge in that direction is
-associated with the addition of the sequence after the overlapped region in $g$,
-up to the end of the $g$.
-
-If we instead walk $g \to f$, we instead use both $g$ and $f$ in
-reverse-complement orientation by walking from the {\em beginning} of read $g$
-to the {\em beginning} of read $f$.  In other words, traversing the edge in that
-direction is associated with the addition of the sequence before the overlapped
-region in $f$, up to the beginning of $f$, and this sequence must be
-reverse-complemented because it is being used opposite its original direction.
+\Subfigure{subfig:bidiedge_interp_a}.  When we walk $f \to g$, both $f$ and $g$
+are used in forward orientation, as indicated by the directions of the adjacent
+arrow heads.  But what this really means is that we are walking from the {\em
+end} of read $f$ to the {\em end} of read $g$.  In other words, traversing the
+edge in that direction is associated with the addition of the sequence after the
+overlapped region in $g$, up to the end of the $g$.
+
+If we instead walk $g \to f$, both $g$ and $f$ are used reverse-complement
+orientation by walking from the {\em beginning} of read $g$ to the {\em
+beginning} of read $f$.  In other words, traversing the edge in that direction
+is associated with the addition of the sequence before the overlapped region in
+$f$, up to the beginning of $f$, and this sequence must be reverse-complemented
+because it is being used opposite its original direction.
 
@@
 start (if we start at the end of a read) and/or the end (if we end at the
-beginning of a read).
+beginning of a read) of the walk.
 
@@
 			\begin{tikzpicture}[>=\ReadArrowType,scale=0.8]
-				\draw[->,very thick] (0, 0) node[anchor=east] {$f.B$}
+				\draw[->,very thick] (-0.2, 0) node[anchor=east] {$f.B$}
 						-> (4, 0) node[anchor=west] {$f.E$};
 				\draw[->,very thick] (2.09, -0.9) node[anchor=east] {$g.B$}
-						-> (6.09, -0.9) node[anchor=west] {$g.E$};
-
-				\draw (0, -0.1) node[anchor=south west]
-					{\hspace*{-1.4mm}{\small \tt
+						-> (6.29, -0.9) node[anchor=west] {$g.E$};
+
+				\draw (-0.2, -0.1) node[anchor=south west]
+					{\hspace*{-1.5mm}{\small \tt
 						%GATTACACAGTGTAGTT
-						GATTACACA\hspace*{0.26cm}$a_1 \dots a_n$
+						GATTACACA\hspace*{1.5mm}$a_1 \dots a_n$
 					}};
 
-				\draw (0, 0.1) node[anchor=north west]
+				\draw (-0.2, 0.1) node[anchor=north west]
 					{\hspace*{-2.6mm}
@@
 				\draw (2.09, -0.9) node[anchor=north west]
-					{\hspace*{12.48mm}
+					{\hspace*{12.88mm}
 						\rotatebox{180}{{\small \tt TACTTCTAA}}};
@@
 a directed graph instead a bidirected graph, which can be helpful for the
-implementation.  The disadvantage is that two vertices are required for each
-read, they may be widely separated in the graph.
+implementation.  The disadvantage is that there are two completely separate
+vertices for each read.
 
@@
 
-The contained reads cannot, however be fully discarded, as they are needed for a
+This implies that a vertex in the bidirected string graph may actually
+correspond to more than one read.
+
+The contained reads cannot be completely discarded because they are needed for a
 later step (see \Section{sec:mapping_contained_reads}).
 
-\subsection{Implementation}
-
-As described in \Subsection{subsec:removing_contained_reads}, before we even
-build the graph, contained reads need to be removed.  This is implemented by the
-program \ProgramName{remove-contained-reads}.  It takes as input two files: the
-set of reads and the set of overlaps that were computed from them, and produces
-as output three files: the set of uncontained reads, the set of overlaps
-excluding those that involve contained reads, and an array that maps from the
-original read indices to the new read indices.
+\subsection{Implementation of assembly graph construction}
+
+As described in \Subsection{subsec:removing_contained_reads}, before the
+assembly graph is even built, contained reads need to be removed.  This is
+implemented by the program \ProgramName{remove-contained-reads}.  It takes as
+input two files: the set of reads and the set of overlaps that were computed
+from them, and produces as output three files: the set of uncontained reads, the
+set of overlaps excluding those that involve contained reads, and an array that
+maps from the original read indices to the new read indices.
 
@@
 \ProgramName{print-string-graph}, was implemented that can print a string graph
-in plain-text format, or print a string graph in the DOT language for input to
-the graph-drawing software {\it GraphViz}\cite{GraphVizPublication}, or print
-some other information about a string graph.
+in plain-text format, print a string graph in the DOT language for input to the
+graph-drawing software {\it GraphViz}\cite{GraphVizPublication}, or print some
+other information about a string graph.
 
@@
 
-\section{Simplifying the graph}
+\section{Simplifying the assembly graph}
 
@@
 reads $f$, $g$, and $h$, $f$ will overlap $h$ as well as $g$.  This is redundant
-because we can, equivalently, walk $f$ to $g$ to $h$ rather than directly
-walking from $f$ to $h$.  \Figure{fig:transitive} shows an example of this.
+because we can, equivalently, walk $f \to g \to h$ rather than directly walking
+from $f \to h$.  \Figure{fig:transitive} shows an example of this.
 
@@
 
-\Algorithm{alg:transitive_reduction} produces pseudocode for a transitive
-reduction algorithm.  Essentially, the algorithm is to through each vertex $v$
-and examine vertices up to 2 edges away from $v$ to identify all transitive
-edges leaving $v$.  These edges are then marked for removal, and removed after
-all vertices in the graph have been considered.  Some optimizations are
-possible; for example, the lengths of the sequences the edges are labeled with
-can be taken into account to allow terminating the search of a vertex $v$'s
-neighborhood before all possible walks of length 2 leaving $v$ have been
-considered.
+\Algorithm{alg:transitive_reduction} provides pseudocode for a transitive
+reduction algorithm similar to that described in \cite{Myers2005}.  Essentially,
+the algorithm is to through each vertex $v$ and examine vertices up to 2 edges
+away from $v$ to identify all transitive edges leaving $v$.  These edges are
+then marked for removal, and they are removed after all vertices in the graph
+have been considered.  Some optimizations are possible; for example, the lengths
+of the sequences the edges are labeled with can be taken into account to allow
+terminating the search of $v$'s neighborhood before all possible walks of length
+2 starting at $v$ have been considered.
 
 Transitive reduction can be performed on either the bidirected or directed
-representations of the string graph, although for simplicity here it is
-described on the directed graph.
+representations of the string graph, although for simplicity
+\Algorithm{alg:transitive_reduction} implies the directed representation.
 
@@
 Collapsing unbranched paths is implemented as the program
-\ProgramName{collapsed-unbranched-paths}.  It takes as input a file containing a
+\ProgramName{collapse-unbranched-paths}.  It takes as input a file containing a
 string graph and produces as output a file containing the string graph with
@@
 
-\section{Examples of graph construction and simplification}
+\section{Examples of assembly graph construction and simplification}
 
 At this point, the assembler has been given a set of reads, has computed exact
-overlaps from them, removed contained reads and all overlaps involving them,
-then built a bidirected string graph that models the ways (via walks through the
-graph) in which the reads can be assembled together when the overlaps are used
-as evidence, then simplifies this graph through transitive reduction and
-collapsing unbranched paths.  Usually, the graph at this point will be much
-simpler than the original graph, although it may still be very complicated,
-depending on the genome and the data.
+overlaps from them, has removed contained reads and all overlaps involving them,
+has built a bidirected string graph that models the ways in which the reads can
+be assembled together when the overlaps are used as evidence, and has simplified
+this graph through transitive reduction and collapsing unbranched paths.
+Usually, the graph at this point will be much simpler than the original graph,
+although it may still be very complicated, depending on the genome and the data.
 
@@
 reduction is expected to leave the entire graph as a single unbranched path,
-provided the original genome was a single, non-circular string, and every base
+provided the original genome was a single, non-circular string and every base
 pair of it was included in at least one read.  Then, collapsing unbranched paths
@@
 any of the loops more than one time (this will also be elaborated on more
-later).
+later), since there are no repeats of substantial length in a hypothetical
+random genome.
 
@@
 	\begin{center}
+		\vspace{1cm}
 		\begin{tikzpicture}
@@
 	\begin{center}
-		\includegraphics[width=0.6\textwidth]{E_coli.pdf}
+		\includegraphics[width=1.0\textwidth]{E_coli.pdf}
 		\caption{Transitively-reduced, collapsed bidirected string graph for
@@
 		\ProgramName{compute-overlaps} would need to improved further to handle
-		larger genomes.  (Also it still does not handle inexact overlaps, which
-		are not considered in this paper.)}
+		larger genomes.}
 		\label{fig:E_coli}
@@
 complicated even after transitive reduction and traversal.  In even worse
-examples, the graph could easily have thousands of vertices and edges.
-Therefore, we need to figure out how to walk through the graph to produce the
-best assembly.
+examples, the graph could easily have thousands of vertices and edges, and it
+may not even be connected.  Therefore, we need to figure out how to walk through
+the graph to produce the best assembly.
 
@@
 However, there are multiple problems with the Eulerian path approach.  First,
-the graph may not be Eulerian.  This can happen if the graph is not strongly
-connected, not balanced (indegree always equal to out
-
-This implies we really would need to solve the
-bidirected equivalent of the Chinese Postman Problem instead.  Even then, the
-graph may not be fully connected, which would require that the final walk take
-the form of multiple discontinuous walks.  An equally severe problem is that
-some overlaps may be spurious, meaning that the two overlapping reads do not
-actually form a valid genomic sequence when combined.  Spurious overlaps can be
-caused by genomic repeats.  Therefore, we need to allow for the possibility that
-a given edge in the bidirected graph is omitted from the assembly if it seems
-like a good idea to do so.
+the graph may not be Eulerian.  This could happen if the graph is not balanced.
+This implies we really would need to solve the bidirected equivalent of the
+Chinese Postman Problem instead.\footnote{The Chinese Postman problem on a
+bidirected graph is formalized in \cite{Medvedev2009}.}  Even then, the graph
+may not be fully connected, which would require that the final walk take the
+form of multiple discontinuous walks.  An equally severe problem is that some
+overlaps may be spurious, meaning that the two overlapping reads do not actually
+form a valid genomic sequence when combined.  Spurious overlaps can be caused by
+genomic repeats.  Therefore, we need to allow for the possibility that an edge
+in the bidirected graph can be omitted from the walk if it seems like a good
+idea to do so.
 
@@
 count of edges in the graph in order for every vertex to be balanced (i.e. have
-the same indegree as out degree), then (3) find a generalized Eulerian tour
-through the graph, traversing each edge the exact number of times that was
-calculated in (2) and also allowing for the tour be discontinuous.
+the same in-degree as out-degree), then (3) find a generalized Eulerian tour
+through the bidirected graph, traversing each edge the exact number of times
+that was calculated in (2) and also allowing for the tour be discontinuous.
 
@@
 
-We would like to use information from the data to restrict the number of times
-each edge may be traversed.  This can be done using a calculation called the
-\KeyTerm{A-statistic}, or \KeyTerm{arrival-rate statistic}, which is the
-logarithm of the probability that an edge represents unique sequence rather than
-duplicate sequence.  The A-statistic $A(e)$ of an edge $e$ in a bidirected
-string graph is:
+We would like to use hints from the original data to restrict the number of
+times each bidirected edge may be traversed.  This can be done using a
+calculation called the \KeyTerm{A-statistic}, or \KeyTerm{arrival-rate
+statistic}, which is the logarithm of the probability that an edge represents
+unique sequence rather than duplicate sequence\cite{Myers2005}.  The A-statistic
+$A(e)$ of an edge $e$ in a bidirected string graph is:
 

### 2012-12-11 presentation/paper.tex
paper: mapping contained reads

@@
 The contained reads cannot be completely discarded because they are needed for a
-later step (see \Section{sec:mapping_contained_reads}).
+later step (see \Subsection{subsec:mapping_contained_reads}).
 
@@
 statistic}, which is the logarithm of the probability that an edge represents
-unique sequence rather than duplicate sequence\cite{Myers2005}.  The A-statistic
+unique sequence rather than duplicate sequence \cite{Myers2005}.  The A-statistic
 $A(e)$ of an edge $e$ in a bidirected string graph is:
@@
 support that edge.  To more precise, $k$ is the number of inner vertices in the
-unbranched path corresponding to $e$ before the unbranched paths were collapsed.
-
-Edges with a sufficiently high A-statistic are very likely to be single-copy
-edges and therefore should be traversed exactly one time.  The remaining edges
-may be traversed any number of times (with the exact number being determined
-during the network flow calculation), except for edges that originally were
-associated with one or more inner vertices, in which case we require the edge be
-traversed at least one time so that the reads corresponding to the inner
-vertices are used in the reconstruction of the genome (as every read was
-presumably sampled from the genome, in the absence of foreign DNA).
-
-Calculation of the A-statistic on each edge in the bidirected string graph is
-implemented by the program {\tt calculate-A-statistics}.
+unbranched path corresponding to $e$ before the unbranched paths were collapsed
+\cite{Myers2005}.
+
+Edges with sufficiently high A-statistic are very likely to be single-copy edges
+and therefore should be traversed exactly one time.  The remaining edges may be
+traversed any number of times (with the exact number being determined during the
+network flow calculation), except for edges that originally were associated with
+one or more inner vertices, in which case we require the edge be traversed at
+least one time so that the reads corresponding to the inner vertices are used in
+the reconstruction of the genome.  This is needed because in the absence of
+foreign DNA, every read was presumably sampled from the genome.
+
+However, $G$, the estimated genome length, is actually unknown, so it must be
+estimated first.  It can be estimated by producing a ``bootstrap'' genome length
+by adding up the lengths of the sequences on every edge in both directions and
+dividing by 2, then calculating the A-statistics using this bootstrap genome
+length and identifying edges that are probably single-copy.  The A-statistics
+can then be re-estimated over just these single-copy edges in order to refine
+the genome length further before calculating the final A-statistic on all the
+edges.  This procedure was outlined in \cite{Myers2005}.
+
+The calculation of the A-statistic on each edge in the bidirected string graph,
+including the genome length bootstrapping algorithm, is implemented by the
+program {\tt calculate-A-statistics}.
 
@@
 
-\label{sec:mapping_contained_reads}
+Because contained reads and their overlaps were removed before constructing the
+assembly graph (see \Subsection{subsec:removing_contained_reads}), the
+calculation of the A-statistic as described above is flawed because contained
+reads need to be counted.  This is especially a problem for high copy-number
+regions because those regions are likely to generate many contained reads.
+Therefore, we need to map the contained reads into the graph before calculating
+the A-statistics.
+
+A contained read may easily be mapped to the vertex corresponding to a read that
+contains it, but this is not sufficient because the A-statistic is calculated
+over the edges, not the vertices.  Instead, we need to map each end of each the
+contained read to the edge, or in general the set of edges, that contains it.
+
+The algorithm for doing so is outlined in \cite{Myers2005} and was implemented
+by the program \ProgramName{map-contained-reads}, which runs before
+\ProgramName{collapse-unbranched-paths}.  Briefly, for each contained read $r$
+the algorithm identifies the uncontained read $R$ that contains $r$ and has the
+shortest distance from the end of $r$ to the end of $R$ (the shortest
+\KeyTerm{overhang}).  This is done for each of the two directions.  The edges in
+the graph starting at the vertex corresponding to that end of $R$ are then
+walked backwards for the length of the overhang to find the set of edges $S$ to
+which the corresponding edge of $r$ may map, and each edge in $S$ is marked as
+containing $\frac{1}{|S|}$ extra inner vertices.  This mapping, therefore, makes
+it possible to account for contained reads when calculating the A-statistic.
+
+\label{subsec:mapping_contained_reads}
 

### 2012-12-11 presentation/paper.tex
paper: minimum-cost circulation

@@
 bound on the number of times each edge may be traversed, then (2) formulate and
-solve a minimum-cost network flow problem that finds the minimum total traversal
+solve a minimum-cost circulation problem that finds the minimum total traversal
 count of edges in the graph in order for every vertex to be balanced (i.e. have
@@
 \subsection{Calculating the A-statistic}
+
+\label{subsec:a_statistics}
 
@@
 traversed any number of times (with the exact number being determined during the
-network flow calculation), except for edges that originally were associated with
-one or more inner vertices, in which case we require the edge be traversed at
-least one time so that the reads corresponding to the inner vertices are used in
-the reconstruction of the genome.  This is needed because in the absence of
-foreign DNA, every read was presumably sampled from the genome.
+minimum-cost circulation calculation), except for edges that originally were
+associated with one or more inner vertices, in which case we require the edge be
+traversed at least one time so that the reads corresponding to the inner
+vertices are used in the reconstruction of the genome.  This is needed because
+in the absence of foreign DNA, every read was presumably sampled from the
+genome.
 
@@
 
-\subsection{Minimum-cost network flow}
-
-Formally, a minimum-cost network flow problem on a directed graph takes as input
-for each edge $e$ an upper bound $c(e) \ge 0$ on flow and a cost per unit flow
-$v(e)$, and for each vertex $v$ a supply (positive) or demand (negative)
-$b(v)$ for flow.  It produces as output a flow $x(e)$ on each edge $e$ such
-that $\sum_e v(e)x(e)$ is minimized subject to $x(e) \in [0, c(e)]$ and
-$\sum_{u \to v}x(u \to v) + b(v) = \sum_{v \to w} x(v \to w)$.\cite{Myers2005}.
-
-We can roughly map our problem onto this definition assigning $c(e) = 0$ when
-$e$ is definitely single-copy and $c(e) = \infty$ otherwise, $b(v)$ the sum
-of the lower bounds on the traversal counts of all bidirected edges incident to
-$v$ with an inward head minus the sum of the lower bounds on the traversal
-counts of all edges incident to $v$ with an outward head, and $v(e) = 1$.  The
-resulting traversal count $t(e)$ for a bidirected edge $e$ is then the lower
-bound on the traversal count for that edge plus $x(e)$, the traversal count
-assigned in the network flow calculation.  The traversal count $t(e)$ represents
-the total number of times the edge may be traversed in a given assembly,
-counting traversals in both directions.
-
-There is, however, a serious complication with this network flow formulation
-which was not addressed in \cite{Myers2005}.  We have a bidirected graph, not a
-directed graph, so it cannot be trivially mapped to a directed network flow
-problem.  A possible workaround for this is to solve directed flow on the
-directed equivalent to the bidirected graph, described in
+\subsection{Minimum-cost circulation}
+
+Formally, a minimum-cost circulation problem on a directed graph takes as input
+for each edge $e$ an lower bound $l(e) \ge 0$ and an upper bound $u(e) \ge 0$ on
+flow and a cost per unit flow $v(e)$, and for each vertex $v$ a supply
+(positive) or demand (negative) $b(v)$ for flow.  It produces as output a flow
+$x(e)$ on each edge $e$ such that $\sum_e v(e)x(e)$ is minimized subject to
+$x(e) \in [l(e), u(e)]$ and $\sum_{u \to v}x(u \to v) + b(v) = \sum_{v \to w}
+x(v \to w)$ \cite{Myers2005}.  In the minimum-cost circulation problem, as
+opposed to less general network flow problems, no nodes are treated specially as
+sources or sinks; instead, flow must be balanced at every node $v$, but only
+after taking into account the supply or demand $b(v)$ at that node.
+
+The A-statistic calculation on each edge $e$, as described in
+\Subsection{subsec:a_statistics}, allows us to assign a lower bound $l(e)$ and
+an upper bound $u(e)$ on the traversal count for that edge.  Note that $u(e)$
+may be $\infty$, as these bounds are very approximate.  We also have $b(v) = 0$
+for all $v$, because the flow at every vertex must be balanced; this corresponds
+to the walk entering each vertex the same number of times that it is exited,
+making the graph as a whole be balanced.
+
+%It produces as output a flow
+%$x(e)$ on each edge $e$ such that $\sum_e v(e)x(e)$ is minimized subject to
+%$x(e) \in [0, c(e)]$ and $\sum_{u \to v}x(u \to v) + b(v) = \sum_{v \to w} x(v
+%\to w)$.\cite{Myers2005}.
+
+%We can roughly map our problem onto this definition assigning $c(e) = 0$ when
+%$e$ is definitely single-copy and $c(e) = \infty$ otherwise, $b(v)$ the sum
+%of the lower bounds on the traversal counts of all bidirected edges incident to
+%$v$ with an inward head minus the sum of the lower bounds on the traversal
+%counts of all edges incident to $v$ with an outward head, and $v(e) = 1$.  The
+%resulting traversal count $t(e)$ for a bidirected edge $e$ is then the lower
+%bound on the traversal count for that edge plus $x(e)$, the traversal count
+%assigned in the network flow calculation.  The traversal count $t(e)$ represents
+%the total number of times the edge may be traversed in a given assembly,
+%counting traversals in both directions.
+
+There are still two serious problems with this circulation formulation that need
+to be addressed.  First, we actually have a bidirected graph, not a directed
+graph.\footnote{The fact that this circulation problem actually needs to be
+formulated and solved on a bidirected graph is addressed in \cite{Medvedev2009},
+but not in \cite{Myers2005}, even though the latter introduced the overall
+fragment string graph algorithm.}
+
+but is a.  We have a
+bidirected graph, not a directed graph, so it cannot be trivially mapped to a
+directed network flow problem.  A possible workaround for this is to solve
+directed flow on the directed equivalent to the bidirected graph, described in
 \Subsection{subsec:directed}.  The calculated traversal counts for a directed
@@
 the other direction of the implied bidirected edge, which would be $g.B \to f.B$
-in this example.  However, gWe can take the average of the two counts and use this as the
-traversal count for the corresponding bidirected edge, and this
+in this example.  However, we can take the average of the two counts and use
+this as the traversal count for the corresponding bidirected edge, and this
 

### 2012-12-12 presentation/paper.tex
paper: circulation, Eulerian path

@@
 					    (0.8,0.1) node[anchor=west] {#2};}
+\newcommand{\BidirectedEdgeInnie}{
+		\tikz[>=triangle 45,baseline]
+		\draw[>-<,thin] (0,0.1) -- (0.8,0.1) node {\ \ } ;}
+\newcommand{\BidirectedEdgeOutie}{
+		\tikz[>=triangle 45,baseline]
+		\draw[<->,thin] (0,0.1) --(0.8,0.1) node {\ \ };}
 
@@
 
-However, there are multiple problems with the Eulerian path approach.  First,
+However, there are multiple problems with this Eulerian path approach.  First,
 the graph may not be Eulerian.  This could happen if the graph is not balanced.
@@
 count of edges in the graph in order for every vertex to be balanced (i.e. have
-the same in-degree as out-degree), then (3) find a generalized Eulerian tour
+the same in-degree as out-degree), then (3) find a generalized Eulerian cycle
 through the bidirected graph, traversing each edge the exact number of times
-that was calculated in (2) and also allowing for the tour be discontinuous.
+that was calculated in (2), but also allowing for the cycle to be logically
+discontinuous by introducing a special vertex and special edges.
 
@@
 
-We would like to use hints from the original data to restrict the number of
-times each bidirected edge may be traversed.  This can be done using a
+We would like to use statistical properties of the original data to restrict the
+number of times each bidirected edge may be traversed.  This can be done using a
 calculation called the \KeyTerm{A-statistic}, or \KeyTerm{arrival-rate
 statistic}, which is the logarithm of the probability that an edge represents
-unique sequence rather than duplicate sequence \cite{Myers2005}.  The A-statistic
-$A(e)$ of an edge $e$ in a bidirected string graph is:
+unique sequence rather than duplicate sequence \cite{Myers2005}.  The
+A-statistic $A(e)$ of an edge $e$ in a bidirected string graph is:
 
@@
 
-However, $G$, the estimated genome length, is actually unknown, so it must be
-estimated first.  It can be estimated by producing a ``bootstrap'' genome length
-by adding up the lengths of the sequences on every edge in both directions and
-dividing by 2, then calculating the A-statistics using this bootstrap genome
-length and identifying edges that are probably single-copy.  The A-statistics
-can then be re-estimated over just these single-copy edges in order to refine
-the genome length further before calculating the final A-statistic on all the
-edges.  This procedure was outlined in \cite{Myers2005}.
+$G$, the estimated genome length, is actually unknown, so it must be estimated
+first.  It can be estimated by producing a ``bootstrap'' genome length by adding
+up the lengths of the sequences on every edge in both directions and dividing by
+2, then calculating the A-statistics using this bootstrap genome length and
+identifying edges that are probably single-copy.  The A-statistics can then be
+re-estimated over just these single-copy edges in order to refine the genome
+length further before calculating the final A-statistic on all the edges.  This
+procedure was outlined in \cite{Myers2005}.
 
@@
 over the edges, not the vertices.  Instead, we need to map each end of each the
-contained read to the edge, or in general the set of edges, that contains it.
+contained read to the edge, or in general the set of edges, that may contain it.
 
@@
 
+\label{subsec:circulation}
+
 Formally, a minimum-cost circulation problem on a directed graph takes as input
@@
 an upper bound $u(e)$ on the traversal count for that edge.  Note that $u(e)$
-may be $\infty$, as these bounds are very approximate.  We also have $b(v) = 0$
-for all $v$, because the flow at every vertex must be balanced; this corresponds
-to the walk entering each vertex the same number of times that it is exited,
-making the graph as a whole be balanced.
+may be $\infty$.  We also have $b(v) = 0$ for all $v$, because the flow at every
+vertex must be balanced; this corresponds to the walk entering each vertex the
+same number of times that it is exited, making the graph as a whole be balanced.
+Furthermore, we have $v(e) = 1$ because no edges are considered to be more
+costly to use than any other in this formulation.\footnote{Note that the goal of
+this assembly algorithm is not to minimize the assembled genome length, but
+rather to maximize the similarity of our assembled genome to the true genome.
+Therefore, there is no need to assign greater costs to edges that represent
+longer DNA sequences (at least in this formulation of the program), because that
+would make it more costly to repeat a longer edge even if there is strong
+evidence that edge corresponds to repeated sequence.}
 
@@
 to be addressed.  First, we actually have a bidirected graph, not a directed
-graph.\footnote{The fact that this circulation problem actually needs to be
-formulated and solved on a bidirected graph is addressed in \cite{Medvedev2009},
-but not in \cite{Myers2005}, even though the latter introduced the overall
-fragment string graph algorithm.}
-
-but is a.  We have a
-bidirected graph, not a directed graph, so it cannot be trivially mapped to a
-directed network flow problem.  A possible workaround for this is to solve
-directed flow on the directed equivalent to the bidirected graph, described in
-\Subsection{subsec:directed}.  The calculated traversal counts for a directed
-edge $e$ between, for example, two vertices $f.E \to g.E$ is not guaranteed to
-be the same as the calculated traversal count for the edge that corresponds to
-the other direction of the implied bidirected edge, which would be $g.B \to f.B$
-in this example.  However, we can take the average of the two counts and use
-this as the traversal count for the corresponding bidirected edge, and this
-
-%vertices $f.X$ and $g.X
-
-\subsection{Generalized Eulerian path}
+graph, so we need to calculate a \KeyTerm{bidirected flow}.\footnote{The fact
+that this circulation problem actually needs to be formulated and solved on a
+bidirected graph is addressed in \cite{Medvedev2009}, but not in
+\cite{Myers2005}, even though the latter introduced the overall fragment string
+graph algorithm.}
+A bidirected flow is not the same as a directed flow because of the
+presence of \BidirectedEdgeInnie and \BidirectedEdgeOutie edges, where the
+magnitude of the flow on that edge is counted as going into both vertices (in
+the case of a \BidirectedEdgeOutie edge) or out of both vertices (in the case of
+an \BidirectedEdgeInnie edge).  This does not imply that the edge itself is a
+source or a sink, but rather than the flow is defined in a peculiar way that
+makes sense on bidirected graphs.
+
+Fortunately, there does exist a way to solve a bidirected flow problem by
+mapping it to a directed flow problem by using the directed equivalent of the
+bidirected graph, described in \Subsection{subsec:directed}.  However, the
+calculated flow on directed edge $e$ between, for example, two vertices $f.E \to
+g.E$ is not guaranteed to be the same as the flow on $g.B \to f.B$, the edge
+that corresponds to the other direction of the implied bidirected edge.  This
+problem can be worked around by taking the average of the two counts and using
+this as the flow on the bidirected edge.  This approximation is equivalent to
+performing monotonization on the binet matrix that represents the linear program
+reduction of bidirected flow to produce a totally unimodular matrix that
+corresponds to a linear program that can be solved to find an integal flow that
+is at worst a 2-approximation to the optimal integral
+flow\cite{Medvedev2009}\cite{Hochbaum2004}.
+
+The second problem with our circulation formulation is that the bidirected
+string graph is not guaranteed to be connected, let alone strongly connected.
+For example, there might be vertices that have in-degree 0 but one or more out
+edges with nonzero lower bound on flow.  Such vertices may represent the last
+read in a genomic region before a coverage gap or the end of the sequence.  Our
+goal is to have such vertices be the endpoints of walks through the bidirected
+graph.  To take this into account in the formulation of the bidirected flow
+problem, we can introduce a special vertex $s$ and unlabeled special edges
+between $s$ and every other vertex.  Each special edge $e$ can have any amount
+of flow ($l(e) = 0,\ u(e) = \infty$), but at great cost ($v(e) = X$, where $X$
+is a very large number).  The addition of these edges will make the graph
+strongly connected, and therefore it will be possible to solve the minimum-cost
+circulation problem to find a flow that corresponds to balancing the in-degree
+and out-degree of every vertex when a certain traversal count $t(e)$ is assigned
+to each edge $e$.
+
+\subsection{Generalized Eulerian cycle}
+
+\label{subsec:eulerian}
+
+Here, a \KeyTerm{generalized Eulerian cycle} means a cycle in a bidirected graph
+$G$ that traverses each edge $e$ exactly $t(e)$ times, where $t(e)$ is the flow
+on edge $e$ as calculated using the method \Subsection{subsec:circulation}.  A
+generalized Eulerian cycle is guaranteed to exist because the graph was made
+strongly connected before solving the minimum-cost circulation problem.  (A more
+thorough explanation of this observation can be found in \cite{Medvedev2009}).
+The cycle corresponds to a consistent assembly of the genome, and it can be
+decomposed into a set of contigs by placing a contig break every time the
+special vertex $s$ is passed through.
+
+It should be possible to find this generalized Eulerian cycle by adapting an
+existing Eulerian path algorithm to work on a bidirected graph.
 
@@
 \Figure{fig:impl_graph} shows how the data flows through the main stages of the
-assembler.
+assembler.  The assembly pipeline is run using a Makefile that describes this
+flow.
 
@@
 contain a list of reads.  The final output of the assembler will eventually be a
-file in plain-text format that gives the list of assembled contigs.  Currently,
-the final stages of the algorithm (\Section{sec:final}, except for the
-calculation of the A-statistics) have not been implemented as of the end of the
-semester of Fall 2012.
+file in plain-text format that gives the list of assembled contigs.  However,
+the final stages of the algorithm (Sections \ref{subsec:circulation} and
+\ref{subsec:eulerian}) have not been implemented as of the end of the semester
+of Fall 2012.
 
@@
 \begin{itemize}
-\item Finish the network flow and generalized Eulerian path parts of the
-algorithm so that the implementation is ``complete'' and can produce a ``final''
-assembly.
+\item Finish the minimum-cost circulation and generalized Eulerian path parts of
+the algorithm so that the implementation is ``complete'' and can produce a
+``final'' assembly.
 \item Improve the performance of \ProgramName{compute-overlaps}.
@@
 Burrows-Wheeler transform to efficiently construct the transitively-reduced
-bidirected string graph directly from the reads\cite{Simpson2010}.
+bidirected string graph directly from the reads \cite{Simpson2010}.
 \item Use mate-pair information to help resolve repeats and connect assembled

### 2012-12-12 presentation/paper.tex
small example

@@
 \usepackage{array}
+\usepackage{grffile}
 %\usepackage{amsthm}
@@
 	\begin{subfigure}[b]{0.3\textwidth}
-		\begin{tikzpicture}=\ReadArrowType]
+		\begin{tikzpicture}[>=\ReadArrowType]
 			\draw[->,style=thick] (0,0) -- (2,0);
@@
 	\begin{subfigure}[b]{0.3\textwidth}
-		\begin{tikzpicture}=\ReadArrowType]
+		\begin{tikzpicture}[>=\ReadArrowType]
 			\draw[->,style=thick] (2,0) -- (0,0);
@@
 
+This section walks through the construction and simplification of the bidirected
+string graph on a very small example and provides a more detailed look at the
+actual data.  The genome is 200 bp long and is given by the following file:
+
+%make -B SAMPLE_SIZE=200 READ_LEN=50 MIN_OVERLAP_LEN=15 out.reduced.mapped.bidigraph.pdf out.reduced.mapped.collapsed.bidigraph.pdf out.bidigraph.pdf COVERAGE=2.5 SEED=9 USE_PIRS=false
+
+\begin{verbatim}
+>random_genome
+ACTCGAATTCCGAATAGATAGAAGTCTGCTGATAATATCGCGCCGGTTCTGATGCGCCTC
+GGGGAACGTGGCGGGTAGAGCTATCGTGTCACTCTCTGTCAGTGGTCTGTGAAAATTGAT
+CGCTACTGTCCCGGTGTAACCAGTCGACCCGCGGCAACTGCGAAGGGCCTCTCGCTCGCA
+GTACTTCTCGGAGCTATCCC
+\end{verbatim}
+
+The reads are the following 10 reads (50bp each, sampled randomly from either
+strand of the above genome):
+
+\begin{verbatim}
+>read_1
+CAGAACCGGCGCGATATTATCAGCAGACTTCTATCTATTCGGAATTCGAG
+>read_2
+GCAGTTGCCGCGGGTCGACTGGTTACACCGGGACAGTAGCGATCAATTTT
+>read_3
+GCGGGTAGAGCTATCGTGTCACTCTCTGTCAGTGGTCTGTGAAAATTGAT
+>read_4
+AGTCTGCTGATAATATCGCGCCGGTTCTGATGCGCCTCGGGGAACGTGGC
+>read_5
+TTCGCAGTTGCCGCGGGTCGACTGGTTACACCGGGACAGTAGCGATCAAT
+>read_6
+TACACCGGGACAGTAGCGATCAATTTTCACAGACCACTGACAGAGAGTGA
+>read_7
+ACAGAGAGTGACACGATAGCTCTACCCGCCACGTTCCCCGAGGCGCATCA
+>read_8
+ACAGACCACTGACAGAGAGTGACACGATAGCTCTACCCGCCACGTTCCCC
+>read_9
+TAGCGATCAATTTTCACAGACCACTGACAGAGAGTGACACGATAGCTCTA
+>read_10
+CTATCGTGTCACTCTCTGTCAGTGGTCTGTGAAAATTGATCGCTACTGTC
+\end{verbatim}
+
+The location on the genome from which each of these reads came is shown below.
+Note that by chance, 36 bp from the right end of the genome are not included in
+any read:
+
+\newcounter{readNumber}
+\newcounter{readPos}
+\newcounter{forwardReads}
+\newcounter{rcReads}
+
+\setcounter{forwardReads}{0}
+\setcounter{rcReads}{1}
+
+% ForwardRead{pos}{len}
+\newcommand{\ForwardRead}[2]{
+	\addtocounter{readNumber}{1}
+	\addtocounter{forwardReads}{-1}
+	\setcounter{readPos}{#1}
+	\addtocounter{readPos}{#2}
+	\draw[->, style=thick] (#1, \arabic{forwardReads}) node[anchor=east]
+									{\arabic{readNumber}.} -- (\arabic{readPos},
+									\arabic{forwardReads});
+}
+
+% RcRead{pos}{len}
+\newcommand{\RcRead}[2]{
+	\addtocounter{readNumber}{1}
+	\addtocounter{rcReads}{1}
+	\setcounter{readPos}{#1}
+	\addtocounter{readPos}{#2}
+	\draw[<-, style=thick] (#1, \arabic{rcReads}) node[anchor=east]
+									{\arabic{readNumber}.} -- (\arabic{readPos},
+									\arabic{rcReads});
+}
+
+\begin{center}
+	\begin{tabular}{|ccc|}
+	\hline
+	{\bf Read number} & {\bf Position} & {\bf Reverse-complement?} \\ \hline
+	1  & 1  & yes \\
+	2  & 111  & yes \\
+	3  & 70  & no \\
+	4  & 22  & no \\
+	5  & 114  & yes \\
+	6  & 88  & yes \\
+	7  & 49  & yes \\
+	8  & 60  & yes \\
+	9  & 75  & yes \\
+	10  & 80  & no \\ \hline
+	\end{tabular}
+
+	\fbox{
+	\begin{tikzpicture}[>=\ReadArrowType,x=0.56mm,y=3.5mm]
+		\draw[<-, style=thick] (0, 1) node[anchor=east] {G.} -- (200, 1);
+		\draw[->, style=thick] (0, 0) node[anchor=east] {G.} -- (200, 0);
+
+		\RcRead{1}{50}
+		\RcRead{111}{50}
+		\ForwardRead{70}{50}
+		\ForwardRead{22}{50}
+		\RcRead{114}{50}
+		\RcRead{88}{50}
+		\RcRead{49}{50}
+		\RcRead{60}{50}
+		\RcRead{75}{50}
+		\ForwardRead{80}{50}
+	\end{tikzpicture}
+	}
+\end{center}
+
+\ProgramName{compute-overlaps} computes the following exact overlaps of minimum
+length 15 bp among the above reads:
+
+\begin{verbatim}
+Overlap { Read 1: [0, 28], Read 4: [0, 28], rc = 1 }
+Overlap { Read 2: [0, 46], Read 5: [3, 49], rc = 0 }
+Overlap { Read 2: [23, 49], Read 6: [0, 26], rc = 0 }
+Overlap { Read 2: [31, 49], Read 10: [31, 49], rc = 1 }
+Overlap { Read 3: [18, 49], Read 6: [18, 49], rc = 1 }
+Overlap { Read 3: [0, 28], Read 7: [0, 28], rc = 1 }
+Overlap { Read 3: [0, 39], Read 8: [0, 39], rc = 1 }
+Overlap { Read 3: [5, 49], Read 9: [5, 49], rc = 1 }
+Overlap { Read 3: [10, 49], Read 10: [0, 39], rc = 0 }
+Overlap { Read 4: [27, 49], Read 7: [27, 49], rc = 1 }
+Overlap { Read 5: [26, 49], Read 6: [0, 23], rc = 0 }
+Overlap { Read 5: [34, 49], Read 10: [34, 49], rc = 1 }
+Overlap { Read 6: [28, 49], Read 8: [0, 21], rc = 0 }
+Overlap { Read 6: [13, 49], Read 9: [0, 36], rc = 0 }
+Overlap { Read 6: [8, 49], Read 10: [8, 49], rc = 1 }
+Overlap { Read 7: [0, 38], Read 8: [11, 49], rc = 0 }
+Overlap { Read 7: [0, 23], Read 9: [26, 49], rc = 0 }
+Overlap { Read 7: [0, 18], Read 10: [0, 18], rc = 1 }
+Overlap { Read 8: [0, 34], Read 9: [15, 49], rc = 0 }
+Overlap { Read 8: [0, 29], Read 10: [0, 29], rc = 1 }
+Overlap { Read 9: [0, 44], Read 10: [0, 44], rc = 1 }
+\end{verbatim}
+
+These overlaps are then used by \ProgramName{build-bidirected-string-graph} to
+build the assembly graph.  Note that there are no contained reads in this
+example, so there are as many vertices as original reads.  Also, in this image
+(and the following images), each edge is labeled with the length of that edge's
+sequence, rather than the sequence itself.  Furthermore, although each edge
+really contains a separate sequence for each direction, for this particular
+example all the reads are the same length and only exact overlaps are
+considered, which makes it so the two sequences of each edge are the same
+length.
+
+\begin{center}
+	\includegraphics[scale=0.7]{example_200bp.bidigraph-crop.pdf}	
+\end{center}
+
+The assembly graph then can be simplified by \ProgramName{transitive-reduction}:
+
+\begin{center}
+	\includegraphics[scale=0.8]{example_200bp.reduced.mapped.bidigraph-crop.pdf}	
+\end{center}
+
+Then, the assembly can be simplified further by
+\ProgramName{collapse-unbranched-paths}:
+
+\begin{center}
+	\includegraphics[scale=0.8]{example_200bp.reduced.mapped.collapsed.bidigraph-crop.pdf}	
+\end{center}
+
+%\begin{figure}
+	%\begin{center}
+		%\begin{subfigure}{0.99\textwidth}
+			%\begin{center}
+				%\includegraphics[scale=0.5]{example.bidigraph-crop.pdf}	
+				%\caption{Bidirected string graph built from 100 bp reads sampled
+				%randomly at 5 $\times$ coverage from a 450 bp genome, using minimum 25
+				%bp overlaps.  Note: edges are labeled with their lengths, not DNA
+				%sequences.}
+			%\end{center}
+		%\end{subfigure}
+		%\begin{subfigure}{0.99\textwidth}
+			%\begin{center}
+				%\includegraphics[scale=0.7]{example.reduced.mapped.bidigraph-crop.pdf}	
+				%\caption{The bidirected string graph from the previous slide, after
+				%transitive reduction.}
+			%\end{center}
+		%\end{subfigure}
+		%\begin{subfigure}{0.99\textwidth}
+			%\begin{center}
+				%\includegraphics[scale=0.7]{example.reduced.mapped.collapsed.bidigraph-crop.pdf}	
+				%\caption{The bidirected string graph from the previous slide, after
+				%collapsing unbranched paths.}
+			%\end{center}
+		%\end{subfigure}
+	%\end{center}
+%\end{figure}
+
 \section{Producing the final assembly}

### 2012-12-12 test/simulate_uniform_reads.pl

@@ -8,16 +8,23 @@ my $help = 0;
 my $read_len = 100;
 my $read_sep = 10;
 my $allow_rc = 0;
-
+my $random_sep = 0;
+my $read_pos_log = undef;
+my $coverage = 0;
+my $seed = 1;
 
 my $USAGE =
 "Usage: simulate_uniform_reads.pl [--help] [--read-len=LEN]
-        [--read-sep=SEP] [--allow-rc] GENOME_FASTA\n";
+        [--read-sep=SEP] [--allow-rc] [--coverage=X]
+        [--read-pos-log=FILE] GENOME_FASTA\n";
 
-my $res = GetOptions("help"        => \$help,
-                     "read-len=n", => \$read_len,
-                     "read-sep=n", => \$read_sep,
-                     "allow-rc",   => \$allow_rc);
+my $res = GetOptions("help"           => \$help,
+                     "read-len=n"     => \$read_len,
+                     "read-sep=n"     => \$read_sep,
+                     "allow-rc"       => \$allow_rc,
+                     "coverage=f"     => \$coverage,
+                     "read-pos-log=s" => \$read_pos_log,
+                     "seed=n"         => \$seed);
 
 if (!$res) {
     die "$USAGE";
@@ -28,11 +35,15 @@ if ($help) {
     exit 0;
 }
 
-srand(1);
+srand($seed);
 
 my $file = $ARGV[0] or die $USAGE;
 
 open FILE, "<", $file or die "Cannot open \"$file\": $!";
+
+if ($read_pos_log) {
+    open READ_POS_LOG, ">", $read_pos_log or die "Cannot open \"$read_pos_log\": $!";
+}
 
 my $seq = "";
 while (<FILE>) {
@@ -43,19 +54,44 @@ while (<FILE>) {
         $seq .= $_;
     }
 }
+$seq =~ tr/acgt/ACGT/;
 
 my $read_no = 1;
-for (my $i = 0; ; $i += $read_sep) {
-    if ($i + $read_len > length($seq)) {
-        last;
-    }
-    my $read = substr($seq, $i, $read_len);
-    $read =~ tr/acgt/ACGT/;
+
+sub sim_read_at_pos {
+    my $pos = shift;
+
+    my $read = substr($seq, $pos, $read_len);
+    my $dir;
+
     if ($allow_rc && int(rand(2)) == 0) {
         $read = reverse $read;
         $read =~ tr/ACGT/TGCA/;
[…]

### 2012-12-12 presentation/paper.tex
paper: updates

@@
 \usepackage{grffile}
+\usepackage{threeparttable}
+%\usepackage{hyperref}
 %\usepackage{amsthm}
@@
 sequencing approach to genome sequencing relies on breaking up a sample of
-genomic DNA into fragments, sequencing these fragments, then using a program
-called a {\em genome assembler} to put together these short sequences to
-reconstruct the original, much longer sequence.  This paper provides an
-introduction and formalization of the genome assembly problem, then introduces a
-model and algorithm for genome assembly that is based on a special type of graph
-called a \KeyTerm{bidirected string graph}.  \end{abstract}
+genomic DNA into short fragments, sequencing these fragments, then using a
+program called a {\em genome assembler} to put together these short sequences to
+reconstruct the original genome.  This paper provides an introduction and
+formalization of the genome assembly problem, then introduces a model and
+algorithm for genome assembly that is based on a special type of graph called a
+\KeyTerm{bidirected string graph}.  \end{abstract}
 
@@
 \subsection{Definition of a genome}
+
+\label{sec:genome}
 
@@
 polyploid organisms, including most plants and animals, contain multiple
-slightly-differing copies of each chromosome.  In this case, the above
-definition will consider each copy of each chromosome to be a completely
-different string in the genome, unless the copies are completely identical.
+slightly-differing copies of each chromosome.  In this case, this definition
+will consider each copy of each chromosome to be a completely different string
+in the genome, unless the copies are completely identical.
 
@@
 	\begin{center}
-		\begin{tabular}{|c|>{\centering\arraybackslash}p{5.2cm}|}
+	\caption{Approximate genome sizes of selected organisms}
+		\label{tab:GenomeSizes}
+		\begin{threeparttable}
+			\begin{tabular}{|l|>{\arraybackslash}p{5.2cm}|}
 				%\hline
@@
 				\hline
-			{\it Variola} virus (smallpox) &  186,000 bp\\
+				{\it Variola} virus (smallpox) \tnote{1} &  186,000 bp \cite{smallpox}\\
 				\hline
@@
 				\hline
-			{\it Zea mays} (domesticated corn) & 2,000,000,000 bp \\
+				{\it Zea mays} (domesticated corn) & 2,400,000,000 bp \cite{maize} \\
 				\hline
@@
 			\end{tabular}
+			\begin{tablenotes}
+				\item [1] Technically, viruses are not organisms, but they still
+				have genomes.  Actually, some viruses have genomes made of RNA,
+				not DNA, but this is not considered in this paper.
+			\end{tablenotes}
+		\end{threeparttable}
 	\end{center}
-	\caption{Approximate genome sizes of selected organisms}
-	\label{tab:GenomeSizes}
 \end{table}
@@
 (\KeyTerm{sequencing}) one or more DNA molecules.  \KeyTerm{Genome sequencing}
-is the use of DNA sequencing to determine the base pair sequence of an entire
-genome.
+is the process of sequencing of all the DNA molecules that make up a genome.
 
@@
 restrict sequences to certain regions of the genome in order to make sequencing
-the genome easier (currently, at the disadvantage of increased cost and
+the genome easier (currently, at the disadvantage of increased cost and much more
 non-computational work).} Following this, the original genome must be
@@
 generally have already been developed by the manufacturer of the sequencing
-machine and work fairly well, and ``genome assembly'' does not cover these
-algorithms, although assemblers still may be tuned for specific sequencing
+machine and are assumed to work fairly well, so ``genome assembly'' does not cover these
+algorithms.  Still, assemblers may be tuned for specific sequencing
 technologies.}
@@
 
-The goal of genome assembly is to reconstruct a genome, as defined earlier,
-given a set of substrings that may originate from any strand of any DNA molecule
-in the genome.  Each substring is called a \KeyTerm{read} and is the sequence of
-some DNA that was sequenced by the sequencing machine.  In general, reads may
-contain errors, including \KeyTerm{substitutions}, where a base pair in the read
-was incorrectly read as some other base pair, as well as \KeyTerm{insertions},
-where a sequence of one or more random base pairs have been inserted into some
-position in the read, and \KeyTerm{deletions}, where a sequence of one or more
-base pairs from the original genome is missing from the interior of the read.
-Reads are expected to be uniformly distributed on the genome with a certain
-coverage $X$ depending on how much data was collected, meaning that each base
-pair in the original genome is expected to be included in $X$ different reads on
-average, allowing for the reads to be from either strand.  However, some
-sequencing technologies produce biased data where the coverage across the genome
-is much more variable than that expected by uniform, random sampling.
+The goal of genome assembly is to reconstruct a genome, as defined in
+\Section{sec:genome}, given a set of substrings that may originate from any
+strand of any DNA molecule in the genome.  Each substring is called a
+\KeyTerm{read} and is the sequence of some DNA that was sequenced by the
+sequencing machine.  In general, reads may contain errors, including
+\KeyTerm{substitutions}, where a base pair in the read was incorrectly read as
+some other base pair, as well as \KeyTerm{insertions}, where a sequence of one
+or more random base pairs have been inserted into some position in the read, and
+\KeyTerm{deletions}, where a sequence of one or more base pairs from the
+original genome is missing from the interior of the read.  Reads are expected to
+be uniformly distributed on the genome with a certain coverage $X$ depending on
+how much data was collected, meaning that each base pair in the original genome
+is expected to be included in $X$ different reads on average, allowing for the
+reads to be from either strand.  However, some sequencing technologies produce
+biased data where the coverage across the genome is much more variable than that
+expected by uniform, random sampling.
 
@@
 It is also possible for reads to come from foreign DNA; for example, if a DNA
-sample is obtained from grinding up plant or animal tissue, there could be some
+sample is obtained by grinding up plant or animal tissue, there could be some
 bacterial cells, each with its own genome, contaminating the sample.  Ideally,

### 2012-12-12 presentation/paper.tex
paper: updates

@@
-\documentclass[12pt]{article}
+\documentclass[10pt]{article}
 \usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
@@
 %\usepackage{amsthm}
+
+%\newcommand{\nodecolor}{blue!20}
+\newcommand{\nodecolor}{black!0}
+%\newcommand{\specialnodecolor}{green!25}
+\newcommand{\specialnodecolor}{black!10}
 
@@
 multiple times in the genome--- possibly adjacent to each other (a
-\KeyTerm{tandem repeat}), but not necessarily.  Repeats may be longer than the
-length of any individual read, which makes the genome more difficult to assemble
-accurately.  However, real genomes almost always contain repeats of substantial
-length, so assemblers must be able to handle repeats.
+\KeyTerm{tandem repeat}), but not necessarily.  The number of times a
+subsequence occurs in the genome is referred to as its \KeyTerm{copy number}.
+Repeats may be longer than the length of any individual read, which makes the
+genome more difficult to assemble accurately.  However, real genomes almost
+always contain repeats of substantial length, so assemblers must be able to
+handle repeats.
+
+The output of an assembler may be a list of \KeyTerm{contigs}, which are
+continuous sequences of DNA that were assembled and therefore are believed to
+exist in the original genome.  Each contig should be much longer than the
+original reads.  Ideally, if the original genome contains $N$ separate strings,
+there should be $N$ contigs, each of which is one of the genomic strings.
+However, in general there will be more than $N$ contigs because it is rarely
+possible to perfectly assemble the genome.
 
@@
 genome assembly is not the same as the shortest common superstring problem, and
-in fact seems to be harder, despite the fact that the shortest common
+in some ways seems to be harder, despite the fact that the shortest common
 superstring problem is already NP-complete\cite{Turner1989}.
@@
 graph algorithm because the graph is built directly from the sequenced fragments
-(or reads), rather than subsequences ($k$-mers) taken from them.
+(reads), rather than subsequences ($k$-mers) taken from them.
 
@@
 {\em not} a reverse-complement overlap, as it can equivalently be stated as the
-last $\LengthVar$ bp of $g$ exactly match the last $\LengthVar$ bp of $f$.
+last $\LengthVar$ bp of $g$ exactly match the first $\LengthVar$ bp of $f$.
 \Figure{fig:overlap_types} shows the three fundamental types of overlaps that
@@
 Overlaps may also be \KeyTerm{inexact}.  An inexact overlap is an overlap where
-overlapped regions of the two reads do not match exactly (in either a forward or
-reverse-complement manner), but instead differ by a number of base-pair
-substitutions, insertions, and/or deletions up to a predetermined upper bound,
-such as 5\%.  However, inexact overlaps will not be considered further in this
-paper.
+the overlapped regions of the two reads do not match exactly (in either a
+forward or reverse-complement manner), but instead differ by a number of
+base-pair substitutions, insertions, and/or deletions up to a predetermined
+upper bound, such as 5\%.  However, inexact overlaps will not be considered
+further in this paper.
 
@@
 spent on each read pair will depend several factors, including the minimum
-overlap length $L$ and the read lengths.  But even if we could somehow spend
+overlap length $L$ and the read lengths.  But even if we somehow could spend
 constant time on each read pair, this naive algorithm would be prohibitively
@@
 After overlaps have been computed from the reads, the next step is to construct
-a graph that models how the reads can be assembled together when the overlaps
-are used as evidence.  To do this, define a graph where each vertex corresponds
-to a read\footnote{Actually, more than one read may map to the same vertex, as
+a graph that models how the reads can be assembled when the overlaps are used as
+evidence.  To do this, define a graph where each vertex corresponds to a
+read\footnote{Actually, more than one read may map to the same vertex, as
 explained in \Subsection{subsec:removing_contained_reads}.} and each edge
@@
 			\begin{tikzpicture}[>=triangle 45]
-				\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+				\tikzstyle{every node} = [circle,fill=\nodecolor,draw=black];
 				\node (f) at (0, 0) {f};
@@
 			\end{tikzpicture}
-			\caption{A possible directed \\representation of the \\string graph.}
+			\caption{A possible directed \\representation of the \\string graph}
 			\label{subfig:possible_directed}
@@
 			\end{tikzpicture}
-			\caption{A possible undirected \\representation of the \\string graph.}
+			\caption{A possible undirected \\representation of the \\string graph}
 			\label{subfig:possible_undirected}
@@
 			\end{tikzpicture}
-			\caption{A possible bidirected \\representation of the \\string graph.}
+			\caption{The bidirected \\representation of the \\string graph}
 			\label{subfig:possible_bidirected}
@@
 		\begin{tikzpicture}[>=triangle 45]
-			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+			\tikzstyle{every node} = [circle,fill=\nodecolor,draw=black];
 			\node (1) at (0.00, 0) {1};
@@
 assembled together consistently\cite{Myers2005}.  In such a walk, each read may
-be used in forward or reverse-complement orientation, depending on the
+be used in either forward or reverse-complement orientation, depending on the
 direction, relative to the walk, of the adjacent arrow heads when the
@@
 			\begin{tikzpicture}[>=triangle 45,scale=1.5]
-				\node[circle,fill=blue!20,draw=black] (f) at (0, 0) {f};
-				\node[circle,fill=blue!20,draw=black] (g) at (2.0, 0) {g};
+				\node[circle,fill=\nodecolor,draw=black] (f) at (0, 0) {f};
+				\node[circle,fill=\nodecolor,draw=black] (g) at (2.0, 0) {g};
 				\draw[>->,style=very thick] (f) edge (g);
@@
 			\begin{tikzpicture}[>=triangle 45,scale=1.5]
-				\node[circle,fill=blue!20,draw=black] (fE) at (0, 0) {f.E};
-				\node[circle,fill=blue!20,draw=black] (gE) at (2.0, 0) {g.E};
-
-				\node[circle,fill=blue!20,draw=black] (fB) at (0, -1) {f.B};
-				\node[circle,fill=blue!20,draw=black] (gB) at (2.0, -1) {g.B};
+				\node[circle,fill=\nodecolor,draw=black] (fE) at (0, 0) {f.E};
+				\node[circle,fill=\nodecolor,draw=black] (gE) at (2.5, 0) {g.E};
+
+				\node[circle,fill=\nodecolor,draw=black] (fB) at (0, -1) {f.B};
+				\node[circle,fill=\nodecolor,draw=black] (gB) at (2.5, -1) {g.B};
 
@@
 				\draw[->,style=very thick] (gB) edge (fB);
-				\draw (0.90, 0.01) node[anchor=south] {\small \tt TTAGAAGTA};
-				\draw (1.10, -1.01) node[anchor=north] {
+				\draw (1.2, 0.01) node[anchor=south] {\small \tt TTAGAAGTA};
+				\draw (1.2, -1.01) node[anchor=north] {
 						\rotatebox{180}{\small \tt
@@
 reads need not be included in the graph because there will always be a walk that
-spells the sequence of each contained read.  In other words, contained reads
-provide redundant information because the sequence of any given contained read
-is already spelled out by an uncontained read.
+spells the sequence of each contained read.
+%In other words, contained reads
+%provide redundant information because the sequence of any given contained read
+%is already spelled out by an uncontained read.
 
 Therefore, prior to mapping the reads to vertices and overlaps to edges, we can
-examine the overlaps to identify contained reads, and include in the initial
+examine the overlaps to identify contained reads, then include in the initial
 construction of the bidirected string graph neither the contained reads nor any
@@
 This implies that a vertex in the bidirected string graph may actually
-correspond to more than one read.
+correspond to more than one read, which makes it meaningful to include a vertex
+more than one time in a walk in the event of a genomic repeat.
 
@@
 uncontained reads and their overlaps, and produce as output the bidirected or
-directed graph, serialized to a file.  As should be expected from what was
-described earlier, \ProgramName{build-bidirected-string-graph} will add one
-vertex for each read and add a bidirected edge for each overlap, while
-\ProgramName{build-directed-string-graph} will add two vertices for each read
-and add two directed edges for each overlap.
+directed graph, serialized to a file.
+\ProgramName{build-bidirected-string-graph} will add one vertex for each read
+and add a bidirected edge for each overlap, while
+\ProgramName{build-directed-string-graph} will add two vertices for each read and
+add two directed edges for each overlap.
 
@@
 in plain-text format, print a string graph in the DOT language for input to the
-graph-drawing software {\it GraphViz}\cite{GraphVizPublication}, or print some
-other information about a string graph.
+graph-drawing software {\it GraphViz}\cite{GraphVizPublication}, or print other
+information about a string graph.
 
@@
 sequences that the edges are labeled with, as this can be done to produce a
-preliminary assembly (especially after the graph simplifications that are about
-to be described).
+preliminary assembly (especially after the graph simplifications described in
+\Section{sec:simplify}).
 
@@
 
+\label{sec:simplify}
+
 \subsection{Transitive reduction}
 
-Before trying to find walks through the string graph, it is very helpful to
-simplify the graph is much as possible.  Very commonly, given three adjacent
+Before trying to find walks through the string graph, it is helpful to
+simplify the graph as much as possible.  Very commonly, given three adjacent
 reads $f$, $g$, and $h$, $f$ will overlap $h$ as well as $g$.  This is redundant
 because we can, equivalently, walk $f \to g \to h$ rather than directly walking
-from $f \to h$.  \Figure{fig:transitive} shows an example of this.
+$f \to h$.  \Figure{fig:transitive} shows an example of this.
 
@@
 		\begin{tikzpicture}[>=triangle 45]
-			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+			\tikzstyle{every node} = [circle,fill=\nodecolor,draw=black];
 			\node (f) at (0, 0) {f};
@@
 reduction algorithm similar to that described in \cite{Myers2005}.  Essentially,
-the algorithm is to through each vertex $v$ and examine vertices up to 2 edges
+the algorithm is to go through each vertex $v$ and examine vertices up to 2 edges
 away from $v$ to identify all transitive edges leaving $v$.  These edges are
@@
 		\begin{tikzpicture}[>=triangle 45]
-			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
+			\tikzstyle{every node} = [circle,fill=\nodecolor,draw=black];
 			\node (1) at (0.0, -1) {1};
@@
 			\node (3) at (1.5, 0) {3};
-			\node[fill=green!25] (4) at (3.25, 0) {4};
+			\node[fill=\specialnodecolor] (4) at (3.25, 0) {4};
 			\node (5) at (5.0, 0) {5};
@@
 
-A completely random genome is very unlikely to have repeats of significant
-length compared to the minimum overlap length.  In this case, transitive
-reduction is expected to leave the entire graph as a single unbranched path,
-provided the original genome was a single, non-circular string and every base
-pair of it was included in at least one read.  Then, collapsing unbranched paths
-will collapse the entire bidirected graph into two vertices connected by a
-single edge.  Finding a walk through the edges of this graph is trivial and will
-produce an assembly of the entire genome.  \Figure{fig:random_example} shows the
-bidirected string graph that might result from such a genome and data.
-
-If we still assume a genome with no repeats of length greater than the overlap
-length, but relax the assumption that the genome is a single, non-circular
-string from which every base pair was included in a read, the graph will instead
-reduce and collapse to a set of components.  Components corresponding to
-non-circular sequence will contain two vertices with an edge between, while
-components corresponding to circular sequence will contain one vertex with an
-edge connecting it to itself (a loop).  In this case (and as will be elaborated
-on later), finding a walk through the bidirected string graph must be
-generalized to allow walking each component separately; but again for this
+A completely random genome is unlikely to have repeats of significant length
+compared to the minimum overlap length.  In this case, transitive reduction is
+expected to leave the entire graph as a single unbranched path, provided the
+original genome was a single, non-circular string and every base pair of it was
+included in at least one read.  Then, collapsing unbranched paths will collapse
+the entire bidirected graph into two vertices connected by a single edge.
+Finding a walk through the edges of this graph is trivial and will produce an
+assembly of the entire genome.  \Figure{fig:random_example} shows the bidirected
+string graph that might result from such a genome and data.
+
+If we still assume a genome with no repeats of length greater than the minimum
+overlap length, but relax the assumption that the genome is a single,
+non-circular string from which every base pair was included in a read, the graph
+will instead reduce and collapse to a set of components.  Components
+corresponding to non-circular sequence will contain two vertices with an edge
+between, while components corresponding to circular sequence will contain one
+vertex with an edge connecting it to itself (a loop).  In this case (and as will
+be elaborated on later), finding a walk through the bidirected string graph must
+be generalized to allow walking each component separately; but again for this
 particular case of a random genome, each such walk is still trivial, and the
@@
 			\begin{tikzpicture}[>=triangle 45]
-				\node[circle,fill=blue!20,draw=black] (A) at (0.0, 0) {A};
-				\node[circle,fill=blue!20,draw=black] (B) at (3.0, 0) {B};
+				\node[circle,fill=\nodecolor,draw=black] (A) at (0.0, 0) {A};
+				\node[circle,fill=\nodecolor,draw=black] (B) at (3.0, 0) {B};
 				\draw[>->,very thick] (A) edge node[anchor=north] {4639100} (B);
@@
 		(a common bacterium) using a minimum overlap length of 250 bp.  This
-		graph was produced by the code that has been discussed, but the actual
+		graph was produced by the code written for this project, but the actual
 		drawing of the picture was done by {\it
@@
 
-The reads are the following 10 reads (50bp each, sampled randomly from either
-strand of the above genome):
+Note that in this file format, a continuous sequence can be split among multiple
+lines.  A new sequence only begins after a line beginning with {\tt >} is
+reached.
+
+The reads are the following 10 reads that are 50bp each and sampled randomly
+from either strand of the above genome:
 
@@
 
-The location on the genome from which each of these reads came is shown below.
-Note that by chance, 36 bp from the right end of the genome are not included in
-any read:
+The location on the genome from which each of these 10 reads came is shown
+below.  Note that by chance, 36 bp on the right end of the genome and 1 bp on
+the left end of the genome are not included in any read:
 
@@
 build the assembly graph.  Note that there are no contained reads in this
-example, so there are as many vertices as original reads.  Also, in this image
-(and the following images), each edge is labeled with the length of that edge's
-sequence, rather than the sequence itself.  Furthermore, although each edge
-really contains a separate sequence for each direction, for this particular
-example all the reads are the same length and only exact overlaps are
-considered, which makes it so the two sequences of each edge are the same
-length.
+example, so there are as many vertices as original reads.  Also, in the below
+graph drawing (as well as following graph drawings), each edge is labeled with
+the length of that edge's sequence rather than the sequence itself.
+Furthermore, although each edge really contains a separate sequence for each
+direction, for this particular example all the reads are the same length and
+only exact overlaps are considered, which makes it so the two sequences of each
+edge are always the same length.
 
@@
 
+The result is the fully simplified bidirected string graph.  In this particular
+example, the final graph is trivial and can be used to assemble the read-sampled
+part of the original genome by walking $2 \to 1$.  Note that there is an implied
+50bp before we start at the end of the read now labeled as 2, so the actual
+assembled sequence is $113 + 50 = 163$ bp long.  The assembler should then
+generate the assembly:
+
+\begin{verbatim}
+>contig_1
+CTCGAATTCCGAATAGATAGAAGTCTGCTGATAATATCGCGCCGGTTCTGATGCGCCTCG
+GGGAACGTGGCGGGTAGAGCTATCGTGTCACTCTCTGTCAGTGGTCTGTGAAAATTGATC
+GCTACTGTCCCGGTGTAACCAGTCGACCCGCGGCAACTGCGAA
+\end{verbatim}
+
 \section{Producing the final assembly}
@@
 may not be fully connected, which would require that the final walk take the
-form of multiple discontinuous walks.  An equally severe problem is that some
-overlaps may be spurious, meaning that the two overlapping reads do not actually
-form a valid genomic sequence when combined.  Spurious overlaps can be caused by
-genomic repeats.  Therefore, we need to allow for the possibility that an edge
-in the bidirected graph can be omitted from the walk if it seems like a good
-idea to do so.
+form of a set of multiple discontinuous walks.  An equally severe problem is
+that some overlaps may be spurious, meaning that the two overlapping reads do
+not actually form a valid genomic sequence when combined.  Spurious overlaps can
+be caused by genomic repeats.  Therefore, we need to allow for the possibility
+that an edge in the bidirected graph can be omitted from the walk if it seems
+like a good idea to do so.
 
@@
 and therefore should be traversed exactly one time.  The remaining edges may be
-traversed any number of times (with the exact number being determined during the
+traversed any number of times (with the exact number being determined by the
 minimum-cost circulation calculation), except for edges that originally were
@@
 length further before calculating the final A-statistic on all the edges.  This
-procedure was outlined in \cite{Myers2005}.
+algorithm is outlined in \cite{Myers2005}.
 
@@
 contains it, but this is not sufficient because the A-statistic is calculated
-over the edges, not the vertices.  Instead, we need to map each end of each the
-contained read to the edge, or in general the set of edges, that may contain it.
+over the edges, not the vertices.  Instead, we need to map each end of each
+contained read to the edge, or in general the set of edges, in which the end
+might be located if it were included as its own vertex.
 
@@
 walked backwards for the length of the overhang to find the set of edges $S$ to
-which the corresponding edge of $r$ may map, and each edge in $S$ is marked as
+which the relevant end of $r$ may map, and each edge in $S$ is marked as
 containing $\frac{1}{|S|}$ extra inner vertices.  This mapping, therefore, makes
@@
 	flow and a cost per unit flow $v(e)$, and for each vertex $v$ a supply
-(positive) or demand (negative) $b(v)$ for flow.  It produces as output a flow
-$x(e)$ on each edge $e$ such that $\sum_e v(e)x(e)$ is minimized subject to
-$x(e) \in [l(e), u(e)]$ and $\sum_{u \to v}x(u \to v) + b(v) = \sum_{v \to w}
-x(v \to w)$ \cite{Myers2005}.  In the minimum-cost circulation problem, as
-opposed to less general network flow problems, no nodes are treated specially as
-sources or sinks; instead, flow must be balanced at every node $v$, but only
-after taking into account the supply or demand $b(v)$ at that node.
+	(positive) or demand (negative) $b(v)$ for flow.  It produces as output a
+	flow $x(e)$ on each edge $e$ such that $\sum_e v(e)x(e)$ is minimized
+	subject to $x(e) \in [l(e), u(e)]$ and $\sum_{u \to v}x(u \to v) + b(v) =
+	\sum_{v \to w} x(v \to w)$ \cite{Myers2005}.  In the minimum-cost
+	circulation problem, as opposed to other network flow problems, no vertices are
+	treated specially as sources or sinks; instead, flow must be balanced at
+	every vertex $v$, but only after taking into account the supply or demand
+	$b(v)$ at that vertex.
 
@@
 may be $\infty$.  We also have $b(v) = 0$ for all $v$, because the flow at every
-vertex must be balanced; this corresponds to the walk entering each vertex the
-same number of times that it is exited, making the graph as a whole be balanced.
-Furthermore, we have $v(e) = 1$ because no edges are considered to be more
-costly to use than any other in this formulation.\footnote{Note that the goal of
-this assembly algorithm is not to minimize the assembled genome length, but
-rather to maximize the similarity of our assembled genome to the true genome.
-Therefore, there is no need to assign greater costs to edges that represent
-longer DNA sequences (at least in this formulation of the program), because that
-would make it more costly to repeat a longer edge even if there is strong
-evidence that edge corresponds to repeated sequence.}
+vertex must be balanced; this corresponds to the desired walk entering each
+vertex the same number of times that it is exited, making the graph as a whole
+be balanced.  Furthermore, we have $v(e) = 1$ because no edges are considered to
+be more costly to use than any other in this formulation.\footnote{Note that the
+goal of this assembly algorithm is not to minimize the assembled genome length,
+but rather to maximize the similarity of our assembled genome to the true
+genome.  Therefore, there is no need to assign greater costs to edges that
+represent longer DNA sequences (at least in this formulation of the problem),
+because that would make it more costly to repeat a longer edge even if there is
+strong evidence that edge corresponds to repeated sequence.}
 
@@
 bidirected graph, described in \Subsection{subsec:directed}.  However, the
-calculated flow on directed edge $e$ between, for example, two vertices $f.E \to
-g.E$ is not guaranteed to be the same as the flow on $g.B \to f.B$, the edge
+calculated flow on a directed edge between, for example, two vertices $f.E
+\to g.E$ is not guaranteed to be the same as the flow on $g.B \to f.B$, the edge
 that corresponds to the other direction of the implied bidirected edge.  This
@@
 reduction of bidirected flow to produce a totally unimodular matrix that
-corresponds to a linear program that can be solved to find an integal flow that
+represents a linear program that can be solved to find an integal flow that
 is at worst a 2-approximation to the optimal integral
-flow\cite{Medvedev2009}\cite{Hochbaum2004}.
+flow on the bidirected graph\cite{Medvedev2009}\cite{Hochbaum2004}.
 
@@
 string graph is not guaranteed to be connected, let alone strongly connected.
-For example, there might be vertices that have in-degree 0 but one or more out
-edges with nonzero lower bound on flow.  Such vertices may represent the last
-read in a genomic region before a coverage gap or the end of the sequence.  Our
-goal is to have such vertices be the endpoints of walks through the bidirected
-graph.  To take this into account in the formulation of the bidirected flow
-problem, we can introduce a special vertex $s$ and unlabeled special edges
-between $s$ and every other vertex.  Each special edge $e$ can have any amount
-of flow ($l(e) = 0,\ u(e) = \infty$), but at great cost ($v(e) = X$, where $X$
-is a very large number).  The addition of these edges will make the graph
-strongly connected, and therefore it will be possible to solve the minimum-cost
-circulation problem to find a flow that corresponds to balancing the in-degree
-and out-degree of every vertex when a certain traversal count $t(e)$ is assigned
-to each edge $e$.
+For example, there might be vertices that have in-degree 0 but have one or more
+adjacent out-edges with nonzero lower bound on flow.  Such vertices may
+represent the last read in a genomic region before a coverage gap or the end of
+the sequence.  Our goal is to have such vertices be the endpoints of logical
+walks through the bidirected graph.  To take this into account in the
+formulation of the bidirected flow problem, we can introduce a special vertex
+$s$ and unlabeled special edges between $s$ and every other vertex.  Each
+special edge $e$ can have any amount of flow ($l(e) = 0,\ u(e) = \infty$), but
+at great cost ($v(e) = X$, where $X$ is a very large number).  The addition of
+these edges will make the graph strongly connected, and therefore it will be
+possible to solve the minimum-cost circulation problem to find a flow that
+corresponds to balancing the in-degree and out-degree of every vertex by
+assigning a certain traversal count $t(e)$ to each edge $e$.
+
+For the implementation of this part of the algorithm, it should be possible to
+make the above-mentioned modifications to the graph, then solve the bidirected
+flow problem using the above-mentioned mapping to a directed flow problem.
 
@@
 $G$ that traverses each edge $e$ exactly $t(e)$ times, where $t(e)$ is the flow
-on edge $e$ as calculated using the method \Subsection{subsec:circulation}.  A
-generalized Eulerian cycle is guaranteed to exist because the graph was made
-strongly connected before solving the minimum-cost circulation problem.  (A more
-thorough explanation of this observation can be found in \cite{Medvedev2009}).
-The cycle corresponds to a consistent assembly of the genome, and it can be
-decomposed into a set of contigs by placing a contig break every time the
-special vertex $s$ is passed through.
+on edge $e$ as calculated using the method in \Subsection{subsec:circulation}.
+A generalized Eulerian cycle is guaranteed to exist in the bidirected graph at
+this point because the graph was made strongly connected before solving the
+minimum-cost circulation problem.  (A more thorough explanation of this
+observation can be found in \cite{Medvedev2009}.)  The cycle corresponds to a
+consistent assembly of the genome, and it can be decomposed into a set of
+logical walks by placing a logical breakpoint every time the special vertex $s$
+is passed through.  Each of these logical walks represents the assembly of a
+separate contig.
 
@@
 assembler.  The assembly pipeline is run using a Makefile that describes this
-flow.
+control flow.
 
@@
 	\caption{The flow of data through the assembly pipeline.  Each node is a
-	program, each arrow represents the output of a program being used as input
-	in another program.  Yellow programs have not yet been implemented as of
-	December 10, 2012.}
+	program and each arrow represents the output of a program being used as
+	input in another program.  Grayed-out programs have not yet been implemented
+	as of December 10, 2012.}
 	\label{fig:impl_graph}

### 2012-12-12 test/fasta_head.pl
fasta_head.pl

@@ -1,0 +1,49 @@
+#!/usr/bin/perl
+
+use strict;
+use Getopt::Long;
+$| = 1;
+
+my $help = 0;
+my $nbytes = -1;
+my $nlines = -1;
+
+my $USAGE = "Usage: fasta_head.pl [--bytes BYTES] [--lines LINES]\n";
+
+my $res = GetOptions("help"        => \$help,
+                     "bytes=n"     => \$nbytes,
+                     "lines=n"     => \$nlines);
+
+if (!$res) {
+    die "$USAGE";
+}
+
+if ($help) {
+    print "$USAGE";
+    exit 0;
+}
+
+if ($nbytes == -1 && $nlines == -1) {
+    $nlines = 10;
+}
+
+my $bytes = 0;
+my $lines = 0;
+
+while (<>) {
+    chomp;
+    if (/^>/) {
+        print "$_\n";
+    } else {
+        if ($lines++ == $nlines) {
+            last;
+        }
+        if ($nbytes != -1 && $bytes + length($_) >= $nbytes) {
+            my $s = substr($_, 0, $nbytes - $bytes);
+            print "$s\n";
+            last;
+        }
+        print "$_\n";
+        $bytes += length($_);
+    }
+}

### 2012-12-12 test/run-example.sh
run-example

@@ -1,0 +1,7 @@
+#!/bin/sh
+
+make GENOME=E_coli.fa   \
+     SAMPLE_SIZE=500000 \
+     COVERAGE=5         \
+     READ_LEN=100       \
+     MIN_OVERLAP_LEN=25

### 2012-12-12 test/run-example.sh
tee assemble.log

@@ -4,4 +4,5 @@ make GENOME=E_coli.fa   \
      SAMPLE_SIZE=500000 \
      COVERAGE=5         \
      READ_LEN=100       \
-     MIN_OVERLAP_LEN=25
+     MIN_OVERLAP_LEN=25 \
+     2>&1 | tee assemble.log

### 2012-12-12 presentation/paper.tex
fasta_head.pl
cleanup

@@
 %\usepackage{amsthm}
+
+\usepackage[nofiglist,notablist]{endfloat}
+\renewcommand{\efloatseparator}{\mbox{}}
 
@@
 				\item [1] Technically, viruses are not organisms, but they still
-				have genomes.  Actually, some viruses have genomes made of RNA,
-				not DNA, but this is not considered in this paper.
+				have genomes.
+				%Actually, some viruses have genomes made of RNA,
+				%not DNA, but this is not considered in this paper.
 			\end{tablenotes}
@@
 Genomes may contain \KeyTerm{repeats}, which are subsequences that occur
-multiple times in the genome--- possibly adjacent to each other (a
-\KeyTerm{tandem repeat}), but not necessarily.  The number of times a
+multiple times in the genome--- possibly adjacent to each other,
+%(a \KeyTerm{tandem repeat})
+but not necessarily.  The number of times a
 subsequence occurs in the genome is referred to as its \KeyTerm{copy number}.
@@
 	reads that were provided as input to the assembler.  The chosen value of $L$
-	must be balanced so that we find can enough overlaps to adequately cover the
+	must be balanced so that we can find enough overlaps to adequately cover the
 	genome (small $L$ better) but still exclude most spurious overlaps (large
@@
 densely from the genome, the running time of the seed-and-extend algorithm rises
-quadratically with respect to read density.  In addition, the number of read
-pairs the algorithm considers for a specific seed also rises quadratically with
+quadraticly with respect to read density.  In addition, the number of read
+pairs the algorithm considers for a specific seed also rises quadraticly with
 the number of times that seed is repeated in the genome due to genomic repeats.
@@
 evidence.  To do this, define a graph where each vertex corresponds to a
-read\footnote{Actually, more than one read may map to the same vertex, as
-explained in \Subsection{subsec:removing_contained_reads}.} and each edge
+read
+%\footnote{Actually, more than one read may map to the same vertex, as explained
+%in \Subsection{subsec:removing_contained_reads}.}
+and each edge
 corresponds to an overlap.  Two vertices are connected by an edge if and only if
@@
 	\end{subfigure}
-	\caption{Full interpetation of an edge in a bidirected string graph}
+	\caption{Full interpretation of an edge in a bidirected string graph}
 	\label{fig:bidiedge_interp}
@@
 directed graph, serialized to a file.
-\ProgramName{build-bidirected-string-graph} will add one vertex for each read
-and add a bidirected edge for each overlap, while
-\ProgramName{build-directed-string-graph} will add two vertices for each read and
-add two directed edges for each overlap.
+\ProgramName{build-bidirected-string-graph} will add one vertex for each
+uncontained read and a bidirected edge for each overlap, while
+\ProgramName{build-directed-string-graph} will add two vertices for each
+uncontained read and two directed edges for each overlap.
 
@@
 string graph and produces as output a file containing the string graph with
-unbranched paths collapsed.  Vertices are currently re-numbered; so, for
-example, vertex 10 after collapsing unbranched paths does not necessarily refer
-to the 10th read in the list of uncontained reads from which the graph was
-constructed.
+unbranched paths collapsed.  The algorithm marks vertices with in-degree 1 and
+out-degree 1 as inner, then iterates through every non-inner vertex to locate
+adjacent inner vertices, which must be the ends of unbranched paths.  Each
+unbranched path is then walked and collapsed.  Vertices are currently
+re-numbered; so, for example, vertex 10 after collapsing unbranched paths does
+not necessarily refer to the 10th read in the list of uncontained reads from
+which the graph was constructed.
 
@@
 			Bidirected string graph for reads uniformly sampled from a 4.6 Mbp
-			random genome after transitive reduction and path collapsing.  (Edge
+			random genome after transitive reduction and unbranched path collapsing.  (Edge
 			labeled with length of sequence, with the length being the same in both
@@
 Note that in this file format, a continuous sequence can be split among multiple
-lines.  A new sequence only begins after a line beginning with {\tt >} is
+lines; a new sequence only begins after a line beginning with {\tt >} is
 reached.
@@
 where $n$ is the total number of reads, $G$ is the estimated genome length,
-$\Delta$ is the length of the edge in bp, and $k$ is the number of reads that
-support that edge.  To more precise, $k$ is the number of inner vertices in the
-unbranched path corresponding to $e$ before the unbranched paths were collapsed
-\cite{Myers2005}.
+$\Delta$ is the average length of $e$'s two sequences in bp, and $k$
+%is the number of reads that support that edge.  To more precise, $k$
+is the number of inner vertices in the unbranched path corresponding to $e$
+before the unbranched paths were collapsed \cite{Myers2005}.
 
@@
 2, then calculating the A-statistics using this bootstrap genome length and
-identifying edges that are probably single-copy.  The A-statistics can then be
-re-estimated over just these single-copy edges in order to refine the genome
-length further before calculating the final A-statistic on all the edges.  This
-algorithm is outlined in \cite{Myers2005}.
+identifying edges that are probably single-copy.  The estimated genome length
+can then be refined using just these single-copy edges before calculating the
+final A-statistic on all the edges.  This algorithm is outlined in
+\cite{Myers2005}.
 
@@
 bidirected graph, described in \Subsection{subsec:directed}.  However, the
-calculated flow on a directed edge between, for example, two vertices $f.E
-\to g.E$ is not guaranteed to be the same as the flow on $g.B \to f.B$, the edge
+calculated flow on a directed edge between, for example, two vertices $f.E \to
+g.E$ is not guaranteed to be the same as the flow on $g.B \to f.B$, the edge
 that corresponds to the other direction of the implied bidirected edge.  This
@@
 performing monotonization on the binet matrix that represents the linear program
-reduction of bidirected flow to produce a totally unimodular matrix that
-represents a linear program that can be solved to find an integal flow that
-is at worst a 2-approximation to the optimal integral
-flow on the bidirected graph\cite{Medvedev2009}\cite{Hochbaum2004}.
+reduction of bidirected flow; this produces a totally unimodular matrix that
+represents a linear program that can be solved to find an integal flow that is
+at worst a 2-approximation to the optimal integral flow on the bidirected
+graph\cite{Medvedev2009}\cite{Hochbaum2004}.
 
@@
 
+\addcontentsline{toc}{section}{References}
 \bibliographystyle{plain}

### 2012-12-12 test/gen_random_genome.pl

@@ -1,4 +1,4 @@
-#!/usr/bin/perl
+#!/usr/bin/env perl
 
 my $USAGE =
 "Usage: gen_random_genome.pl LENGTH\n";

### 2012-12-12 test/run-example.sh

@@ -1,4 +1,4 @@
-#!/bin/sh
+#!/usr/bin/env bash
 
 make GENOME=E_coli.fa   \
      SAMPLE_SIZE=500000 \

### 2012-12-13 presentation/paper.tex
Figures throughout
paper: finalize

@@
 
-\usepackage[nofiglist,notablist]{endfloat}
-\renewcommand{\efloatseparator}{\mbox{}}
+%\usepackage[nofiglist,notablist]{endfloat}
+%\renewcommand{\efloatseparator}{\mbox{}}
 
@@
 Based on these biological facts, a genome is formally defined, for the purposes
-of this paper, as a set of possibly circular, dual-stranded strings of the
-symbols \Base{A}, \Base{C}, \Base{G}, and \Base{T}, where dual-stranded has the
-meaning indicated above (the two strands run in opposite directions, and the
+of this paper, as a set of possibly circular, double-stranded strings of the
+symbols \Base{A}, \Base{C}, \Base{G}, and \Base{T}, where double-stranded has
+the meaning indicated above (the two strands run in opposite directions, and the
 pairings \Base{A}---\Base{T} and \Base{C}---\Base{G} always hold).
@@
 which are then sequenced.\footnote{This specifically refers to the
-\KeyTerm{whole-genome shotgun sequencing} approach to genome assembly rather
+\KeyTerm{whole-genome shotgun sequencing} approach to genome sequencing rather
 than alternatives such as \KeyTerm{BAC-to-BAC} sequencing that are able to
@@
 read in the overlap is used in the original direction and the other read is used
-in the reverse-complement direction.  Note that an overlap from condition (4) is
-{\em not} a reverse-complement overlap, as it can equivalently be stated as the
-last $\LengthVar$ bp of $g$ exactly match the first $\LengthVar$ bp of $f$.
-\Figure{fig:overlap_types} shows the three fundamental types of overlaps that
-may arise from this definition.
-
-A read may overlap with itself; for example, a read may contain a suffix that
-matches its prefix of the same length.  Such an overlap is called a
-\KeyTerm{self-overlap}.
-
-Overlaps may also be \KeyTerm{inexact}.  An inexact overlap is an overlap where
-the overlapped regions of the two reads do not match exactly (in either a
-forward or reverse-complement manner), but instead differ by a number of
-base-pair substitutions, insertions, and/or deletions up to a predetermined
-upper bound, such as 5\%.  However, inexact overlaps will not be considered
-further in this paper.
-
-An overlap $o$ is fully described by the 7-tuple $(f,\ f_{begin},\ f_{end},\ g,
-\ g_{begin},\ g_{end},\ rc)$, where $f$ and $g$ are the two reads involved in the
-overlap, $f_{begin}$ and $f_{end}$ are the starting and ending indices of the
-overlapped region on the $f$ read, $g_{begin}$ and $g_{end}$ are the starting
-and ending indices of the overlapped region on the $g$ read, and $rc$ is a
-boolean that indicates whether the overlap is reverse-complement or not.  An
-overlap therefore has two overlapped regions--- one on each read.  For exact
-overlaps, the two overlapped regions are the same length; in other words,
-$g_{end} - g_{begin} = f_{end} - f_{begin}$.
-
-An \KeyTerm{extreme point} occurs when an overlapped region extends to the
-beginning or the end of the corresponding read.  An overlap always contains at
-least two extreme points.
-
-A \KeyTerm{containing overlap} is an overlap in which the overlapped region of
-one read, the \KeyTerm{contained read}, is the entire read.  A special case of a
-containing overlap is an overlap where the overlapped regions include the
-entirety of both reads.  For an exact overlap, this can only happen if the two
-reads are either identical or exact reverse-complements of each other.
+in the reverse-complement direction.  Note that an overlap from the last
+condition is {\em not} a reverse-complement overlap, as it can equivalently be
+stated as the last $\LengthVar$ bp of $g$ exactly match the first $\LengthVar$
+bp of $f$.  \Figure{fig:overlap_types} shows the three fundamental types of
+overlaps that may arise from this definition.
 
@@
 \end{figure}
+
+A read may overlap with itself; for example, a read may contain a suffix that
+matches its prefix of the same length.  Such an overlap is called a
+\KeyTerm{self-overlap}.
+
+Overlaps may also be \KeyTerm{inexact}.  An inexact overlap is an overlap where
+the overlapped regions of the two reads do not match exactly (in either a
+forward or reverse-complement manner), but instead differ by a number of
+base-pair substitutions, insertions, and/or deletions up to a predetermined
+upper bound, such as 5\%.  However, inexact overlaps will not be considered
+further in this paper.
+
+An overlap $o$ is fully described by the 7-tuple $(f,\ f_{begin},\ f_{end},\ g,
+\ g_{begin},\ g_{end},\ rc)$, where $f$ and $g$ are the two reads involved in the
+overlap, $f_{begin}$ and $f_{end}$ are the starting and ending indices of the
+overlapped region on the $f$ read, $g_{begin}$ and $g_{end}$ are the starting
+and ending indices of the overlapped region on the $g$ read, and $rc$ is a
+boolean that indicates whether the overlap is reverse-complement or not.  An
+overlap therefore has two overlapped regions--- one on each read.  For exact
+overlaps, the two overlapped regions are the same length; in other words,
+$g_{end} - g_{begin} = f_{end} - f_{begin}$.
+
+An \KeyTerm{extreme point} occurs when an overlapped region extends to the
+beginning or the end of the corresponding read.  An overlap always contains at
+least two extreme points.
+
+A \KeyTerm{containing overlap} is an overlap in which the overlapped region of
+one read, the \KeyTerm{contained read}, is the entire read.  A special case of a
+containing overlap is an overlap where the overlapped regions include the
+entirety of both reads.  For an exact overlap, this can only happen if the two
+reads are either identical or exact reverse-complements of each other.
 
@@
 that two overlapping reads actually came from adjacent positions, we {\em can}
-say for certain that two reads that do not overlap did not come from adjacent
+be certain that two reads that do not overlap did not come from adjacent
 positions.  Regardless of this, \Figure{fig:overlap_examples} shows how two
@@
 
-Now that the idea of an exact overlap has been formalized, what we would like to
-do is compute all pairwise exact overlaps of some minimum length $L$ among the
-	reads that were provided as input to the assembler.  The chosen value of $L$
-	must be balanced so that we can find enough overlaps to adequately cover the
-	genome (small $L$ better) but still exclude most spurious overlaps (large
-	$L$ better).
+Now that the idea of an exact overlap has been formalized, our goal is to
+compute all pairwise exact overlaps of some minimum length $L$ among the reads
+that were provided as input to the assembler.  The chosen value of $L$ must be
+balanced so that we can find enough overlaps to adequately cover the genome
+(small $L$ better) but still exclude most spurious overlaps (large $L$ better).
 
 The naive way to compute overlaps is to consider every pair of reads in the
-input, including only distinct pairs but also including the special cases of
-each read paired with itself, and search for each of the types of overlaps
-between the two reads in each pair.  If there are $N$ reads in the input, this
-algorithm will need to consider $\frac{N^2}{2} + N$ pairs of reads. The time
-spent on each read pair will depend several factors, including the minimum
-overlap length $L$ and the read lengths.  But even if we somehow could spend
-constant time on each read pair, this naive algorithm would be prohibitively
-slow because the running time rises quadraticly with respect to the number of
-reads, and there can easily be millions of reads in the input.
+input, including only distinct pairs but also including each read paired with
+itself, and search for each of the types of overlaps between the two reads in
+each pair.  If there are $N$ reads in the input, this algorithm will need to
+consider $\frac{N^2}{2} + N$ pairs of reads. The time spent on each read pair
+will depend several factors, including the minimum overlap length $L$ and the
+read lengths.  But even if we somehow could spend constant time on each read
+pair, this naive algorithm would be prohibitively slow because the running time
+rises quadraticly with respect to the number of reads, and there can easily be
+millions of reads in the input.
 
@@
 maximal matched region is compared against the definition of an overlap.  In
-particular, it must be at least as long as the minimum overlap length, it must
-have least 2 extreme points over the two reads, and it cannot be the prefix or
-suffix of both reads if the match does not extend to the opposite end of at
-least one of the reads.  If all of these conditions are met, this maximal matched
-region is translated into an overlap and added to the set of overlaps if not
-already present.
+particular, it must have length greater than or equal to $L$, it must have least
+2 extreme points over the two reads, and it cannot be the prefix or suffix of
+both reads if the match does not extend to the opposite end of at least one of
+the reads.  If all of these conditions are met, this maximal matched region is
+translated into an overlap and added to the set of overlaps if not already
+present.
 
@@
 implemented as the program \ProgramName{compute-overlaps}.  It takes as input a
-file that contains a list of reads, as well as a parameter that specifies the
-minimum overlap length $L$.  It produces as output (via a data structure
-serialized to a file) all exact pairwise overlaps of minimum length $L$ among
-reads in the input file, including reverse-complement overlaps and
-self-overlaps.  The running time of the algorithm was discussed above.  The
-memory usage is linear with respect to the total length of the reads and the
-number of overlaps, which likely could be improved.
+file that contains a list of reads and a parameter that specifies the minimum
+overlap length $L$.  It produces as output, via a data structure serialized to a
+file, all exact pairwise overlaps of minimum length $L$ among the input reads,
+including reverse-complement overlaps and self-overlaps.  The running time of
+the algorithm was discussed above.  The memory usage is linear with respect to
+the total length of the reads and the number of overlaps, which likely could be
+improved.
 
@@
 \Figure{fig:bidiwalk_example}, in order to check whether the walk is valid or
-not, we must consider every vertex in the possible walk except for the very
-first and last vertices.  For each such vertex, we need to verify that each time
-that vertex is passed through as part of a walk, the arrow heads adjacent to
-that vertex that are used in the relevant part of the walk are pointed in the
-same direction {\em relative to the walk}. If there are no such pairs of heads
-that point towards each other or point away from each other, the walk is valid.
-The red circles in \Figure{fig:bidiwalk_example} show the two pairs of heads
-that need to be checked for the possible walk $1 \to 2 \to 3 \to 4$.
+not, we must consider every vertex in the possible walk except for the first and
+last vertices.  For each such vertex, we need to verify that each time that
+vertex is passed through as part of a walk, the arrow heads adjacent to that
+vertex that are used in the relevant part of the walk are pointed in the same
+direction {\em relative to the walk}. If there are no such pairs of heads that
+point towards each other or point away from each other, the walk is valid.  The
+red circles in \Figure{fig:bidiwalk_example} show the two pairs of heads that
+need to be checked for the possible walk $1 \to 2 \to 3 \to 4$.
 
@@
 
-After transitive reduction, a non-repetitive region of the genome will typically
+After transitive reduction, a non-repetitive region of the genome typically will
 be represented in the string graph by a chain of vertices that is part of a
@@
 This replacement edge is labeled with the concatenation of the sequences that
-the edges are labeled with.  In the bidirected case, this concatenation must be
-done for both directions in which the unbranched path might have been traversed.
-\Figure{fig:collapse_unbranched_paths} shows an example of how a short
-unbranched path can be collapsed.
-
-Collapsing unbranched paths is implemented as the program
-\ProgramName{collapse-unbranched-paths}.  It takes as input a file containing a
-string graph and produces as output a file containing the string graph with
-unbranched paths collapsed.  The algorithm marks vertices with in-degree 1 and
-out-degree 1 as inner, then iterates through every non-inner vertex to locate
-adjacent inner vertices, which must be the ends of unbranched paths.  Each
-unbranched path is then walked and collapsed.  Vertices are currently
-re-numbered; so, for example, vertex 10 after collapsing unbranched paths does
-not necessarily refer to the 10th read in the list of uncontained reads from
-which the graph was constructed.
+the original edges were labeled with.  In the bidirected case, this
+concatenation must be done for both directions in which the unbranched path
+might have been traversed.  \Figure{fig:collapse_unbranched_paths} shows an
+example of how a short unbranched path can be collapsed.
 
@@
 	\end{center}
-	\caption{Vertex 4 an \KeyTerm{inner vertex} that is part of an unbranched
-	path that can be collapsed.  Edge labels are not shown.}
+	\caption{Vertex 4 is an inner vertex that is part of an unbranched path that
+	can be collapsed.  Edge labels are not shown.}
 	\label{fig:collapse_unbranched_paths}
 \end{figure}
+
+Collapsing unbranched paths is implemented as the program
+\ProgramName{collapse-unbranched-paths}.  It takes as input a file containing a
+string graph and produces as output a file containing the string graph with
+unbranched paths collapsed.  The algorithm marks vertices with in-degree 1 and
+out-degree 1 as inner, then iterates through every non-inner vertex to locate
+adjacent inner vertices, which must be the ends of unbranched paths.  Each
+unbranched path is then walked and collapsed.  Vertices are currently
+re-numbered; so, for example, vertex 10 after collapsing unbranched paths does
+not necessarily refer to the 10th read in the list of uncontained reads from
+which the graph was constructed.
 
@@
 			\begin{tikzpicture}[>=triangle 45]
-				\node[circle,fill=\nodecolor,draw=black] (A) at (0.0, 0) {A};
-				\node[circle,fill=\nodecolor,draw=black] (B) at (3.0, 0) {B};
+				\node[circle,fill=\nodecolor,draw=black] (A) at (0.0, 0) {1};
+				\node[circle,fill=\nodecolor,draw=black] (B) at (3.0, 0) {2};
 				\draw[>->,very thick] (A) edge node[anchor=north] {4639100} (B);
@@
 of reasons, such as gene duplications.} \Figure{fig:E_coli} shows an example of
-the simplified bidirected string graph for a real genome.  It is apparent that
-even when using perfect, fairly long reads that cover the entire genome, the
-repeats in the genome cause branches to appear in the graph when the repeats are
-too long to be resolved by the reads at this stage in the assembly.
+the simplified bidirected string graph for reads sampled from a real genome.  It
+is apparent that even when using perfect, fairly long reads that cover the
+entire genome, the repeats in the genome cause branches to appear in the graph
+when the repeats are too long to be resolved by the reads at this stage in the
+assembly algorithm.
 
@@
 string graph on a very small example and provides a more detailed look at the
-actual data.  The genome is 200 bp long and is given by the following file:
+actual data.  The genome is 200 bp long and is given by the following plain-text
+file:
 
@@
 Note that in this file format, a continuous sequence can be split among multiple
-lines; a new sequence only begins after a line beginning with {\tt >} is
+lines, and a new sequence only begins after a line beginning with {\tt >} is
 reached.
@@
 	\begin{tikzpicture}[>=\ReadArrowType,x=0.56mm,y=3.5mm]
-		\draw[<-, style=thick] (0, 1) node[anchor=east] {G.} -- (200, 1);
-		\draw[->, style=thick] (0, 0) node[anchor=east] {G.} -- (200, 0);
+		\draw[<-, style=thick] (0, 1) node[anchor=east] {$G'$.} -- (200, 1);
+		\draw[->, style=thick] (0, 0) node[anchor=east] {$G$.} -- (200, 0);
 
@@
 build the assembly graph.  Note that there are no contained reads in this
-example, so there are as many vertices as original reads.  Also, in the below
-graph drawing (as well as following graph drawings), each edge is labeled with
-the length of that edge's sequence rather than the sequence itself.
-Furthermore, although each edge really contains a separate sequence for each
-direction, for this particular example all the reads are the same length and
-only exact overlaps are considered, which makes it so the two sequences of each
-edge are always the same length.
+example, so there are as many vertices as original reads.  Also, in the
+following graph drawings, each edge is labeled with the length of that edge's
+sequence rather than the sequence itself.  Furthermore, although each edge
+really contains a separate sequence for each direction, for this particular
+example all the reads are the same length and only exact overlaps are
+considered, which makes it so the two sequences of each edge are always the same
+length.
 
@@
 part of the original genome by walking $2 \to 1$.  Note that there is an implied
-50bp before we start at the end of the read now labeled as 2, so the actual
-assembled sequence is $113 + 50 = 163$ bp long.  The assembler should then
-generate the assembly:
+50bp before we start at the end of the read now labeled as 2 (which was
+originally read 5), so the actual assembled sequence is $113 + 50 = 163$ bp
+long.  The assembler should then generate the assembly:
 
@@
 contained read to the edge, or in general the set of edges, in which the end
-might be located if it were included as its own vertex.
-
-The algorithm for doing so is outlined in \cite{Myers2005} and was implemented
-by the program \ProgramName{map-contained-reads}, which runs before
+might be located if it were included as its own vertex.  Note that this mapping
+makes more sense if we consider it in terms of the directed analogue of the
+bidirected graph introduced in \Subsection{subsec:directed}.
+
+The algorithm for doing this mapping is outlined in \cite{Myers2005} and was
+implemented by the program \ProgramName{map-contained-reads}, which runs before
 \ProgramName{collapse-unbranched-paths}.  Briefly, for each contained read $r$
@@
 flow and a cost per unit flow $v(e)$, and for each vertex $v$ a supply
-	(positive) or demand (negative) $b(v)$ for flow.  It produces as output a
-	flow $x(e)$ on each edge $e$ such that $\sum_e v(e)x(e)$ is minimized
-	subject to $x(e) \in [l(e), u(e)]$ and $\sum_{u \to v}x(u \to v) + b(v) =
-	\sum_{v \to w} x(v \to w)$ \cite{Myers2005}.  In the minimum-cost
-	circulation problem, as opposed to other network flow problems, no vertices are
-	treated specially as sources or sinks; instead, flow must be balanced at
-	every vertex $v$, but only after taking into account the supply or demand
-	$b(v)$ at that vertex.
+(positive) or demand (negative) $b(v)$ for flow.  It produces as output a flow
+$x(e)$ on each edge $e$ such that $\sum_e v(e)x(e)$ is minimized subject to
+$x(e) \in [l(e), u(e)]$ and $\sum_{u \to v}x(u \to v) + b(v) = \sum_{v \to w}
+x(v \to w)$ \cite{Myers2005}.  In the minimum-cost circulation problem, as
+opposed to other network flow problems, no vertices are treated specially as
+sources or sinks; instead, ingoing and outgoing flow must be balanced at every
+vertex $v$, but only after taking into account the supply or demand $b(v)$ at
+that vertex.
 
@@
 vertex the same number of times that it is exited, making the graph as a whole
-be balanced.  Furthermore, we have $v(e) = 1$ because no edges are considered to
-be more costly to use than any other in this formulation.\footnote{Note that the
+be balanced.  Furthermore, we have $v(e) = 1$ because no edges are considered
+more costly to use than any other in this formulation.\footnote{Note that the
 goal of this assembly algorithm is not to minimize the assembled genome length,
@@
 presence of \BidirectedEdgeInnie and \BidirectedEdgeOutie edges, where the
-magnitude of the flow on that edge is counted as going into both vertices (in
-the case of a \BidirectedEdgeOutie edge) or out of both vertices (in the case of
+magnitude of the flow on that edge is incoming to both both vertices (in
+the case of a \BidirectedEdgeOutie edge) or outgoing from both vertices (in the case of
 an \BidirectedEdgeInnie edge).  This does not imply that the edge itself is a
@@
 at great cost ($v(e) = X$, where $X$ is a very large number).  The addition of
-these edges will make the graph strongly connected, and therefore it will be
-possible to solve the minimum-cost circulation problem to find a flow that
-corresponds to balancing the in-degree and out-degree of every vertex by
-assigning a certain traversal count $t(e)$ to each edge $e$.
+these edges will make the graph strongly connected and possible to solve the
+minimum-cost circulation problem to find a flow that corresponds to balancing
+the in-degree and out-degree of every vertex by assigning a certain traversal
+count $t(e)$ to each edge $e$.
 
@@
 minimum-cost circulation problem.  (A more thorough explanation of this
-observation can be found in \cite{Medvedev2009}.)  The cycle corresponds to a
-consistent assembly of the genome, and it can be decomposed into a set of
-logical walks by placing a logical breakpoint every time the special vertex $s$
-is passed through.  Each of these logical walks represents the assembly of a
-separate contig.
+observation can be found in \cite{Medvedev2009}.)  The cycle corresponds to an
+assembly of the genome that is consistent with the original reads, and it can be
+decomposed into a set of logical walks by placing a logical breakpoint every
+time the special vertex $s$ is passed through.  Each of these logical walks
+represents the assembly of a separate contig.
 
@@
 Burrows-Wheeler transform to efficiently construct the transitively-reduced
-bidirected string graph directly from the reads \cite{Simpson2010}.
+bidirected string graph directly from the original reads \cite{Simpson2010}.
 \item Use mate-pair information to help resolve repeats and connect assembled

### 2012-12-13 src/AnyStringGraph.h
Prepare network flow

@@ -75,6 +75,7 @@ public:
 	DISPATCH0(transitive_reduction);
 	DISPATCH0(calculate_A_statistics);
 	DISPATCH0(collapse_unbranched_paths);
+	DISPATCH0(min_cost_circulation);
 	DISPATCH1(print_stats, std::ostream &);
 	DISPATCH1(write, const char *);
 	DISPATCH1(extract_edge_seqs, BaseVecVec &);

### 2012-12-13 src/BidirectedStringGraph.h

@@ -277,6 +277,7 @@ public:
 	}
 
 	void transitive_reduction();
+	void min_cost_circulation();
 	void collapse_unbranched_paths();
 	void calculate_A_statistics();
 	void print_stats(std::ostream & os) const;

### 2012-12-13 src/DirectedStringGraph.cc

@@ -5,6 +5,7 @@
 #include <boost/archive/binary_oarchive.hpp>
 #include "BidirectedStringGraph.h"
 #include <math.h>
+#include <lemon/network_simplex.h>
 
 const char DirectedStringGraph::magic[] =
 	{'D', 'i', 'g', 'r', 'a', 'p', 'h', '\0', '\0', '\0'};
@@ -675,7 +676,7 @@ void DirectedStringGraph::calculate_A_statistics()
 		for (size_t j = 0; j < n_edges; j++) {
 			DirectedStringGraphEdge & e = _edges[j];
 			size_t edge_len = _edges[j].length();
-			unsigned edge_reads = _edges[j].get_num_inner_vertices();
+			unsigned edge_reads = _edges[j].get_mapped_read_count();
 			float A_statistic = (global_arrival_rate * edge_len) -
 					     (edge_reads * M_LN2);
 			e.set_A_statistic(A_statistic);
@@ -699,5 +700,33 @@ void DirectedStringGraph::calculate_A_statistics()
 		info("num_unique_edges = %zu", num_unique_edges);
 		info("num_optional_edges = %zu", num_optional_edges);
 		info("num_required_edges = %zu", num_required_edges);
+	}
+}
+
+void DirectedStringGraph::min_cost_circulation()
+{
+	foreach (DirectedStringGraphEdge & e, edges()) {
+		if (e.get_A_statistic() > 0) {
+			e.set_flow_bounds(1, 1);
+		} else if (e.get_num_inner_vertices() > 0) {
+			e.set_flow_bounds(1, DirectedStringGraphEdge::INFINITE_FLOW);
+		} else {
+			e.set_flow_bounds(0, DirectedStringGraphEdge::INFINITE_FLOW);
+		}
+		e.set_cost_per_unit_flow(1);
+	}
+
+	v_idx_t n_verts = num_vertices();
+	v_idx_t special_v_idx = n_verts;
+	_vertices.resize(n_verts + 1);
+	for (v_idx_t v_idx = 0; v_idx < n_verts; v_idx++) {
+		DirectedStringGraphEdge * e;
+
+		e = &add_unlabeled_edge(v_idx, special_v_idx);
+		e->set_flow_bounds(0, DirectedStringGraphEdge::INFINITE_FLOW);
+		e->set_cost_per_unit_flow(10000000);
+		e = &add_unlabeled_edge(special_v_idx, v_idx);
[…]

### 2012-12-13 src/DirectedStringGraph.h

@@ -40,9 +40,13 @@ class DirectedStringGraphEdge : public StringGraphEdge {
 public:
 	typedef unsigned int v_idx_t;
 private:
-	v_idx_t _v1_idx;
-	v_idx_t _v2_idx;
-	BaseVec _seq;
+	v_idx_t  _v1_idx;
+	v_idx_t  _v2_idx;
+	BaseVec  _seq;
+	unsigned _flow_lower_bound;
+	unsigned _flow_upper_bound;
+	unsigned _cost_per_unit_flow;
+	unsigned _traversal_count;
 
 	// Serialize or deserialize the directed string graph edge to/from a
 	// stream.
@@ -56,6 +60,8 @@ private:
 		ar & _seq;
 	}
 public:
+	static const unsigned INFINITE_FLOW = std::numeric_limits<unsigned>::max();
+
 	// Return a reference to the sequence associated with this edge of the
 	// directed string graph.
 	BaseVec & get_seq() { return _seq; }
@@ -92,6 +98,18 @@ public:
 	void set_v1_idx(const v_idx_t v1_idx) { _v1_idx = v1_idx; }
 
 	void set_v2_idx(const v_idx_t v2_idx) { _v2_idx = v2_idx; }
+
+	void set_flow_bounds(const unsigned flow_lower_bound,
+			     const unsigned flow_upper_bound)
+	{
+		_flow_lower_bound = flow_lower_bound;
+		_flow_upper_bound = flow_upper_bound;
+	}
+
+	void set_cost_per_unit_flow(const unsigned cost_per_unit_flow)
+	{
+		_cost_per_unit_flow = cost_per_unit_flow;
+	}
 
 	// Print this directed string graph edge.
 	void print(std::ostream & os = std::cout, const v_idx_t v_idx = 0,
@@ -159,6 +177,19 @@ private:
 		edge_idx_t edge_idx = this->push_back_edge(e);
 		_vertices[v1_idx].add_edge_idx(edge_idx);
 	}
+
+	DirectedStringGraphEdge &
+	add_unlabeled_edge(const v_idx_t v1_idx,
+			   const v_idx_t v2_idx)
+	{
+		DirectedStringGraphEdge e;
+		e.set_v_indices(v1_idx, v2_idx);
+		edge_idx_t edge_idx = this->push_back_edge(e);
+		_vertices[v1_idx].add_edge_idx(edge_idx);
+		return _edges[edge_idx];
+	}
+
+				
 public:
[…]

### 2012-12-13 src/Kmer.h

@@ -171,6 +171,5 @@ public:
 	{
 		for (unsigned i = 0; i < _K; i++)
 			os << BaseUtils::bin_to_ascii(kmer[i]);
-		return os;
 	}
 };

### 2012-12-13 src/StringGraph.h

@@ -26,6 +26,7 @@ protected:
 	}
 
 	void print(std::ostream & os) const {
+		os << "_mapped_read_count=" << _mapped_read_count << '\t';
 		os << "_num_inner_vertices=" << _num_inner_vertices << '\t';
 		os << "_A_statistic=" << _A_statistic << '\t';
 	}

### 2012-12-13 src/DirectedStringGraph.cc
network flow

@@ -6,6 +6,7 @@
 #include "BidirectedStringGraph.h"
 #include <math.h>
 #include <lemon/network_simplex.h>
+#include <lemon/smart_graph.h>
 
 const char DirectedStringGraph::magic[] =
 	{'D', 'i', 'g', 'r', 'a', 'p', 'h', '\0', '\0', '\0'};
@@ -729,4 +730,22 @@ void DirectedStringGraph::min_cost_circulation()
 		e->set_flow_bounds(0, DirectedStringGraphEdge::INFINITE_FLOW);
 		e->set_cost_per_unit_flow(10000000);
 	}
+
+	n_verts++;
+	edge_idx_t n_edges = num_edges();
+
+	lemon::SmartDigraph G;
+	G.reserveNode(n_verts);
+	G.reserveArc(n_edges);
+	{
+		std::vector<lemon::SmartDigraph::Node> nodes(n_verts);
+		for (v_idx_t v_idx = 0; v_idx < n_verts; v_idx++) {
+			nodes[v_idx] = G.addNode();
+		}
+		foreach (DirectedStringGraphEdge & e, edges()) {
+			G.addArc(nodes[e.get_v1_idx()], nodes[e.get_v2_idx()]);
+		}
+	}
+	lemon::ArcMap lower_map(G);
+	lemon::NetworkSimplex<lemon::SmartDigraph> simplex(G);
 }

### 2012-12-13 src/BidirectedStringGraph.cc
min-cost-circulation (in progress)

@@ -330,3 +330,8 @@ void BidirectedStringGraph::calculate_A_statistics()
 {
 	unimplemented();
 }
+
+void BidirectedStringGraph::min_cost_circulation()
+{
+	unimplemented();
+}

### 2012-12-13 src/DirectedStringGraph.cc

@@ -706,7 +706,8 @@ void DirectedStringGraph::calculate_A_statistics()
 
 void DirectedStringGraph::min_cost_circulation()
 {
-	foreach (DirectedStringGraphEdge & e, edges()) {
+	info("Initializing lower and upper flow bounds on %zu edges", num_edges());
+	foreach (DirectedStringGraphEdge & e, _edges) {
 		if (e.get_A_statistic() > 0) {
 			e.set_flow_bounds(1, 1);
 		} else if (e.get_num_inner_vertices() > 0) {
@@ -718,34 +719,70 @@ void DirectedStringGraph::min_cost_circulation()
 	}
 
 	v_idx_t n_verts = num_vertices();
-	v_idx_t special_v_idx = n_verts;
-	_vertices.resize(n_verts + 1);
+	_vertices.resize(n_verts + 2);
+	_vertices[n_verts].set_special();
+	_vertices[n_verts + 1].set_special();
+
+	info("Adding special vertex and edges");
 	for (v_idx_t v_idx = 0; v_idx < n_verts; v_idx++) {
 		DirectedStringGraphEdge * e;
 
-		e = &add_unlabeled_edge(v_idx, special_v_idx);
-		e->set_flow_bounds(0, DirectedStringGraphEdge::INFINITE_FLOW);
-		e->set_cost_per_unit_flow(10000000);
-		e = &add_unlabeled_edge(special_v_idx, v_idx);
-		e->set_flow_bounds(0, DirectedStringGraphEdge::INFINITE_FLOW);
-		e->set_cost_per_unit_flow(10000000);
+		//v_idx_t special_v_idx = n_verts | (v_idx & 1);
+		for (v_idx_t special_v_idx = n_verts;
+		     special_v_idx < n_verts + 2;
+		     special_v_idx++)
+		{
+			e = &add_unlabeled_edge(v_idx, special_v_idx);
+			e->set_flow_bounds(0, DirectedStringGraphEdge::INFINITE_FLOW);
+			e->set_cost_per_unit_flow(DirectedStringGraphEdge::INFINITE_COST);
+			e->set_special();
+
+			e = &add_unlabeled_edge(special_v_idx, v_idx);
+			e->set_flow_bounds(0, DirectedStringGraphEdge::INFINITE_FLOW);
+			e->set_cost_per_unit_flow(DirectedStringGraphEdge::INFINITE_COST);
+			e->set_special();
+		}
[…]

### 2012-12-13 src/DirectedStringGraph.h

@@ -43,10 +43,9 @@ private:
 	v_idx_t  _v1_idx;
 	v_idx_t  _v2_idx;
 	BaseVec  _seq;
-	unsigned _flow_lower_bound;
-	unsigned _flow_upper_bound;
-	unsigned _cost_per_unit_flow;
-	unsigned _traversal_count;
+	int _flow_lower_bound;
+	int _flow_upper_bound;
+	int _cost_per_unit_flow;
 
 	// Serialize or deserialize the directed string graph edge to/from a
 	// stream.
@@ -60,7 +59,8 @@ private:
 		ar & _seq;
 	}
 public:
-	static const unsigned INFINITE_FLOW = std::numeric_limits<unsigned>::max();
+	static const int INFINITE_FLOW = 1000000;
+	static const int INFINITE_COST = 1000000;
 
 	// Return a reference to the sequence associated with this edge of the
 	// directed string graph.
@@ -99,14 +99,17 @@ public:
 
 	void set_v2_idx(const v_idx_t v2_idx) { _v2_idx = v2_idx; }
 
-	void set_flow_bounds(const unsigned flow_lower_bound,
-			     const unsigned flow_upper_bound)
+	void set_flow_bounds(const int flow_lower_bound,
+			     const int flow_upper_bound)
 	{
 		_flow_lower_bound = flow_lower_bound;
 		_flow_upper_bound = flow_upper_bound;
 	}
 
-	void set_cost_per_unit_flow(const unsigned cost_per_unit_flow)
+	int get_flow_lower_bound() const { return _flow_lower_bound; }
+	int get_flow_upper_bound() const { return _flow_upper_bound; }
+
+	void set_cost_per_unit_flow(const int cost_per_unit_flow)
 	{
 		_cost_per_unit_flow = cost_per_unit_flow;
 	}
@@ -149,7 +152,11 @@ public:
 			os << _seq;
 		else
 			os << _seq.length();
-		os << "\" ];\n";
+		os << "\" ";
+		if (_traversal_count != 0) {
+			os << "color=red ";
+		}
+		os << "];\n";
 	}
 };
 

### 2012-12-13 src/StringGraph.h

@@ -17,18 +17,24 @@ protected:
 	float _mapped_read_count;
 	unsigned _num_inner_vertices;
 	float _A_statistic;
+	int _traversal_count;
+	bool _is_special;
 
 	StringGraphEdge()
 	{
 		_mapped_read_count = 1.0;
 		_num_inner_vertices = 0;
 		_A_statistic = 0.0;
+		_traversal_count = 0;
+		_is_special = false;
 	}
 
 	void print(std::ostream & os) const {
 		os << "_mapped_read_count=" << _mapped_read_count << '\t';
 		os << "_num_inner_vertices=" << _num_inner_vertices << '\t';
 		os << "_A_statistic=" << _A_statistic << '\t';
+		os << "_traversal_count=" << _traversal_count << '\t';
+		os << "_is_special=" << _is_special << '\t';
 	}
 
 	friend class boost::serialization::access;
@@ -38,6 +44,8 @@ protected:
 		ar & _mapped_read_count;
 		ar & _num_inner_vertices;
 		ar & _A_statistic;
+		ar & _is_special;
+		ar & _traversal_count;
 	}
 public:
 	void increment_mapped_read_count(float n) { _mapped_read_count += n; }
@@ -50,6 +58,15 @@ public:
 
 	float get_A_statistic() const { return _A_statistic; }
 	void set_A_statistic(float f) { _A_statistic = f; }
+
+	void set_special() { _is_special = true; }
+	bool is_special() const { return _is_special; }
+
+	void set_traversal_count(const int traversal_count)
+	{
+		_traversal_count = traversal_count;
+	}
+
 };
 
 // Base class for vertices of the string graph.
@@ -61,6 +78,7 @@ public:
 protected:
 	// vector of the indices of the edges that go out from this vertex.
 	std::vector<edge_idx_t> _edge_indices;
+	bool _is_special;
 
 	// Serialize or deserialize the vertex to/from a stream.
 	friend class boost::serialization::access;
@@ -68,9 +86,10 @@ protected:
 	void serialize(Archive & ar, unsigned version)
 	{
 		ar & _edge_indices;
+		ar & _is_special;
 	}
 
-	StringGraphVertex() { }
+	StringGraphVertex() { _is_special = false; }
 public:
 	// Add a new edge index to the list of edge indices associated with this
 	// string graph vertex.
@@ -78,6 +97,9 @@ public:
 	{
 		_edge_indices.push_back(edge_idx);
 	}
+
[…]

### 2012-12-13 src/min-cost-circulation.cc

@@ -1,0 +1,25 @@
+#include "AnyStringGraph.h"
+
+DEFINE_USAGE(
+"Usage: min-cost-circulation GRAPH_FILE OUT_GRAPH_FILE\n"
+"\n"
+"Solves a minimum-cost circulation problem on a directed or bidirected\n"
+"string graph.\n"
+"\n"
+"Input:\n"
+"      GRAPH_FILE:   A directed or bidirected string graph in binary format.\n"
+"\n"
+"Output:\n"
+"      OUT_GRAPH_FILE:  The resulting graph with each edge labeled with a\n"
+"                       traversal count.\n"
+);
+
+int main(int argc, char *argv[])
+{
+	USAGE_IF(argc != 3);
+	info("Loading string graph from \"%s\"", argv[1]);
+	AnyStringGraph graph(argv[1]);
+	graph.min_cost_circulation();
+	info("Writing string graph to \"%s\"", argv[2]);
+	graph.write(argv[2]);
+}

### 2012-12-13 src/BidirectedStringGraph.cc
color

@@ -226,6 +226,8 @@ void BidirectedStringGraph::build_from_digraph(const DirectedStringGraph & digra
 							 v_w.get_mapped_read_count()) / 2);
 				e.set_A_statistic((w_v.get_A_statistic() +
 						   v_w.get_A_statistic()) / 2);
+				e.set_traversal_count((w_v.get_traversal_count() +
+						       v_w.get_traversal_count()));
 				//if (w_v.get_num_inner_vertices() !=
 				    //v_w.get_num_inner_vertices())
 				//{

### 2012-12-13 src/BidirectedStringGraph.h

@@ -239,6 +239,9 @@ public:
 			} else {
 				os << " label=\"" << length() << "\"";
 			}
+			if (_traversal_count != 0) {
+				os << " color=red ";
+			}
 			os << " ];\n";
 		}
 	}

### 2012-12-13 src/StringGraph.h

@@ -66,6 +66,7 @@ public:
 	{
 		_traversal_count = traversal_count;
 	}
+	int get_traversal_count() const { return _traversal_count; }
 
 };
 

### 2012-12-13 src/DirectedStringGraph.cc
DirectedStringGraph::min_cost_circulation()

@@ -706,24 +706,16 @@ void DirectedStringGraph::calculate_A_statistics()
 
 void DirectedStringGraph::min_cost_circulation()
 {
-	info("Initializing lower and upper flow bounds on %zu edges", num_edges());
-	foreach (DirectedStringGraphEdge & e, _edges) {
-		if (e.get_A_statistic() > 0) {
-			e.set_flow_bounds(1, 1);
-		} else if (e.get_num_inner_vertices() > 0) {
-			e.set_flow_bounds(1, DirectedStringGraphEdge::INFINITE_FLOW);
-		} else {
-			e.set_flow_bounds(0, DirectedStringGraphEdge::INFINITE_FLOW);
-		}
-		e.set_cost_per_unit_flow(1);
-	}
+	static const int INFINITE_FLOW = 1000000;
+	static const int HUGE_COST = 1000000;
+	static const float SINGLE_COPY_THRESHOLD = 17.0;
 
+	info("Adding special vertex and edges");
 	v_idx_t n_verts = num_vertices();
 	_vertices.resize(n_verts + 2);
 	_vertices[n_verts].set_special();
 	_vertices[n_verts + 1].set_special();
 
-	info("Adding special vertex and edges");
 	for (v_idx_t v_idx = 0; v_idx < n_verts; v_idx++) {
 		DirectedStringGraphEdge * e;
 
@@ -733,13 +725,9 @@ void DirectedStringGraph::min_cost_circulation()
 		     special_v_idx++)
 		{
 			e = &add_unlabeled_edge(v_idx, special_v_idx);
-			e->set_flow_bounds(0, DirectedStringGraphEdge::INFINITE_FLOW);
-			e->set_cost_per_unit_flow(DirectedStringGraphEdge::INFINITE_COST);
 			e->set_special();
 
 			e = &add_unlabeled_edge(special_v_idx, v_idx);
-			e->set_flow_bounds(0, DirectedStringGraphEdge::INFINITE_FLOW);
-			e->set_cost_per_unit_flow(DirectedStringGraphEdge::INFINITE_COST);
 			e->set_special();
 		}
 	}
@@ -747,13 +735,15 @@ void DirectedStringGraph::min_cost_circulation()
 	n_verts += 2;
 	edge_idx_t n_edges = num_edges();
 
-	info("Creating lemon::SmartDigraph with %zu nodes and %zu arcs",
[…]

### 2012-12-13 src/DirectedStringGraph.h

@@ -43,9 +43,6 @@ private:
 	v_idx_t  _v1_idx;
 	v_idx_t  _v2_idx;
 	BaseVec  _seq;
-	int _flow_lower_bound;
-	int _flow_upper_bound;
-	int _cost_per_unit_flow;
 
 	// Serialize or deserialize the directed string graph edge to/from a
 	// stream.
@@ -59,9 +56,6 @@ private:
 		ar & _seq;
 	}
 public:
-	static const int INFINITE_FLOW = 1000000;
-	static const int INFINITE_COST = 1000000;
-
 	// Return a reference to the sequence associated with this edge of the
 	// directed string graph.
 	BaseVec & get_seq() { return _seq; }
@@ -98,21 +92,6 @@ public:
 	void set_v1_idx(const v_idx_t v1_idx) { _v1_idx = v1_idx; }
 
 	void set_v2_idx(const v_idx_t v2_idx) { _v2_idx = v2_idx; }
-
-	void set_flow_bounds(const int flow_lower_bound,
-			     const int flow_upper_bound)
-	{
-		_flow_lower_bound = flow_lower_bound;
-		_flow_upper_bound = flow_upper_bound;
-	}
-
-	int get_flow_lower_bound() const { return _flow_lower_bound; }
-	int get_flow_upper_bound() const { return _flow_upper_bound; }
-
-	void set_cost_per_unit_flow(const int cost_per_unit_flow)
-	{
-		_cost_per_unit_flow = cost_per_unit_flow;
-	}
 
 	// Print this directed string graph edge.
 	void print(std::ostream & os = std::cout, const v_idx_t v_idx = 0,

### 2012-12-14 src/BidirectedStringGraph.cc
BidirectedStringGraph::eulerian_path() (in progress)

@@ -237,6 +237,10 @@ void BidirectedStringGraph::build_from_digraph(const DirectedStringGraph & digra
 				//}
 				e.set_num_inner_vertices((w_v.get_num_inner_vertices() +
 							 v_w.get_num_inner_vertices()) / 2);
+				if (v_w.is_special()) {
+					assert(w_v.is_special());
+					e.set_special();
+				}
 
 				edge_idx_t edge_idx = this->push_back_edge(e);
 				_vertices[v1_idx].add_edge_idx(edge_idx);
@@ -336,4 +340,62 @@ void BidirectedStringGraph::calculate_A_statistics()
 void BidirectedStringGraph::min_cost_circulation()
 {
 	unimplemented();
+}
+
+void BidirectedStringGraph::eulerian_path(std::vector<edge_idx_t> & path) const
+{
+	v_idx_t n_verts = num_vertices();
+	edge_idx_t n_edges = num_edges();
+
+	info("Finding Eulerian path in bidirected graph");
+	info("num_vertices = %lu", n_verts);
+	info("num_edges = %lu", n_edges);
+
+	unsigned long total_traversal_count = 0;
+	unsigned long num_special_edges = 0;
+	unsigned long special_traversal_count = 0;
+	v_idx_t start_v_idx;
+	bool found_start_v_idx = false;
+	foreach (const BidirectedStringGraphEdge & e, _edges) {
+		if (e.is_special()) {
+			num_special_edges++;
+			special_traversal_count += e.get_traversal_count();
+		}
+		total_traversal_count += e.get_traversal_count();
+		if (!found_start_v_idx) {
+			if (e.v1_outward()) {
+				start_v_idx = e.get_v1_idx();
+				found_start_v_idx = true;
+			} else if (e.v2_outward()) {
+				start_v_idx = e.get_v2_idx();
+				found_start_v_idx = true;
+			}
+		}
+	}
+	info("total_traversal_count = %lu", total_traversal_count);
+	info("num_special_edges = %lu", num_special_edges);
+	info("special_traversal_count = %lu", special_traversal_count);
+
+	if (!found_start_v_idx) {
+		info("WARNING: Empty Eulerian path!");
+		return;
+	}
+	info("start_v_idx = %lu", start_v_idx);
+
+	// Start with empty path and empty stack
+	path.clear();
+	path.reserve(total_traversal_count);
+
[…]

### 2012-12-14 src/BidirectedStringGraph.h

@@ -327,4 +327,6 @@ public:
 			bvv.push_back(e.get_seq_2_to_1());
 		}
 	}
+
+	void eulerian_path(std::vector<edge_idx_t> & path) const;
 };

### 2012-12-14 src/BidirectedStringGraph.cc
BidirectedStringGraph::eulerian_path() (in progress)

@@ -395,7 +395,45 @@ void BidirectedStringGraph::eulerian_path(std::vector<edge_idx_t> & path) const
 	};
 	std::vector<stack_elem> stack;
 
-	std::vector<edge_idx_t> in_idx(n_verts, 0);
-	std::vector<edge_idx_t> out_idx(n_verts, 0);
+	std::vector<edge_idx_t> in_indices(n_verts, 0);
+	std::vector<edge_idx_t> out_indices(n_verts, 0);
 	std::vector<bool> visited(n_edges, false);
+	int dir = IN;
+	v_idx_t v_idx = start_v_idx;
+	while (1) {
+		const BidirectedStringGraphVertex & v = _vertices[v_idx];
+		edge_idx_t edge_idx;
+		bool pop_stack = true;
+		const BidirectedStringGraphEdge * e;
+		if (dir == IN) {
+			for (edge_idx = in_indices[v_idx];
+			     edge_idx < v.degree();
+			     edge_idx++)
+			{
+				e = &_edges[edge_idx];
+				if (e->this_v_outward(v_idx)) {
+					pop_stack = false;
+					break;
+				}
+			}
+			in_indices[v_idx] = edge_idx;
+		} else {
+			for (edge_idx = out_indices[v_idx];
+			     edge_idx < v.degree();
+			     edge_idx++)
+			{
+				e = &_edges[edge_idx];
+				if (e->this_v_inward(v_idx)) {
+					pop_stack = false;
+					break;
+				}
+			}
+			out_indices[v_idx] = edge_idx;
+		}
+		if (pop_stack) {
+			if (stack.size() == 0)
+				break;
+		} else {
+		}
+	}
 }

### 2012-12-14 src/BidirectedStringGraph.h

@@ -25,6 +25,8 @@ public:
 		unimplemented();
 	}
 
+	size_t degree() const { return _edge_indices.size(); }
+
 	// Print a bidirected string graph vertex in DOT format
 	void print_dot(std::ostream & os, size_t v_idx) const
 	{
@@ -141,7 +143,7 @@ public:
 	bool v1_inward() const { return !(v1_outward()); }
 	bool v2_outward() const { return !(v2_inward()); }
 
-	bool this_v_outward(v_idx_t this_v_idx) const
+	bool this_v_outward(const v_idx_t this_v_idx) const
 	{
 		v_idx_t v1_idx, v2_idx;
 		get_v_indices(v1_idx, v2_idx);
@@ -153,7 +155,12 @@ public:
 		}
 	}
 
-	bool other_v_outward(v_idx_t this_v_idx) const
+	bool this_v_inward(const v_idx_t this_v_idx) const
+	{
+		return !this_v_outward(this_v_idx);
+	}
+
+	bool other_v_outward(const v_idx_t this_v_idx) const
 	{
 		v_idx_t v1_idx, v2_idx;
 		get_v_indices(v1_idx, v2_idx);
@@ -166,7 +173,7 @@ public:
 	}
 
 
-	void set_dirs(unsigned dirs)
+	void set_dirs(const unsigned dirs)
 	{
 		_data = (_data & ~(3ULL << 62)) | (uint64_t(dirs) << 62);
 	}

### 2012-12-14 src/BidirectedStringGraph.cc
BidirectedStringGraph::eulerian_path() (in progress)

@@ -342,7 +342,7 @@ void BidirectedStringGraph::min_cost_circulation()
 	unimplemented();
 }
 
-void BidirectedStringGraph::eulerian_path(std::vector<edge_idx_t> & path) const
+void BidirectedStringGraph::eulerian_cycle(std::vector<size_t> & cycle) const
 {
 	v_idx_t n_verts = num_vertices();
 	edge_idx_t n_edges = num_edges();
@@ -383,8 +383,8 @@ void BidirectedStringGraph::eulerian_path(std::vector<edge_idx_t> & path) const
 	info("start_v_idx = %lu", start_v_idx);
 
 	// Start with empty path and empty stack
-	path.clear();
-	path.reserve(total_traversal_count);
+	cycle.clear();
+	cycle.reserve(total_traversal_count);
 
 	static const int IN = 0;
 	static const int OUT = 1;
@@ -401,29 +401,28 @@ void BidirectedStringGraph::eulerian_path(std::vector<edge_idx_t> & path) const
 	int dir = IN;
 	v_idx_t v_idx = start_v_idx;
 	while (1) {
-		const BidirectedStringGraphVertex & v = _vertices[v_idx];
 		edge_idx_t edge_idx;
 		bool pop_stack = true;
-		const BidirectedStringGraphEdge * e;
+		const BidirectedStringGraphVertex & v = _vertices[v_idx];
 		if (dir == IN) {
-			for (edge_idx = in_indices[v_idx];
-			     edge_idx < v.degree();
-			     edge_idx++)
+			for (edge_idx_t edge_idx_idx = in_indices[v_idx];
+			     edge_idx_idx < v.degree();
+			     edge_idx_idx++)
 			{
-				e = &_edges[edge_idx];
-				if (e->this_v_outward(v_idx)) {
+				edge_idx = v.edge_indices()[edge_idx_idx];
+				if (_edges[edge_idx].this_v_outward(v_idx)) {
 					pop_stack = false;
 					break;
 				}
 			}
 			in_indices[v_idx] = edge_idx;
 		} else {
-			for (edge_idx = out_indices[v_idx];
-			     edge_idx < v.degree();
-			     edge_idx++)
+			for (edge_idx_t edge_idx_idx = out_indices[v_idx];
+			     edge_idx_idx < v.degree();
+			     edge_idx_idx++)
 			{
-				e = &_edges[edge_idx];
[…]

### 2012-12-14 src/BidirectedStringGraph.h

@@ -335,5 +335,5 @@ public:
 		}
 	}
 
-	void eulerian_path(std::vector<edge_idx_t> & path) const;
+	void eulerian_cycle(std::vector<size_t> & cycle) const;
 };

### 2012-12-14 src/BidirectedStringGraph.cc
BidirectedStringGraph::eulerian_cycle() (in progress)

@@ -397,7 +397,7 @@ void BidirectedStringGraph::eulerian_cycle(std::vector<size_t> & cycle) const
 
 	std::vector<edge_idx_t> in_indices(n_verts, 0);
 	std::vector<edge_idx_t> out_indices(n_verts, 0);
-	std::vector<bool> visited(n_edges, false);
+	std::vector<unsigned> times_traversed(n_edges, 0);
 	int dir = IN;
 	v_idx_t v_idx = start_v_idx;
 	while (1) {
@@ -410,7 +410,11 @@ void BidirectedStringGraph::eulerian_cycle(std::vector<size_t> & cycle) const
 			     edge_idx_idx++)
 			{
 				edge_idx = v.edge_indices()[edge_idx_idx];
-				if (_edges[edge_idx].this_v_outward(v_idx)) {
+				const BidirectedStringGraphEdge & e = _edges[edge_idx];
+
+				if (times_traversed[edge_idx] < e.get_traversal_count() &&
+				    e.this_v_outward(v_idx))
+				{
 					pop_stack = false;
 					break;
 				}
@@ -422,7 +426,10 @@ void BidirectedStringGraph::eulerian_cycle(std::vector<size_t> & cycle) const
 			     edge_idx_idx++)
 			{
 				edge_idx = v.edge_indices()[edge_idx_idx];
-				if (_edges[edge_idx].this_v_inward(v_idx)) {
+				const BidirectedStringGraphEdge & e = _edges[edge_idx];
+				if (times_traversed[edge_idx] < e.get_traversal_count() &&
+				    e.this_v_inward(v_idx))
+				{
 					pop_stack = false;
 					break;
 				}
@@ -430,6 +437,8 @@ void BidirectedStringGraph::eulerian_cycle(std::vector<size_t> & cycle) const
 			out_indices[v_idx] = edge_idx;
 		}
 		if (pop_stack) {
+			// v does not have an edge that can be traversed.
+			//
 			// Pop (v_idx, edge_idx, dir) from stack and add
 			// edge_idx to cycle.  If stack is empty, the cycle is
 			// complete.
@@ -443,6 +452,8 @@ void BidirectedStringGraph::eulerian_cycle(std::vector<size_t> & cycle) const
 			_edges[edge_idx].print(std::cout, 0, false);
 			cycle.push_back(edge_idx);
 		} else {
+			// v has an edge that can be traversed.
+			//
[…]

### 2012-12-14 src/bidigraph-eulerian-cycle.cc

@@ -1,0 +1,35 @@
+#include "BidirectedStringGraph.h"
+
+DEFINE_USAGE(
+"Usage: bidigraph-eulerian-cycle BIDIGRAPH_FILE OUT_CYCLE_FILE\n"
+"\n"
+"Finds an Eulerian cycle in a bidirected graph.\n"
+"\n"
+"Input:\n"
+"      BIDIGRAPH_FILE:   A bidirected string graph in binary format.\n"
+"\n"
+"Output:\n"
+"      OUT_CYCLE_FILE:  The resulting Eulerian cycle as a vector of edge\n"
+"                       indices (in binary format)\n"
+);
+
+int main(int argc, char *argv[])
+{
+	USAGE_IF(argc != 3);
+	info("Loading bidirected graph graph from \"%s\"", argv[1]);
+	BidirectedStringGraph graph(argv[1]);
+
+	std::vector<size_t> cycle;
+
+	graph.eulerian_cycle(cycle);
+
+	info("Writing Eulerian cycle to \"%s\"", argv[2]);
+	{
+		std::ofstream out(argv[2]);
+		boost::archive::binary_oarchive ar(out);
+		ar << cycle;
+		out.close();
+		if (!out)
+			fatal_error_with_errno("Error writing to \"%s\"", argv[2]);
+	}
+}

### 2012-12-14 src/BidirectedStringGraph.cc
BidirectedStringGraph::eulerian_cycle() (in progress)

@@ -342,8 +342,83 @@ void BidirectedStringGraph::min_cost_circulation()
 	unimplemented();
 }
 
+void BidirectedStringGraph::assert_eulerian_cycle_valid(const std::vector<size_t> & cycle) const
+{
+	unsigned long total_traversal_count = 0;
+	foreach (const BidirectedStringGraphEdge & e, _edges) {
+		total_traversal_count += e.get_traversal_count();
+	}
+
+	info("cycle.size() = %zu", cycle.size());
+	assert(cycle.size() == total_traversal_count);
+	if (cycle.empty())
+		return;
+
+	v_idx_t cur_v_idx = _edges[cycle[0]].get_v1_idx();
+	bool inward = _edges[cycle[0]].v1_inward();
+	for (size_t i = 1; i < cycle.size(); i++) {
+		const BidirectedStringGraphEdge & e = _edges[i];
+		if (e.get_v1_idx() == e.get_v2_idx()) {
+			// Loops
+			if (e.v1_inward() != inward) {
+				inward = e.v2_inward();
+				cur_v_idx = e.get_v2_idx();
+			} else {
+				assert(e.v2_inward() != inward);
+				inward = e.v1_inward();
+				cur_v_idx = e.get_v1_idx();
+			}
+		} else {
+			// Non-loops
+			if (e.get_v1_idx() == cur_v_idx) {
+				assert(e.v1_inward() != inward);
+				inward = e.v2_inward();
+				cur_v_idx = e.get_v2_idx();
+			} else {
+				assert(e.get_v2_idx() == cur_v_idx);
+				assert(e.v2_inward() != inward);
+				inward = e.v1_inward();
+				cur_v_idx = e.get_v1_idx();
+			}
+		}
+	}
+	assert(inward != _edges[cycle[0]].v2_inward());
+	assert(cur_v_idx == _edges[cycle[0]].get_v2_idx());
+}
+
+// XXX this only checks if in-degree == out-degree for all vertices.  The graph
+// must also be strongly connected for an eulerian cycle to exist.
+void BidirectedStringGraph::assert_eulerian_cycle_possible() const
+{
+	for (v_idx_t v_idx = 0; v_idx < num_vertices(); v_idx++) {
+		const BidirectedStringGraphVertex & v = _vertices[v_idx];
+		unsigned long in_degree = 0;
+		unsigned long out_degree = 0;
+		foreach (const edge_idx_t edge_idx, v.edge_indices()) {
[…]

### 2012-12-14 src/BidirectedStringGraph.h

@@ -9,6 +9,7 @@
 #include <math.h>
 
 class DirectedStringGraph;
+class BidirectedStringGraphEdge;
 
 // A vertex of a bidirected string graph.
 class BidirectedStringGraphVertex : public StringGraphVertex {
@@ -88,7 +89,7 @@ public:
 	const BaseVec & get_seq_2_to_1() const { return _seq_2_to_1; }
 	BaseVec::size_type length() const { return _seq_1_to_2.size(); }
 
-	v_idx_t get_other_v_idx(v_idx_t this_v_idx) const
+	v_idx_t get_other_v_idx(const v_idx_t this_v_idx) const
 	{
 		v_idx_t v1_idx, v2_idx;
 		get_v_indices(v1_idx, v2_idx);
@@ -110,19 +111,19 @@ public:
 		return (_data >> 31) & 0x7fffffffULL;
 	}
 
-	void set_v1_idx(v_idx_t v1_idx)
+	void set_v1_idx(const v_idx_t v1_idx)
 	{
 		_data &= ~0x7fffffffULL;
 		_data |= v1_idx;
 	}
 
-	void set_v2_idx(v_idx_t v2_idx)
+	void set_v2_idx(const v_idx_t v2_idx)
 	{
 		_data &= ~(0x7fffffffULL << 31);
 		_data |= (uint64_t(v2_idx) << 31);
 	}
 
-	void set_v_indices(v_idx_t v1_idx, v_idx_t v2_idx)
+	void set_v_indices(const v_idx_t v1_idx, const v_idx_t v2_idx)
 	{
 		_data = (_data & (3ULL << 62)) | (uint64_t(v2_idx) << 31) | v1_idx;
 	}
@@ -142,6 +143,31 @@ public:
 	bool v2_inward() const { return (get_dirs() & 0x1) != 0; }
 	bool v1_inward() const { return !(v1_outward()); }
 	bool v2_outward() const { return !(v2_inward()); }
+
+	bool is_loop() const
+	{
+		v_idx_t v1_idx, v2_idx;
+		get_v_indices(v1_idx, v2_idx);
+		return (v1_idx == v2_idx);
+	}
+
+	bool v_outward(const v_idx_t v_idx) const
+	{
+		v_idx_t v1_idx, v2_idx;
+		get_v_indices(v1_idx, v2_idx);
+		assert2(v_idx == v1_idx || v_idx == v2_idx);
+		return ((v_idx == v1_idx && v1_outward()) ||
[…]

### 2012-12-14 src/BidirectedStringGraph.cc
BidirectedStringGraph::eulerian_cycle() (in progress)

@@ -344,43 +344,71 @@ void BidirectedStringGraph::min_cost_circulation()
 
 void BidirectedStringGraph::assert_eulerian_cycle_valid(const std::vector<size_t> & cycle) const
 {
-	unsigned long total_traversal_count = 0;
-	foreach (const BidirectedStringGraphEdge & e, _edges) {
-		total_traversal_count += e.get_traversal_count();
+	{
+		std::vector<edge_idx_t> times_traversed(num_edges(), 0);
+		foreach (const size_t edge_idx, cycle) {
+			times_traversed[edge_idx]++;
+		}
+		for (size_t i = 0; i < num_edges(); i++) {
+			assert(times_traversed[i] == _edges[i].get_traversal_count());
+		}
 	}
-
-	info("cycle.size() = %zu", cycle.size());
-	assert(cycle.size() == total_traversal_count);
 	if (cycle.empty())
 		return;
 
-	v_idx_t cur_v_idx = _edges[cycle[0]].get_v1_idx();
-	bool inward = _edges[cycle[0]].v1_inward();
-	for (size_t i = 1; i < cycle.size(); i++) {
-		const BidirectedStringGraphEdge & e = _edges[i];
-		if (e.get_v1_idx() == e.get_v2_idx()) {
-			// Loops
-			if (e.v1_inward() != inward) {
-				inward = e.v2_inward();
-				cur_v_idx = e.get_v2_idx();
-			} else {
-				assert(e.v2_inward() != inward);
-				inward = e.v1_inward();
-				cur_v_idx = e.get_v1_idx();
-			}
-		} else {
-			// Non-loops
-			if (e.get_v1_idx() == cur_v_idx) {
-				assert(e.v1_inward() != inward);
-				inward = e.v2_inward();
-				cur_v_idx = e.get_v2_idx();
-			} else {
-				assert(e.get_v2_idx() == cur_v_idx);
-				assert(e.v2_inward() != inward);
-				inward = e.v1_inward();
-				cur_v_idx = e.get_v1_idx();
-			}
+	const BidirectedStringGraphEdge & first_edge = _edges[cycle[0]];
+	v_idx_t cur_v_idx = first_edge.get_v1_idx();
+
+	// Direction of head through which cur_v_idx was entered
+	bool inward = !first_edge.v1_inward();
+
+	if (cycle.size() > 1) {
+		v_idx_t second_v_idx = first_edge.get_v2_idx();
+		// We want to go around the cycle in the direction indicated.
[…]

### 2012-12-14 src/BidirectedStringGraph.h

@@ -205,8 +205,9 @@ public:
 	}
 
 	// Print a bidirected string graph edge
-	void print(std::ostream & os, const v_idx_t v_idx,
-		   const bool print_seqs) const
+	void print(std::ostream & os = std::cout,
+		   const v_idx_t v_idx = 0,
+		   const bool print_seqs = false) const
 	{
 		v_idx_t read_1_idx = get_v1_idx();
 		v_idx_t read_2_idx = get_v2_idx();

### 2012-12-14 src/BidirectedStringGraph.cc
BidirectedStringGraph::eulerian_cycle()

@@ -342,8 +342,15 @@ void BidirectedStringGraph::min_cost_circulation()
 	unimplemented();
 }
 
-void BidirectedStringGraph::assert_eulerian_cycle_valid(const std::vector<size_t> & cycle) const
+// Verifies an Eulerian cycle that was computed on this bidirected graph.
+void BidirectedStringGraph::assert_eulerian_cycle_valid
+			(const std::vector<size_t> & cycle) const
 {
+
+	info("Verifying Eulerian cycle of length %zu on bidirected graph...",
+	     cycle.size());
+
+	// Make sure every edge was traversed the correct number of times.
 	{
 		std::vector<edge_idx_t> times_traversed(num_edges(), 0);
 		foreach (const size_t edge_idx, cycle) {
@@ -353,43 +360,45 @@ void BidirectedStringGraph::assert_eulerian_cycle_valid(const std::vector<size_t
 			assert(times_traversed[i] == _edges[i].get_traversal_count());
 		}
 	}
+
+	// An empty cycle is valid provided that no edges needed to be
+	// traversed.
 	if (cycle.empty())
 		return;
 
 	const BidirectedStringGraphEdge & first_edge = _edges[cycle[0]];
-	v_idx_t cur_v_idx = first_edge.get_v1_idx();
 
-	// Direction of head through which cur_v_idx was entered
+	// Index of the current vertex, and the direction of the head relative
+	// through which it was entered (relative to itself)
+	v_idx_t cur_v_idx = first_edge.get_v1_idx();
 	bool inward = !first_edge.v1_inward();
 
 	if (cycle.size() > 1) {
-		v_idx_t second_v_idx = first_edge.get_v2_idx();
-		// We want to go around the cycle in the direction indicated.
-		// So, for the first edge between vertices A and B, check to
-		// make sure the second edge in the cycle includes vertex B with
-		// the direction of the adjacent head consistent with the first
-		// head.  If not, flip vertices B and A so we go the other way
-		// around the cycle.
+		// We want to go around the cycle in the direction that it was
+		// provided in.  So, for the first edge between vertices A and
+		// B, check to make sure the second edge in the cycle includes
+		// vertex B with the direction of the adjacent head consistent
[…]

### 2012-12-14 src/BidirectedStringGraph.cc
Document BidirectedStringGraph::eulerian_cycle()

@@ -459,6 +459,21 @@ void BidirectedStringGraph::assert_eulerian_cycle_possible() const
 	}
 }
 
+//
+// Finds a generalized Eulerian cycle in this bidirected graph.
+//
+// A cycle in a bidirected graph is defined as a valid walk that begins and ends
+// at the same vertex in such a way that it would be valid to continue the walk
+// in the same way it started.  A generalized Eulerian cycle in a bidirected
+// graph is a cycle that traverses each edge e in the graph in either direction
+// exactly some specified number t(e) >= 0 times.  Here, t(e) is provided by
+// calling BidirectedStringGraphEdge::get_traversal_count().
+//
+// If a generalized Eulerian cycle can be found in this bidirected graph, the
+// indices of the edges in the cycle, in order, are returned in the vector
+// @cycle, which is resized to the length of the cycle.  Otherwise, this
+// function aborts the program with an error message.
+//
 void BidirectedStringGraph::eulerian_cycle(std::vector<size_t> & cycle) const
 {
 	assert_eulerian_cycle_possible();

### 2012-12-14 src/BidirectedStringGraph.cc
assert_eulerian_cycle_possible()

@@ -455,7 +455,14 @@ void BidirectedStringGraph::assert_eulerian_cycle_possible() const
 					out_degree += e.get_traversal_count();
 			}
 		}
-		assert(in_degree == out_degree);
+		if (in_degree != out_degree) {
+			std::cerr << _vertices[v_idx] << std::endl;
+			foreach (const edge_idx_t edge_idx, v.edge_indices()) {
+				std::cerr << _edges[edge_idx] << std::endl;
+			}
+			fatal_error("Vertex %u: in_degree(=%u) != out_degree(=%u)",
+				    v_idx + 1, in_degree, out_degree);
+		}
 	}
 }
 

### 2012-12-14 src/BidirectedStringGraph.h

@@ -234,6 +234,14 @@ public:
 		os << '\r';
 	}
 
+	// Print the string graph vertex.
+	friend std::ostream & operator<<(std::ostream & os,
+					 const BidirectedStringGraphEdge & e)
+	{
+		e.print(os, e.get_v1_idx());
+		return os;
+	}
+
 	// Print a bidirected string graph edge in DOT format
 	void print_dot(std::ostream & os, const v_idx_t v_idx,
 		       const bool print_seqs) const

### 2012-12-17 presentation/paper.tex
circle colors
paper: add missing "be"

@@
 \author{Eric Biggers}
+\date{December 12, 2012}
 
@@
 point towards each other or point away from each other, the walk is valid.  The
-red circles in \Figure{fig:bidiwalk_example} show the two pairs of heads that
-need to be checked for the possible walk $1 \to 2 \to 3 \to 4$.
+circles in \Figure{fig:bidiwalk_example} show the two pairs of heads that need
+to be checked for the possible walk $1 \to 2 \to 3 \to 4$.
 
@@
 			\draw[<-<,very thick] (3) edge (4);
-			\draw[color=red,thick] (2.00, 0) circle (0.8);
-			\draw[color=red,thick] (4.00, 0) circle (0.8);
+			\draw[color=black!70] (2.00, 0) circle (0.8);
+			\draw[color=black!70] (4.00, 0) circle (0.8);
 		\end{tikzpicture}
@@
 		reads, seeding on 128-mers.  The running time and memory usage of
-		\ProgramName{compute-overlaps} would need to improved further to handle
+		\ProgramName{compute-overlaps} would need to be improved further to handle
 		larger genomes.}

### 2012-12-17 presentation/paper.tex
paper: fix typo

@@
 reduction of bidirected flow; this produces a totally unimodular matrix that
-represents a linear program that can be solved to find an integal flow that is
+represents a linear program that can be solved to find an integral flow that is
 at worst a 2-approximation to the optimal integral flow on the bidirected

### 2014-03-24 docs/hw6/hw6.tex
Cleanup repository

\documentclass[letterpaper,12pt]{article}
\usepackage[left=3cm,right=3cm,top=3cm,bottom=3cm]{geometry}
\usepackage{enumerate}

\title{MATH-469 Homework 6}
\author{Eric Biggers}

\begin{document}
\maketitle

The goal of my project is to implement an algorithm that can assemble a genome,
given a set of reads that came from it.  For the purposes of this project, a
genome is defined as a sequence of A's, T's, C's, and G's, along with the
corresponding opposite strand that matches T's with A's and G's with C's.  The
input to the algorithm is a set of {\em reads} that are substrings from the
genome.  Each read may be from either the ``forward'' strand of the genome or
the ``reverse'' strand of the genome, but the reads will always be directed
forwards along the ``forward'' strand or backwards along the ``reverse'' strand
(this corresponds to the sequencing reactions always proceeding from the 5' to
the 3' end of each strand).

The reads will not be required to be of any particular length, other than long
enough to actually compute some overlaps, but the longer they are, the better
the algorithm should work.  The algorithm will not account for any pairing
information that may exist among reads (e.g. reads that came from the same DNA
fragment or insert of known size).  Also, I will initially assume that the reads
do not contain any sequencing errors and that there is no foreign DNA in the
sample.

The overall description of the algorithm is as follows:

\begin{enumerate}
\setcounter{enumi}{-1}
\item For testing the algorithm, the initial step is to simulate reads from a
sample genome.  An existing program can be used for this.
\item Compute all overlaps of some length $\ell$ among the reads.  An overlap is
where the bases at one end of a read match the bases at one end of another read.
The bases may match literally, or they may match the reverse-complement
sequence.  The length $\ell$ is chosen so that the probability of two random DNA
sequences having an overlap of length $\ell$ purely by chance is sufficiently
low; this will depend on the genome size.

The original algorithm accounts for erroneous data by looking not for perfect
overlaps, but rather for overlaps where the edit distance between the overlapped
regions is less than a certain fraction of the overlap.  (i.e., the bases might
be different by up to a certain threshold, such as 2.5\%, and still be
considered a valid overlap).  I will initially work with error-free data and
compute only perfect overlaps, but this could be something to look into
implementing later.

The naive approach to computing overlaps requires comparing all $N$ of the reads
to every other read--- an $O(N^2)$ approach that would likely be a significant
bottleneck in the overall algorithm.  I plan to speed up the running time of
this step of the algorithm by using a seed-and-extend approach where only reads
sharing a $k$-mer are compared, where $k \le \ell$ (for perfect overlaps).  This
can be done by placing each $k$-mer, along with its location, in a hash table
indexed by the $k$-mer.

\item Build a graph where each vertex corresponds to either the beginning or the
end of a read, and the overlaps are used to connect vertices.  (Edges are
labeled with DNA sequences.)  The exact edges that should be added to the graph
given an overlap $o$ are described in the original paper.  Basically, the edges
are always labeled with sequences oriented {\em outwards} from the overlapped
region, and this sequence may be either forward or reverse-complement.  Reads
that are fully contained by another read are not included in the graph, as they
theoretically do not contain any information not provided by the longer read.
The main challenge in this step is probably to represent the graph in a
memory-efficient way.
\item Delete all transitive edges from each vertex (i.e. edges $f \to h$ where
there are also edges $f \to g \to h$), since these edges are redundant.  The
algorithm for this step is provided in the original paper, and pseudocode is
provided, so it should be fairly easy to implement this step, although again it
might be tricky to keep memory usage from being too high.
\item Collapse unbranched paths in the graph, since the interior vertices are
redundant.  This should be relatively easy.  For example, branchless paths could
be extended from all vertices of degree 2.
\item Use a calculation called the A-statistic to compute the probabilitity that
each unbranched path contains unique sequence (i.e. should be traversed only 1
time).  The A-statistic for an unbranched path of length $\Delta$ that contains
$k$ internal vertices is $\Delta(n/G) - k \ln 2$, where $n$ is the total number
of reads and $G$ is the estimated genome size.  This step sounds
straightforward, but there are a couple things to consider.  First, when
collapsing unbranched paths in the previous step, the number of vertices in the
uncollapsed path must be remembered.  Secondly, the genome size must be
estimated, as it is not known.  And finally, reads that are fully contained in a
longer read should be mapped back into the graph to increase the accurancy of
the A-statistic calculations.
\item Calculate a traversal count for each edge in the graph to make the graph
Eulerian.  In the original paper, this is formulated as a minimum cost network
flow problem, so it should theoretically be possible to just set up the problem
and plug it into existing code, such as that in the LEMON library.  However,
since the actual problem is not actually a fully general minimum cost network
flow problem, there could eventually be some optimizations made by writing
stand-alone code.
\item Compute an Eulerian tour of the graph.  There may be several, but one
should be the original sequence.  I'm planning to do this using the stack-based
algorithm.
\item When doing evalutions of this algorithm, the last step is to compare the
assembly with the original genome.  This can be done with existing tools such as
the {\tt dnadiff} script included with the {\em MUMmer} sequence alignment
software.  This comparison will indicate how similar the assembled genome is to
the original genome.  Ideally, they should be identical.
\end{enumerate}

The actual implementation of this algorithm will be done in C++.  I will
initially implement the entire algorithm in one executable program, but I may
eventually split it into several executable problems if it makes sense to do so
(this is only possible once all the intermediate data structures can be
serialized and deserialized to external files).  Regardless of this, each step
of the algorithm will be split into one or more subroutines.  Logging will be
enabled by default.

\end{document}
